### 2017-07-26 ms/results.tex
model and results revised by UL & MWF

\magnification=1170
\hsize=6.5truein
\hoffset=.05truein
%\vsize=8.5truein
%\voffset=.25truein
\hfuzz=14pt
\parskip=5pt
\baselineskip=16pt

\def\harr#1#2{\smash{\mathop{\hbox to .2in{\rightarrowfill}}\limits^{\scriptstyle#1}_{\scriptstyle#2}}}

\def\var{\varepsilon}

\def\cl{\hbox{\tenbf L}}
\def\ci{\hbox{\tenbf I}}

\font\tit=cmbx10 at 14pt
\ \vskip1.1in
\centerline{\tit Vertical and Oblique Transmission under}\smallskip
\centerline{\tit Fluctuating Selection}
\vskip.3in

\centerline{Yoav Ram$^1$, Uri Liberman$^{2}$, and Marcus W. Feldman$^1$}

\vskip.5truein

%\centerline{$*$ Corresponding author}
\bigskip
\centerline{$^1$Department of Biology}
%\vskip-7pt
\centerline{Stanford University}
%\vskip-7pt
\centerline{Stanford, CA 94305-5020}
\bigskip

\centerline{$^2$School of Mathematical Sciences}
%\vskip-7pt
\centerline{Tel Aviv University}
%\vskip-7pt
\centerline{Tel Aviv, Israel 69978}
%\vskip-7pt
\centerline{uril@tauex.tau.ac.il}
\vfil
\centerline{July 24, 2017, v.2}

\break

\noindent{\bf THE MODEL}
\smallskip

Consider an infinite population whose members are characterized by their phenotype $\phi$, which can be of two types $\phi=A$ or $\phi=B$. Each individual has a probability $x$ $(0\le x\le 1)$ of exhibiting the phenotype $A$ and $(1-x)$ of exhibiting $B$.
We follow the evolution of $x$ over discrete non-overlapping generations. In each generation individuals are subject to  selection where the fitnesses of $A$ and $B$ are $w_A$ and $w_B$, respectively.

An offspring inherits its phenotype, namely its probability of exhibiting the phenotype $A$ {\sl vertically} from  parent of type $A$ with probability $\rho$ and {\sl diagonally} from a random individual in the parental population with probability $(1-\rho)$. Therfore, given that the parent phenotype is $\phi$ and assuming {\sl uni-parental inheritance} (Zefferman 2016), the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is
$$P(\phi' =A|\phi) =\left\{\matrix{(1-\rho)x+\rho &\hbox{\tenrm if}\ \phi=A\cr 
\noalign{\smallskip}
 (1-\rho)x & \hbox{\tenrm if}\ \phi =B\cr}\right.\eqno(1)$$
 where $x=P(\phi=A)$ in the parent's generation before selection.

 Since $x$ is also the frequency of  phenotype $A$ before selection and $(1-x)$ that of $B$ before selection, after one generation $x'=P(\phi'=A)$ is given by the recursion equation
 $$x'={w_A\over w}x\bigl[(1-\rho)x +\rho\bigr] +{w_B\over w}(1-x)\bigl[(1-\rho)x\bigr],\eqno(2)$$
 where $w$ is the {\sl mean fitness}, namely
 $$w=w_A x+w_B(1-x).\eqno(3)$$
 Equation (2) can be rewritten as
 $$x'=x\cdot{x(1-\rho)(w_A-w_B) +\rho w_A +(1-\rho)w_B\over x(w_A-w_B) +w_B}.\eqno(4)$$

 In what follows we explore the evolution of the recursion (4), namely the equilibria and their stability properties, in the cases of {\sl constant environments} and {\sl changing environments}.
 \bigskip
 \bigskip

\noindent{\bf CONSTANT ENVIRONMENT}
\smallskip

When the environment is constant, the fitness parameters $w_A$ and $w_B$ do not change between generations, and we have the following result.

\proclaim Result 1. If $0<\rho\le 1$ and both $w_A$ and $w_B$ are positive with $w_A\ne w_B$, then fixation in the phenotype $A(B)$ is globally stable when $w_A>w_B$ $(w_A<w_B)$.\par

\noindent{\it Proof}.
If we rewrite (4) as $x'=x\cdot F(x)$, it is easily seen that $F(1)=1$, and for $\rho>0$ and $0<x<1$
$$\eqalign{
&F(x)>1\quad\hbox{\tenrm when}\quad w_A>w_B,\cr
&F(x)<1\quad\hbox{\tenrm when}\quad w_A<w_B.}\eqno(5)$$
Hence, as $w_A>0$ and $w_B>0$, both fixations in $A$ or in $B$  ($x^*=1$ for fixation in $A$ and $x^*=0$ for fixation in $B$) are equilibrium points of (4). Moreover, if $x_t$ is the value of $x$ at the $t$-th generation $(t=0,1,2,\dots)$, from (4)  and (5) we have for any $0<x_0<1$ and all $t=0,1,2,\dots$,
$$\eqalign{
&x^{t+1}>x^t\quad\hbox{\tenrm when}\quad w_A>w_B,\cr
&x^{t+1}<x^t\quad\hbox{\tenrm when}\quad w_A<w_B.}\eqno(6)$$

\noindent Hence, as $x^*=1$ or $x^*=0$ are the only equilibrium points, then
$$\eqalign{
&\lim_{t\to\infty}x_t=1,\ \ \hbox{\tenrm for all}\ 0<x_0\le 1,\ \ \hbox{\tenrm when}\ w_A>w_B,\cr
&\lim_{t\to\infty}x_t=0,\ \ \hbox{\tenrm for all}\ 0\le x_0< 1,\ \ \hbox{\tenrm when}\ w_A<w_B.}\eqno(7)$$
Thus fixation of the favored phenotype is globally stable, and ultimately all individuals have the favored phenotype.
\bigskip

\noindent{\it Remarks}

\item{1.} If $\rho=0$, there is perfect oblique transmission, and the recursion (4) reduces to $x'=x$; the phenotype frequencies do not change over time.
\item{2.} If $w_A=w_B>0$, again $x'=x$, and no stable equilibrium exists.
\item{3.}With {\sl extreme selection}, where for example $w_A=1$ and $w_B=0$ (or $w_B=1$ and $w_A=0$), individuals of the disfavored phenotype cannot reproduce and the recursion (4) reduces to
$$x'=(1-\rho)x+\rho.\eqno(8)$$
Therefore
$$x_t=1-(1-\rho)^t(1-x_0).\eqno(9)$$
Hence, unless $\rho=0$ we have $x_t\,\harr{}{t\to\infty}\,1$ for all initial values $0\le x_0\le 1$ and ultimately the favored phenotype fixes.\par

Figure 1(a) shows the transformation (4) and Figure 1(b) represents its dynamics for different vertical transmission rates $\rho$ when $w_A=1$ and $w_B=0.9$. Figure 2 shows the time to fixation $\tau$ (i.e., time to reach $x_t=1-\var$, for $\var$ small, which is $\tau=\log{\var\over1-x_0}/\log(1-\rho)\simeq\log\var/\log(1-\rho)$ for $x_0\approx0$) with extreme selection $(w_A=1, w_B=0)$ for different values of $\rho$ and initial frequency $x_0$. We see that greater vertical transmission leads to significantly lower fixation time.
\bigskip
\bigskip

\noindent{\bf PERIODICALLY CHANGING ENVIRONMENT}
\smallskip

Next we consider the case where the environment is not constant but rather changes periodically, such that the favored phenotype changes after a fixed number of generations. Simple examples are $A1B1=ABABAB\dots$, in which the favored genotype switches every generation, or $A2B1=AABAABAAB\dots$, where every two generations in which selection favors $A$ are followed by a single generation in which selection favors $B$.

 In general, $AkBl$ denotes a selection regime in which the period is of $(k+l)$ generations with $k$ generations favoring phenotype $A$ followed by $l$ generations favoring $B$.

 Let $W$ be the fitness of the favored phenotype and $w$ be that of the other phenotype where $0<w<W$. Rewrite equation (4) as $x'=F_A(x)=xf_A(x)$ when $A$ is favored and $x'=F_B(x)=xf_B(x)$ when $B$ is favored. Then
  $$\eqalign{
  &f_A(x) = {x(1-\rho)(W-w) +\rho W +(1-\rho)w\over x(W-w) +w},\cr\noalign{\smallskip}
  &f_B(x) = {x(1-\rho)(w-W) +\rho w +(1-\rho)W\over x(w-W) +W}.}\eqno(10)$$

  \noindent If $x_t$ denotes the frequency of exhibiting the phenotype $A$ at generation $t$ starting with $x_0$ initially, then as we are interested in the values of $x_t$ for $t=n(k+l)$ with $n=0,1,\dots$ at the end of complete periods then in fact
  $$x_{(n+1)(k+l)}=F(x_{n(k+l)}),\quad n=0,1,2,\dots,\eqno(11)$$
  where $F$ is the composed function
  $$F=\underbrace{F_B\circ F_B \circ\cdots\circ F_B}_{l\ \hbox{\sevenrm times}} \circ\underbrace{F_A\circ F_A \circ\cdots\circ F_A}_{k\ \hbox{\sevenrm times}}.\eqno(12)$$
  Clearly, since $F_A(0) =F_B(0) =0$ and $F_A(1) =F_B(1) =1$, both fixations in $A$ or in $B$ are equilibrium points. An interesting question is when these fixations are locally stable. We concentrate on $x^*=0$, the fixation of the phenotype $B$. As $x'=F_A(x) =xf_A(x)$ for $k$ generations and $x'=F_B(x) =xf_B(x)$ for $l$ generations, the linear approximation of $F(x)$ ``near'' $x=0$ is
   $$F(x) \simeq\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^lx.\eqno(13)$$
   Hence the local stability of $x^*=0$ is determined by the product $\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l$, and $x^*=0$  is locally stable if the product in (13) is less than 1 and unstable if it is larger than 1.

   From (10) we have
   $$\eqalign{
   &f_A(0) = 1+\rho{W-w\over w},\cr\noalign{\smallskip}
   &f_B(0) = 1+\rho{w-W\over W}.}\eqno(14)$$
   We start with the case $k=l$.

   \proclaim Result 2. If $k=l$ and $0<w<W$ with $0<\rho<1$, fixation of $B$ is unstable.\par

   \noindent {\it Proof.} The local stability of $x^*=0$, the fixation of $B$, is determined by the product
   $$\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^k=\left[\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right)\right]^k.\eqno(15)$$
   Observe that
   $$\eqalign{
   \left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right) &= \left(1-\rho+\rho{W\over w}\right)\left(1-\rho+\rho{w\over W}\right)\cr
   &= \left(1-\rho\right)^2+ \rho^2 +\rho(1-\rho){W^2+w^2\over Ww}.}\eqno(16)$$
   Thus,
   $$\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right) =1+\rho(1-\rho){\left(W-w\right)^2\over wW}>1.\eqno(17)$$
   Since $0<\rho<1$ and $0<w<W$,  the $B$ fixation is unstable.
  \bigskip

\noindent{\it Conclusions}

\item{1.}As $k=l$ and the above result holds also when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is not stable. Thus neither phenotype can become extinct, and we have a {\sl protected polymorphism} (Prout 1968).

\item{2.}Observe that for general $k$, $l$, the condition for local stability of fixation in $A$ is
$$\bigl[f_A(0)\bigr]^l \bigl[f_B(0)\bigr]^k <1, \eqno(18)$$
and that of $B$ is
$$\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l <1. \eqno(19)$$
Therefore, following Result 2, 
$$\bigl[f_A(0)\bigr]^{k+l}\bigl[f_B(0)\bigr]^{k+l}>1,\eqno(20)$$
and it is impossible that both fixations are stable. Further, since by (14)  $f_A(0)>1$ and $0<f_B(0)<1$ when $0<w<W$, by choosing $k$ and $l$ appropriately, fixation on $A$ or fixation on $B$ (but not both) can be stable. In addition, we can have both fixations unstable giving the following result.

\proclaim Result 3. With $0<\rho<1$; $0<w<W$ in the case of $AkBl$ periodically changing environments, both fixations may be unstable, producing a protected polymorphism.\par

\noindent{\it Proof}. Let $a=1+\rho{W-w\over w}$ and $b=1+\rho{w-W\over w}$. Then since $0<\rho <1$ and $0<w<W$, we have $a>1$ and $0<b<1$. Following (13), fixation in $B$ is not stable if $a^kb^l>1$, and similarly  fixation in $A$ is unstable if $a^lb^k>1$. Therefore both fixations are unstable if
$$a^kb^l>1\quad\hbox{\tenrm and}\quad a^lb^k>1,\eqno(21)$$
or equivalently if
$$k\log a +l\log b>0\quad\hbox{\tenrm and}\quad l\log a+k\log b>0.\eqno(22)$$
Now (22) holds if and only if
$$k\,{\log(1/b)\over\log a} < l < k\,{\log a\over\log(1/b)}.\eqno(23)$$ 
Inequalities (23) are consistent if and only if $\log(1/b)<\log a$, i.e., $ab>1$. But by (17),
$$ab=\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over w}\right)>1.\eqno(24)$$

\noindent{\it Remark}.
It should be pointed out that the local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations $A$ was favored $k$ times and $B$ was favored $l$ times, and not their order in the cycle.

When neither fixation in $A$ or $B$ is stable, we have a case of protected polymorphism, and we expect to have one or more polymorphic equilibria. For the simple case of $A1B1$ periodically changing environment we have

\proclaim Result 4. In the case $A1B1$ with $0<\rho<1$ and $0<w<W$ the two fixations are unstable and there exists a unique stable polymorphism.

\noindent{\it Proof}. Let $x$ be the initial frequency of $A$ and $x'$ its frequency  after one cycle of the $A1B1$ selection. Then $x' =F_B\bigl(F_A(x)\bigr)$ where by (10)
$$\eqalign{
&F_A(x)=x{x(1-\rho)(W-w) +\rho W+(1-\rho)w\over x(W-w) +w},\cr\noalign{\smallskip}
&F_B(y) = y{y(1-\rho)(w-W) +\rho w+(1-\rho)W\over y(w-W)+W}.}\eqno(25)$$
The equilibrium equation is $x=F_B\bigl(F_A(x)\bigr)$, which reduces to a fourth degree polynomial equation in $x$. Since the two fixations in $B$ and $A$ are equilibria corresponding to the two solutions $x=0$ and $x=1$, the other equilibria correspond to solutions of a quadratic equation $G(x) =Ax^2 +Bx +C=0$ with
$$A=1,\quad B=-{W(1-\rho)-w(3-\rho)\over (2-\rho)(W-w)},\quad C={-w\over (2-\rho)(W-w)}.\eqno(26)$$
As $0<\rho<1$ and $0<w<W$ we have
$$G(0) = {-w\over (2-\rho)(W-w)}<0\eqno(27)$$
and
$$G(1)=A +B+ C =1-{W(1-\rho) -w(3-\rho)\over (2-\rho)(W-w)} -{w\over (2-\rho)(W-w)},\eqno(28)$$
or
$$G(1) ={W\over (2-\rho)(W-w)} >0.\eqno(29)$$
Also, as $A=1$, $G(x)\to\infty$ when $x\to\pm\infty$, the quadratic equation $G(x)=0$ has two real roots, one negative and one positive $x^*$ satisfying $0<x^*<1$. The latter determines a unique polymorphism. Let $H(x) =F_B\bigl(F_A(x)\bigr)$. Then 
 $$H(0)=0,\quad H(x^*)=x^*,\quad H(1)=1.\eqno(30)$$
 Also 
 $$F'_A(x) = {x^2(1-\rho)\left(W-w\right)^2 +2xw(1-\rho)(W-w) +w\left[\rho W +(1-\rho)w\right]\over \left[x(W-w) +w\right]^2}\eqno(31)$$
 and
  $$F'_B(x) = {x^2(1-\rho)\left(w-W\right)^2 +2xW(1-\rho)(w-W) +W\left[\rho w +(1-\rho)W\right]\over \left[x(w-W) +W\right]^2}.\eqno(32)$$
  As $0<\rho<1$ and $0<w<W$, we have $F'_A(x)>0$ for  $0\le x\le 1$. Observe that the numerator of $F'_B(x)$ is linear in $\rho$; its value when $\rho=1$ is $wW>0$, and when $\rho=0$ it is
  $$x^2\left(w-W\right)^2 +2xW(w-W) +W^2 =\left[x(w-W) +W\right]^2 >0.\eqno(33)$$
  Hence $F'_{B(x)}>0$ for all $0\le x\le 1$, and $H'(x) =F'_B\bigl(F_A(x)\bigr)F'_A(x)$ is positive when $0\le x\le 1$. Thus $H(x)$ is a monotone increasing function of $x$ for $0\le x\le 1$.

  As the two fixations are unstable, combining all the above properties, $H(x)$ has the following graph.

  \bigskip\centerline{[GRAPH HERE]}\bigskip

  \noindent Therefore $x^*$, the unique polymorphism, is globally stable. That is, starting from any initial value $0<x_0 <1$ we have  $x_t\,\harr{}{t\to\infty}\,x^*$.

 \bigskip
 \bigskip

 \noindent{\bf RANDOMLY CHANGING ENVIRONMENT}
 \smallskip

 Now consider the  case where the environment changes according to a stochastic process. Without loss of generality, assume that the fitness parameters at generation $t$ ($t=0,1,2,\dots$) are $1+s_t$ for phenotype $A$ and $1$ for phenotype $B$, such that the random variables $s_t$ for $t=0,1,2,\dots$ are independent and identically distributed. Also we assume that there are positive constants $C$ and $D$ such that $P(-1+C<s_t <D)=1$.

 Corresponding to (4), the recursion equation is
 $$x_{t+1}=x_t {1+\rho s_t +x_t(1-\rho)s_t\over 1+x_ts_t}\quad t=0,1,2,\dots\; .\eqno(34)$$

 As $\{x_t\}$ for $t=0,1,2,\dots$ is a sequence of random variables, the notion of stability of the two fixation states needs clarification. Following Karlin and Liberman (1975) we make the following definition.

 \noindent{\it Definition}: ``stochastic local stability''.
 A constant equilibrium state $p^*$ is said to be {\sl stochastically locally stable} if for any $\var>0$ there exists a $\delta>0$ such that $|x_0-p^*|<\delta$ implies
 $$P\left(\lim_{t\to\infty}x_t =p^*\right)\ge 1-\var.\eqno(35)$$
 Thus stochastic local stability holds for $p^*$ provided for any initial $x_0$ sufficiently near $p^*$ the process $x_t$ converges to $p^*$ with high probability.

 In our case there are two constant equilibria $x^*=0$ and $x^*=1$ corresponding to fixation in $B$ and $A$, respectively. Following Karlin and Liberman (1975) we can characterize the stochastic local stability of these fixations as follows.

 \proclaim Result 5. Suppose $E\left[\log (1+\rho s_t)\right]>0$. Then $x^*=0$, the fixation in $B$, is not stochastically locally stable. In fact $P\left(\lim_{t\to\infty}x_t=0\right)=0$.

 \noindent{\it Proof}. Rewrite recursion (34) as
 $${x_t+1\over x_t} =(1+\rho s_t)\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right].\eqno(36)$$
 Then
 $$\log x_{t+1} -\log x_t =\log(1+\rho s_t) +\log\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right].\eqno(37)$$
 Summation yields
 $${1\over t}\left[\log x_t-\log x_0\right] ={1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n) +{1\over t}\sum_{n=0}^{t-1}\log\left[1-x_n{\rho s_n(1+s_n)\over (1+\rho s_n)(1+x_ns_n)}\right].\eqno(38)$$
Let $\mu=E\left[\log(1+\rho s_t)\right]$. As $\{s_t\}_{t\ge 0}$ are independent and identically distributed random variables, the {\sl strong law of large numbers} applies and
$$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)=\mu\eqno(39)$$
almost surely.

Let $\zeta$ be such that ${1\over t}\sum_{n=0}^{t-1}\log[1+\rho s_n(\zeta)]=\mu$ and assume that $\lim_{t\to\infty}x_t(\zeta)=0$. As the random variables $\{s_t\}_{t\ge 0}$ are uniformally bounded,
$$x_t(\zeta){\rho s_t(\zeta)[1+s_t(\zeta)]\over [1+\rho s_t(\zeta)][1+x_t(\zeta)s_t(\zeta)]}\;\;\harr{}{t\to\infty}\;\;0\eqno(40)$$
and
$$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log\left[1-x_n(\zeta){\rho x_n(\zeta)[1+s_n(\zeta)]\over [1+\rho s_n(\zeta)][1+x_t(\zeta)s_n(\zeta)]}\right]=0.\eqno(41)$$
Thus (38) implies that
$$\lim_{t\to\infty}{1\over t}\left[\log x_t(\zeta) -\log x_0(\zeta)\right]=\mu.\eqno(42)$$
If $\mu =E\left[\log(1+s_t)\right]>0$, then from (42) we deduce that $\lim_{t\to\infty}x_t(\zeta)=\infty$, a contradiction.
Therefore when $\mu>0$, $P\left(\lim_{t\to\infty}x_t=0\right)=0$, and fixation of $B$ ($x^*=0$) is stochastically locally unstable.

 Thus by Result 5, for $x^*=0$ to be stochastically locally stable it is necessary that $E[\log(1+\rho s_t)]\le 0$. In fact, the strict inequality is sufficient.

 \proclaim Result 6. Suppose $E[\log(1+\rho s_t)]<0$. Then $x^*=0$ is stochastically locally stable.

 \noindent{\it Proof}. Let $\mu=E[\log(1+\rho s_t)]$. Then as $\{s_t\}_{t\ge 0}$ are independent and identically distributed random variables, the strong law of large number applies and almost surely
 $$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)=\mu<0.\eqno(43)$$
 Appealing to the Egoroff Theorem, for any $\var>0$ there exists $T$ such that 
 $$P\left({1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)<{\mu\over 2}\ \hbox{\tenrm for all}\ t\ge T\right)\ge 1-\var.\eqno(44)$$
 As $0\le \rho\le 1$ and the $\{s_t\}_{t\ge 0}$ are uniformly bounded, we can find a $\delta'>0$ such that
 $$x_t<\delta'\Longrightarrow\left|\log\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right]\right|<-{\mu\over 4}.\eqno(45)$$
 Also, as $0\le x_t\le 1$ for all $t$,
 $$x_{t+1}=x_t{1+\rho s_t +x_t(1-\rho)s_t\over 1+x_ts_t}< Kx_t,\eqno(46)$$
 where $K$ is independent of $t$. It follows that there exists a $\delta$ with $0<\delta<\delta'$ such that
 $$x_o<\delta\Longrightarrow x_t<\delta'\ \hbox{\tenrm for all}\ t=0,1,2,\dots,T-1.\eqno(47)$$
 Let $\xi$ be a realization of the evolutionary process such that
 $${1\over t}\sum_{n=0}^{t-1}\log[1+\rho s_n(\xi)]<{\mu\over 2}\ \hbox{\tenrm for all}\  t\ge T\eqno(48)$$
 and assume $x_0<\delta$. Then
 $$\eqalign{
 {1\over T}&[\log x_T(\xi) -\log x_0(\xi)] \cr
 &={1\over T}\sum_{n=0}^{T-1}\log[1+\rho s_n(\xi)]+{1\over T}\sum_{n=0}^{T-1}\log[1-x_n(\xi){\rho s_n(\xi)[1+s_n(\xi)]\over [1+\rho s_n(\xi)][1+x_n(\xi)s_n(\xi)]}\cr\noalign{\smallskip}
 &<{\mu\over 2}- {\mu\over 4} ={\mu\over 4}<0, }\eqno(49)$$
 and therefore $x_T(\xi) <x_0(\xi) <\delta'$. Invoking induction we get that for $t\ge T$
 $${1\over t}\log{x_t(\xi)\over x_0}\le {\mu\over 4},\eqno(50)$$
 or for all $t\ge T$
 $$x_t(\xi)\le x_0\exp\left({\mu\over 4}t\right).\eqno(51)$$

 As $\mu<0$, this implies that $x_t(\xi)\,\harr{}{t\to\infty}\,0$. Therefore we have shown that for given $\var>0$ there is a $\delta>0$ such that if $0<x_0<\delta$, then $P\left(\lim_{t\to\infty}x_t =0\right)\ge 1-\var$, and therefore $x^*=0$, the fixation in $B$ is stochastically locally stable.
\smallskip

 \noindent{\it Remark}. 
 Using the general notation for the fitness parameters $w_A$ and $w_B$, the stochastic local stability of the $B$ fixation is determined by the sign of $E\left[\log\left(1-\rho+\rho{w_A\over w_B}\right)\right]$ and that of the $A$ fixation by the sign of $E\left[\log\left(1-\rho+\rho{w_B\over w_A}\right)\right]$. When one of these signs is negative, there is stochastic local stability, and when it is positive, with probability one convergence to that fixation does not occur. Following (16),
 $$\left(1-\rho +\rho{w_A\over w_B}\right)\left(1-\rho+\rho{w_B\over w_A}\right) >1.\eqno(52)$$
 Therefore, as in the case of periodically changing environments $AkBl$, it is impossible that the two fixations are both stochastically locally stable.
 \bigskip
 \bigskip

 \noindent{\bf EVOLUTIONARY STABILITY OF OBLIQUE TRANSMISSION}
 \smallskip

 An interesting question concerns the evolution of oblique transmission itself. For example, is there an evolutionarily stable rate of oblique transmission? To answer this question we use a modifier model.

 Suppose that the vertical transmission rate is determined by a genetic locus with two possible alleles $m$ and $M$. Let the vertical transmission rates determined by $m$ and $M$ be $\rho$ and $P$, respectively. Thus there are four pheno-genotypes $mA$, $mB$, $MA$, $MB$ whose frequencies at a given generation are denoted by $x_1$, $x_2$, $x_3$, $x_4$, respectively. As the fitnesses are determined by the two phenotypes $A$ and $B$, and the modifier locus is selectively neutral, we have the following table.
 $$\matrix{
 \hbox{\tenrm pheno-genotype}&mA&mB&MA&MB\cr\noalign{\smallskip}
 \hbox{\tenrm frequency}&x_1&x_2&x_3&x_4\cr\noalign{\smallskip}
 \hbox{\tenrm fitness}&w_A&w_B&w_A&w_B\cr\noalign{\smallskip}
 \hbox{\tenrm vertical transmission rate}&\rho&\rho&P&P\cr}\eqno(53)$$
 Following the rationale leading to equation (2), the next generation pheno-genotype frequencies $x'_1$, $x'_2$, $x'_3$, $x'_4$ are
 $$\eqalign{
 \overline wx'_1 &= w_Ax_1\bigl[(1-\rho)(x_1+x_3) +\rho\bigr] +w_Bx_2(1-\rho)(x_1+x_3)\cr
 \overline wx'_2 &= w_BAx_1(1-\rho)(x_2+x_4) +w_Bx_2\bigl[(1-\rho)(x_2+x_4)+\rho\bigr]\cr 
  \overline wx'_3 &= w_Ax_3\bigl[(1-P)(x_1+x_3) +P\bigr] +w_Bx_4(1-P)(x_1+x_3)\cr
\overline wx'_4 &= w_Bx_3(1-P)(x_2+x_4) +w_Bx_4\bigl[(1-P)(x_2+x_4)+P\bigr], 
}\eqno(54)$$
with $\overline w$, the mean fitness, given by
$$\overline w =w_A(x_1+x_3) +w_B(x_2+x_4).\eqno(55)$$
Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} to invasion by allele $M$.

Because we assume the initial existence of a stable polymorphic equilibrium that depends on $\rho$ (the rate determined by $m$), we cannot assume a {\sl constant environment}, which always leads to fixation of the favored type, independent of $\rho$. We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$. Specifically, following (54) with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
$$\underline x''=T_2(T_1\underline x),\eqno(56)$$
where $\underline x'=T_1\underline x$ is given by (54) with $w_A=W$, $w_B=w$, and $\underline x''=T_2\underline x'$ is given by (54) with $w_A=w$, $w_B=W$. Here $\underline x$, $\underline x'$, $\underline x''$ are the frequency vectors.

From the analysis of the $A1B1$ case, we know that when only the $m$ allele is present with associated rate $\rho$, with $0<\rho<1$ and $0<w<W$, a unique stable equilibrium $\underline x^* =(x_1^*,x_2^*,0,0)$ exists. $x_1^*$ is the only positive root of the quadratic equation $G(x) =Ax^2 +Bx +C=0$ with $A,B,C$ specified in (26). Solving $G(x)=0$ gives
$$x_1^* ={1\over 2} -{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}\over 2\cdot(2-\rho)(W-w)},\eqno(57)$$
 and it is easily seen that
 $${\sqrt{Ww}-w\over W-w}<x_1^* <{1\over 2}.\eqno(58)$$

 The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ derived from (54) and given by
 $$\overline w^*\cl_1 =\left[\matrix{W\bigl[(1-P)x_1^* +P\bigr] & w(1-P)x_1^*\cr\noalign{\medskip}
 W(1-P)x_2^* & w\bigl[(1-P)x_2^* +P\bigr]\cr}\right]\eqno(59)$$
 and
 $$\overline w^{**}\cl_2 =\left[\matrix{w\bigl[(1-P)x_1^{**} +P\bigr] & W(1-P)x_1^{**}\cr\noalign{\medskip}
 w(1-P)x_2^{**} & W\bigl[(1-P)x_2^{**} +P\bigr]\cr}\right],\eqno(60)$$
where $\underline x^{**} =T_1\underline x^*$ and
$$\overline w^* =Wx_1^* +wx_2^*,\qquad \overline w^{**} =wx_1^{**} +Wx_2^{**}.\eqno(61)$$
Due to the symmetry between the two phenotypes $A$ and $B$ in the $A1B1$ case, we have $x_1^{**} =x_2^*$ and $x_2^{**} =x_1^*$ so that $\overline w^{**} =\overline w^*$ and in fact
$$\overline w^*\cl_2 = \left[\matrix{w\bigl[(1-P)x_2^* +P\bigr] & W(1-P)x_2^*\cr\noalign{\medskip}
 w(1-P)x_1^* & W\bigl[(1-P)x_1^* +P\bigr]\cr}\right].\eqno(62)$$
  \smallskip

   \noindent{\it Remark}. 
 Observe that as $\underline x^* =T_2(T_1\underline x^*)$ with $x_3^*=x_4^*=0$, from (59) and (60) with $P=\rho$ we have
 $$\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right]= \cl_2\cdot\cl_1\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right] =\cl\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right].\eqno(63)$$
 Hence when $P=\rho$ one of the eigenvalues of $\cl$ is 1.
 \smallskip

 In general $\cl=\cl_2\cdot\cl_1$, and using (59) and (62) we get
 $$\eqalign{
 &\left(\overline w^*\right)^2\cl =\cr\noalign{\smallskip}
 &\left[\matrix{Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[w(1-P)x_1^*\bigr]^2 & W(1-P)\bigl[P+x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr]\cr
 \noalign{\medskip}
 w(1-P)\bigl[1-x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr] & Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[W(1-P)x_2^*\bigr]^2
 }\right].}\eqno(64)$$
 The external stability of $\underline x^*$ is determined by the eigenvalues of $\cl$, namely the roots of its characteristic polynomial $R(\lambda) =\det(\cl-\lambda\ci)$, with $\ci$ the $2\times 2$ identity matrix. From (64), $R(\lambda)=a_2\lambda^2 +a_1\lambda +a_0$, where
 $$a_0={P^2W^2w^2\over\left(\overline w^*\right)^4},\qquad a_1=-{2PWw +\left(1-P\right)^2\left[Wx_2^* +wx_1^*\right]^2\over \left(\overline w^*\right)^2},\qquad a_2=1.\eqno(65)$$

 We can prove the following result.

 \proclaim Result 7. $\cl$ has two positive eigenvalues and\hfil\break
  {\hglue.82truein}(i) when $P>\rho$ the two eigenvalues are less than 1,\hfil\break
  {\hglue.79truein}(ii) when $P<\rho$ the largest eigenvalue is larger than 1.\par

  \noindent{\it Proof}.  As $\cl$ is a positive matrix by the Perron-Frobenius theory, $\cl$ has a positive eigenvalue, and as $a_0>0$ and $a_2=1$ the product of the two eigenvalues of $\cl$ is positive. Thus in fact $\cl$ has two positive eigenvalues. Let $R(1)=R(1;P)$, then from (65)
  $$R(1;P) = {W^2w^2-\left(\overline w^*\widetilde w^*\right)^2\over\left(\overline w^*\right)^4}P^2 +2P{\left(\widetilde w^*\right)^2 -Ww\over \left(\overline w^*\right)^2} +{\left(\overline w^*\right)^2-\left(\widetilde w^*\right)^2\over\left(\overline w^*\right)^2},\eqno(66)$$
where $\widetilde w^* =Wx_2^* +wx_1^*$.

By (58) $\bigl(\sqrt{Ww}-w\bigr)/(W-w) <x_1^* <{1\over 2}$, from which it is easily seen that
$$\widetilde w^* >\overline w^*,\qquad \overline w^*>\sqrt{Ww}.\eqno(67)$$

 When $P=\rho$ one of the eigenvalues of $\cl$ is 1; hence $R(1;\rho)=0$. Another root of $R(1;P)=0$ is $\bigl[(\overline w^*)^2 +\overline w^*\widetilde w^*\bigr]/\bigl[Ww +\overline w^*\widetilde w^*\bigr]$, which by (67) is larger than 1.  As $R(1;0)=\bigl[(\overline w^*)^2 -(\widetilde w^*)^2\bigr]/(\overline w^*)^2 <0$ by (67), we deduce that when $0<P<\rho$, $R(1;P)<0$, whereas when $\rho<P<1$, $R(1;P)>0$.
 Hence, when $P<\rho$, $R(1)<0$, and since $a_2=1$, $R(+\infty)>0$, we conclude that $R(\lambda)=0$ has a positive root larger than 1 and the largest positive eigenvalue of $\cl$ is larger than 1.

 When $P>\rho$, we have $R(1)>0$ and also $R(0)=a_0>0$. As $R(\lambda)=0$ has two positive roots and as $a_2>0$, $R(\lambda)$ is convex, either the two positive roots are less than 1 or both larger than one. But the product of the two roots is $P^2W^2w^2/(\overline w^*)^2<1$ by (67), and we conclude that when $P>\rho$ the two positive eigenvalues of $\cl$ are less than 1.
 \bigskip
 \bigskip

 \noindent{\bf CONCLUSION}
 \smallskip

 An allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$. Thus with the $A1B1$ environment cycle, evolution tends to reduce vertical transmission. There is therefore a {\sl Reduction Principle for Vertical Transmission} (Altenberg et al. 2017).

 \end\bye

### 2017-07-26 ms/results.tex
revised by YR

@@
 \vfil
-\centerline{July 24, 2017, v.2}
+\centerline{July 26, 2017, v.3}
 
@@
 
-Consider an infinite population whose members are characterized by their phenotype $\phi$, which can be of two types $\phi=A$ or $\phi=B$.
-Each individual has a probability $x$ $(0\le x\le 1)$ of exhibiting the phenotype $A$ and $(1-x)$ of exhibiting $B$.
-We follow the evolution of $x$ over discrete non-overlapping generations. In each generation individuals are subject to  selection where the fitnesses of $A$ and $B$ are $w_A$ and $w_B$, respectively.
+Consider an infinite population whose members are characterized by their phenotype $\phi$, which can be $\phi=A$ or $\phi=B$, with associated frequencies $x$ and $1-x$, respectively.
+We follow the evolution of $x$ over discrete non-overlapping generations.
+In each generation individuals are subject to selection where the fitness values of $A$ and $B$ are $w_A$ and $w_B$, respectively, such that the frequency of phenotype $A$ after selection is
+$$
+x^S = {w_A\over \overline w}x,
+\eqno(0)$$
+where $\overline w$ is the {\sl mean fitness}, namely
+ $$\overline w = w_A x+w_B(1-x).\eqno(3)$$
 
-An offspring inherits its phenotype, namely its probability of exhibiting the phenotype $A$ {\sl vertically} from  parent of type $A$ with probability $\rho$ and {\sl diagonally} from a random individual in the parental population with probability $(1-\rho)$.
-Therfore, given that the parent phenotype is $\phi$ and assuming {\sl uni-parental inheritance} (Zefferman 2016), the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is
-$$P(\phi' =A|\phi) =\left\{\matrix{(1-\rho)x+\rho &\hbox{\tenrm if}\ \phi=A\cr 
+An offspring inherits its phenotype from its parent, via {\sl vertical transmission}, with probability $\rho$,
+and from a random individual in the parental population, via {\sl oblique transmission}, with probability $1-\rho$.
+Therefore, given that the parent phenotype is $\phi$ and assuming {\sl uni-parental inheritance} (Zefferman 2016), 
+the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is
+$$P(\phi' =A|\phi) =\left\{\matrix{(1-\rho)x+\rho, &\hbox{\tenrm if}\ \phi=A\cr 
 \noalign{\smallskip}
- (1-\rho)x & \hbox{\tenrm if}\ \phi =B\cr}\right.\eqno(1)$$
+(1-\rho)x, & \hbox{\tenrm if}\ \phi =B\cr}\right.\eqno(1)$$
 where $x=P(\phi=A)$ in the parent's generation before selection.
 
- Since $x$ is also the frequency of  phenotype $A$ before selection and $(1-x)$ that of $B$ before selection, after one generation $x'=P(\phi'=A)$ is given by the recursion equation
- $$x'={w_A\over w}x\bigl[(1-\rho)x +\rho\bigr] +{w_B\over w}(1-x)\bigl[(1-\rho)x\bigr],\eqno(2)$$
- where $w$ is the {\sl mean fitness}, namely
- $$w=w_A x+w_B(1-x).\eqno(3)$$
- Equation (2) can be rewritten as
- $$x'=x\cdot{x(1-\rho)(w_A-w_B) +\rho w_A +(1-\rho)w_B\over x(w_A-w_B) +w_B}.\eqno(4)$$
+Therefore, the frequency of phenotype $A$ after one generation is given by the recursion equation
+$$
+x' = \rho  {w_A\over \overline w} x + (1-\rho) x
+.\eqno(2)$$
+which can be rewritten as 
+$$
+x' = 
+x \cdot \Bigl(1 - \rho + \rho {w_A \over w_A x + w_B (1-x)}\Bigr) = 
+x \cdot \Bigl(1 + \rho (1-x) {w_A-w_B \over \overline w} \Bigr)
+.\eqno(4)$$
 
@@
 \noindent{\it Proof}.
-If we rewrite (4) as $x'=x\cdot F(x)$, it is easily seen that $F(1)=1$, and for $\rho>0$ and $0<x<1$
+If we rewrite (4) as $x'=x\cdot F(x)$, it can be seen that $F(1)=1$, and for $\rho>0$ and $0<x<1$
 $$\eqalign{
@@
   $$\eqalign{
-  &f_A(x) = {x(1-\rho)(W-w) +\rho W +(1-\rho)w\over x(W-w) +w},\cr\noalign{\smallskip}
-  &f_B(x) = {x(1-\rho)(w-W) +\rho w +(1-\rho)W\over x(w-W) +W}.}\eqno(10)$$
+  &f_A(x) = {1 + \rho (1-x) {W-w \over Wx+w(1-x)}},\cr\noalign{\smallskip}
+  &f_B(x) = {1 + \rho (1-x) {w-W \over wx+W(1-x)}}.}\eqno(10)$$
 
-  \noindent If $x_t$ denotes the frequency of exhibiting the phenotype $A$ at generation $t$ starting with $x_0$ initially, then as we are interested in the values of $x_t$ for $t=n(k+l)$ with $n=0,1,\dots$ at the end of complete periods then in fact
+  \noindent If $x_t$ denotes the frequency of phenotype $A$ at generation $t$ starting with $x_0$ initially,
+  then as we are interested in the values of $x_t$ for $t=n(k+l)$ with $n=0,1,\dots$ at the end of complete periods then in fact
   $$x_{(n+1)(k+l)}=F(x_{n(k+l)}),\quad n=0,1,2,\dots,\eqno(11)$$
@@
   Clearly, since $F_A(0) =F_B(0) =0$ and $F_A(1) =F_B(1) =1$, both fixations in $A$ or in $B$ are equilibrium points.
-An interesting question is when these fixations are locally stable.
+An interesting question is when these fixations are locally stable (.
 We concentrate on $x^*=0$, the fixation of the phenotype $B$.
@@
    Observe that
-   $$\eqalign{
-   \left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right) &= \left(1-\rho+\rho{W\over w}\right)\left(1-\rho+\rho{w\over W}\right)\cr
-   &= \left(1-\rho\right)^2+ \rho^2 +\rho(1-\rho){W^2+w^2\over Ww}.}\eqno(16)$$
-   Thus,
    $$\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right) =1+\rho(1-\rho){\left(W-w\right)^2\over wW}>1.\eqno(17)$$
-   Since $0<\rho<1$ and $0<w<W$,  the $B$ fixation is unstable.
+   Since $0<\rho<1$ and $0<w<W$,  fixation of $B$ is unstable.
   \bigskip
@@
 
-\proclaim Result 3. With $0<\rho<1$; $0<w<W$ in the case of $AkBl$ periodically changing environments, both fixations may be unstable, producing a protected polymorphism.\par
+\proclaim Result 3. With $0<\rho<1$ and $0<w<W$ in the case of $AkBl$ periodically changing environments, both fixations may be unstable, producing a protected polymorphism.\par
 
 \noindent{\it Proof}.
-Let $a=1+\rho{W-w\over w}$ and $b=1+\rho{w-W\over w}$.
+Let $a=1+\rho{W-w\over w}$ and $b=1+\rho{w-W\over W}$.
 Then since $0<\rho <1$ and $0<w<W$, we have $a>1$ and $0<b<1$.
@@
 $$k\,{\log(1/b)\over\log a} < l < k\,{\log a\over\log(1/b)}.\eqno(23)$$ 
-Inequalities (23) are consistent if and only if $\log(1/b)<\log a$, i.e., $ab>1$. But by (17),
-$$ab=\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over w}\right)>1.\eqno(24)$$
+Inequalities (23) are consistent if and only if $\log(1/b)<\log a$, i.e., $ab>1$.
+Indeed, by (17),
+$$ab=\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right)>1.\eqno(24)$$
 
@@
 $$\eqalign{
-&F_A(x)=x{x(1-\rho)(W-w) +\rho W+(1-\rho)w\over x(W-w) +w},\cr\noalign{\smallskip}
-&F_B(y) = y{y(1-\rho)(w-W) +\rho w+(1-\rho)W\over y(w-W)+W}.}\eqno(25)$$
+&F_A(x) = x \Bigl(1 + \rho (1-x){W-w \over Wx+w(1-x)}\Bigr),\cr\noalign{\smallskip}
+&F_B(y) = y \Bigl(1 + \rho (1-y){w-W \over wy+W(1-y)}\Bigr).}\eqno(25)$$
 The equilibrium equation is $x=F_B\bigl(F_A(x)\bigr)$, which reduces to a fourth degree polynomial equation in $x$.
 Since the two fixations in $B$ and $A$ are equilibria corresponding to the two solutions $x=0$ and $x=1$, the other equilibria correspond to solutions of a quadratic equation $G(x) =Ax^2 +Bx +C=0$ with
-$$A=1,\quad B=-{W(1-\rho)-w(3-\rho)\over (2-\rho)(W-w)},\quad C={-w\over (2-\rho)(W-w)}.\eqno(26)$$
+$$A=1,\quad B={W+w\over (2-\rho)(W-w)}-1,\quad C={-w\over (2-\rho)(W-w)}.\eqno(26)$$
 As $0<\rho<1$ and $0<w<W$ we have
-$$G(0) = {-w\over (2-\rho)(W-w)}<0\eqno(27)$$
+$$G(0) = {-w\over (2-\rho)(W-w)}<0,\eqno(27)$$
 and
-$$G(1)=A +B+ C =1-{W(1-\rho) -w(3-\rho)\over (2-\rho)(W-w)} -{w\over (2-\rho)(W-w)},\eqno(28)$$
-or
-$$G(1) ={W\over (2-\rho)(W-w)} >0.\eqno(29)$$
-Also, as $A=1$, $G(x)\to\infty$ when $x\to\pm\infty$, the quadratic equation $G(x)=0$ has two real roots, one negative and one positive $x^*$ satisfying $0<x^*<1$.
-The latter determines a unique polymorphism.
-Let $H(x) =F_B\bigl(F_A(x)\bigr)$.
-Then
+$$G(1)=A +B+ C =
+{W\over (2-\rho)(W-w)} >0,\eqno(29)$$
+and $G(x)$ must have two real roots.
+Also, since $A=1$, $G(x)$ is convex, and therefore there is one negative root, and one positive root $x^*$ which satisfies $0<x^*<1$.
+The latter root determines a unique polymorphism.
+
+Let $H(x) =F_B\bigl(F_A(x)\bigr)$. Then 
  $$H(0)=0,\quad H(x^*)=x^*,\quad H(1)=1.\eqno(30)$$
@@
 Without loss of generality, assume that the fitness parameters at generation $t$ ($t=0,1,2,\dots$) are $1+s_t$ for phenotype $A$ and $1$ for phenotype $B$, such that the random variables $s_t$ for $t=0,1,2,\dots$ are independent and identically distributed.
-Also we assume that there are positive constants $C$ and $D$ such that $P(-1+C<s_t <D)=1$.
+Also we assume that there are positive constants $C$ and $D$ such that $P(C<s_t <D)=1$.
 
@@
  \noindent{\it Definition}: ``stochastic local stability''.
- A constant equilibrium state $p^*$ is said to be {\sl stochastically locally stable} if for any $\var>0$ there exists a $\delta>0$ such that $|x_0-p^*|<\delta$ implies
- $$P\left(\lim_{t\to\infty}x_t =p^*\right)\ge 1-\var.\eqno(35)$$
- Thus stochastic local stability holds for $p^*$ provided for any initial $x_0$ sufficiently near $p^*$ the process $x_t$ converges to $p^*$ with high probability.
+ A constant equilibrium state $x^*$ is said to be {\sl stochastically locally stable} if for any $\var>0$ there exists a $\delta>0$ such that $|x_0-x^*|<\delta$ implies
+ $$P\left(\lim_{t\to\infty}x_t =x^*\right)\ge 1-\var.\eqno(35)$$
+ Thus stochastic local stability holds for $x^*$ provided for any initial $x_0$ sufficiently near $x^*$ the process $x_t$ converges to $x^*$ with high probability.
 
@@
  \noindent{\it Remark}. 
- Using the general notation for the fitness parameters $w_A$ and $w_B$, the stochastic local stability of the $B$ fixation is determined by the sign of $E\left[\log\left(1-\rho+\rho{w_A\over w_B}\right)\right]$ and that of the $A$ fixation by the sign of $E\left[\log\left(1-\rho+\rho{w_B\over w_A}\right)\right]$.
+ Using the general notation for the fitness parameters $w_A$ and $w_B$, the stochastic local stability of the $B$ fixation is determined by the sign of $E\left[\log\left(1+\rho{w_A-w_B\over w_B}\right)\right]$ and that of the $A$ fixation by the sign of $E\left[\log\left(1+\rho{w_B-w_A\over w_A}\right)\right]$.
 When one of these signs is negative, there is stochastic local stability, and when it is positive, with probability one convergence to that fixation does not occur.
-Following (16),
- $$\left(1-\rho +\rho{w_A\over w_B}\right)\left(1-\rho+\rho{w_B\over w_A}\right) >1.\eqno(52)$$
+Following (17),
+ $$\left(1+\rho{w_A-w_B\over w_B}\right)\left(1+\rho{w_B-w_A\over w_A}\right) >1.\eqno(52)$$
  Therefore, as in the case of periodically changing environments $AkBl$, it is impossible that the two fixations are both stochastically locally stable.
@@
 Let the vertical transmission rates determined by $m$ and $M$ be $\rho$ and $P$, respectively.
-Thus there are four pheno-genotypes $mA$, $mB$, $MA$, $MB$ whose frequencies at a given generation are denoted by $x_1$, $x_2$, $x_3$, $x_4$, respectively.
-As the fitnesses are determined by the two phenotypes $A$ and $B$, and the modifier locus is selectively neutral, we have the following table.
+Thus there are four pheno-genotypes $mA$, $mB$, $MA$, and $MB$ whose frequencies at a given generation are denoted by $x_1$, $x_2$, $x_3$, and $x_4$, respectively.
+As the fitness values are determined by the two phenotypes $A$ and $B$, and the modifier locus is selectively neutral, we have the following table.
  $$\matrix{
@@
  \hbox{\tenrm vertical transmission rate}&\rho&\rho&P&P\cr}\eqno(53)$$
- Following the rationale leading to equation (2), the next generation pheno-genotype frequencies $x'_1$, $x'_2$, $x'_3$, $x'_4$ are
+ Following the rationale leading to equation (2), the next generation pheno-genotype frequencies $x'_1$, $x'_2$, $x'_3$, and $x'_4$ are
  $$\eqalign{
- \overline wx'_1 &= w_Ax_1\bigl[(1-\rho)(x_1+x_3) +\rho\bigr] +w_Bx_2(1-\rho)(x_1+x_3)\cr
- \overline wx'_2 &= w_BAx_1(1-\rho)(x_2+x_4) +w_Bx_2\bigl[(1-\rho)(x_2+x_4)+\rho\bigr]\cr 
-  \overline wx'_3 &= w_Ax_3\bigl[(1-P)(x_1+x_3) +P\bigr] +w_Bx_4(1-P)(x_1+x_3)\cr
-\overline wx'_4 &= w_Bx_3(1-P)(x_2+x_4) +w_Bx_4\bigl[(1-P)(x_2+x_4)+P\bigr], 
+ x'_1 &= \rho {w_A \over \overline w} x_1 + (1-\rho)(x_1+x_3)\cr
+ x'_2 &= \rho {w_B \over \overline w} x_1 + (1-\rho)(x_2+x_4)\cr 
+ x'_3 &= P {w_A \over \overline w} x_3 + (1-P)(x_1+x_3)\cr
+ x'_4 &= P {w_B \over \overline w} x_4 + (1-P)(x_2+x_4), 
 }\eqno(54)$$
@@
 $$\overline w =w_A(x_1+x_3) +w_B(x_2+x_4).\eqno(55)$$
-Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} to invasion by allele $M$.
+Starting with a stable equilibrium where only the modifier allele $m$ is present, we check its {\sl external stability} to invasion by the modifier allele $M$.
 
@@
 $$x_1^* ={1\over 2} -{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}\over 2\cdot(2-\rho)(W-w)},\eqno(57)$$
- and it is easily seen that
+ and it can be seen that
  $${\sqrt{Ww}-w\over W-w}<x_1^* <{1\over 2}.\eqno(58)$$
@@
 $$\overline w^* =Wx_1^* +wx_2^*,\qquad \overline w^{**} =wx_1^{**} +Wx_2^{**}.\eqno(61)$$
-Due to the symmetry between the two phenotypes $A$ and $B$ in the $A1B1$ case, we have $x_1^{**} =x_2^*$ and $x_2^{**} =x_1^*$ so that $\overline w^{**} =\overline w^*$ and in fact
+Due to the symmetry between the two phenotypes $A$ and $B$ in the $A1B1$ case, we have $x_1^{**} =1-x_1^*=x_2^*$ and $x_2^{**} =1-x_2^*=x_1^*$ so that $\overline w^{**} =\overline w^*$ and in fact $\cl_2$ can be written in terms of $x_1^*$ and $x_2^*$:
 $$\overline w^*\cl_2 = \left[\matrix{w\bigl[(1-P)x_2^* +P\bigr] & W(1-P)x_2^*\cr\noalign{\medskip}
@@
  Observe that as $\underline x^* =T_2(T_1\underline x^*)$ with $x_3^*=x_4^*=0$, from (59) and (60) with $P=\rho$ we have
- $$\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right]= \cl_2\cdot\cl_1\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right] =\cl\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right].\eqno(63)$$
- Hence when $P=\rho$ one of the eigenvalues of $\cl$ is 1.
+ $$\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right]=\cl\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right].\eqno(63)$$
+ Hence when $P=\rho$, one of the eigenvalues of $\cl$ is 1.
  \smallskip
 
- In general $\cl=\cl_2\cdot\cl_1$, and using (59) and (62) we get
+ In general, using (59) and (62) we get
  $$\eqalign{
@@
 
-  \noindent{\it Proof}.
-As $\cl$ is a positive matrix by the Perron-Frobenius theory, $\cl$ has a positive eigenvalue, and as $a_0>0$ and $a_2=1$ the product of the two eigenvalues of $\cl$ is positive.
-Thus in fact $\cl$ has two positive eigenvalues.
-Let $R(1)=R(1;P)$, then from (65)
-  $$R(1;P) = {W^2w^2-\left(\overline w^*\widetilde w^*\right)^2\over\left(\overline w^*\right)^4}P^2 +2P{\left(\widetilde w^*\right)^2 -Ww\over \left(\overline w^*\right)^2} +{\left(\overline w^*\right)^2-\left(\widetilde w^*\right)^2\over\left(\overline w^*\right)^2},\eqno(66)$$
+  \noindent{\it Proof}.  As $\cl$ is a positive matrix, by the Perron-Frobenius theory, $\cl$ has a positive eigenvalue.
+  Also, the product of the two eigenvalues of $\cl$  is $a_0>0$.
+  Thus $\cl$ has two positive eigenvalues.
+  In addition, $R(\lambda)$ is convex because $a_2=1$.
+
+  Let $R(1)=R(1;P)$ be a polynomial in $P$, then from (65)
+  $$R(1;P) = -{\left(\overline w^*\widetilde w^*\right)^2 - \left(Ww\right)^2 \over\left(\overline w^*\right)^4}P^2 +
+  2P{\left(\widetilde w^*\right)^2 -Ww\over \left(\overline w^*\right)^2} -
+  {\left(\widetilde w^*\right)^2 - \left(\overline w^*\right)^2 \over\left(\overline w^*\right)^2},\eqno(66)$$
 where $\widetilde w^* =Wx_2^* +wx_1^*$.  
+By (58), $\bigl(\sqrt{Ww}-w\bigr)/(W-w) <x_1^* <{1\over 2}$, from which it can be seen that
+$$\widetilde w^* >\overline w^* >\sqrt{Ww}.\eqno(67)$$
 
-By (58) $\bigl(\sqrt{Ww}-w\bigr)/(W-w) <x_1^* <{1\over 2}$, from which it is easily seen that
-$$\widetilde w^* >\overline w^*,\qquad \overline w^*>\sqrt{Ww}.\eqno(67)$$
+ When $P=\rho$, one of the eigenvalues of $\cl$ is 1; hence $R(1;\rho)=0$.
+ Another root of $R(1;P)$ is $\bigl[(\overline w^*)^2 +\overline w^*\widetilde w^*\bigr]/\bigl[Ww +\overline w^*\widetilde w^*\bigr]$, which by (67) is larger than 1.
+ As $R(1;0)= - \bigl[(\widetilde w^*)^2 - (\overline w^*)^2\bigr]/(\overline w^*)^2 <0$ by (67), and we deduce that when $0<P<\rho$, $R(1;P)<0$, whereas when $\rho<P<1$, $R(1;P)>0$.
 
- When $P=\rho$ one of the eigenvalues of $\cl$ is 1; hence $R(1;\rho)=0$.
-Another root of $R(1;P)=0$ is $\bigl[(\overline w^*)^2 +\overline w^*\widetilde w^*\bigr]/\bigl[Ww +\overline w^*\widetilde w^*\bigr]$, which by (67) is larger than 1.
-As $R(1;0)=\bigl[(\overline w^*)^2 -(\widetilde w^*)^2\bigr]/(\overline w^*)^2 <0$ by (67), we deduce that when $0<P<\rho$, $R(1;P)<0$, whereas when $\rho<P<1$, $R(1;P)>0$.
- Hence, when $P<\rho$, $R(1)<0$, and since $a_2=1$, $R(+\infty)>0$, we conclude that $R(\lambda)=0$ has a positive root larger than 1 and the largest positive eigenvalue of $\cl$ is larger than 1.
+ Hence, when $P<\rho$, we have $R(1)<0$ and $R(0)=a_0>0$, and we conclude that $R(\lambda)$ has a positive root larger than 1 and the therefore largest eigenvalue of $\cl$ is larger than 1.
 
- When $P>\rho$, we have $R(1)>0$ and also $R(0)=a_0>0$.
-As $R(\lambda)=0$ has two positive roots and as $a_2>0$, $R(\lambda)$ is convex, either the two positive roots are less than 1 or both larger than one.
-But the product of the two roots is $P^2W^2w^2/(\overline w^*)^2<1$ by (67), and we conclude that when $P>\rho$ the two positive eigenvalues of $\cl$ are less than 1.
+ When $P>\rho$, we have $R(1)>0$ and $R(0)>0$. Also, because $R(\lambda)$ is convex and has two positive roots, these roots are either both less than 1 or both larger than 1.
+ But the product of the two roots is $a_0=P^2{(Ww)^2 \over (\overline w^*)^4}<1$ by (67), and we conclude that when $P>\rho$, both roots are less than 1 and therefore the largest eigenvalue of $\cl$ is less than 1.
  \bigskip
@@
 
- \noindent{\bf CONCLUSION}
+ \proclaim Conclusion.
  \smallskip
 
- An allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
+ A modifier allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
 Thus with the $A1B1$ environment cycle, evolution tends to reduce vertical transmission.

### 2017-07-27 ms/results.tex
fix modifier model

@@
 \vfil
-\centerline{July 26, 2017, v.3}
+\centerline{July 27, 2017, v.1}
 
@@
  \hbox{\tenrm vertical transmission rate}&\rho&\rho&P&P\cr}\eqno(53)$$
- Following the rationale leading to equation (2), the next generation pheno-genotype frequencies $x'_1$, $x'_2$, $x'_3$, and $x'_4$ are
+ Following the rationale leading to equation (2), the pheno-genotype frequencies after selection are
  $$\eqalign{
- x'_1 &= \rho {w_A \over \overline w} x_1 + (1-\rho)(x_1+x_3)\cr
- x'_2 &= \rho {w_B \over \overline w} x_1 + (1-\rho)(x_2+x_4)\cr 
- x'_3 &= P {w_A \over \overline w} x_3 + (1-P)(x_1+x_3)\cr
- x'_4 &= P {w_B \over \overline w} x_4 + (1-P)(x_2+x_4), 
+ x_1^S = {w_A \over \overline w}x_1\cr
+ x_2^S = {w_B \over \overline w}x_2\cr
+ x_3^S = {w_A \over \overline w}x_3\cr
+x_4^S = {w_B \over \overline w}x_4,
 }\eqno(54)$$
-with $\overline w$, the mean fitness, given by
-$$\overline w =w_A(x_1+x_3) +w_B(x_2+x_4).\eqno(55)$$
+with the mean fitness $\overline w$ given by $\overline w=\sum_{i=1}^{4}{x_i^S}$,
+and the frequencies of the phone-genotypes at the next generations are
+ $$\eqalign{
+ x'_1 &= \rho x_1^S + (1-\rho)(x_1^S + x_2^S)(x_1+x_3)\cr
+ x'_2 &= \rho x_2^S + (1-\rho)(x_1^S + x_2^S)(x_2+x_4)\cr 
+ x'_3 &= P x_3^S + (1-P)(x_3^S+x_4^S)(x_1+x_3)\cr
+ x'_4 &= P x_4^S + (1-P)(x_3^S+x_4^S)(x_2+x_4), 
+}\eqno(55)$$
+
 Starting with a stable equilibrium where only the modifier allele $m$ is present, we check its {\sl external stability} to invasion by the modifier allele $M$.
@@
 We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$.
-Specifically, following (54) with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
+Specifically, following (54-55) with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
 $$\underline x''=T_2(T_1\underline x),\eqno(56)$$
-where $\underline x'=T_1\underline x$ is given by (54) with $w_A=W$, $w_B=w$, and $\underline x''=T_2\underline x'$ is given by (54) with $w_A=w$, $w_B=W$.
+where $\underline x'=T_1\underline x$ is given by (54-55) with $w_A=W$, $w_B=w$, and $\underline x''=T_2\underline x'$ is given by (54-55) with $w_A=w$, $w_B=W$.
 Here $\underline x$, $\underline x'$, $\underline x''$ are the frequency vectors.
@@
 
- The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ derived from (54) and given by
+ The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ derived from (54-55) and given by
  $$\overline w^*\cl_1 =\left[\matrix{W\bigl[(1-P)x_1^* +P\bigr] & w(1-P)x_1^*\cr\noalign{\medskip}

### 2017-07-30 ms/results.tex
old version of eq 2 in 4, of 54-55

@@
 \vfil
-\centerline{July 27, 2017, v.1}
+\centerline{July 27, 2017, v.2}
 
@@
 
-Therefore, the frequency of phenotype $A$ after one generation is given by the recursion equation
+Therefore, the change in frequency of phenotype $A$ after one generation due to selection ($ {w_A\over \overline w}$) and vertical ($\rho$) and oblique ($1-\rho$) transmission is given by the recursion equation
 $$
 x' = \rho  {w_A\over \overline w} x + (1-\rho) x
-.\eqno(2)$$
+,\eqno(2)$$
 which can be rewritten as 
-$$
+$$\eqalign{
 x' = 
-x \cdot \Bigl(1 - \rho + \rho {w_A \over w_A x + w_B (1-x)}\Bigr) = 
-x \cdot \Bigl(1 + \rho (1-x) {w_A-w_B \over \overline w} \Bigr)
-.\eqno(4)$$
+\bigl(1-x\bigr) {w_A\over \overline w} \bigl((1-\rho)x + \rho\bigr) + \bigl(1-x\bigr) {w_B\over \overline w} \bigl(1-\rho\bigr)x = \cr
+x \cdot \Bigl(1 - \rho + \rho {w_A \over \overline w}\Bigr) = \cr
+x \cdot \Bigl(1 + \rho \bigl(1-x\bigr) {w_A-w_B \over \overline w} \Bigr)
+.}\eqno(4)$$
 
@@
  \hbox{\tenrm vertical transmission rate}&\rho&\rho&P&P\cr}\eqno(53)$$
- Following the rationale leading to equation (2), the pheno-genotype frequencies after selection are
+ Following the rationale leading to equation (2), the pheno-genotype frequencies after one generations are described by
  $$\eqalign{
- x_1^S = {w_A \over \overline w}x_1\cr
- x_2^S = {w_B \over \overline w}x_2\cr
- x_3^S = {w_A \over \overline w}x_3\cr
-x_4^S = {w_B \over \overline w}x_4,
+ \overline w x_1' = w_A x_1\bigr[(1-\rho)(x_1+x_3)+\rho \bigl] + w_B x_2(1-\rho)(x_1+x_3)\cr
+ \overline w x_2' = w_A x_1(1-\rho)(x_2+x_4) + w_B x_2 \bigl[(1-\rho)(x_2+x_4) + \rho \bigr]\cr
+ \overline w x_3' = w_A x_2\bigr[(1-P)(x_1+x_3)+P \bigl] + w_B x_4(1-\rho)(x_1+x_3)\cr
+ \overline w x_2' = w_A x_3(1-P)(x_2+x_4) + w_B x_4 \bigl[(1-P)(x_2+x_4) + P \bigr]
 }\eqno(54)$$
-with the mean fitness $\overline w$ given by $\overline w=\sum_{i=1}^{4}{x_i^S}$,
-and the frequencies of the phone-genotypes at the next generations are
- $$\eqalign{
- x'_1 &= \rho x_1^S + (1-\rho)(x_1^S + x_2^S)(x_1+x_3)\cr
- x'_2 &= \rho x_2^S + (1-\rho)(x_1^S + x_2^S)(x_2+x_4)\cr 
- x'_3 &= P x_3^S + (1-P)(x_3^S+x_4^S)(x_1+x_3)\cr
- x'_4 &= P x_4^S + (1-P)(x_3^S+x_4^S)(x_2+x_4), 
-}\eqno(55)$$
+with the mean fitness $\overline w$ given by
+$$\overline w=w_A (x_1+x_3) + w_B (x_2+x_4).\eqno(55)$$
 

### 2017-08-15 ms/figures.tex
md->tex; revised with figures

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{lineno}
\linenumbers

\begin{document}

\begin{figure}
\centering
\includegraphics{../figures/lk_phase_plane.pdf}
\caption{Ratios of selection periods \(\frac{k}{l}\) that lead to
fixation of phenotype \emph{A} (red) or polymorphism of phenotypes
\emph{A} and \emph{B} (blue). \emph{k} and \emph{l} are the number of
generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
The boundary is set at $-\log{(1+\rho \frac{s}{1-s}}) / \log{(1-\rho s})$, see eq. 23.
In all cases, \emph{W} = 1, \emph{w} = \emph{1-s}.}\label{fig:lk_phase_plane}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/env_A1B1.pdf}
\caption{Frequency of phenotype \emph{A} after every two generations in
selection regime \emph{A1B1}. The orange line is the finite population model
(average of 100 simulations). The blue line is the infinite population model
(eq. 25), and the green line is the equilibrium solution (eq. 56). Parameters: \emph{W}=1,
\emph{N}=10,000, initial value \(x=0.5\).}\label{fig:env_A1B1}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/A1B1_modifier_invasions.pdf}
\caption{Consecutive fixation of modifiers that reduce the vertical
transmission rate in environmental regime \emph{A1B1}. The figure shows
results of numerical simulations of evolution with two modifier alleles
(eq. 53). When a modifier allele fixes
(frequency\textgreater{}99.9\%), a new modifier allele is introduced
with a vertical transmission rate one order of magnitude lower (vertical
dashed lines). \textbf{(A,D,G)} The frequency of phenotype \emph{A} in
the population over time. \textbf{(B,E,H)} The frequency of the invading
modifier allele over time. \textbf{(C,F,I)} The population mean
fitness over time; insets zoom in to show that the mean fitness 
decreases slightly with each invasion. Parameters: vertical transmission rate of
the initial resident modifier allele, \(\rho_0 =0.1\); fitness values:
\emph{W}=1, \emph{w}=0.1 (\textbf{A-C}), 0.5 (\textbf{D-F}), and 0.9
(\textbf{G-I}). The x-axis is on a log-scale, as each sequential invasion
takes an order of magnitude longer to
complete.}\label{fig:A1B1_modifier_invasions}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/AkBl_equilibrium.pdf}
\caption{Stability in a fluctuating environment.
\textbf{(A)} Stable frequency of phenotype $A$ after every two generations and \textbf{(B)} stable mean fitness in selection regime \emph{A1B1} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
\textbf{(C)} Stable population mean fitness in selection regime \emph{AkBl} as a
function of the vertical transmission rate \(\rho\) and different ratios of selection periods \(k=l\). \emph{k} and \emph{l} are the number of generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
The geometric average of the population mean fitness over $k+l$ generations is shown. 
Here \emph{w=0.9}.
In all cases, \emph{W=1}.
}\label{fig:AkBl_equilibrium}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/env_A1B2.pdf}
\caption{Frequency of phenotype \emph{A} after every three generation in
selection regime \emph{A1B2}. Comparison of dynamics starting with
different initial frequency of phenotype \(A\) (0.01-0.99).
See also Fig. 1 and 6. Parameters: \emph{W}=1, \emph{N=10,000},
initial population \(x=0.5\).}\label{fig:env_A1B2}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/A1B2_polymorphism.pdf}
\caption{Stable frequency of phenotype \emph{A} and population mean fitness in selection regime
\emph{A1B2} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
In the grey area \(B\) reached fixation and the frequency of \(A\) is 0.
\textbf{(A)} Stable frequency of phenotype $A$ at the end of each period of three generations.
\textbf{(B)} Geometric average of the stable mean fitness over the three generations period: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
In all cases, \emph{W=1}.}\label{fig:A1B2_polymorphism}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/stochastic_env_x_t.pdf}
\caption{Stochastic local stability. The figure shows the frequency of
phenotype \emph{A} after \(10^6\) generations in a very large populationt 
evolving in a stochastic environment (eq. 33). The
fitness of phenotypes \emph{A} and \emph{B} are \(1+s_t\) and \(1\),
where \(s_t\) is \emph{s} with probability \emph{p} and \emph{-s} with
probability \emph{1-p}. The white line marks combinations of \emph{p}
and \emph{s} for which \(\mathbb{E}[\log{(1+\rho s_t)}]=0\); according
to our analysis, we expect that below this line \(x^*=0\) will be
stochastically locally stable. Parameters: \(x_0=0.1\);
\(\rho=0.1\).}\label{fig:stochastic_env_x_t}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=\dimexpr \textheight - 8\baselineskip\relax]{../figures/stoch_modifier_invasions.png}
\caption{Consecutive fixation of modifiers that reduce the vertical
transmission rate under randomly changing selection. The figure shows
results of numerical simulations of evolution with two modifier alleles
(eq. 33). The
fitness of phenotypes \emph{A} and \emph{B} are 1 and 0.1 with probability 0.5 or 0.1 and 1 otherwise.
When a modifier allele fixes
(frequency\textgreater{}99.9\%), a new modifier allele is introduced
with a vertical transmission rate one order of magnitude lower (vertical
dashed lines). \textbf{(A)} The frequency of phenotype \emph{A} in the
population over time. \textbf{(B)} The frequency of the invading
modifier allele over time. \textbf{(C)} The population mean fitness over time. Parameters: vertical transmission rate of the initial
resident modifier allele, \(\rho_0 =0.1\); fitness values: \emph{W=1}
and \emph{w=0.1} with probability 0.5 and \emph{w=0.1} and \emph{W=1}
otherwise. The x-axis is on a log-scale, as each sequential invasion takes
an order of magnitude longer to
complete.}\label{fig:stoch_modifier_invasions}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/stoch_two_iid_rvs.png}
\caption{Effect of vertical transmission rate on phenotype polymorphism
in randomly changing environments. Dynamics of the frequency of phenotype
\(A\) over time starting
at \(x_0=10^{-5}\) when the fitness of phenotypes \(A\) and \(B\) are
identically and independently distributed random variable. As the vertical transmission rate \(\rho\) increases
from 0.001 to 0.5 the frequency reaches a polymorphic distribution with \(E(X_t/N)\to0.5\) faster, but
the variance also increases. \(w_A\) and \(w_B\) are both exponential
random variables with expected value 2.}\label{fig:stoch_two_iid_rvs}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/fixation_prob_time.pdf}
\caption{Fixation probability and time in a finite population.
\textbf{(A)} Fixation probability \(u(x)\) of phenotype \(A\) (eq. 88), and \textbf{(B)}
Expected time to fixation \(T(x)\) of phenotype \(A\) (eq. 90), conditioned on its
fixation, starting with a single copy in a population of size \(N\). The
figure compares two estimates: simulations (blue circles) and diffusion
equation approximation (green solid line). Parameters: Selection coefficient, \(s=w_A-w_B=0.1\);
Population size, \(N=10,000\).}\label{fixation_prob_time}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/lk_fix_prob.pdf}
\caption{Fixation in a finite population with different ratios of selection periods \(\frac{k}{l}\). Fixation probability of phenotype $A$ when starting with a single copy in a population of size $N$: $u(1/N) = (1-\exp(-2 \rho \frac{k-l}{k+l}(W-w))/(1-\exp(-2 N \rho \frac{k-l}{k+l}(W-w))$ (see eqs. 115-116).
\emph{k} and \emph{l} are the number of
generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by
selection. In all cases, \emph{W} = 1, \emph{w} = \emph{1-s}, Population size, \(N=10,000\).} \label{lk_fix_prob}
\end{figure}

\end{document}

### 2017-08-24 ms/figures.tex
figure captions aug 16

@@
 \centering
+\includegraphics{../figures/AkBl_equilibrium.pdf}
+\caption{Stability in a fluctuating environment.
+\textbf{(A)} Stable frequency of phenotype $A$ in the \emph{A1B1} selection regime and \textbf{(B)} stable mean fitness in selection regime \emph{A1B1} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
+\textbf{(C)} Stable population mean fitness in selection regime \emph{AkBl} as a
+function of the vertical transmission rate \(\rho\) and different ratios of selection periods \(k=l\).
+\emph{k} and \emph{l} are the number of generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
+The geometric average of the population mean fitness over $k+l$ generations is shown. 
+Here \emph{w=0.9}.
+In all cases, \emph{W=1}.
+}\label{fig:AkBl_equilibrium}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/stochastic_env_x_t.pdf}
+\caption{Stochastic local stability. The figure shows the frequency of
+phenotype \emph{A} after \(10^6\) generations in a very large populationt 
+evolving in a stochastic environment (eq. 33). The
+fitness of phenotypes \emph{A} and \emph{B} are \(1+s_t\) and \(1\),
+where \(s_t\) is \emph{s} with probability \emph{p} and \emph{-s} with
+probability \emph{1-p}. The white line marks combinations of \emph{p}
+and \emph{s} for which \(\mathbb{E}[\log{(1+\rho s_t)}]=0\); according
+to our analysis, we expect that below this line \(x^*=0\) will be
+stochastically locally stable. Parameters: \(x_0=0.1\);
+\(\rho=0.1\).}\label{fig:stochastic_env_x_t}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/stoch_two_iid_rvs.png}
+\caption{Effect of vertical transmission rate on phenotype polymorphism
+in randomly changing environments. Dynamics of the frequency of phenotype
+\(A\) over time starting
+at \(x_0=10^{-5}\) when the fitness of phenotypes \(A\) and \(B\) are
+identically and independently distributed random variable. As the vertical transmission rate \(\rho\) increases
+from 0.001 to 0.5 the frequency reaches a polymorphic distribution with \(E(X_t/N)\to0.5\) faster, but
+the variance also increases. \(w_A\) and \(w_B\) are both exponential
+random variables with expected value 2.}\label{fig:stoch_two_iid_rvs}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics{../figures/A1B1_modifier_invasions.pdf}
@@
 complete.}\label{fig:A1B1_modifier_invasions}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/AkBl_equilibrium.pdf}
-\caption{Stability in a fluctuating environment.
-\textbf{(A)} Stable frequency of phenotype $A$ after every two generations and \textbf{(B)} stable mean fitness in selection regime \emph{A1B1} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
-\textbf{(C)} Stable population mean fitness in selection regime \emph{AkBl} as a
-function of the vertical transmission rate \(\rho\) and different ratios of selection periods \(k=l\).
-\emph{k} and \emph{l} are the number of generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
-The geometric average of the population mean fitness over $k+l$ generations is shown. 
-Here \emph{w=0.9}.
-In all cases, \emph{W=1}.
-}\label{fig:AkBl_equilibrium}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/env_A1B2.pdf}
-\caption{Frequency of phenotype \emph{A} after every three generation in
-selection regime \emph{A1B2}. Comparison of dynamics starting with
-different initial frequency of phenotype \(A\) (0.01-0.99).
-See also Fig. 1 and 6. Parameters: \emph{W}=1, \emph{N=10,000},
-initial population \(x=0.5\).}\label{fig:env_A1B2}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/A1B2_polymorphism.pdf}
-\caption{Stable frequency of phenotype \emph{A} and population mean fitness in selection regime
-\emph{A1B2} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
-In the grey area \(B\) reached fixation and the frequency of \(A\) is 0.
-\textbf{(A)} Stable frequency of phenotype $A$ at the end of each period of three generations.
-\textbf{(B)} Geometric average of the stable mean fitness over the three generations period: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
-In all cases, \emph{W=1}.}\label{fig:A1B2_polymorphism}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/stochastic_env_x_t.pdf}
-\caption{Stochastic local stability. The figure shows the frequency of
-phenotype \emph{A} after \(10^6\) generations in a very large populationt 
-evolving in a stochastic environment (eq. 33). The
-fitness of phenotypes \emph{A} and \emph{B} are \(1+s_t\) and \(1\),
-where \(s_t\) is \emph{s} with probability \emph{p} and \emph{-s} with
-probability \emph{1-p}. The white line marks combinations of \emph{p}
-and \emph{s} for which \(\mathbb{E}[\log{(1+\rho s_t)}]=0\); according
-to our analysis, we expect that below this line \(x^*=0\) will be
-stochastically locally stable. Parameters: \(x_0=0.1\);
-\(\rho=0.1\).}\label{fig:stochastic_env_x_t}
 \end{figure}
@@
 \centering
-\includegraphics{../figures/stoch_two_iid_rvs.png}
-\caption{Effect of vertical transmission rate on phenotype polymorphism
-in randomly changing environments. Dynamics of the frequency of phenotype
-\(A\) over time starting
-at \(x_0=10^{-5}\) when the fitness of phenotypes \(A\) and \(B\) are
-identically and independently distributed random variable. As the vertical transmission rate \(\rho\) increases
-from 0.001 to 0.5 the frequency reaches a polymorphic distribution with \(E(X_t/N)\to0.5\) faster, but
-the variance also increases. \(w_A\) and \(w_B\) are both exponential
-random variables with expected value 2.}\label{fig:stoch_two_iid_rvs}
+\includegraphics{../figures/A1B2_polymorphism.pdf}
+\caption{Stable frequency of phenotype \emph{A} and population mean fitness in selection regime
+\emph{A1B2} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
+In the grey area \(B\) reached fixation and the frequency of \(A\) is 0.
+\textbf{(A)} Stable frequency of phenotype $A$ at the end of each period of three generations.
+\textbf{(B)} Geometric average of the stable mean fitness over the three generations period: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
+In all cases, \emph{W=1}.}\label{fig:A1B2_polymorphism}
 \end{figure}

### 2017-08-24 ms/figures_supp.tex
figure captions aug 16

--- 2017-08-15 ms/figures.tex
@@
 \centering
-\includegraphics{../figures/lk_phase_plane.pdf}
-\caption{Ratios of selection periods \(\frac{k}{l}\) that lead to
-fixation of phenotype \emph{A} (red) or polymorphism of phenotypes
-\emph{A} and \emph{B} (blue). \emph{k} and \emph{l} are the number of
-generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
-The boundary is set at $-\log{(1+\rho \frac{s}{1-s}}) / \log{(1-\rho s})$, see eq. 23.
-In all cases, \emph{W} = 1, \emph{w} = \emph{1-s}.}\label{fig:lk_phase_plane}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/env_A1B1.pdf}
-\caption{Frequency of phenotype \emph{A} after every two generations in
-selection regime \emph{A1B1}. The orange line is the finite population model
-(average of 100 simulations). The blue line is the infinite population model
-(eq. 25), and the green line is the equilibrium solution (eq. 56). Parameters: \emph{W}=1,
-\emph{N}=10,000, initial value \(x=0.5\).}\label{fig:env_A1B1}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/A1B1_modifier_invasions.pdf}
-\caption{Consecutive fixation of modifiers that reduce the vertical
-transmission rate in environmental regime \emph{A1B1}. The figure shows
-results of numerical simulations of evolution with two modifier alleles
-(eq. 53). When a modifier allele fixes
-(frequency\textgreater{}99.9\%), a new modifier allele is introduced
-with a vertical transmission rate one order of magnitude lower (vertical
-dashed lines). \textbf{(A,D,G)} The frequency of phenotype \emph{A} in
-the population over time. \textbf{(B,E,H)} The frequency of the invading
-modifier allele over time. \textbf{(C,F,I)} The population mean
-fitness over time; insets zoom in to show that the mean fitness 
-decreases slightly with each invasion. Parameters: vertical transmission rate of
-the initial resident modifier allele, \(\rho_0 =0.1\); fitness values:
-\emph{W}=1, \emph{w}=0.1 (\textbf{A-C}), 0.5 (\textbf{D-F}), and 0.9
-(\textbf{G-I}). The x-axis is on a log-scale, as each sequential invasion
-takes an order of magnitude longer to
-complete.}\label{fig:A1B1_modifier_invasions}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/AkBl_equilibrium.pdf}
-\caption{Stability in a fluctuating environment.
-\textbf{(A)} Stable frequency of phenotype $A$ after every two generations and \textbf{(B)} stable mean fitness in selection regime \emph{A1B1} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
-\textbf{(C)} Stable population mean fitness in selection regime \emph{AkBl} as a
-function of the vertical transmission rate \(\rho\) and different ratios of selection periods \(k=l\).
-\emph{k} and \emph{l} are the number of generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
-The geometric average of the population mean fitness over $k+l$ generations is shown. 
-Here \emph{w=0.9}.
-In all cases, \emph{W=1}.
-}\label{fig:AkBl_equilibrium}
-\end{figure}
-
-\begin{figure}
-\centering
 \includegraphics{../figures/env_A1B2.pdf}
@@
 
-\begin{figure}
-\centering
-\includegraphics{../figures/A1B2_polymorphism.pdf}
-\caption{Stable frequency of phenotype \emph{A} and population mean fitness in selection regime
-\emph{A1B2} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
-In the grey area \(B\) reached fixation and the frequency of \(A\) is 0.
-\textbf{(A)} Stable frequency of phenotype $A$ at the end of each period of three generations.
-\textbf{(B)} Geometric average of the stable mean fitness over the three generations period: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
-In all cases, \emph{W=1}.}\label{fig:A1B2_polymorphism}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/stochastic_env_x_t.pdf}
-\caption{Stochastic local stability. The figure shows the frequency of
-phenotype \emph{A} after \(10^6\) generations in a very large populationt 
-evolving in a stochastic environment (eq. 33). The
-fitness of phenotypes \emph{A} and \emph{B} are \(1+s_t\) and \(1\),
-where \(s_t\) is \emph{s} with probability \emph{p} and \emph{-s} with
-probability \emph{1-p}. The white line marks combinations of \emph{p}
-and \emph{s} for which \(\mathbb{E}[\log{(1+\rho s_t)}]=0\); according
-to our analysis, we expect that below this line \(x^*=0\) will be
-stochastically locally stable. Parameters: \(x_0=0.1\);
-\(\rho=0.1\).}\label{fig:stochastic_env_x_t}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=\dimexpr \textheight - 8\baselineskip\relax]{../figures/stoch_modifier_invasions.png}
-\caption{Consecutive fixation of modifiers that reduce the vertical
-transmission rate under randomly changing selection. The figure shows
-results of numerical simulations of evolution with two modifier alleles
-(eq. 33). The
-fitness of phenotypes \emph{A} and \emph{B} are 1 and 0.1 with probability 0.5 or 0.1 and 1 otherwise.
-When a modifier allele fixes
-(frequency\textgreater{}99.9\%), a new modifier allele is introduced
-with a vertical transmission rate one order of magnitude lower (vertical
-dashed lines). \textbf{(A)} The frequency of phenotype \emph{A} in the
-population over time. \textbf{(B)} The frequency of the invading
-modifier allele over time. \textbf{(C)} The population mean fitness over time. Parameters: vertical transmission rate of the initial
-resident modifier allele, \(\rho_0 =0.1\); fitness values: \emph{W=1}
-and \emph{w=0.1} with probability 0.5 and \emph{w=0.1} and \emph{W=1}
-otherwise. The x-axis is on a log-scale, as each sequential invasion takes
-an order of magnitude longer to
-complete.}\label{fig:stoch_modifier_invasions}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/stoch_two_iid_rvs.png}
-\caption{Effect of vertical transmission rate on phenotype polymorphism
-in randomly changing environments. Dynamics of the frequency of phenotype
-\(A\) over time starting
-at \(x_0=10^{-5}\) when the fitness of phenotypes \(A\) and \(B\) are
-identically and independently distributed random variable. As the vertical transmission rate \(\rho\) increases
-from 0.001 to 0.5 the frequency reaches a polymorphic distribution with \(E(X_t/N)\to0.5\) faster, but
-the variance also increases. \(w_A\) and \(w_B\) are both exponential
-random variables with expected value 2.}\label{fig:stoch_two_iid_rvs}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/fixation_prob_time.pdf}
-\caption{Fixation probability and time in a finite population.
-\textbf{(A)} Fixation probability \(u(x)\) of phenotype \(A\) (eq. 88), and \textbf{(B)}
-Expected time to fixation \(T(x)\) of phenotype \(A\) (eq. 90), conditioned on its
-fixation, starting with a single copy in a population of size \(N\). The
-figure compares two estimates: simulations (blue circles) and diffusion
-equation approximation (green solid line). Parameters: Selection coefficient, \(s=w_A-w_B=0.1\);
-Population size, \(N=10,000\).}\label{fixation_prob_time}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/lk_fix_prob.pdf}
-\caption{Fixation in a finite population with different ratios of selection periods \(\frac{k}{l}\).
-Fixation probability of phenotype $A$ when starting with a single copy in a population of size $N$: $u(1/N) = (1-\exp(-2 \rho \frac{k-l}{k+l}(W-w))/(1-\exp(-2 N \rho \frac{k-l}{k+l}(W-w))$ (see eqs.
-115-116).
-\emph{k} and \emph{l} are the number of
-generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by
-selection. In all cases, \emph{W} = 1, \emph{w} = \emph{1-s}, Population size, \(N=10,000\).} \label{lk_fix_prob}
-\end{figure}
-
 \end{document}

### 2017-08-24 ms/figures.tex
s.tex
aug 22 update with YR changes

@@
 \begin{document}
+\pagenumbering{gobble} % remove page numbers
 
+%Figure 1
 \begin{figure}
@@
 generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
-The boundary is set at $-\log{(1+\rho \frac{s}{1-s}}) / \log{(1-\rho s})$, see eq. 23.
-In all cases, \emph{W} = 1, \emph{w} = \emph{1-s}.}\label{fig:lk_phase_plane}
+In all cases, \emph{W} = 1, \emph{w} = \(1-s\).}\label{fig:lk_phase_plane}
 \end{figure}
 
+%Figure 2
 \begin{figure}
@@
 \caption{Frequency of phenotype \emph{A} after every two generations in
-selection regime \emph{A1B1}. The orange line is the finite population model
-(average of 100 simulations). The blue line is the infinite population model
-(eq. 25), and the green line is the equilibrium solution (eq. 56). Parameters: \emph{W}=1,
-\emph{N}=10,000, initial value \(x=0.5\).}\label{fig:env_A1B1}
+selection regime \(A1B1\). The orange line is the finite population model
+(eqs. 52-53; average of 100 simulations). The blue line is the infinite population model
+(eq.\ 23), and the green line is the  solution of \(G(x)=0\) (see eq.\ 24).
+Parameters: \(W=1\); For the finite population model (orange lines), population size is \(N=10,000\) and initial frequency of \(A\) is \(x_0=0.5\).}\label{fig:env_A1B1}
 \end{figure}
 
+%Figure 3
 \begin{figure}
 \centering
-\includegraphics{../figures/AkBl_equilibrium.pdf}
-\caption{Stability in a fluctuating environment.
-\textbf{(A)} Stable frequency of phenotype $A$ in the \emph{A1B1} selection regime and \textbf{(B)} stable mean fitness in selection regime \emph{A1B1} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
-\textbf{(C)} Stable population mean fitness in selection regime \emph{AkBl} as a
-function of the vertical transmission rate \(\rho\) and different ratios of selection periods \(k=l\).
-\emph{k} and \emph{l} are the number of generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
-The geometric average of the population mean fitness over $k+l$ generations is shown. 
-Here \emph{w=0.9}.
+\includegraphics{../figures/A1B1_equilibrium.pdf}
+\caption{Properties of $A1B1$ environment.
+\textbf{(A)} Stable frequency of phenotype $A$ in the \emph{A1B1} selection regime and \textbf{(B)} stable mean fitness in selection regime \emph{A1B1} as functions of the vertical transmission rate \(\rho\) and the fitness ratio \(w/W\) between the favored and unfavored phenotypes.
 In all cases, \emph{W=1}.
@@
 
+%Figure 4
 \begin{figure}
@@
 \caption{Stochastic local stability. The figure shows the frequency of
-phenotype \emph{A} after \(10^6\) generations in a very large populationt 
-evolving in a stochastic environment (eq. 33). The
+phenotype \emph{A} after \(10^6\) generations in a very large population 
+evolving in a stochastic environment (eq.\ 31). The
 fitness of phenotypes \emph{A} and \emph{B} are \(1+s_t\) and \(1\),
 where \(s_t\) is \emph{s} with probability \emph{p} and \emph{-s} with
-probability \emph{1-p}. The white line marks combinations of \emph{p}
-and \emph{s} for which \(\mathbb{E}[\log{(1+\rho s_t)}]=0\); according
-to our analysis, we expect that below this line \(x^*=0\) will be
-stochastically locally stable. Parameters: \(x_0=0.1\);
-\(\rho=0.1\).}\label{fig:stochastic_env_x_t}
+probability \(1-p\). The white line marks combinations of \emph{p}
+and \emph{s} for which \(E[\log{(1+\rho s_t)}]=0\); according
+to Result 6, below this line fixation of phenotype \emph{B} (\(x^*=0\)) will be
+stochastically locally stable. Parameters: initial frequency of \(A\) is \(x_0=0.1\); vertical transmission rate is \(\rho=0.1\).}\label{fig:stochastic_env_x_t}
 \end{figure}
 
+%Figure 5
 \begin{figure}
@@
 \includegraphics{../figures/stoch_two_iid_rvs.png}
-\caption{Effect of vertical transmission rate on phenotype polymorphism
-in randomly changing environments. Dynamics of the frequency of phenotype
-\(A\) over time starting
-at \(x_0=10^{-5}\) when the fitness of phenotypes \(A\) and \(B\) are
-identically and independently distributed random variable. As the vertical transmission rate \(\rho\) increases
-from 0.001 to 0.5 the frequency reaches a polymorphic distribution with \(E(X_t/N)\to0.5\) faster, but
-the variance also increases. \(w_A\) and \(w_B\) are both exponential
-random variables with expected value 2.}\label{fig:stoch_two_iid_rvs}
+\caption{Effect of vertical transmission rate $\rho$ on phenotype polymorphism in a randomly changing environment.
+Dynamics of the frequency of phenotype \(A\) over time starting at \(x_0=10^{-5}\) when the fitness of phenotypes \(A\) and \(B\) are identically and independently distributed random variables.
+As the vertical transmission rate \(\rho\) increases from 0.001 to 0.5 the frequency reaches a polymorphic distribution with \(E(x_t)\to0.5\) faster,
+but the variance also increases.
+The fitness of phenotypes \(A\) and \(B\), \(w_A\) and \(w_B\), are both exponential random variables with expected value 2.}\label{fig:stoch_two_iid_rvs}
 \end{figure}
 
+%Figure 6
 \begin{figure}
@@
 results of numerical simulations of evolution with two modifier alleles
-(eq. 53). When a modifier allele fixes
+(eq.\ 35). When a modifier allele fixes
 (frequency\textgreater{}99.9\%), a new modifier allele is introduced
@@
 the initial resident modifier allele, \(\rho_0 =0.1\); fitness values:
-\emph{W}=1, \emph{w}=0.1 (\textbf{A-C}), 0.5 (\textbf{D-F}), and 0.9
+$W=1$, $w=0.1$ (\textbf{A-C}), 0.5 (\textbf{D-F}), and 0.9
 (\textbf{G-I}). The x-axis is on a log-scale, as each sequential invasion
@@
 
+%Figure 7
 \begin{figure}
@@
 \caption{Consecutive fixation of modifiers that reduce the vertical
-transmission rate under randomly changing selection. The figure shows
+transmission rate $\rho$ under symmetric randomly changing selection. The figure shows
 results of numerical simulations of evolution with two modifier alleles
-(eq. 33). The
-fitness of phenotypes \emph{A} and \emph{B} are 1 and 0.1 with probability 0.5 or 0.1 and 1 otherwise.
-When a modifier allele fixes
+(eq.\ 33). When a modifier allele fixes
 (frequency\textgreater{}99.9\%), a new modifier allele is introduced
@@
 population over time. \textbf{(B)} The frequency of the invading
-modifier allele over time. \textbf{(C)} The population mean fitness over time. Parameters: vertical transmission rate of the initial
-resident modifier allele, \(\rho_0 =0.1\); fitness values: \emph{W=1}
-and \emph{w=0.1} with probability 0.5 and \emph{w=0.1} and \emph{W=1}
-otherwise. The x-axis is on a log-scale, as each sequential invasion takes
+modifier allele over time. Parameters: vertical transmission rate of the initial
+resident modifier allele, \(\rho_0 =0.1\); fitness values: $W=1$
+and $w=0.1$ with probability 0.5 and $w=0.1$ and $W=1$
+also with probability 0.5. The x-axis is on a log-scale, as each sequential invasion takes
 an order of magnitude longer to
@@
 
+%Figure 8
 \begin{figure}
@@
 \caption{Stable frequency of phenotype \emph{A} and population mean fitness in selection regime
-\emph{A1B2} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
-In the grey area \(B\) reached fixation and the frequency of \(A\) is 0.
+$A1B2$ as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
+In the grey area \(B\) reaches fixation and the frequency of \(A\) is 0.
 \textbf{(A)} Stable frequency of phenotype $A$ at the end of each period of three generations.
-\textbf{(B)} Geometric average of the stable mean fitness over the three generations period: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
+\textbf{(B)} Geometric average of the stable mean fitness over the three-generation period: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
 In all cases, \emph{W=1}.}\label{fig:A1B2_polymorphism}
@@
 
+%Figure 9
+\begin{figure}
+\centering
+\includegraphics{../figures/rho_evol_stability.pdf}
+\caption{Evolutionarily stable vertical transmission rate in $AkBk$ selection regime. 
+The evolutionary stable rate $\rho^*$ was determined as in Carja et al.\ (2014) using simulations of eq.\ 35: for each $k$ from 1 to 50, ten initial rates were randomly chosen between 0 and 1 for the resident modifier $m$; after the population evolved with only $m$ for 1,000 generations towards an equilibrium, invader modifier alleles $M$ were consecutively introduced at frequency $10^{-4}$; their value for $\rho$ was a product of the resident rate and a random number from an exponential distribution with a mean of 1; if an invader increased in frequency after $5,000 \times 2k$ generations, the invader was considered successful and became the resident modifier for the next invasion, and allowed to evolve for 1,000 generations to reach a new equilibrium before the next invasion; if the invader was unsuccessful, a new invader was drawn.
+After at least 500 invasion attempts and 50 unsuccessful ones, the evolutionary process stopped.
+The markers show the average for ten such processes for each $k$ value.
+95\% confidence intervals (calculated with bootstrap, 1,000 resamples) are too small to be seen.
+The dashed line shows the relation $\rho=1-1/(k-1)$ (Carja et al.
+2014), which is close to the simulation results for \(w=0.5\).
+In all cases, $W=1$.}\label{fig:rho_evol_stability}
+\end{figure}
+
+%Figure 10
+\begin{figure}
+\centering
+\includegraphics{../figures/AkBk_stable_wbar_slice.pdf}
+\caption{Vertical transmission rates that maximize population mean fitness in the \emph{AkBk} selection regime.
+Populations evolved with different vertical transmission rates \(\rho\) and in different \emph{AkBk} selection regimes for \(1000 \times 2k\) generations, and the geometric average of the population mean fitness in the final \(2k\) generations was calculated.
+For each \(k\), the rate \(\hat{\rho}\) that maximizes the geometric average is shown.
+In all cases, $W=1$. See Fig. S3 for a broader view.} \label{AkBk_stable_wbar_slice}
+\end{figure}
+
+%Figure 11
 \begin{figure}
@@
 \caption{Fixation probability and time in a finite population.
-\textbf{(A)} Fixation probability \(u(x)\) of phenotype \(A\) (eq. 88), and \textbf{(B)}
-Expected time to fixation \(T(x)\) of phenotype \(A\) (eq. 90), conditioned on its
+\textbf{(A)} Fixation probability \(u(x)\) of phenotype \(A\) (eq.\ 70), and \textbf{(B)}
+Expected time to fixation \(T(x)\) of phenotype \(A\) (eq.\ 72), conditioned on its
 fixation, starting with a single copy in a population of size \(N\). The
@@
 
+%Figure 12
 \begin{figure}
@@
 generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by
-selection. In all cases, \emph{W} = 1, \emph{w} = \emph{1-s}, Population size, \(N=10,000\).} \label{lk_fix_prob}
+selection. In all cases, $W = 1$, $w = 1-s$, Population size, \(N=10,000\).} \label{lk_fix_prob}
 \end{figure}

### 2017-08-24 ms/figures_supp.tex
s.tex
aug 22 update with YR changes

@@
 
+\newcommand{\beginsupplement}{%
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}%
+     }
+
 % set default figure placement to htbp
@@
 \begin{document}
+\pagenumbering{gobble}
+\beginsupplement
 
@@
 \includegraphics{../figures/env_A1B2.pdf}
-\caption{Frequency of phenotype \emph{A} after every three generation in
-selection regime \emph{A1B2}. Comparison of dynamics starting with
-different initial frequency of phenotype \(A\) (0.01-0.99).
-See also Fig. 1 and 6. Parameters: \emph{W}=1, \emph{N=10,000},
-initial population \(x=0.5\).}\label{fig:env_A1B2}
+\caption{Frequency of phenotype \emph{A} after every three generations in
+selection regime $A1B2$. Comparison of dynamics starting with
+different initial frequency of phenotype \(A\) (0.01--0.99).
+See also Figure 1 and Figure S2. In all cases, $W=1$.}\label{fig:env_A1B2}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/AkBk_x0.pdf}
+\caption{Convergence of the frequency of phenotype \emph{A} to a stable polymorphism in selection regime \emph{AkBk}.
+Comparison of dynamics starting with different initial frequencies of phenotype \(A\) (0.01--0.99), and different $k$, $\rho$ and $w$ values.
+The lines show the $x$ frequency of phenotype \emph{A} at the end of each period, after every $2k$ generations.
+In all cases, $W=1$.}\label{fig:AkBk_x0}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/AkBk_stable_wbar.pdf}
+\caption{Stable population mean fitness in selection regime \emph{AkBk} as a function of the vertical transmission rate \(\rho\) and the number \(k\) of generations in which phenotypes \emph{A} and \emph{B} are favored by selection, for different selection intensities.
+Colors represent the geometric average of the population mean fitness over \(2k\) generations after \(1,000 \times 2k\) generations of evolution.
+Blue markers show the maximum average mean fitness for each period \(k\).
+For example, with \(w=0.1\), \(\hat{\rho}=0\) maximizes the average fitness for \(k \le 11\), then \(\hat{\rho}\) increases to \(\hat{\rho} \approx 0.24\), and then continues to decrease as \(k\) increases, down to \(\hat{\rho} \approx 0.15\) for \(k=50\).
+Contour lines represent \(\rho\) and \(k\) combinations that produce the same average mean fitness. 
+In all cases, $W=1$.}\label{fig:AkBk_stable_wbar}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/env_A3B10.pdf}
+\caption{Frequency of phenotype \emph{A} after every thirteen generations in
+selection regime $A3B10$. Comparison of dynamics starting with
+different initial frequency of phenotype \(A\) (0.01--0.99).
+In all cases, $W=1$.}\label{fig:env_A3B10}
 \end{figure}

### 2017-08-24 ms/ms.tex
aug 24 update from MWF

\magnification=1170
\hsize=6.5truein
\hoffset=.05truein
%\vsize=8.5truein
%\voffset=.25truein
\hfuzz=14pt
\parskip=5pt
\baselineskip=16pt

\def\harr#1#2{\smash{\mathop{\hbox to .2in{\rightarrowfill}}\limits^{\scriptstyle#1}_{\scriptstyle#2}}}

\def\var{\varepsilon}

\def\ref{\par\noindent\hangindent.35truein}

\def\cl{\hbox{\tenbf L}}
\def\ci{\hbox{\tenbf I}}

\font\tit=cmbx10 at 14pt
\ \vskip1.1in
\centerline{\tit Vertical and Oblique Transmission under}\smallskip
\centerline{\tit Fluctuating Selection}
\vskip.3in

\centerline{Yoav Ram$^1$, Uri Liberman$^{2}$, and Marcus W. Feldman$^1$}

\vskip.5truein

%\centerline{$*$ Corresponding author}
\bigskip
\centerline{$^1$Department of Biology}
%\vskip-7pt
\centerline{Stanford University}
%\vskip-7pt
\centerline{Stanford, CA 94305-5020}
\bigskip

\centerline{$^2$School of Mathematical Sciences}
%\vskip-7pt
\centerline{Tel Aviv University}
%\vskip-7pt
\centerline{Tel Aviv, Israel 69978}
%\vskip-7pt
%\centerline{uril@tauex.tau.ac.il}
\vfil
\centerline{August 23, 2017, v.1}

\break
\noindent{\bf INTRODUCTION}
\smallskip

Cavalli-Sforza and Feldman (1981) distinguished two forms of non-parental phenotypic transmission in the context of cultural evolution. Horizontal transmission  occurs when a trait is passed between members of the same generation and is analogous to transmission of an infectious agent. Oblique transmission to offspring is from non-parental members of the parental generation. Evolution under either of these is expected to be more rapid than under purely vertical, i.e., parent-to-offspring, transmission (Lycett and Gowlett 2008; Bergstrom and Dugatkin 2012, Ch. 19.4).

Oblique transmission occurs via some mechanism of social learning, which may include imitation or active teaching. There has been an interesting debate over the past thirty years concerning the conditions under which social learning would have an advantage over individual learning or vertical (including genetic) transmission. This debate is usually couched in terms of the mode and tempo of environmental fluctuations that would affect fitness and hence evolution (Rogers 1988; Boyd and Richerson 1988, 1995; Feldman et al. 1996; Wakano et al. 2004: Aoki et al. 2005; Wakano and Aoki 2007; Aoki and Feldman 2014). Mathematical analyses of models of competition between individual and social learning have generally shown that social learning has an advantage when the environment does not fluctuate too frequently. On the other hand, when the environmental changes are very frequent, individual learning is favored, while innate (genetic) determination of the trait does best when periods between environmental change are long on average.

In some situations, oblique transmission of biological material is possible. An example could be the transmission of microbes occasioned by the sharing or manipulation of food or other consumable resources during a social interaction. Although transmission of the microbiome in humans is likely to be mostly vertical (Rosenberg and Zilber-Rosenberg 2016), in other organisms there is multi-generational food sharing during which symbionts from the parental cohort may be transmitted obliquely to younger individuals (Theis et al.\ 2016). In such cases, fluctuations in the resource type or availability may have fitness effects that depend on features of the transmitted microbiome. This ecological perspective on community transmission is stressed by van Opstal and Bordenstein (2015), who emphasize the ``need to consider the relative roles of vertical and horizontal transmission of microbial communities.''

Another perspective on the evolutionary consequences of fluctuating environments (and, as a result, fluctuating fitnesses) has derived from the phenomenon of phenotypic switching (Balaban et al.\ 2004; Kussell and Leibler 2005; Thattai and van Oudenaarden 2004; Salath\'e et al.\ 2009; Liberman et al.\ 2011; Gaal et al.\ 2010). In these studies, mutation caused the organism to switch phenotypes (usually treated as haploid genotypes), and the problem has usually been posed in terms of the optimal rate of mutation when the phenotypic fitnesses fluctuate over time. These models did not include social learning, and the evolution was regarded as a mode of bet-hedging against future environmental change. Optimal (that is, evolutionarily stable) mutation rates depend on many features of the fluctuations, for example,  degree of fitness symmetry, strength of selection, and variance in the period of fluctuation (Salath\'e et al.\ 2009).

In a recent analysis of evolution in fluctuating environments, Xue and Leibler (2016) allowed an organism to absorb information about the distribution of possible environments by learning the phenotypes of members of its parental lineage from previous generations. They describe this as ``positive feedback that enhances the probability of the offspring to express the same phenotype as the parent.'' In this formulation there was ``reinforcement of the parent phenotype'' in an offspring, such as might occur through epigenetic inheritance. Although their analysis was not couched in terms of oblique and vertical transmission, as defined by Cavalli-Sforza and Feldman (1981), we have been stimulated by their analysis to develop a model in which oblique transmission, at a rate dependent on the trait frequency in the parental generation, occurs in addition to classical vertical transmission. We then ask how fluctuations in fitnesses interact with the rate of oblique transmission to affect evolutionary dynamics and how the rate of oblique transmission itself might evolve.

In our formulation, both the parental phenotype and the distribution of phenotypes in the whole population contribute to an offspring's phenotype. Using conventional modifier theory (Feldman and Liberman 1986), we show that in a cyclic selection regime with cycles of periods 1 or 2, an allele reducing the rate of vertical transmission is expected to increase in frequency when rare and in so doing to increase the mean fitness of the population. However, for cycles of greater length, interesting non-monotonicities emerge, both in the uninvadable rate of vertical transmission, and the rate that maximizes geometric mean fitness. We develop the models in very large populations with random fitnesses and also in the case where drift occurs via sampling from generation to generation in a finite population.

\vfil\break

\noindent{\bf THE MODEL}
\smallskip

Consider an infinite population whose members are characterized by their phenotype $\phi$, which can be of two types $\phi=A$ or $\phi=B$, with associated frequencies $x$ and $(1-x)$, respectively. We follow the evolution of $x$ over discrete non-overlapping generations. In each generation individuals are subject to  selection where the fitnesses of $A$ and $B$ are $w_A$ and $w_B$, respectively.

An offspring inherits its phenotype from its parent via {\sl vertical transmission} with probability $\rho$ and from a random individual in the parental population via {\sl oblique transmission} with probability $(1-\rho)$. Therefore, given that the parent phenotype is $\phi$ and assuming {\sl uni-parental inheritance} (Zefferman 2016), the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is
$$P(\phi' =A|\phi) =\left\{\matrix{(1-\rho)x+\rho &\hbox{\tenrm if}\ \phi=A\cr 
\noalign{\smallskip}
 (1-\rho)x & \hbox{\tenrm if}\ \phi =B\cr}\right.\eqno(1)$$
 where $x=P(\phi=A)$ in the parent's generation before selection.

 Therefore the frequency $x'$ of  phenotype $A$ after one generation is given by the recursion equation
 $$x'= \rho{w_Ax\over \overline w} +(1-\rho)x=     {w_A\over \overline w}x\bigl[(1-\rho)x +\rho\bigr] +{w_B\over \overline w}(1-x)\bigl[(1-\rho)x\bigr],\eqno(2)$$
 where $\overline w$ is the {\sl mean fitness}, namely
 $$\overline w=w_A x+w_B(1-x).\eqno(3)$$
 Equation (2) can be rewritten as
 $$x' =x\left[1+\rho(1-x){w_A-w_B\over \overline w}\right]    =x\cdot{x(1-\rho)(w_A-w_B) +\rho w_A +(1-\rho)w_B\over x(w_A-w_B) +w_B}.\eqno(4)$$

 In what follows we explore the evolution of the recursion (4), namely the equilibria and their stability properties, in the cases of {\sl constant environments} and {\sl changing environments}.
 \bigskip
 \bigskip

\noindent{\bf CONSTANT ENVIRONMENT}
\smallskip

When the environment is constant, the fitness parameters $w_A$ and $w_B$ do not change between generations, and we have the following result.

\proclaim Result 1. If $0<\rho\le 1$ and both $w_A$ and $w_B$ are positive with $w_A\ne w_B$, then fixation in the phenotype $A(B)$ is globally stable when $w_A>w_B$ $(w_A<w_B)$.\par

\noindent{\it Proof}.
If we rewrite (4) as $x'=x\cdot F(x)$, it can be seen that $F(1)=1$, and for $\rho>0$ and $0<x<1$
$$\eqalign{
&F(x)>1\quad\hbox{\tenrm when}\quad w_A>w_B,\cr
&F(x)<1\quad\hbox{\tenrm when}\quad w_A<w_B.}\eqno(5)$$
Hence, as $w_A>0$ and $w_B>0$, both fixations in $A$ or in $B$  ($x^*=1$ for fixation in $A$ and $x^*=0$ for fixation in $B$) are equilibrium points of (4). Moreover, if $x_t$ is the value of $x$ at the $t$-th generation $(t=0,1,2,\dots)$, from (4)  and (5) we have for any $0<x_0<1$ and all $t=0,1,2,\dots$,
$$\eqalign{
&x^{t+1}>x^t\quad\hbox{\tenrm when}\quad w_A>w_B,\cr
&x^{t+1}<x^t\quad\hbox{\tenrm when}\quad w_A<w_B.}\eqno(6)$$

\noindent Thus, as $x^*=1$ or $x^*=0$ are the only equilibrium points, then
$$\eqalign{
&\lim_{t\to\infty}x_t=1,\ \ \hbox{\tenrm for all}\ 0<x_0\le 1,\ \ \hbox{\tenrm when}\ w_A>w_B,\cr
&\lim_{t\to\infty}x_t=0,\ \ \hbox{\tenrm for all}\ 0\le x_0< 1,\ \ \hbox{\tenrm when}\ w_A<w_B.}\eqno(7)$$
Therefore fixation of the favored phenotype is globally stable, and ultimately all individuals have the favored phenotype.

If $\rho=0$ or $w_A=w_B>0$, the recursion (4) reduces to $x'=x$; the phenotype frequencies do not change over time.
With {\sl extreme selection},  for example $w_A=1$ and $w_B=0$ (or $w_B=1$ and $w_A=0$), recursion (4) reduces to
$x'=(1-\rho)x+\rho$, and therefore 
$x_t=1-(1-\rho)^t(1-x_0)$.
Hence, for $\rho>0$,  $x_t\,\harr{}{t\to\infty}\,1$ for all initial values $0\le x_0\le 1$, and  the favored phenotype fixes.
\bigskip
\bigskip

\noindent{\bf PERIODICALLY CHANGING ENVIRONMENT}
\smallskip

Next we consider the case where the environment is not constant but rather changes periodically, such that the favored phenotype changes after a fixed number of generations. Simple examples are $A1B1=ABABAB\dots$, in which the favored genotype switches every generation, or $A2B1=AABAABAAB\dots$, where every two generations in which selection favors $A$ are followed by a single generation in which selection favors $B$.

 In general, $AkBl$ denotes a selection regime in which the period is of $(k+l)$ generations with $k$ generations favoring phenotype $A$ followed by $l$ generations favoring $B$.

 Let $W$ be the fitness of the favored phenotype and $w$ be that of the other phenotype where $0<w<W$. Rewrite equation (4) as $x'=F_A(x)=xf_A(x)$ when $A$ is favored and $x'=F_B(x)=xf_B(x)$ when $B$ is favored. Then
  $$\eqalign{
  &f_A(x) = {x(1-\rho)(W-w) +\rho W +(1-\rho)w\over x(W-w) +w} = 1+\rho(1-x){W-w\over Wx +w(1-x)},\cr\noalign{\smallskip}
  &f_B(x) = {x(1-\rho)(w-W) +\rho w +(1-\rho)W\over x(w-W) +W} = 1+\rho(1-x){w-W\over wx+W(1-x)}.}\eqno(8)$$

  \noindent If $x_t$ denotes the frequency of exhibiting the phenotype $A$ at generation $t$ starting with $x_0$ initially, then as we are interested in the values of $x_t$ for $t=n(k+l)$ with $n=0,1,\dots$ at the end of complete periods then in fact
  $$x_{(n+1)(k+l)}=F(x_{n(k+l)}),\quad n=0,1,2,\dots,\eqno(9)$$
  where $F$ is the composed function
  $$F=\underbrace{F_B\circ F_B \circ\cdots\circ F_B}_{l\ \hbox{\sevenrm times}} \circ\underbrace{F_A\circ F_A \circ\cdots\circ F_A}_{k\ \hbox{\sevenrm times}}.\eqno(10)$$
  Clearly, since $F_A(0) =F_B(0) =0$ and $F_A(1) =F_B(1) =1$, both fixations in $A$ or in $B$ are equilibrium points. An interesting question is when these fixations are locally stable. We concentrate on $x^*=0$, the fixation of the phenotype $B$. As $x'=F_A(x) =xf_A(x)$ for $k$ generations and $x'=F_B(x) =xf_B(x)$ for $l$ generations, the linear approximation of $F(x)$ ``near'' $x=0$ is
   $$F(x) \simeq\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^lx.\eqno(11)$$
   Hence the local stability of $x^*=0$ is determined by the product $\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l$, and $x^*=0$  is locally stable if this product  is less than 1 and unstable if it is larger than 1.

   From (8) we have
   $$\eqalign{
   &f_A(0) = 1+\rho{W-w\over w},\cr\noalign{\smallskip}
   &f_B(0) = 1+\rho{w-W\over W}.}\eqno(12)$$
   We start with the case $k=l$.

   \proclaim Result 2. If $k=l$ and $0<w<W$ with $0<\rho<1$, fixation of $B$ is unstable.\par

   \noindent {\it Proof.} The local stability of $x^*=0$, the fixation of $B$, is determined by the product
   $$\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^k=\left[\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right)\right]^k.\eqno(13)$$
   Observe that
   $$\eqalign{
   \left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right) &= \left(1-\rho+\rho{W\over w}\right)\left(1-\rho+\rho{w\over W}\right)\cr
   &= \left(1-\rho\right)^2+ \rho^2 +\rho(1-\rho){W^2+w^2\over Ww}.}\eqno(14)$$
   Thus,
   $$\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right) =1+\rho(1-\rho){\left(W-w\right)^2\over wW}>1.\eqno(15)$$
   Since $0<\rho<1$ and $0<w<W$,  the $B$ fixation is unstable.
  \bigskip

\noindent{\it Conclusions}

\item{1.}As $k=l$ and the above result holds also when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is not stable. Thus neither phenotype can become extinct, and we have a {\sl protected polymorphism} (Prout 1968).

\item{2.}Observe that for general $k$, $l$, the condition for local stability of fixation in $A$ is
$$\bigl[f_A(0)\bigr]^l \bigl[f_B(0)\bigr]^k <1, \eqno(16)$$
and that of $B$ is
$$\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l <1. \eqno(17)$$
Therefore, following Result 2, 
$$\bigl[f_A(0)\bigr]^{k+l}\bigl[f_B(0)\bigr]^{k+l}>1,\eqno(18)$$
and it is impossible that both fixations are stable. Further, since by (12)  $f_A(0)>1$ and $0<f_B(0)<1$ when $0<w<W$, by choosing $k$ and $l$ appropriately, fixation on $A$ or fixation on $B$ (but not both) can be stable. In addition, we can have both fixations unstable giving the following result.

\proclaim Result 3. With $0<\rho<1$ and $0<w<W$ in the case of $AkBl$ periodically changing environments, both fixations may be unstable, producing a protected polymorphism.\par

\noindent{\it Proof}. Let $a=1+\rho{W-w\over w}$ and $b=1+\rho{w-W\over W}$. Then since $0<\rho <1$ and $0<w<W$, we have $a>1$ and $0<b<1$. Following (11), fixation in $B$ is not stable if $a^kb^l>1$, and similarly  fixation in $A$ is unstable if $a^lb^k>1$. Therefore both fixations are unstable if
$$a^kb^l>1\quad\hbox{\tenrm and}\quad a^lb^k>1,\eqno(19)$$
or equivalently if
$$k\log a +l\log b>0\quad\hbox{\tenrm and}\quad l\log a+k\log b>0.\eqno(20)$$
Now (20) holds if and only if
$$k\,{\log(1/b)\over\log a} < l < k\,{\log a\over\log(1/b)}.\eqno(21)$$ 
Inequalities (21) are consistent if and only if $\log(1/b)<\log a$, i.e., $ab>1$. But by (15),
$$ab=\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right)>1.\eqno(22)$$

\noindent{\it Remark}.
It should be pointed out that the local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations $A$ was favored $k$ times and $B$ was favored $l$ times, and not their order in the cycle.

When neither fixation in $A$ or $B$ is stable, we have a case of protected polymorphism, and we expect to have one or more polymorphic equilibria. Figure 1 shows the relationship between $k,l$ and $\rho$ that supports polymorphism of $A$ and $B$, or fixation, for different values of $W$ and $w$.

\medskip
\centerline{[FIGURE 1 HERE]}
For the simple case of $A1B1$ periodically changing environment we have the following.

\proclaim Result 4. In the case $A1B1$ with $0<\rho<1$ and $0<w<W$ the two fixations are unstable and there exists a unique stable polymorphism.

\noindent{\it Proof}. Let $x$ be the initial frequency of $A$ and $x'$ its frequency  after one cycle of the $A1B1$ selection. Then $x' =F_B\bigl(F_A(x)\bigr)$ where by (8)
$$\eqalign{
&F_A(x)=x{x(1-\rho)(W-w) +\rho W+(1-\rho)w\over x(W-w) +w},\cr\noalign{\smallskip}
&F_B(y) = y{y(1-\rho)(w-W) +\rho w+(1-\rho)W\over y(w-W)+W}.}\eqno(23)$$
The equilibrium equation is $x=F_B\bigl(F_A(x)\bigr)$, which reduces to a fourth degree polynomial equation in $x$. Since the two fixations in $B$ and $A$ are equilibria corresponding to the two solutions $x=0$ and $x=1$, the other equilibria correspond to solutions of a quadratic equation $G(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with
$$\alpha_2=1,\quad \alpha_1={W+w\over (2-\rho)(W-w)}-1,\quad \alpha_0={-w\over (2-\rho)(W-w)}.\eqno(24)$$
As $0<\rho<1$ and $0<w<W$ we have
$$G(0) = {-w\over (2-\rho)(W-w)}<0\eqno(25)$$
and
$$G(1) ={W\over (2-\rho)(W-w)} >0.\eqno(26)$$
Also, as $\alpha_2=1$ and $\alpha_0<0$, the quadratic equation $G(x)=0$ has two real roots, one negative and one positive $x^*$ satisfying $0<x^*<1$. The latter determines a unique polymorphism. Let $H(x) =F_B\bigl(F_A(x)\bigr)$. Then 
 $$H(0)=0,\quad H(x^*)=x^*,\quad H(1)=1.\eqno(27)$$
 Also 
 $$F'_A(x) = {x^2(1-\rho)\left(W-w\right)^2 +2xw(1-\rho)(W-w) +w\left[\rho W +(1-\rho)w\right]\over \left[x(W-w) +w\right]^2}\eqno(28)$$
 and
  $$F'_B(x) = {x^2(1-\rho)\left(w-W\right)^2 +2xW(1-\rho)(w-W) +W\left[\rho w +(1-\rho)W\right]\over \left[x(w-W) +W\right]^2}.\eqno(29)$$
  As $0<\rho<1$ and $0<w<W$, we have $F'_A(x)>0$ for  $0\le x\le 1$. Observe that the numerator of $F'_B(x)$ is linear in $\rho$; its value when $\rho=1$ is $wW>0$, and when $\rho=0$ it is
  $$x^2\left(w-W\right)^2 +2xW(w-W) +W^2 =\left[x(w-W) +W\right]^2 >0.\eqno(30)$$
  Hence $F'_{B}(x)>0$ for all $0\le x\le 1$, and $H'(x) =F'_B\bigl(F_A(x)\bigr)F'_A(x)$ is positive when $0\le x\le 1$. Thus $H(x)$ is a monotone increasing function of $x$ for $0\le x\le 1$.
Thus, starting from any initial value $0<x_0 <1$ we have  $x_t\,\harr{}{t\to\infty}\,x^*$. Figure 2 illustrates examples of how the frequency of $A$ changes over time in the $A1B1$ regime of cycling selection, while Figure 3A shows the equilibrium frequencies $x^*$ as a function of $\rho$, $W$, and $w$ in this regime.

For more general cyclic fitness regimes, the polynomial that gives the equilibria is of higher order, and it is conceivable that more than one stable polymorphism could exist for given values of $\rho$, $W$, and $w$. We investigated this numerically, and as shown in Figure S1, the $A1B2$ regime also appears to allow a single stable polymorphism. For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplemental Figure S2 shows that only a single stable equilibrium appears to exist.
\medskip
\centerline{[FIGURES 2, 3 HERE]}  
 \bigskip

 \noindent{\bf RANDOMLY CHANGING ENVIRONMENT}
 \smallskip

 We now consider the  case where the environment changes according to a stochastic process. Without loss of generality, assume that the fitness parameters at generation $t$ ($t=0,1,2,\dots$) are $1+s_t$ for phenotype $A$ and $1$ for phenotype $B$, where the random variables $s_t$ for $t=0,1,2,\dots$ are independent and identically distributed. Also we assume that there are positive constants $C$ and $D$ such that $P(-1+C<s_t <D)=1$.

 Corresponding to (4), with $w_A=1+s_t$, $w_B=1$ the recursion equation is
 $$x_{t+1}=x_t {1+\rho s_t +x_t(1-\rho)s_t\over 1+x_ts_t}\quad t=0,1,2,\dots\; .\eqno(31)$$

\noindent As $\{x_t\}$ for $t=0,1,2,\dots$ is a sequence of random variables, the notion of stability of the two fixation states needs clarification. Following Karlin and Lieberman (1974, 1975) we make the following definition.

 \noindent{\it Definition}: ``stochastic local stability''.
 A constant equilibrium state $x^*$ is said to be {\sl stochastically locally stable} if for any $\var>0$ there exists a $\delta>0$ such that $|x_0-x^*|<\delta$ implies
 $$P\left(\lim_{t\to\infty}x_t =x^*\right)\ge 1-\var.\eqno(32)$$
 Thus stochastic local stability holds for $x^*$ provided for any initial $x_0$ sufficiently near $x^*$ the process $x_t$ converges to $x^*$ with high probability.

 In our case there are two constant equilibria $x^*=0$ and $x^*=1$ corresponding to fixation in $B$ and $A$, respectively. We can characterize the stochastic local stability of these fixations as follows.

 \proclaim Result 5. Suppose $E\left[\log (1+\rho s_t)\right]>0$. Then $x^*=0$, the fixation in $B$, is not stochastically locally stable. In fact $P\left(\lim_{t\to\infty}x_t=0\right)=0$.

  \proclaim Result 6. Suppose $E[\log(1+\rho s_t)]<0$. Then $x^*=0$ is stochastically locally stable. In particular, if $E(s_t)\le 0$, $x^*=0$ is stochastically locally stable.

 The proofs of Results 5 and  6 can be found in Supplementary Materials SP1 and SP2.

 Using the general notation for the fitness parameters $w_A$ and $w_B$, the stochastic local stability of the $B$ fixation is determined by the sign of $E\left[\log\left(1-\rho+\rho{w_A\over w_B}\right)\right]$ and that of the $A$ fixation by the sign of $E\left[\log\left(1-\rho+\rho{w_B\over w_A}\right)\right]$. For example, if the first sign  is negative, fixation in $B$ is stochastically locally stable, and when it is positive, with probability one convergence to fixation in $B$ does not occur. It is also true  that if $E(w_A/w_B)\le 1$, then fixation of $B$ is stochastically locally stable. Following (14),
 $$\left(1-\rho +\rho{w_A\over w_B}\right)\left(1-\rho+\rho{w_B\over w_A}\right) >1.\eqno(33)$$
 Therefore, as in the case of periodically changing environments $AkBl$, it is impossible that both fixations are simultaneously stochastically locally stable.  It is possible, however, that neither fixation is  stochastically locally stable, in which case we expect the population to converge to a polymorphic distribution. Figure 4 illustrates how the properties of $s_t$ in recursion (31) affect the frequency of phenotype $A$, and in particular the stochastic local stability of fixation in $B$. Figure 5 shows the dynamics of the frequency of $A$ in a case where $w_A$ and $w_B$ are identically distributed and independent; in this case the expectation of the stationary distribution is 1/2, and its variance increases as $\rho$ increases.
\medskip
\centerline{[FIGURES 4, 5 HERE]}  
 \bigskip

 \noindent{\bf EVOLUTIONARY STABILITY OF OBLIQUE TRANSMISSION}
 \smallskip

 An interesting question concerns the evolution of oblique transmission itself. For example, is there an evolutionarily stable rate of oblique transmission? To answer this question we use a modifier model.

 Suppose that the vertical transmission rate is determined by a genetic locus with two possible alleles $m$ and $M$. Let the vertical transmission rates determined by $m$ and $M$ be $\rho$ and $P$, respectively. Thus there are four pheno-genotypes $mA$, $mB$, $MA$, $MB$ whose frequencies at a given generation are denoted by $x_1$, $x_2$, $x_3$, $x_4$, respectively. As the fitnesses are determined by the two phenotypes $A$ and $B$, and the modifier locus is selectively neutral, we have the following table.
 $$\matrix{
 \hbox{\tenrm pheno-genotype}&mA&mB&MA&MB\cr\noalign{\smallskip}
 \hbox{\tenrm frequency}&x_1&x_2&x_3&x_4\cr\noalign{\smallskip}
 \hbox{\tenrm fitness}&w_A&w_B&w_A&w_B\cr\noalign{\smallskip}
 \hbox{\tenrm vertical transmission rate}&\rho&\rho&P&P\cr}\eqno(34)$$
 Following the rationale leading to equation (2), the next generation pheno-genotype frequencies $x'_1$, $x'_2$, $x'_3$, $x'_4$ are
 $$\eqalign{
 \overline wx'_1 &= w_Ax_1\bigl[(1-\rho)(x_1+x_3) +\rho\bigr] +w_Bx_2(1-\rho)(x_1+x_3)\cr
 \overline wx'_2 &= w_Ax_1(1-\rho)(x_2+x_4) +w_Bx_2\bigl[(1-\rho)(x_2+x_4)+\rho\bigr]\cr 
  \overline wx'_3 &= w_Ax_3\bigl[(1-P)(x_1+x_3) +P\bigr] +w_Bx_4(1-P)(x_1+x_3)\cr
\overline wx'_4 &= w_Ax_3(1-P)(x_2+x_4) +w_Bx_4\bigl[(1-P)(x_2+x_4)+P\bigr], 
}\eqno(35)$$
with $\overline w$, the mean fitness, given by
$$\overline w =w_A(x_1+x_3) +w_B(x_2+x_4).\eqno(36)$$
Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} to invasion by allele $M$.

Because we assume the initial existence of a stable polymorphic equilibrium that depends on $\rho$ (the rate determined by $m$), we cannot assume a {\sl constant environment}, which always leads to fixation of the favored type, independent of $\rho$. We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$. Specifically, following (35) with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
$$\underline x''=T_2(T_1\underline x),\eqno(37)$$
where $\underline x'=T_1\underline x$ is given by (35) with $w_A=W$, $w_B=w$, and $\underline x''=T_2\underline x'$ is given by (35) with $w_A=w$, $w_B=W$. Here $\underline x$, $\underline x'$, $\underline x''$ are the frequency vectors.

For the $A1B1$ case, when only the $m$ allele is present with associated rate $\rho$, with $0<\rho<1$ and $0<w<W$, a unique stable equilibrium $\underline x^* =(x_1^*,x_2^*,0,0)$ exists. $x_1^*$ is the only positive root of the quadratic equation $G(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with $\alpha_2,\alpha_1,\alpha_0$ specified in (24). Solving $G(x)=0$ gives
$$x_1^* ={1\over 2} -{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}\over 2\cdot(2-\rho)(W-w)},\eqno(38)$$
 and it is easily seen that
 $${\sqrt{Ww}-w\over W-w}<x_1^* <{1\over 2}.\eqno(39)$$

 The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ derived from (35) and given by
 $$\overline w^*\cl_1 =\left[\matrix{W\bigl[(1-P)x_1^* +P\bigr] & w(1-P)x_1^*\cr\noalign{\medskip}
 W(1-P)x_2^* & w\bigl[(1-P)x_2^* +P\bigr]\cr}\right]\eqno(40)$$
 and
 $$\overline w^{**}\cl_2 =\left[\matrix{w\bigl[(1-P)x_1^{**} +P\bigr] & W(1-P)x_1^{**}\cr\noalign{\medskip}
 w(1-P)x_2^{**} & W\bigl[(1-P)x_2^{**} +P\bigr]\cr}\right],\eqno(41)$$
where $\underline x^{**} =T_1\underline x^*$ and
$$\overline w^* =Wx_1^* +wx_2^*,\qquad \overline w^{**} =wx_1^{**} +Wx_2^{**}.\eqno(42)$$
Due to the symmetry between the two phenotypes $A$ and $B$ in the $A1B1$ case, we have $x_1^{**} =x_2^*$ and $x_2^{**} =x_1^*$ so that $\overline w^{**} =\overline w^*$ and in fact
$$\overline w^*\cl_2 = \left[\matrix{w\bigl[(1-P)x_2^* +P\bigr] & W(1-P)x_2^*\cr\noalign{\medskip}
 w(1-P)x_1^* & W\bigl[(1-P)x_1^* +P\bigr]\cr}\right].\eqno(43)$$
  \smallskip

   \noindent{\it Remark}. 
 Observe that as $\underline x^* =T_2(T_1\underline x^*)$ with $x_3^*=x_4^*=0$, from (40) and (41) with $P=\rho$ we have
 $$\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right]= \cl_2\cdot\cl_1\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right] =\cl\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right].\eqno(44)$$
 Hence when $P=\rho$ one of the eigenvalues of $\cl$ is 1.
 \smallskip

 In general $\cl=\cl_2\cdot\cl_1$, and using (40) and (43) we get
 $$\eqalign{
 &\left(\overline w^*\right)^2\cl =\cr\noalign{\smallskip}
 &\left[\matrix{Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[w(1-P)x_1^*\bigr]^2 & W(1-P)\bigl[P+x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr]\cr
 \noalign{\medskip}
 w(1-P)\bigl[1-x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr] & Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[W(1-P)x_2^*\bigr]^2
 }\right].}\eqno(45)$$
 The external stability of $\underline x^*$ is determined by the eigenvalues of $\cl$, namely the roots of its characteristic polynomial $R(\lambda) =\det(\cl-\lambda\ci)$, with $\ci$ the $2\times 2$ identity matrix. From (45), $R(\lambda)=a_2\lambda^2 +a_1\lambda +a_0$, where
 $$a_0={P^2W^2w^2\over\left(\overline w^*\right)^4},\qquad a_1=-{2PWw +\left(1-P\right)^2\left[Wx_2^* +wx_1^*\right]^2\over \left(\overline w^*\right)^2},\qquad a_2=1.\eqno(46)$$

 We can prove the following result.

 \proclaim Result 7. $\cl$ has two positive eigenvalues and\hfil\break
  {\hglue.82truein}(i) when $P>\rho$ the two eigenvalues are less than 1,\hfil\break
  {\hglue.79truein}(ii) when $P<\rho$ the largest eigenvalue is larger than 1.\par

  \noindent{\it Proof}.  As $\cl$ is a positive matrix by the Perron-Frobenius theory, $\cl$ has a positive eigenvalue, and as $a_0>0$ and $a_2=1$ the product of the two eigenvalues of $\cl$ is positive. Thus in fact $\cl$ has two positive eigenvalues. Let $R(1)=R(1;P)$, then from (46)
  $$R(1;P) = {W^2w^2-\left(\overline w^*\widetilde w^*\right)^2\over\left(\overline w^*\right)^4}P^2 +2P{\left(\widetilde w^*\right)^2 -Ww\over \left(\overline w^*\right)^2} +{\left(\overline w^*\right)^2-\left(\widetilde w^*\right)^2\over\left(\overline w^*\right)^2},\eqno(47)$$
where $\widetilde w^* =Wx_2^* +wx_1^*$.

By (39) $\bigl(\sqrt{Ww}-w\bigr)/(W-w) <x_1^* <{1\over 2}$, from which it is easily seen that
$$\sqrt{Ww}< \overline w^*<\widetilde w^* .\eqno(48)$$

 When $P=\rho$ one of the eigenvalues of $\cl$ is 1; hence $R(1;\rho)=0$. Another root of $R(1;P)=0$ is $\bigl[(\overline w^*)^2 +\overline w^*\widetilde w^*\bigr]/\bigl[Ww +\overline w^*\widetilde w^*\bigr]$, which by (48) is larger than 1.  As $R(1;0)=\bigl[(\overline w^*)^2 -(\widetilde w^*)^2\bigr]/(\overline w^*)^2 <0$ by (48), we deduce that when $0<P<\rho$, $R(1;P)<0$, whereas when $\rho<P<1$, $R(1;P)>0$.
 Hence, when $P<\rho$, $R(1)<0$, and since $a_2=1$, $R(+\infty)>0$, we conclude that $R(\lambda)=0$ has a positive root larger than 1 and the largest positive eigenvalue of $\cl$ is larger than 1.

 When $P>\rho$, we have $R(1)>0$ and also $R(0)=a_0>0$. As $R(\lambda)=0$ has two positive roots and as $a_2>0$, $R(\lambda)$ is convex, either the two positive roots are less than 1 or both larger than one. But the product of the two roots is $P^2W^2w^2/(\overline w^*)^2<1$ by (48); thus when $P>\rho$ the two positive eigenvalues of $\cl$ are less than 1.

It is natural to also ask how the population mean fitness is affected by modification of $\rho$. We have the following result.

\proclaim Result 8. If $W>w$ and $0\le\rho\le 1$, then the mean fitness at the stable equilibrium in the $A1B1$ environment is a decreasing function of $\rho$.\par

\noindent{\it Proof.} In $A1B1$ the stable frequency of phenotype $A$ is by (38)
$$x^* ={1\over 2} -{W+w-Z\over 2(2-\rho)(W-w)},\eqno(49)$$
where $Z =\sqrt{(1-\rho)^2(W-w)^2 +4Ww}$.   The mean fitness at the stable equilibrium  is\hfil\break $\overline w^*\cdot\overline w^{**}$, and as $\overline w^* =\overline w^{**}$  by equation (42), this allows us to reduce the problem to properties of $\overline w^*$.  Now since $W>w$,  $\overline w^*$ is an increasing linear function of $x^*$:
$$\overline w^* =x^*W +(1-x^*)w = x^*(W-w) +w.\eqno(50)$$
Thus $\overline w^*$ is decreasing in $\rho$ if $dx^*/d\rho$ is negative. Using (49),
$${dx^*\over d\rho} =- {(1-\rho)(W-w)\over 2(2-\rho)Z} -{W+w-Z\over 2(2-\rho)(W-w)} = {x^*-{1\over 2}\over 2-\rho} -{(1-\rho)(W-w)\over 2(2-\rho)Z}.\eqno(51)$$
From (39) $0<x^*\le{1\over 2}$ and therefore $dx^*/d\rho <0$, which completes the proof.

 We conclude that in the $A1B1$ selection regime an allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$. Thus in this case, evolution tends to reduce vertical transmission, and hence increase the rate of oblique transmission. The evolutionary dynamics of the reduction in $\rho$ under the $A1B1$ cycling regime are shown in Figure 6, which also illustrates the change in phenotype frequencies over time. In the case of identically distributed random fitnesses $w_A$ and $w_B$, Figure 7 shows  an example of the success of modifiers that reduce $\rho$.

 Figure 3B illustrates the decrease with increasing $\rho$ of the geometric mean fitness (over time) at a polymorphic equilibrium  in the $A1B1$ regime. Figure 8 illustrates an example of the $A1B2$ selection regime: panel A shows the values of selection and $\rho$ that allow polymorphism of phenotypes $A$ and $B$, and Panel B shows that the mean fitness at this equilibrium is decreasing as $\rho$ increases. Figure 7C shows how the mean fitness fluctuates as a function of $\rho$ in the case of symmetric randomly fluctuating fitnesses.

 Thus in  cycling fitness  $A1B1$ and  $A2B2$ regimes and the symmetric random fitness regime of Figure 7, there is a {\sl reduction principle} for the rate of vertical transmission (Altenberg et al.\ 2017). For the $A2B2$ selection regime, numerical analysis (not shown) confirms the reduction principle for $\rho$.
  \medskip
\centerline{[FIGURES 6, 7,  8 HERE]}

 We investigated the dynamics of modifiers of $\rho$ in $AkBk$ regimes for $k\ge 3$. The reduction principle appears to fail for all $k\ge 3$. For each value of $k\ge 3$, there is a value of $\rho$ below which modifiers that increase $\rho$ succeed and above which modifiers that reduce $\rho$ invade. For $k\le8$, the evolutionarily stable value of $\rho$, $\rho^*$  increases rapidly with $k$, but then the increase  becomes much slower. Note that for $k$ large enough, during the $Ak$ part of the cycle, phenotype $A$ becomes extremely close to fixation for even moderate values of $w$ relative to $W$. At this stage, any effect of the $M/m$ difference would be very difficult to detect computationally. The variation of $\rho^*$ with $k$ for selection regimes $AkBk$ is shown in Figure 9 for $k\le50$, $W=1$ with $w=0.1$ and $0.5$.

 The value of $\rho$ that maximizes the geometric mean fitness, denoted by $\hat\rho$, also varies with $k$ in the $AkBk$ regimes. It is somewhat surprising, given the modifier findings for $k\ge 3$, that $\hat\rho$ is actually zero for $k\le10$ and, depending on the fitnesses, is larger than zero for $k>0$. This is shown in Figure 10, where for $W=1$ with $w=0.1$ and $0.5$, $\hat\rho$ is noticeably positive. For $W=1$, $w=0.9$, however, $\hat\rho=0$ maximizes the geometric mean fitness. More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Supplementary Figure S3. 
\medskip
\centerline{[FIGURES 9,10 HERE]}  
 \bigskip

 \noindent{\bf FINITE POPULATION SIZE}
 \smallskip

 In order to include the effect of random drift due to finite population in the above deterministic model, we use the Wright-Fisher model. Let $X_t$ denote the number of individuals with phenotype $A$ in a population of fixed size $N$ at the $t$-th generation, and suppose $X_t=Nx$. Also, let $x'$ represent the frequency of the phenotype $A$ in the infinite population model in the next generation, namely (see (2)),
 $$x'={w_Ax\over\overline w}\rho +(1-\rho)x.\eqno(52)$$
 Then, according to the Wright-Fisher model, $X_{t+1}$, the number of individuals of phenotype $A$ at generation $(t+1)$, is determined by the probability
 $$P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\matrix{N\cr j\cr}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},\eqno(53)$$
 for $j=0,1,2,\dots,N$.
 Thus the fluctuations in the numbers of phenotypes $A$ and $B$ in the  population of size $N$ are generated by the Wright-Fisher Markov chain process where, given that $X_t=Nx$, $X_{t+1}$ has a binomial distribution with parameters $(N,x')$.

 This Markov chain process has two absorbing states, $X_t =N$ and $X_t=0$, corresponding to the two fixations in $A$ and $B$, respectively, and we are interested in the fixation probabilities and the time to fixation of these two absorbing states as functions of the initial frequency $x$ and also of $\rho$, $w_A$, and $w_B$.

 To these ends we use a {\sl diffusion approximation}  to the process $\{X_t\}$. Specifically, we compute the mean $\mu(x)$ and the variance $\sigma^2(x)$ of the change in {\sl one generation} in the frequency of phenotype $A$ given that at the beginning of the generation $X_t=Nx$.

 To compute $\mu(x)$, observe that by (52)   
 $$\eqalignno{
 x' -x &={w_Ax\over\overline w}\rho +(1-\rho)x -x =\rho x\left[{w_A\over \overline w}-1\right]\cr
 &= \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},&(54)\cr}$$
 since $\overline w =w_A x +w_B(1-x)$.
 For the diffusion approximation, it is essential that the differential selection does not have a large effect per individual in each time period  $\Delta t$ $\bigl(\Delta t\simeq{1\over N}\bigr)$. That is, we assume
 $$w_A\simeq 1+{s_1\over N},\qquad w_B\simeq 1+{s_2\over N}.\eqno(55)$$
 Therefore, up to terms of order smaller than ${1\over N}$,
 $${w_A-w_B\over w_Ax +w_B(1-x)}\simeq {{s_1-s_2\over N}\over 1+ {s_1\over N}x +{s_2\over N}(1-x)} \simeq {s_1 -s_2\over N}.\eqno(56)$$
 Let $s=s_1-s_2$, then
 $$x'-x\simeq {1\over N}\rho sx(1-x)\eqno(57)$$
 up to terms of order small than ${1\over N}$. Since one generation corresponds to $\Delta t\simeq {1\over N}$, we  conclude that
 $$\mu(x) =\rho sx(1-x);\quad 0\le x\le 1.\eqno(58)$$

 We now compute $\sigma^2(x)$. From the binomial distribution of $X_{t+1}$ with parameters $N$ and $x'$, the variance of $X_{t+1} -X_t$ given $X_t =Nx$ is $Nx'(1-x')$. Following (52) we have
 $$\eqalignno{
 x'(1-x') &=x\left[{w_A\over \overline w}\rho +(1-\rho)\right](1-x)\left[{w_B\over\overline w}\rho +(1-p)\right]&(59)\cr
 &=x(1-x)\left[{w_Aw_B\over\left(\overline w\right)^2}\rho^2 +{w_A +w_B\over\overline w}\rho(1-\rho) +\left(1-\rho\right)^2\right].&(60)}$$
 Assuming again that $w_A\simeq 1+{s_1\over N}$, $w_B\simeq 1+{s_2\over N}$, up to terms of order smaller than ${1\over N}$ we have
 $${w_A\over\overline w}\simeq 1+{s_1-s_2\over N}(1-x),\qquad {w_B\over\overline w} \simeq 1 +{s_2-s_1\over N}x.\eqno(61)$$
 Consequently
 $$\eqalign{
 {w_A +w_B\over\overline w} &\simeq 2 +{s_1-s_2\over N}(1-2x)\cr\noalign{\smallskip}
 {w_A\cdot w_B\over\left(\overline w\right)^2} &\simeq 1+{s_1-s_2\over N} (1-2x).}
 \eqno(62)$$
 Thus (62) combined with (60) imply that, up to terms of order smaller than ${1\over N}$,
 $$x'(1-x') =x(1-x)\left[1+ \rho{s\over N}(1-2x)\right],\eqno(63)$$
 and the variance, $V$, of the frequency of the phenotype $A$ given that $X_t =Nx$ is
 $$\eqalign{
 V\left({X_{t+1}-X_t\over N}\mid X_t =Nx\right) &={1\over N^2}V\bigl(X_{t+1}\mid X_t=Nx\bigr)\cr \noalign{\smallskip}
 &={x'(1-x')\over N},}\eqno(64)$$
 and in one generation ($\Delta t\simeq{1\over N}$),
 $$\sigma^2(x)\cdot{1\over N} \simeq{1\over N} x(1-x)\left[1 +\rho{s\over N}(1-2x)\right],\eqno(65)$$
 which up to terms of order smaller than ${1\over N}$ gives
 $$\sigma^2(x) =x(1-x).\eqno(66)$$

 We  can compute now the fixation probabilities and the expected time to fixation. Let $u(x)$ be the probability of fixation in $A$ given that the initial frequency  of $A$ is $x$, and let $T(x)$ be the expected time until the population becomes fixed if the initial frequency of $A$ is $x$. Then (following, e.g., Karlin and Levikson 1974) we have
 $$u(x) ={\int_0^xs(\eta)d\eta\over\int_0^1s(\eta)d\eta},\ \ \hbox{\tenrm where}\ \ s(\eta)=\exp\left[-\int^\eta{2\mu(\xi)\over\sigma^2(\xi)}d\xi\right],\eqno(67)$$
 and
 $$T(x)=\int_0^1G(x,\xi)d\xi,\eqno(68)$$
 where
 $$G(x,\xi)=\left\{{2u(x)\bigl[u(1) -u(\xi)\bigr]{C\over s(\xi)\sigma^2(\xi)},\quad 0\le x\le\xi\le 1 \atop
 2\bigl[u(1) -u(x)\bigr]u(\xi){C\over s(\xi)\sigma^2(\xi)},\quad 0\le\xi\le x\le 1}\right.\eqno(69)$$
 and $C=\int_0^1s(\eta)d\eta$.

 In our case, $\mu(\xi)=\rho s\xi(1-\xi)$ and $\sigma^2(\xi)=\xi(1-\xi)$. Therefore $2\mu(\xi)/
 \sigma^2(\xi) =2\rho s$, $s(\eta) =e^{-2\rho s\eta}$, and it can be easily seen that
 $$u(x) ={1-e^{-2\rho sx}\over 1-e^{-2\rho s}}.\eqno(70)$$
 Now
 $$C=\int_0^1s(\eta)d\eta =\int_0^1e^{-2\rho s\eta}d\eta  ={1-e^{-2\rho s}\over 2\rho s},\eqno(71)$$
Therefore  $T(x)$ is given by
 $$T(x)={1-u(x)\over\rho s}\int_0^x{e^{2\rho s\xi}-1\over \xi(1-\xi)}d\xi +{u(x)\over\rho s}\int_x^1{1-e^{-2\rho s(1-\xi)}\over \xi(1-\xi)}d\xi,\eqno(72)$$
 where $u(x)$ is given in (70), and in generations, $T(x)$ is multiplied by $N$.
  Unfortunately the integrals in (72) cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ (Karlin and Levikson 1974), and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.

  It is important to note that the fixation probability $u(x)$ is a monotone increasing function of $\rho$ when $s>0$. In fact we have the following result.

  \proclaim Result 9. When $s>0$ so that the phenotype $A$ is favored, the fixation probability $u(x)$ is monotone increasing in $\rho$.\par

 The proof of Result 9 is in Supplementary Materials SP3.  
  Figure 11 compares the fixation probability and time to fixation derived numerically from  simulating the Wright-Fisher Markov chain with the diffusion-derived values of $u(x)$ and $T(x)$. The fit is seen to be very good. Note that when $N$ is large, the Wright-Fisher model exhibits persistent fluctuation around the deterministic expectation, as shown by the orange diagram in Figure 2B, D, F.
  \medskip

  \centerline{[FIGURE 11 HERE]}

 We can also develop a diffusion approximation for the case of a cycling environment. Suppose that selection changes in cycles of length $n$ such that within the cycle the fitness parameters are $w_A^t$, $w_B^t$ for $t=1,2,\dots,n$. Also let
 $${1\over N}s_t \simeq w_A^t -w_B^t,\qquad S_t =\sum_{i=1}^t s_i,\qquad t=1,2,\dots,n.\eqno(73)$$
 Following Karlin and Levikson (1974) we have the following result.

 \proclaim Result 10. The mean $\mu(x)$ and variance $\sigma^2(x)$ of the change in the frequency of $A$ in one generation for the diffusion approximation in the case of a cycling environment $AkBl$, where $k+l=n$, are
 $$\eqalign{
 \mu(x) &= \rho S_n x(1-x)\cr \sigma^2(x) &= nx(1-x).}\eqno(74)$$

 The proof of Result 10, based on induction on $n$, is given in Supplemental Materials SP4.

 Using the moments in eqs.\ (74), the fixation probability $u(x)$ and the expected time $T(x)$ to fixation can be computed. Comparing to (70) and (72), with $s$  replaced by $S_n/n$, we find
 $$u(x) ={1-e^{-2\rho{S_n\over n}x}\over 1-e^{-2\rho{S_n\over n}}},\eqno(75)$$
 and $T(x)$ can be computed similarly.

 In the special case of the $AkBl$ cycling environment, we write $n=k+l$, and if $w_A=W$, $w_B=w$ for $k$ generations, and $w_A=w$, $w_B=W$ for $l$ generations, we have
 $$S_n=S_{k+l} =(k-l)(W-w).\eqno(76).$$
 Figure 12 illustrates how $(k-l)$, $\rho$, and $(W-w)$, which enter the formula for $u(x)$ in (75), interact to affect fixation probabilities.

 The orange curve in the second column of Figure 2 shows that as $\rho$ increases in the finite-population $A1B1$ cycling regime, the population fluctuates more around the initial frequency of phenotype $A$. For high rates of oblique transmission (lower $\rho$), the first panel in the second column shows that polymorphism is maintained for 20,000 periods with $N=10,000$. 
 \medskip
\centerline{[FIGURE 12 HERE]}  
  \bigskip

 \noindent{\bf DISCUSSION}
 \smallskip

Non-chromosomal modes of phenotypic transmission are receiving increasing attention (Whiten et al. 2017; Jaenisch and Bird 2003; Allis and Jenuwein 2016), especially with respect to their potential role in adaptation and maintenance of diversity (Rivoire and Leibler 2014). Here we have focused on a dichotomous phenotype transmitted through a combination of parental and non-parental transmission. In addition to the roles that these transmission modes play in the dynamics of phenotypic diversity in large and small populations, we have also investigated a genetic model for the evolution of the transmission mode itself.

Our model differs markedly from that of Xue and Leibler (2016), who took the individual phenotypic distribution (i.e., the probability that an individual develops one of a set of phenotypes) to be the inherited trait. In our model, the transmitted trait is the phenotype itself. Thus, with two phenotypic states $A$ and $B$, we track the frequency $x$ of $A$, whereas Xue and Leibler focus on the dynamics of the per-individual probability $\pi_A$ of learning the phenotype $A$. One interpretation of our model is as a mean-value approximation to the model of Xue and Leibler, where $x$, the state in our model, is the average of the population distribution of individual phenotype probabilities.

In a constant environment, the higher the vertical transmission rate $\rho$, the faster is the approach to fixation of the favored phenotype: $A$ if $w_A>w_B$ or $B$ if $w_B>w_A$. Here $1-\rho$, the oblique transmission rate, represents the added chance that an offspring becomes $A$ by learning from the parent's population after learning from its parents who have undergone selection (eq.\ 2). This simple phenotypic model does not allow a polymorphism to be achieved in a constant environment, but with more oblique transmission, approach to fixation is retarded.

 With fluctuating environments, the dynamics of the phenotype frequencies are, in general, much more complicated. In particular, with deterministically cycling symmetric fitness values (the $AkBl$ model), it is impossible for fixation in $A$ and $B$ to both be stable. If $k=l$, for example, neither fixation is stable, and from numerical iteration it appears that there is a simple stable polymorphic equilibrium (with phenotypes $A$ and $B$ present; see Result 3). In the $A1B1$ case this  polymorphism is globally stable. In the $AkBl$ case, bounds on $l/k$ that determine the instability of both fixations, and hence the protection of polymorphism are given by inequalities (20), which depend on both the fitness differences and the rate $\rho$ of vertical transmission.

 In deterministic one-locus, two-allele diploid population genetic models with cycling fitness regimes, Haldane and Jayakar (1963) first showed the relevance of the geometric mean of genotypic fitnesses (cf.\ inequalities (16) and (17)) for the maintenance (or loss) of polymorphism. However, with equal homozygote fitnesses, which alternated in strength as a two-generation cycle (cf.\ $A1B1$), Karlin and Liberman (1975) extended the Haldane-Jayakar results and found conditions under which both allelic fixations and polymorphic equilibrium could all be stable, with the evolution depending on initial allele frequencies, as well as the homozygote fitness differences between alternate generations. Our model appears not to produce such dependence on the initial conditions.

 When the fitnesses $w_A$ and $w_B$ are treated as random variables, rather than varying cyclically, stochastic local stability is the appropriate analog to local stability in the case of cyclic fitness variation. While fixations in $A$ and $B$ cannot both be stable in this case, both may be unstable and a polymorphic distribution may result. The variance of this distribution is greater for larger values of $\rho$. This is because the  stochastic local stability conditions involve $E\bigl\{\log[1-\rho+\rho(w_A/w_B)]\bigr\}$ and the effect of the variance of $(w_A/w_B)$ will clearly increase as $\rho$ increases. In the finite population case, a greater level of vertical transmission has the effect of enhancing the role of random drift by increasing the probability $u(x)$ of fixation and reducing the expected time to fixation.

 We have shown that the rate of vertical transmission tends to decrease when it is under the control of a genetic modifier. This is true in the deterministic case with cycling environments, and also in the random selection case, when the fitnesses of $A$ and $B$ are independent and identically distributed. In light of these results, together with the finite population result that fixation probability is higher, and time to fixation faster with greater $\rho$, we can conclude that increased oblique transmission in our model leads to greater phenotypic diversity; i.e., it is more likely to result in  phenotypic polymorphism. This can be seen in Figure 1, where it is also clear that the existence and stability of a polymorphism is more sensitive to $(1-\rho)$ than to the fitness difference between $A$ and $B$.

 The mean fitness of the population evaluated at a polymorphism of phenotypes $A$ and $B$ is shown to be an increasing function of the level of oblique transmission in the $AkBk$ (Figure 3C) and $A1B2$ (Figure 8) environments. We can compare this result with the asymptotic growth rate (AGR) of Xue and Leibler (2016), whose parameter  $\eta$ is the rate at which an individual learns from its parental lineage. For $k$ small enough, the $AkBk$ environment shows the AGR decreasing with $\eta$, but larger values of $k$ entail that the AGR has a maximum for an intermediate value of $\eta$.

\bigskip \centerline{GET numerical results on $T(x)$ as a function of $\rho$}

\vfil
 \break

 \noindent{\bf References}

\ref Allis, C. D., and T. Jenuwein. 2016. The molecular hallmarks of epigenetic control. {\it Nat.\ Rev.\ Genet.} {\bf 17}: 487--500.

\ref Altenberg, L.,  U. Liberman, and M. W. Feldman. 2017. A unified reduction principle for the evolution of mutation, migration, and recombination. {\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 114}: E2392--E2400.

\ref Aoki, K., and M. W. Feldman. 2014. Evolution of learning strategies in temporally and spatially variable environments: a review of theory. {\it Theor.\ Pop.\ Biol.} {\bf 91}: 3--19.

\ref Aoki, K., J.Y. Wakano, and M. W. Feldman. 2005. The emergence of social learning in
a temporally changing environment: a theoretical model. {\it
Curr.\ Anthropol.} {\bf 46}: 334--340.

\ref Balaban, N. Q., J. Merrin, R. Chait, L. Kowalik, and S. Leibler. 2004. Bacterial persistence as a phenotypic switch. {\it Science} {\bf 305}: 1622--1625.

\ref Bergstrom, C. T., and L. A. Dugatkin. 2012. {\it Evolution}. New York: W. W. Norton \& Co., Inc.

\ref Boyd, R., and P. J. Richerson. 1988. An evolutionary model of social learning: the effect of spatial and temporal variation. Pp. 29--48 {\it in} Zentall, T., and B.G. Galef (eds.) {\it Social Learning}. Hillsdale, NJ: Lawrence Erlbaum Associates, Publishers.

\ref Boyd, R., and P. J. Richerson. 1995. Why does culture increase human adaptability? {\it Ethol.\ Sociobiol.} {\bf 16}: 125--143.

\ref Carja, O., U. Liberman, and  M. W. Feldman. 2014. Evolution in changing environments: modifiers of mutation, recombination, and migration. {\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 111}: 17935--17940.

\ref Cavalli-Sforza, L. L. and M. W. Feldman.  1981. {\it Cultural
Transmission and Evolution:  A Quantitative Approach}.  Princeton, N.J.: Princeton University
Press.

\ref Feldman, M. W. and U. Liberman.  1986. An evolutionary reduction
principle for genetic modifiers.  {\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 83}:
4824--4827.

\ref Feldman, M. W., K. Aoki, and J. Kumm. 1996. Individual
versus social learning:  Evolutionary analysis in a fluctuating environment.   {\it
Anthropological Science} {\bf 104}: 209--231.

\ref Ga\'al, B., J. W. Pitchford, and A. J. Wood. 2010. Exact results for the evolution of stochastic switching in variable asymmetric environments. {\it Genetics} {\bf 184}: 1113--1119.

\ref Haldane, J. B. S., and S. D. Jayakar. 1963. Polymorphism due to selection of varying direction. {\it J. Genet.} {\bf 58}: 237--242.

\ref Jaenisch, R., and A. Bird. 2003. Epigenetic regulation of gene expression: how the genome integrates intrinsic and environmental signals. {\it Nat.\ Genet.} {\bf 33}: 245--254.

\ref Karlin, S., and B. Levikson. 1974. Temporal fluctuation in selection intensities: case of small population size. {\it Theor.\ Popul.\ Biol.} {\bf 6}: 383--412.

\ref Karlin, S., and U. Lieberman. 1974. Random temporal variation in selection intensities: case of large population size. {Theor.\ Popul. Biol.} {\bf 6}: 355--382.

\ref Karlin, S., and U. Liberman. 1975. Random temporal variation in selection intensities: one-locus two-allele model. {\it J. Math.\ Biol.} {\bf 2}: 1--17.

\ref Liberman, U., J. Van Cleve, and M. W. Feldman. 2011. On the evolution of mutation in changing environments: recombination and phenotypic switching. {\it Genetics}  {\bf 187}: 837--851.

\ref Lycett, S. J., and J. A. J. Gowlett. 2008. On questions surrounding the Acheulean `tradition'. {\it World Archaeol.} {\bf 40}: 295--315.

\ref Rivoire, O., and S. Leibler. 2014. A model for the generation and transmission of variations in evolution. {\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 111}: E1940--E1949.

\ref Rogers, A. R. 1988. Does biology constrain culture? {\it Am.\ Anthropol.} {\bf 90}: 819--831.

\ref Rosenberg, E., and I. Zilber-Rosenberg. 2016. Microbes drive evolution of animals and plants: the hologenome concept. {\it mBio} {\bf 7}: e01395-15.

\ref Salath\'e, M., J. Van Cleve, and M. W. Feldman. 2009. Evolution of stochastic switching rates in asymmetric fitness landscapes. {\it Genetics} {\bf 182}: 1159--1164.

\ref Thattai, M., and A. van Oudenaarden. 2004. Stochastic gene expression in fluctuating environments. {\it Genetics} {\bf 167}: 523--530.

\ref Theis, K. R., N. M. Dheilly, J. L. Klassen, R. M. Brucker, J. F. Baines, T. C. G. Bosch, J. F. Cryan, S. F. Gilbert, C. J. Goodnight, E. A. Lloyd, J. Sapp, P. Vandenkoornhuyse, I. Zilber-Rosenberg, E. Rosenberg, and S. R. Bordenstein. 2016. Getting the hologenome concept right: an eco-evolutionary framework for hosts and their microbiomes. {\it mSystems} {\bf 1}: e00028-16.

\ref van Opstal, E. J., and S. R. Bordenstein. 2015. Rethinking heritability of the microbiome. {\it Science} {\bf 349}: 1172--1173.

\ref Wakano, J. Y., and K. Aoki. 2007. A mixed strategy model for the emergence and intensification of social learning in a periodically changing natural environment. {\it Theor.\ Popul.\ Biol.} {\bf 70}: 486--497.

\ref Wakano, J. Y., K. Aoki, and M. W. Feldman. 2004. Evolution of social learning: a
mathematical analysis. {\it Theor.\ Pop.\ Biol.} {\bf 66}: 249--258.

\ref Whiten, A., F. Ayala, M. W. Feldman, and K. N. Laland. 2017. The extension of biology through culture. {\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 114}: 7775--7781.

\ref Xue, B.-K., and S. Leibler. 2016. Evolutionary learning of adaptation to varying environments through a transgenerational feedback. {\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 113}: 11266--11271.
 \vfil
 \break

 \centerline{\bf SUPPLEMENTAL MATERIALS}
 \bigskip

 \noindent{\bf SP1. Proof of Result 5}
 \medskip

 \noindent{\it Proof of Result 5}. Rewrite recursion (31) as
 $${x_t+1\over x_t} =(1+\rho s_t)\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right].\eqno(\hbox{\tenrm S}1)$$
 Then
 $$\log x_{t+1} -\log x_t =\log(1+\rho s_t) +\log\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right].\eqno(\hbox{\tenrm S}2)$$
 Summation yields
 $${1\over t}\left[\log x_t-\log x_0\right] ={1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n) +{1\over t}\sum_{n=0}^{t-1}\log\left[1-x_n{\rho s_n(1+s_n)\over (1+\rho s_n)(1+x_ns_n)}\right].\eqno(\hbox{\tenrm S}3)$$
Let $\mu=E\left[\log(1+\rho s_t)\right]$. As $\{s_t\}_{t\ge 0}$ are independent and identically distributed random variables, the {\sl strong law of large numbers} applies and
$$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)=\mu\eqno(\hbox{\tenrm S}4)$$
almost surely.

Let $\zeta$ be such that ${1\over t}\sum_{n=0}^{t-1}\log[1+\rho s_n(\zeta)]=\mu$ and assume that $\lim_{t\to\infty}x_t(\zeta)=0$. As the random variables $\{s_t\}_{t\ge 0}$ are uniformally bounded,
$$x_t(\zeta){\rho s_t(\zeta)[1+s_t(\zeta)]\over [1+\rho s_t(\zeta)][1+x_t(\zeta)s_t(\zeta)]}\;\;\harr{}{t\to\infty}\;\;0\eqno(\hbox{\tenrm S}5)$$
and
$$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log\left[1-x_n(\zeta){\rho x_n(\zeta)[1+s_n(\zeta)]\over [1+\rho s_n(\zeta)][1+x_t(\zeta)s_n(\zeta)]}\right]=0.\eqno(\hbox{\tenrm S}6)$$
Thus (S3) implies that
$$\lim_{t\to\infty}{1\over t}\left[\log x_t(\zeta) -\log x_0(\zeta)\right]=\mu.\eqno(\hbox{\tenrm S}7)$$
If $\mu =E\left[\log(1+s_t)\right]>0$, then from (S7) we deduce that $\lim_{t\to\infty}x_t(\zeta)=\infty$, a contradiction.
Therefore when $\mu>0$, $P\left(\lim_{t\to\infty}x_t=0\right)=0$, and fixation of $B$ ($x^*=0$) is stochastically locally unstable.

 Thus by Result 5, for $x^*=0$ to be stochastically locally stable it is necessary that $E[\log(1+\rho s_t)]\le 0$. In fact, the strict inequality is sufficient.

 Figure 4 presents a numerical example of the dynamics of recursion (31) with a specific random selection coefficient $s_t$.
\bigskip
\bigskip

 \noindent{\bf SP2. Proof of Result 6}
 \medskip

Let $\mu=E[\log(1+\rho s_t)]$. Then as $\{s_t\}_{t\ge 0}$ are independent and identically distributed random variables, the strong law of large number applies and almost surely
 $$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)=\mu<0.\eqno(\hbox{\tenrm S}8)$$
 Appealing to the Egoroff Theorem, for any $\var>0$ there exists $T$ such that 
 $$P\left({1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)<{\mu\over 2}\ \hbox{\tenrm for all}\ t\ge T\right)\ge 1-\var.\eqno(\hbox{\tenrm S}9)$$
 As $0\le \rho\le 1$ and the $\{s_t\}_{t\ge 0}$ are uniformly bounded, we can find a $\delta'>0$ such that
 $$x_t<\delta'\Longrightarrow\left|\log\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right]\right|<-{\mu\over 4}.\eqno(\hbox{\tenrm S}10)$$
 Also, as $0\le x_t\le 1$ for all $t$,
 $$x_{t+1}=x_t{1+\rho s_t +x_t(1-\rho)s_t\over 1+x_ts_t}< Kx_t,\eqno(\hbox{\tenrm S}11)$$
 where $K$ is independent of $t$. It follows that there exists a $\delta$ with $0<\delta<\delta'$ such that
 $$x_o<\delta\Longrightarrow x_t<\delta'\ \hbox{\tenrm for all}\ t=0,1,2,\dots,T-1.\eqno(\hbox{\tenrm S}12)$$
 Let $\xi$ be a realization of the evolutionary process such that
 $${1\over t}\sum_{n=0}^{t-1}\log[1+\rho s_n(\xi)]<{\mu\over 2}\ \hbox{\tenrm for all}\  t\ge T\eqno(\hbox{\tenrm S}13)$$
 and assume $x_0<\delta$. Then
 $$\eqalign{
 {1\over T}&[\log x_T(\xi) -\log x_0(\xi)] \cr
 &={1\over T}\sum_{n=0}^{T-1}\log[1+\rho s_n(\xi)]+{1\over T}\sum_{n=0}^{T-1}\log[1-x_n(\xi){\rho s_n(\xi)[1+s_n(\xi)]\over [1+\rho s_n(\xi)][1+x_n(\xi)s_n(\xi)]}\cr\noalign{\smallskip}
 &<{\mu\over 2}- {\mu\over 4} ={\mu\over 4}<0, }\eqno(\hbox{\tenrm S}14)$$
 and therefore $x_T(\xi) <x_0(\xi) <\delta'$. Invoking induction we get that for $t\ge T$
 $${1\over t}\log{x_t(\xi)\over x_0}\le {\mu\over 4},\eqno(\hbox{\tenrm S}15)$$
 or for all $t\ge T$
 $$x_t(\xi)\le x_0\exp\left({\mu\over 4}t\right).\eqno(\hbox{\tenrm S}16)$$

 As $\mu<0$, this implies that $x_t(\xi)\,\harr{}{t\to\infty}\,0$. Therefore we have shown that for given $\var>0$ there is a $\delta>0$ such that if $0<x_0<\delta$, then $P\left(\lim_{t\to\infty}x_t =0\right)\ge 1-\var$, and therefore $x^*=0$, the fixation in $B$ is stochastically locally stable. The second statement of Result 5 follows from the convexity of the log function and Jensen's inequality.
\bigskip
\bigskip

 \noindent{\bf SP3. Proof of Result 9}
 \medskip

Without loss of generality and for the ease of representation, we will show that for $t>0$, 
  $$v (x;t) ={1-e^{-tx}\over 1-e^{-t}}\eqno(\hbox{\tenrm S}17)$$
  is monotone increasing as a function of $t$. Observe that
  $${\partial v\over\partial t} = {\left(1-e^{-t}\right)xe^{-tx} -\left(1-e^{-tx}\right)e^{-t}\over \left(1-e^{-t}\right)^2}.\eqno(\hbox{\tenrm S}18)$$

   \noindent For the monotonicity we have to show that
  $$f(x;t) =\left(1-e^{-t}\right)xe^{-tx} -\left(1 -e^{-tx}\right)e^{-t} \ge 0\eqno(\hbox{\tenrm S}19)$$ 
  when $t>0$ and $0\le x\le 1$. Note that $f(0;t)=0$ and $f(1;t)=0$. Also
  $${\partial f\over\partial x} =\left(1-e^{-t}\right)\left(e^{-tx}-txe^{-tx}\right) -te^{-tx}e^{-t},\eqno(\hbox{\tenrm S}20)$$
  or
  $${\partial f\over\partial x} =e^{-tx}\left[\left(1-e^{-t}\right)\left(1-tx\right) -te^{-t}\right] =e^{-tx}g(x;t),\eqno(\hbox{\tenrm S}21)$$
 say, where for fixed $t$, $g(x;t)$ is a linear function of $x$, which vanishes at $x_0 =(1-e^{-t} -te^{-t})/t(1-e^{-t})$. If $t>0$, $e^t >1+t$, so $1>e^{-t}(1+t)$ and $x_0>0$. Also if $t>0$, $e^{-t}>1-t$, and so $1-e^{-t} -te^{-t} <t(1-e^{-t})$ and $x_0<1$. Since $g(0,t) =1 -e^{-t} -te^{-t} >0$ and $g(1;t) =(1-e^{-t})(1-t) -te^{-t} <0$ for $t>0$, we deduce that ${\partial f\over\partial x}(x,t)>0$ for $ 0<x<x_0$ and ${\partial f\over\partial x}(x,t)<0$ for $x_0<x<1$ for all $t>0$. These facts, combined with $f(0,t) =f(1;t) =0$, prove that $f(x;t)\ge 0$ for $0\le x\le 1$ (in fact, $f(x;t)>0$ for $0<x<1$), and inequality (S19) is satisfied as desired.
\bigskip
\bigskip

 \noindent{\bf SP4. Proof of Result 10}
 \medskip

 The proof is based on induction on $n$, where in order to prove (74), we show that if $X_t$ is the number of individuals with phenotype $A$ at stage $t$ of the cycle, and $x$ is the initial frequency of $A$, then
 $$E\left({X_t\over N} -x\right) \simeq {1\over N}\rho S_t x(1-x),\qquad V\left({X_t\over N}\right)\simeq {1\over N}tx(1-x),\eqno(\hbox{\tenrm S}22)$$
 where $N$ is the size of the population. When $t=1$, (S22) coincides with the constant environment case. Assuming (S22), we go to $t+1$. Now $X_{t+1}$ given $X_t=Ny$ has a binomial distribution with parameters $(N,y')$. Hence
 $$E\left({X_{t+1}\over N} -{X_t\over N} \mid X_t=Ny\right) =y'-y.\eqno(\hbox{\tenrm S}23)$$
 Following (57), $y'-y\simeq (1/N)\rho s_{t+1}y(1-y)$, and so
 $$E\left({X_{t+1}\over N} -{X_t\over N}\mid X_t\right)\simeq {1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right).\eqno(\hbox{\tenrm S}24)$$
 Observe that
 $$\eqalign{
 E\left[{X_t\over N}\left(1-{X_t\over N}\right)\right] &= E\left({X_t\over N}\right) -E\left[\left({X_t\over N}\right)^2\right]\cr\noalign{\smallskip}
    &= E\left({X_t\over N}\right) -V\left({X_t\over N}\right) -\left[E\left({X_t\over N}\right)\right]^2.\cr}\eqno(\hbox{\tenrm S}25)$$
 By the induction assumption, $V(X_t/N)\simeq (1/N)tx(1-x)$, and ignoring terms of order $1/N^2$ we have
 $$E\left({X_{t+1}\over N} -{X_t\over N}\right)\simeq {1\over N}\rho s_{t+1} E\left({X_t\over N}\right)\left[1-E\left({X_t\over N}\right)\right].
\eqno(\hbox{\tenrm S}26)$$
Applying (S22) we have
$$\eqalign{
E\left({X_t\over N}\right)  &\simeq x+{1\over N}\rho S_tx(1-x)\cr\noalign{\smallskip}
1-E\left({X_t\over N}\right) &\simeq 1-x-{1\over N}\rho S_tx(1-x),} \eqno(\hbox{\tenrm S}27)$$
and ignoring terms $O(1/N^2)$, we find
$$E\left({X_{t+1}\over N} -{X_t\over N}\right) \simeq {1\over N}\rho s_{t+1}x(1-x).\eqno(\hbox{\tenrm S}28)$$
Thus
$$\eqalign{
E\left({X_{t+1}\over N} -x\right) &= E\left({X_{t+1}\over N} -{X_t\over N}\right) +E\left({X_t\over N} -x\right)\cr\noalign{\smallskip}
&\simeq {1\over N}\rho s_{t+1}x(1-x) +{1\over N}\rho S_t x(1-x),}\eqno(\hbox{\tenrm S}29)$$
and since $S_t +s_{t+1} =S_{t+1}$,
 $$E\left({X_{t+1}\over N} -x\right)\simeq	{1\over N}\rho S_{t+1}x(1-x)\eqno(\hbox{\tenrm S}30)$$
 as desired.

 We now compute $V(X_{t+1}/N)$ using the induction assumption and the formula
 $$V\left({X_{t+1}\over N}\right) =E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] +V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right],\eqno(\hbox{\tenrm S}31)$$
 where by (S22)
 $$E\left({X_{t+1}\over N}\mid X_t\right)\simeq {X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\eqno(\hbox{\tenrm S}32)$$
 and
 $$V\left({X_{t+1}\over N}\mid X_t\right)\simeq{1\over N}{X_t\over N}\left(1-{X_t\over N}\right).\eqno(\hbox{\tenrm S}33)$$
 Here we used the fact that by (63) $y'(1-y')\simeq y(1-y)$. Now
 $$E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] \simeq {1\over N}E\left[{X_t\over N}\left(1-{X_t\over N}\right)\right]\simeq {1\over N}x(1-x),\eqno(\hbox{\tenrm S}34)$$
 where we use the same computations as led from (S25) to (S28).
 $$V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right] =V\left[{X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right].\eqno(\hbox{\tenrm S}35)$$
 Since $(X_t/N)\bigl[1-(X_t/N)\bigr]$ is a random variable taking values in [0,1],  its variance is less than 1/4 and
 $$V\left[{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right] \le {1\over 4N^2}\rho^2 s_{t+1}^2.\eqno(\hbox{\tenrm S}36)$$
 We ignore terms $O(1/N^2)$ so that the random variable $(1/N)\rho s_{t+1}(X_t/N)\bigl[1-(X_t/N)\bigr]$ is almost constant. As a result,
 $$V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right]\simeq V\left({X_t\over N}\right) \simeq {1\over N}tx(1-x),\eqno(\hbox{\tenrm S}37)$$
 by the induction assumption. Combining (S34) and (S37) gives
 $$V\left({X_{t+1}\over N}\right)\simeq {1\over N}x(1-x) +{1\over N}tx(1-x) ={1\over N}(t+1)x(1-x)\eqno(\hbox{\tenrm S}38)$$
 as expected.

 \end\bye

### 2017-08-24 ms/figures.tex
updated fig 4

@@
 \includegraphics{../figures/stochastic_env_x_t.pdf}
-\caption{Stochastic local stability. The figure shows the frequency of
-phenotype \emph{A} after \(10^6\) generations in a very large population 
-evolving in a stochastic environment (eq.\ 31). The
-fitness of phenotypes \emph{A} and \emph{B} are \(1+s_t\) and \(1\),
-where \(s_t\) is \emph{s} with probability \emph{p} and \emph{-s} with
-probability \(1-p\). The white line marks combinations of \emph{p}
-and \emph{s} for which \(E[\log{(1+\rho s_t)}]=0\); according
-to Result 6, below this line fixation of phenotype \emph{B} (\(x^*=0\)) will be
-stochastically locally stable. Parameters: initial frequency of \(A\) is \(x_0=0.1\); vertical transmission rate is \(\rho=0.1\).}\label{fig:stochastic_env_x_t}
+\caption{Stochastic local stability. 
+The figure shows the frequency of phenotype \emph{A} after \(10^6\) generations in a very large population evolving in a stochastic environment (eq.\ 31).
+The fitness of phenotypes \emph{A} and \emph{B} are \(w_A=1+s_t\) and \(w_B=1\),
+where \(s_t\) is \emph{s} with probability \emph{p} and \emph{-s} with probability \(1-p\).
+The gray lines mark combinations of \emph{p} and \emph{s} for which \(E[\log{(1-\rho+\rho\frac{w_A}{w_B})}]=0\) and \(E[\log{(1-\rho+\rho\frac{w_B}{w_A})}]=0\).
+According to Result 6, between these lines fixation of either phenotype will be stochastically locally unstable, and indeed we have a stationary polymorphism between the lines, with phenotype frequencies neither 0 or 1.
+Parameters: initial frequency of \(A\) is \(x_0=1/10,000\) and the vertical transmission rate is \(\rho=0.1\).}\label{fig:stochastic_env_x_t}
 \end{figure}

### 2017-08-25 ms/ms.tex
aug 25 update

@@
 \vfil
-\centerline{August 23, 2017, v.1}
+\centerline{August 24, 2017, v.1}
 
@@
 Using conventional modifier theory (Feldman and Liberman 1986), we show that in a cyclic selection regime with cycles of periods 1 or 2, an allele reducing the rate of vertical transmission is expected to increase in frequency when rare and in so doing to increase the mean fitness of the population.
-However, for cycles of greater length, interesting non-monotonicities emerge, both in the uninvadable rate of vertical transmission, and the rate that maximizes geometric mean fitness.
-We develop the models in very large populations with random fitnesses and also in the case where drift occurs via sampling from generation to generation in a finite population.
+However, for cycles of greater length, interesting non-monotonicities emerge, both in the uninvadable rate of vertical transmission, and the rate that maximizes the geometric time average of the population mean fitness, which we will refer to as the ``geometric mean fitness.'' We develop the models in very large populations with cyclic selection and with random fitnesses and also in the case where drift occurs via sampling from generation to generation in a finite population.
 
@@
 
-\proclaim Result 1. If $0<\rho\le 1$ and both $w_A$ and $w_B$ are positive with $w_A\ne w_B$, then fixation in the phenotype $A(B)$ is globally stable when $w_A>w_B$ $(w_A<w_B)$.\par
+\proclaim Result 1. If $0<\rho\le 1$ and both $w_A$ and $w_B$ are positive with $w_A\ne w_B$, then fixation in the phenotype $A\ (B)$ is globally stable when $w_A>w_B$ $(w_A<w_B)$.\par
 
@@
 The equilibrium equation is $x=F_B\bigl(F_A(x)\bigr)$, which reduces to a fourth degree polynomial equation in $x$.
-Since the two fixations in $B$ and $A$ are equilibria corresponding to the two solutions $x=0$ and $x=1$, the other equilibria correspond to solutions of a quadratic equation $G(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with
+Since the two fixations in $B$ and $A$ are equilibria corresponding to the two solutions $x=0$ and $x=1$, the other equilibria correspond to solutions of a quadratic equation $Q(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with
 $$\alpha_2=1,\quad \alpha_1={W+w\over (2-\rho)(W-w)}-1,\quad \alpha_0={-w\over (2-\rho)(W-w)}.\eqno(24)$$
 As $0<\rho<1$ and $0<w<W$ we have
-$$G(0) = {-w\over (2-\rho)(W-w)}<0\eqno(25)$$
+$$Q(0) = {-w\over (2-\rho)(W-w)}<0\eqno(25)$$
 and
-$$G(1) ={W\over (2-\rho)(W-w)} >0.\eqno(26)$$
-Also, as $\alpha_2=1$ and $\alpha_0<0$, the quadratic equation $G(x)=0$ has two real roots, one negative and one positive $x^*$ satisfying $0<x^*<1$.
+$$Q(1) ={W\over (2-\rho)(W-w)} >0.\eqno(26)$$
+Also, as $\alpha_2=1$ and $\alpha_0<0$, the quadratic equation $Q(x)=0$ has two real roots, one negative and one positive $x^*$ satisfying $0<x^*<1$.
 The latter determines a unique polymorphism.
@@
 For more general cyclic fitness regimes, the polynomial that gives the equilibria is of higher order, and it is conceivable that more than one stable polymorphism could exist for given values of $\rho$, $W$, and $w$.
-We investigated this numerically, and as shown in Figure S1, the $A1B2$ regime also appears to allow a single stable polymorphism.
+We investigated this numerically, and as shown in Supplemental Figure S1, the $A1B2$ regime also appears to allow a single stable polymorphism.
 For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplemental Figure S2 shows that only a single stable equilibrium appears to exist.
@@
 \noindent As $\{x_t\}$ for $t=0,1,2,\dots$ is a sequence of random variables, the notion of stability of the two fixation states needs clarification.
-Following Karlin and Lieberman (1974, 1975) we make the following definition.
+Following Karlin and Lieberman (1974) and Karlin and Liberman (1975) we make the following definition.
 
@@
 
- \proclaim Result 5. Suppose $E\left[\log (1+\rho s_t)\right]>0$. Then $x^*=0$, the fixation in $B$, is not stochastically locally stable. In fact $P\left(\lim_{t\to\infty}x_t=0\right)=0$.
+ \proclaim Result 5. Suppose $E\left[\log (1+\rho s_t)\right]>0$. Then $x^*=0$, the fixation of phenotype $B$, is not stochastically locally stable. In fact $P\left(\lim_{t\to\infty}x_t=0\right)=0$.
 
-  \proclaim Result 6. Suppose $E[\log(1+\rho s_t)]<0$. Then $x^*=0$ is stochastically locally stable. In particular, if $E(s_t)\le 0$, $x^*=0$ is stochastically locally stable.
+  \proclaim Result 6.
+Suppose $E[\log(1+\rho s_t)]<0$.
+Then $x^*=0$, the fixation of phenotype $B$, is stochastically locally stable.
+In particular, if $E(s_t)\le 0$, $x^*=0$ is stochastically locally stable.
 
@@
  Using the general notation for the fitness parameters $w_A$ and $w_B$, the stochastic local stability of the $B$ fixation is determined by the sign of $E\left[\log\left(1-\rho+\rho{w_A\over w_B}\right)\right]$ and that of the $A$ fixation by the sign of $E\left[\log\left(1-\rho+\rho{w_B\over w_A}\right)\right]$.
-For example, if the first sign  is negative, fixation in $B$ is stochastically locally stable, and when it is positive, with probability one convergence to fixation in $B$ does not occur.
+For example, if the first sign  is negative, fixation in $B$ is stochastically locally stable, and when it is positive, with probability one convergence to fixation in phenotype $B$ does not occur.
 It is also true  that if $E(w_A/w_B)\le 1$, then fixation of $B$ is stochastically locally stable.
@@
 It is possible, however, that neither fixation is  stochastically locally stable, in which case we expect the population to converge to a polymorphic distribution.
-Figure 4 illustrates how the properties of $s_t$ in recursion (31) affect the frequency of phenotype $A$, and in particular the stochastic local stability of fixation in $B$.
+Figure 4 illustrates how the properties of $s_t$ in recursion (31) affect the frequency of phenotype $A$, and in particular the stochastic local stability of fixation in phenotype $B$.
 Figure 5 shows the dynamics of the frequency of $A$ in a case where $w_A$ and $w_B$ are identically distributed and independent; in this case the expectation of the stationary distribution is 1/2, and its variance increases as $\rho$ increases.
@@
 
- Suppose that the vertical transmission rate is determined by a genetic locus with two possible alleles $m$ and $M$.
+ Suppose that the vertical transmission rate is controlled by a genetic locus with two possible alleles $m$ and $M$.
 Let the vertical transmission rates determined by $m$ and $M$ be $\rho$ and $P$, respectively.
@@
 $$\overline w =w_A(x_1+x_3) +w_B(x_2+x_4).\eqno(36)$$
-Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} to invasion by allele $M$.
+Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} (e.g., Feldman and Liberman 1986; Altenberg et al.\ 2017) to invasion by allele $M$.
 
@@
 
-For the $A1B1$ case, when only the $m$ allele is present with associated rate $\rho$, with $0<\rho<1$ and $0<w<W$, a unique stable equilibrium $\underline x^* =(x_1^*,x_2^*,0,0)$ exists.
+For the $A1B1$ case, when only the $m$ allele is present with associated rate $\rho$, with $0<\rho<1$ and $0<w<W$, a unique stable equilibrium $\underline x^* =(x_1^*,1-x_1^*,0,0)$ exists.
 $x_1^*$ is the only positive root of the quadratic equation $G(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with $\alpha_2,\alpha_1,\alpha_0$ specified in (24).
@@
 $$x_1^* ={1\over 2} -{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}\over 2\cdot(2-\rho)(W-w)},\eqno(38)$$
- and it is easily seen that
+ and it can be seen that
  $${\sqrt{Ww}-w\over W-w}<x_1^* <{1\over 2}.\eqno(39)$$
 
- The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ derived from (35) and given by
+ The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ near $x^*$, which is derived from (35) and given by
  $$\overline w^*\cl_1 =\left[\matrix{W\bigl[(1-P)x_1^* +P\bigr] & w(1-P)x_1^*\cr\noalign{\medskip}
@@
  w(1-P)x_2^{**} & W\bigl[(1-P)x_2^{**} +P\bigr]\cr}\right],\eqno(41)$$
-where $\underline x^{**} =T_1\underline x^*$ and
+where $\underline x^{**} =T_1\underline x^*$, $x_1^{**}=x_2^*=1-x_1^*$, $x_2^{**}=x_1^*$,  and
 $$\overline w^* =Wx_1^* +wx_2^*,\qquad \overline w^{**} =wx_1^{**} +Wx_2^{**}.\eqno(42)$$
@@
  \proclaim Result 7. $\cl$ has two positive eigenvalues and\hfil\break
-  {\hglue.82truein}(i) when $P>\rho$ the two eigenvalues are less than 1,\hfil\break
-  {\hglue.79truein}(ii) when $P<\rho$ the largest eigenvalue is larger than 1.\par
+  {\hglue.82truein}(i) when $P>\rho$ the two eigenvalues are less than $1$,\hfil\break
+  {\hglue.79truein}(ii) when $P<\rho$ the largest eigenvalue is larger than $1$,\hfil\break
+  {\hglue.76truein}(iii) when $P=\rho$, the largest eigenvalue is $1$.\par
 
@@
 $$x^* ={1\over 2} -{W+w-Z\over 2(2-\rho)(W-w)},\eqno(49)$$
-where $Z =\sqrt{(1-\rho)^2(W-w)^2 +4Ww}$.
+where $Z =\sqrt{(1-\rho)^2(W-w)^2 +4Ww}>0$.
 The mean fitness at the stable equilibrium  is\hfil\break $\overline w^*\cdot\overline w^{**}$, and as $\overline w^* =\overline w^{**}$  by equation (42), this allows us to reduce the problem to properties of $\overline w^*$.
@@
 The reduction principle appears to fail for all $k\ge 3$.
-For each value of $k\ge 3$, there is a value of $\rho$ below which modifiers that increase $\rho$ succeed and above which modifiers that reduce $\rho$ invade.
+For each value of $k\ge 3$, there is a value of $\rho$ below which modifiers that increase $\rho$ invade and above which modifiers that reduce $\rho$ invade.
 For $k\le8$, the evolutionarily stable value of $\rho$, $\rho^*$  increases rapidly with $k$, but then the increase  becomes much slower.
@@
 At this stage, any effect of the $M/m$ difference would be very difficult to detect computationally.
-The variation of $\rho^*$ with $k$ for selection regimes $AkBk$ is shown in Figure 9 for $k\le50$, $W=1$ with $w=0.1$ and $0.5$.
+The variation of $\rho^*$ with $k$ for selection regimes $AkBk$ is shown in Figure 9 for $k\le50$, $W=1$ with $w=0.1$, $0.5$, and $0.9$.
 
  The value of $\rho$ that maximizes the geometric mean fitness, denoted by $\hat\rho$, also varies with $k$ in the $AkBk$ regimes.
-It is somewhat surprising, given the modifier findings for $k\ge 3$, that $\hat\rho$ is actually zero for $k\le10$ and, depending on the fitnesses, is larger than zero for $k>0$.
+It is somewhat surprising, given the modifier findings for $k\ge 3$, that $\hat\rho$ is actually zero for $k\le11$ and, depending on the fitnesses, is larger than zero for $k>11$.
 This is shown in Figure 10, where for $W=1$ with $w=0.1$ and $0.5$, $\hat\rho$ is noticeably positive.
@@
  $$x'={w_Ax\over\overline w}\rho +(1-\rho)x.\eqno(52)$$
- Then, according to the Wright-Fisher model, $X_{t+1}$, the number of individuals of phenotype $A$ at generation $(t+1)$, is determined by the probability
+ Then, according to the Wright-Fisher model (Ewens 2004), $X_{t+1}$, the number of individuals of phenotype $A$ at generation $(t+1)$, is determined by the probability
  $$P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\matrix{N\cr j\cr}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},\eqno(53)$$
@@
 However, with equal homozygote fitnesses, which alternated in strength as a two-generation cycle (cf.\ $A1B1$), Karlin and Liberman (1975) extended the Haldane-Jayakar results and found conditions under which both allelic fixations and polymorphic equilibrium could all be stable, with the evolution depending on initial allele frequencies, as well as the homozygote fitness differences between alternate generations.
-Our model appears not to produce such dependence on the initial conditions.
+Our haploid model appears not to produce such dependence on the initial conditions.
 
  When the fitnesses $w_A$ and $w_B$ are treated as random variables, rather than varying cyclically, stochastic local stability is the appropriate analog to local stability in the case of cyclic fitness variation.
-While fixations in $A$ and $B$ cannot both be stable in this case, both may be unstable and a polymorphic distribution may result.
+While fixations in phenotypes $A$ and $B$ cannot both be stable in this case, both may be unstable and a polymorphic distribution may result.
 The variance of this distribution is greater for larger values of $\rho$.
 This is because the  stochastic local stability conditions involve $E\bigl\{\log[1-\rho+\rho(w_A/w_B)]\bigr\}$ and the effect of the variance of $(w_A/w_B)$ will clearly increase as $\rho$ increases.
-In the finite population case, a greater level of vertical transmission has the effect of enhancing the role of random drift by increasing the probability $u(x)$ of fixation and reducing the expected time to fixation.
+In the finite population case, a greater level of vertical transmission makes selection more effective, increasing the probability $u(x)$ of fixation and reducing the expected time to fixation.
 
  We have shown that the rate of vertical transmission tends to decrease when it is under the control of a genetic modifier.
-This is true in the deterministic case with cycling environments, and also in the random selection case, when the fitnesses of $A$ and $B$ are independent and identically distributed.
+This is true in the deterministic case with cycling environments, and also in the random selection case, when the fitnesses of $A$ and $B$ are identically distributed and independent between generations.
 In light of these results, together with the finite population result that fixation probability is higher, and time to fixation faster with greater $\rho$, we can conclude that increased oblique transmission in our model leads to greater phenotypic diversity; i.e., it is more likely to result in  phenotypic polymorphism.
@@
 
- The mean fitness of the population evaluated at a polymorphism of phenotypes $A$ and $B$ is shown to be an increasing function of the level of oblique transmission in the $AkBk$ (Figure 3C) and $A1B2$ (Figure 8) environments.
+ The mean fitness of the population evaluated at a polymorphism of phenotypes $A$ and $B$ is shown to be an increasing function of the level of oblique transmission in the $A1B1$ (Figure 3B, Figures 6C, 6F, 6I) and $A1B2$ (Figure 8B) environments.
 We can compare this result with the asymptotic growth rate (AGR) of Xue and Leibler (2016), whose parameter  $\eta$ is the rate at which an individual learns from its parental lineage.
 For $k$ small enough, the $AkBk$ environment shows the AGR decreasing with $\eta$, but larger values of $k$ entail that the AGR has a maximum for an intermediate value of $\eta$.
+[FINISH WRITEUP]
 
@@
 Press.
+
+\ref Ewens, W. 2004. {\it Mathematical Population Genetics}, 2nd Edition. New York: Springer.
 
@@
  \break
+ \centerline{\bf TABLE 1}
+ \smallskip
+
+ $$\vbox{\halign{\hfil #&&\qquad\hfil#\hfil\cr
+ &&&&evolutionarily&fitness\cr
+ &&&&stable&optimal\cr
+ &$k$&$l$&$w$&$\rho^*$&$\hat\rho$\cr
+ 0&3&10&0.1&0.5286296&0.01010101\cr
+ 1&5&30&0.1&0.24641176&0.01010101\cr
+ 2&3&10&0.5&0.38577372&0.02020202\cr
+ 3&5&30&0.5&0.46443287&0.01010101\cr
+ 4&3&10&0.9&0.9681569&0.05050505\cr
+ 5&5&30&0.9&0.64959375&0.02020202\cr
+6&12&12&0.1&0.84855774&0.24242424\cr
+7&30&30&0.5&0.96077135&0\cr
+8&1&1&0.1&0&0\cr
+9&1&2&0.1&0&0.01010101\cr
+10&2&2&0.1&0.00015691&0\cr
+11&1&1&0.5&0&0\cr
+12&1&2&0.5&0&0.22222222\cr
+13&2&2&0.5&5.68E-05&0\cr
+14&1&1&0.9&0&0\cr
+15&1&2&0.9&0&0.36363636\cr
+16&2&2&0.9&5.06E-05&0\cr}}$$
+ \vfil\break
 

### 2017-08-25 ms/figures.tex
update

--- 2017-08-24 ms/figures.tex
@@
-\documentclass[]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage[]{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{longtable,booktabs}
-% Fix footnotes in tables (requires footnote package)
-\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
-\usepackage{lineno}
-\linenumbers
-
-\begin{document}
-
 \begin{figure}
 \centering
-\includegraphics{../figures/lk_phase_plane.pdf}
-\caption{Ratios of selection periods \(\frac{k}{l}\) that lead to
-fixation of phenotype \emph{A} (red) or polymorphism of phenotypes
-\emph{A} and \emph{B} (blue). \emph{k} and \emph{l} are the number of
-generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
-The boundary is set at $-\log{(1+\rho \frac{s}{1-s}}) / \log{(1-\rho s})$, see eq. 23.
-In all cases, \emph{W} = 1, \emph{w} = \emph{1-s}.}\label{fig:lk_phase_plane}
+\includegraphics{figures/lk_phase_plane.pdf}
+\caption{Ratios of selection periods \(\frac{l}{k}\) that lead to
+fixation of phenotype \emph{B} (red) or polymorphism of phenotypes
+\emph{A} and \emph{B} (blue). \emph{l} and \emph{k} are the number of
+generations in which phenotypes \emph{B} and \emph{A}, respectively, are favored by
+selection. In all cases, \emph{W} = 1.}\label{fig:lk_phase_plane}
 \end{figure}
@@
 \centering
-\includegraphics{../figures/env_A1B1.pdf}
+\includegraphics{figures/env_A1B1.pdf}
 \caption{Frequency of phenotype \emph{A} after every two generations in
-selection regime \emph{A1B1}. The orange line is the finite population model
-(average of 100 simulations). The blue line is the infinite population model
-(eq. 25), and the green line is the equilibrium solution (eq. 56). Parameters: \emph{W}=1,
+selection regime \emph{A1B1}. The right column is the finite population model
+(orange; average of 100 simulations). The left column is the infinite population model
+(blue; @eq:recurrenceA1B1), and the equilibrium solution (dashed green;
+@eq:recurrenceA1B1\_solution\_x\_star). Parameters: \emph{W}=1,
 \emph{N}=10,000, initial value \(x=0.5\).}\label{fig:env_A1B1}
@@
 \centering
-\includegraphics{../figures/AkBl_equilibrium.pdf}
-\caption{Stability in a fluctuating environment.
-\textbf{(A)} Stable frequency of phenotype $A$ in the \emph{A1B1} selection regime and \textbf{(B)} stable mean fitness in selection regime \emph{A1B1} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
-\textbf{(C)} Stable population mean fitness in selection regime \emph{AkBl} as a
-function of the vertical transmission rate \(\rho\) and different ratios of selection periods \(k=l\).
-\emph{k} and \emph{l} are the number of generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
-The geometric average of the population mean fitness over $k+l$ generations is shown. 
-Here \emph{w=0.9}.
-In all cases, \emph{W=1}.
-}\label{fig:AkBl_equilibrium}
+\includegraphics{figures/A1B1_modifier_invasions.pdf}
+\caption{Consecutive fixation of modifiers that reduce the vertical
+transmission rate in environmental regime \emph{A1B1}. The figure shows
+results of numerical simulations of evolution with two modifier alleles
+(@eq:recurrence\_modifiers). When a modifier allele fixes
+(frequency\textgreater{}99.9\%), a new modifier allele is introduced
+with a vertical transmission rate one order of magnitude lower (vertical
+dashed lines). \textbf{(A,E,I)} The frequency of phenotype \emph{A} in
+the population over time. \textbf{(B,F,J)} The frequency of the invading
+modifier allele over time. \textbf{(C,G,K)} The population mean vertical
+transmission rate over time. \textbf{(D, H, L)} The population mean
+fitness over time; insets zoom in to show that the mean fitness 
+decreases slightly with each invasion. Parameters: vertical transmission rate of
+the initial resident modifier allele, \(\rho_0 =0.1\); fitness values:
+\emph{W}=1, \emph{w}=0.1 (\textbf{A-D}), 0.5 (\textbf{E-H}), and 0.9
+(\textbf{I-L}). The x-axis is on a log-scale, as each sequential invasion
+takes an order of magnitude longer to
+complete.}\label{fig:A1B1_modifier_invasions}
 \end{figure}
@@
 \centering
-\includegraphics{../figures/stochastic_env_x_t.pdf}
+\includegraphics{figures/A1B1_stable_x.pdf}
+\caption{Stable frequency of phenotype \emph{A}
+(@eq:recurrenceA1B1\_solution\_x\_star) in selection regime \emph{A1B1}
+as a function of the vertical transmission rate \(\rho\) and the fitness
+ratio \(W/w\) between the favored and unfavored phenotypes. Here
+\emph{W=1}.}\label{fig:A1B1_stable_x}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{figures/A1B1_mean_fitness.pdf}
+\caption{Stable mean fitness in selection regime \emph{A1B1} as a
+function of the vertical transmission rate \(\rho\) and the fitness
+ratio \(W/w\) between the favored and unfavored phenotypes. Here 
+\emph{W=1}.}\label{fig:A1B1_mean_fitness}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{figures/env_A1B2.pdf}
+\caption{Frequency of phenotype \emph{A} after every three generation in
+selection regime \emph{A1B2}. Comparison of dynamics starting with
+different initial frequency of phenotype \(A\) (0.01-0.99). Fixation occurs if the final population in 100 simulations with finite
+population size \(N\) has zero frequency of \(A\) at the end of the run. See also @Fig:lk\_phase\_plane
+and @Fig:A1B2\_polymorphism\_x. Parameters: \emph{W}=1, \emph{N=10,000},
+initial population \(x=0.5\).}\label{fig:env_A1B2}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{figures/A1B2_polymorphism_x.pdf}
+\caption{Stable frequency of phenotype \emph{A} in selection regime
+\emph{A1B2} as a function of the vertical transmission rate \(\rho\) and
+the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
+In the grey area \(B\) reached fixation and the frequency of \(A\) is 0.
+Here \emph{W=1}.}\label{fig:A1B2_polymorphism_x}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{figures/stochastic_env_x_t.pdf}
 \caption{Stochastic local stability. The figure shows the frequency of
-phenotype \emph{A} after \(10^6\) generations in a very large populationt 
-evolving in a stochastic environment (eq. 33). The
+phenotype \emph{A} after \(10^6\) generations in a very large population
+evolving in a stochastic environment (@eq:recurrence\_random\_env). The
 fitness of phenotypes \emph{A} and \emph{B} are \(1+s_t\) and \(1\),
@@
 \centering
-\includegraphics{../figures/stoch_two_iid_rvs.png}
+\includegraphics{figures/stoch_modifier_invasions.png}
+\caption{Consecutive fixation of modifiers that reduce the vertical
+transmission rate under randomly changing selection. The figure shows
+results of numerical simulations of evolution with two modifier alleles
+(@eq:recurrence\_modifiers). When a modifier allele fixes
+(frequency\textgreater{}99.9\%), a new modifier allele is introduced
+with a vertical transmission rate one order of magnitude lower (vertical
+dashed lines). \textbf{(A)} The frequency of phenotype \emph{A} in the
+population over time. \textbf{(B)} The frequency of the invading
+modifier allele over time. \textbf{(C)} The population mean vertical
+transmission rate over time. Parameters: vertical transmission rate of the initial
+resident modifier allele, \(\rho_0 =0.1\); fitness values: \emph{W=1}
+and \emph{w=0.1} with probability 0.5 and \emph{w=0.1} and \emph{W=1}
+otherwise. The x-axis is on a log-scale, as each sequential invasion takes
+an order of magnitude longer to
+complete.}\label{fig:stoch_modifier_invasions}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{figures/stoch_two_iid_rvs.png}
 \caption{Effect of vertical transmission rate on phenotype polymorphism
@@
 \centering
-\includegraphics{../figures/A1B1_modifier_invasions.pdf}
-\caption{Consecutive fixation of modifiers that reduce the vertical
-transmission rate in environmental regime \emph{A1B1}. The figure shows
-results of numerical simulations of evolution with two modifier alleles
-(eq. 53). When a modifier allele fixes
-(frequency\textgreater{}99.9\%), a new modifier allele is introduced
-with a vertical transmission rate one order of magnitude lower (vertical
-dashed lines). \textbf{(A,D,G)} The frequency of phenotype \emph{A} in
-the population over time. \textbf{(B,E,H)} The frequency of the invading
-modifier allele over time. \textbf{(C,F,I)} The population mean
-fitness over time; insets zoom in to show that the mean fitness 
-decreases slightly with each invasion. Parameters: vertical transmission rate of
-the initial resident modifier allele, \(\rho_0 =0.1\); fitness values:
-\emph{W}=1, \emph{w}=0.1 (\textbf{A-C}), 0.5 (\textbf{D-F}), and 0.9
-(\textbf{G-I}). The x-axis is on a log-scale, as each sequential invasion
-takes an order of magnitude longer to
-complete.}\label{fig:A1B1_modifier_invasions}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=\dimexpr \textheight - 8\baselineskip\relax]{../figures/stoch_modifier_invasions.png}
-\caption{Consecutive fixation of modifiers that reduce the vertical
-transmission rate under randomly changing selection. The figure shows
-results of numerical simulations of evolution with two modifier alleles
-(eq. 33). The
-fitness of phenotypes \emph{A} and \emph{B} are 1 and 0.1 with probability 0.5 or 0.1 and 1 otherwise.
-When a modifier allele fixes
-(frequency\textgreater{}99.9\%), a new modifier allele is introduced
-with a vertical transmission rate one order of magnitude lower (vertical
-dashed lines). \textbf{(A)} The frequency of phenotype \emph{A} in the
-population over time. \textbf{(B)} The frequency of the invading
-modifier allele over time. \textbf{(C)} The population mean fitness over time. Parameters: vertical transmission rate of the initial
-resident modifier allele, \(\rho_0 =0.1\); fitness values: \emph{W=1}
-and \emph{w=0.1} with probability 0.5 and \emph{w=0.1} and \emph{W=1}
-otherwise. The x-axis is on a log-scale, as each sequential invasion takes
-an order of magnitude longer to
-complete.}\label{fig:stoch_modifier_invasions}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/A1B2_polymorphism.pdf}
-\caption{Stable frequency of phenotype \emph{A} and population mean fitness in selection regime
-\emph{A1B2} as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
-In the grey area \(B\) reached fixation and the frequency of \(A\) is 0.
-\textbf{(A)} Stable frequency of phenotype $A$ at the end of each period of three generations.
-\textbf{(B)} Geometric average of the stable mean fitness over the three generations period: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
-In all cases, \emph{W=1}.}\label{fig:A1B2_polymorphism}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/fixation_prob_time.pdf}
+\includegraphics{figures/fixation_prob_time.pdf}
 \caption{Fixation probability and time in a finite population.
-\textbf{(A)} Fixation probability \(u(x)\) of phenotype \(A\) (eq. 88), and \textbf{(B)}
-Expected time to fixation \(T(x)\) of phenotype \(A\) (eq. 90), conditioned on its
+\textbf{(A)} Fixation probability \(u(x)\) of phenotype \(A\) and \textbf{(B)}
+Expected time to fixation \(T(x)\) of phenotype \(A\), conditioned on its
 fixation, starting with a single copy in a population of size \(N\). The
@@
 \end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/lk_fix_prob.pdf}
-\caption{Fixation in a finite population with different ratios of selection periods \(\frac{k}{l}\).
-Fixation probability of phenotype $A$ when starting with a single copy in a population of size $N$: $u(1/N) = (1-\exp(-2 \rho \frac{k-l}{k+l}(W-w))/(1-\exp(-2 N \rho \frac{k-l}{k+l}(W-w))$ (see eqs.
-115-116).
-\emph{k} and \emph{l} are the number of
-generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by
-selection. In all cases, \emph{W} = 1, \emph{w} = \emph{1-s}, Population size, \(N=10,000\).} \label{lk_fix_prob}
-\end{figure}
-
-\end{document}

### 2017-08-25 ms/figures_supp.tex
update

@@
 \centering
+\includegraphics{../figures/env_A1B1.pdf}
+\caption{Frequency of phenotype \emph{A} after every two generations in
+selection regime \(A1B1\). The orange line is the finite population model
+(eqs. 52-53; average of 100 simulations). The blue line is the infinite population model
+(eq.\ 23), and the green line is the  solution of \(Q(x)=0\) (see eq.\ 24).
+Parameters: \(W=1\); For the finite population model (orange lines), population size is \(N=10,000\) and initial frequency of \(A\) is \(x_0=0.5\).}\label{fig:env_A1B1}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics{../figures/env_A1B2.pdf}

### 2017-08-25 ms/ms.tex
update

@@
 \vfil
-\centerline{August 24, 2017, v.1}
+\centerline{August 25, 2017, v.2}
 
@@
 In some situations, oblique transmission of biological material is possible.
-An example could be the transmission of microbes occasioned by the sharing or manipulation of food or other consumable resources during a social interaction.
+In bacteria, phenotypes might be determined by heritable mobile genetic elements such as phages (Zinder and Lederberg 1952), plasmids (Lederberg and Tatum 1946), integrons (Mazel 2006), and transposons (Salyers et al.\ 2004).
+Similarly, some phenotypes are determined by genes that are commonly converted by uptake of foreign DNA, i.e., transformation (Milkman and Bridges 1990).
+In these cases, inheritance of a phenotype may combine vertical transmission from the parent cell, and oblique transmission from other cells, even if the latter did not originally evolve for that purpose (Redfield 1993).
+
+In some animals, transmission of microbes occasioned by the sharing or manipulation of food or other consumable resources during a social interaction.
 Although transmission of the microbiome in humans is likely to be mostly vertical (Rosenberg and Zilber-Rosenberg 2016), in other organisms there is multi-generational food sharing during which symbionts from the parental cohort may be transmitted obliquely to younger individuals (Theis et al.\ 2016).
@@
 Thus, starting from any initial value $0<x_0 <1$ we have  $x_t\,\harr{}{t\to\infty}\,x^*$.
-Figure 2 illustrates examples of how the frequency of $A$ changes over time in the $A1B1$ regime of cycling selection, while Figure 3A shows the equilibrium frequencies $x^*$ as a function of $\rho$, $W$, and $w$ in this regime.
+Supplementary Figure S1 illustrates examples of how the frequency of $A$ changes over time in the $A1B1$ regime of cycling selection, while Figure 2A shows the equilibrium frequencies $x^*$ as a function of $\rho$, $W$, and $w$ in this regime.
 
 For more general cyclic fitness regimes, the polynomial that gives the equilibria is of higher order, and it is conceivable that more than one stable polymorphism could exist for given values of $\rho$, $W$, and $w$.
-We investigated this numerically, and as shown in Supplemental Figure S1, the $A1B2$ regime also appears to allow a single stable polymorphism.
-For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplemental Figure S2 shows that only a single stable equilibrium appears to exist.
+We investigated this numerically, and as shown in Supplemental Figure S2, the $A1B2$ regime also appears to allow a single stable polymorphism.
+For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplemental Figure S3 shows that only a single stable equilibrium appears to exist.
 \medskip
-\centerline{[FIGURES 2, 3 HERE]}  
+\centerline{[FIGURE 2 HERE]}  
  \bigskip
@@
 It is possible, however, that neither fixation is  stochastically locally stable, in which case we expect the population to converge to a polymorphic distribution.
-Figure 4 illustrates how the properties of $s_t$ in recursion (31) affect the frequency of phenotype $A$, and in particular the stochastic local stability of fixation in phenotype $B$.
-Figure 5 shows the dynamics of the frequency of $A$ in a case where $w_A$ and $w_B$ are identically distributed and independent; in this case the expectation of the stationary distribution is 1/2, and its variance increases as $\rho$ increases.
+Figure 3 illustrates how the properties of $s_t$ in recursion (31) affect the frequency of phenotype $A$, and in particular the stochastic local stability of fixation in phenotype $B$.
+Figure 4 shows the dynamics of the frequency of $A$ in a case where $w_A$ and $w_B$ are identically distributed and independent; in this case the expectation of the stationary distribution is 1/2, and its variance increases as $\rho$ increases.
 \medskip
-\centerline{[FIGURES 4, 5 HERE]}  
+\centerline{[FIGURES 3, 4 HERE]}  
  \bigskip
@@
 Thus in this case, evolution tends to reduce vertical transmission, and hence increase the rate of oblique transmission.
-The evolutionary dynamics of the reduction in $\rho$ under the $A1B1$ cycling regime are shown in Figure 6, which also illustrates the change in phenotype frequencies over time.
-In the case of identically distributed random fitnesses $w_A$ and $w_B$, Figure 7 shows  an example of the success of modifiers that reduce $\rho$.
+The evolutionary dynamics of the reduction in $\rho$ under the $A1B1$ cycling regime are shown in Figure 5, which also illustrates the change in phenotype frequencies over time.
+In the case of identically distributed random fitnesses $w_A$ and $w_B$, Figure 6 shows  an example of the success of modifiers that reduce $\rho$.
 
- Figure 3B illustrates the decrease with increasing $\rho$ of the geometric mean fitness (over time) at a polymorphic equilibrium  in the $A1B1$ regime.
-Figure 8 illustrates an example of the $A1B2$ selection regime: panel A shows the values of selection and $\rho$ that allow polymorphism of phenotypes $A$ and $B$, and Panel B shows that the mean fitness at this equilibrium is decreasing as $\rho$ increases.
-Figure 7C shows how the mean fitness fluctuates as a function of $\rho$ in the case of symmetric randomly fluctuating fitnesses.
+ Figure 2B illustrates the decrease with increasing $\rho$ of the geometric mean fitness (over time) at a polymorphic equilibrium  in the $A1B1$ regime.
+Figure 7 illustrates an example of the $A1B2$ selection regime: panel A shows the values of selection and $\rho$ that allow polymorphism of phenotypes $A$ and $B$, and Panel B shows that the mean fitness at this equilibrium is decreasing as $\rho$ increases.
 
- Thus in  cycling fitness  $A1B1$ and  $A2B2$ regimes and the symmetric random fitness regime of Figure 7, there is a {\sl reduction principle} for the rate of vertical transmission (Altenberg et al.\ 2017).
-For the $A2B2$ selection regime, numerical analysis (not shown) confirms the reduction principle for $\rho$.
+ Thus in  cycling fitness  $A1B1$ and  $A2B2$ regimes and the symmetric random fitness regime of Figure 6, there is a {\sl reduction principle} for the rate of vertical transmission (Feldman and Liberman 1986; Altenberg et al.\ 2017).
+For the $A2B2$ selection regime, numerical analysis (Figure 8) confirms the reduction principle for $\rho$.
   \medskip
-\centerline{[FIGURES 6, 7,  8 HERE]}
+\centerline{[FIGURES 5, 6,  7 HERE]}
 
@@
 At this stage, any effect of the $M/m$ difference would be very difficult to detect computationally.
-The variation of $\rho^*$ with $k$ for selection regimes $AkBk$ is shown in Figure 9 for $k\le50$, $W=1$ with $w=0.1$, $0.5$, and $0.9$.
+The variation of $\rho^*$ with $k$ for selection regimes $AkBk$ is shown in Figure 8 for $k\le50$, $W=1$ with $w=0.1$, $0.5$, and $0.9$.
 
@@
 It is somewhat surprising, given the modifier findings for $k\ge 3$, that $\hat\rho$ is actually zero for $k\le11$ and, depending on the fitnesses, is larger than zero for $k>11$.
-This is shown in Figure 10, where for $W=1$ with $w=0.1$ and $0.5$, $\hat\rho$ is noticeably positive.
+This is shown in Figure 9, where for $W=1$ with $w=0.1$ and $0.5$, $\hat\rho$ is noticeably positive.
 For $W=1$, $w=0.9$, however, $\hat\rho=0$ maximizes the geometric mean fitness.
-More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Supplementary Figure S3.
+More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Supplementary Figure S4.
 \medskip
-\centerline{[FIGURES 9,10 HERE]}  
+\centerline{[FIGURES 8, 9 HERE]}  
  \bigskip
@@
  The proof of Result 9 is in Supplementary Materials SP3.  
-  Figure 11 compares the fixation probability and time to fixation derived numerically from  simulating the Wright-Fisher Markov chain with the diffusion-derived values of $u(x)$ and $T(x)$.
+  Figure 10 compares the fixation probability and time to fixation derived numerically from  simulating the Wright-Fisher Markov chain with the diffusion-derived values of $u(x)$ and $T(x)$.
 The fit is seen to be very good.
-Note that when $N$ is large, the Wright-Fisher model exhibits persistent fluctuation around the deterministic expectation, as shown by the orange diagram in Figure 2B, D, F.
+Note that when $N$ is large, the Wright-Fisher model exhibits persistent fluctuation around the deterministic expectation, as shown by the orange diagram in Supplemental Figure S1.
   \medskip
 
-  \centerline{[FIGURE 11 HERE]}
+  \centerline{[FIGURE 10 HERE]}
 
@@
  $$S_n=S_{k+l} =(k-l)(W-w).\eqno(76).$$
- Figure 12 illustrates how $(k-l)$, $\rho$, and $(W-w)$, which enter the formula for $u(x)$ in (75), interact to affect fixation probabilities.
+ Figure 11 illustrates how $(k-l)$, $\rho$, and $(W-w)$, which enter the formula for $u(x)$ in (75), interact to affect fixation probabilities.
 
- The orange curve in the second column of Figure 2 shows that as $\rho$ increases in the finite-population $A1B1$ cycling regime, the population fluctuates more around the initial frequency of phenotype $A$.
-For high rates of oblique transmission (lower $\rho$), the first panel in the second column shows that polymorphism is maintained for 20,000 periods with $N=10,000$.
  \medskip
-\centerline{[FIGURE 12 HERE]}  
+\centerline{[FIGURE 11 HERE]}  
   \bigskip
@@
 
- The mean fitness of the population evaluated at a polymorphism of phenotypes $A$ and $B$ is shown to be an increasing function of the level of oblique transmission in the $A1B1$ (Figure 3B, Figures 6C, 6F, 6I) and $A1B2$ (Figure 8B) environments.
+ The mean fitness of the population evaluated at a polymorphism of phenotypes $A$ and $B$ is shown to be an increasing function of the level of oblique transmission in the $A1B1$ (Figure 2B, Figures 5C, 5F, 5I) and $A1B2$ (Figure 7B) environments.
 We can compare this result with the asymptotic growth rate (AGR) of Xue and Leibler (2016), whose parameter  $\eta$ is the rate at which an individual learns from its parental lineage.
@@
 [FINISH WRITEUP]
-
-\bigskip \centerline{GET numerical results on $T(x)$ as a function of $\rho$}
 
@@
 
+\ref Lederberg, J., and E. L. Tatum. 1946. Gene recombination in {\it Escherichia coli}. {\it Nature} {\bf 158}: 558.
+
 \ref Liberman, U., J. Van Cleve, and M. W. Feldman. 2011. On the evolution of mutation in changing environments: recombination and phenotypic switching. {\it Genetics}  {\bf 187}: 837--851.
@@
 \ref Lycett, S. J., and J. A. J. Gowlett. 2008. On questions surrounding the Acheulean `tradition'. {\it World Archaeol.} {\bf 40}: 295--315.
+
+\ref Mazel, D. Integrons: agents of bacterial evolution. {\it Nat.\ Rev.\ Microbiol.} {\bf 4}: 608--620.
+
+\ref Milkman, R., and M. M. Bridges. Molecular evolution of the {\it Escherichia coli} chromosome. III. Clonal frames. {\it Genetics} {\bf 126} 505--517.
+
+\ref Redfield, R. J. Evolution of natural transformation: testing the DNA repair hypothesis in {\it Bacillus subtilis} and {\it Haemophilus influenzae}. {\it Genetics} {\bf 133}: 755--761.
 
@@
 \ref Salath\'e, M., J. Van Cleve, and M. W. Feldman. 2009. Evolution of stochastic switching rates in asymmetric fitness landscapes. {\it Genetics} {\bf 182}: 1159--1164.
+
+\ref Salyers, A.
+A., G.
+Whittle, and N.
+B.
+Shoemaker.
+2004.
+Conjugal and mobilizable transposons.
+Pp./ 125--143 {\it in} Miller, R., and M.
+Day (eds.) {\it Microbial Evolution}.
+Washington, D.C.: ASM Press.
 
@@
 \ref Xue, B.-K., and S. Leibler. 2016. Evolutionary learning of adaptation to varying environments through a transgenerational feedback. {\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 113}: 11266--11271.
+
+\ref Zinder, N. D., and J. Lederberg. 1952.  Genetic exchange in {\it Salmonella}. {\it J. Bacteriol.} {\bf 64}: 679--699.
  \vfil
@@
 
- Figure 4 presents a numerical example of the dynamics of recursion (31) with a specific random selection coefficient $s_t$.
+ Figure 3 presents a numerical example of the dynamics of recursion (31) with a specific random selection coefficient $s_t$.
 \bigskip

### 2017-09-12 ms/sum_binomials.tex
found a new signa^2

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amsmath}
\usepackage{amssymb}

%SetFonts

%SetFonts

\date{\today}							% Activate to display a given date or no date

\begin{document}

\section*{Finite population model}

Consider a haploid population with non-overlapping generations of constant finite size $N$.
We focus on a trait with two phenotypes: $A$ and $B$, with fitness values $w_A$ and $w_B$.
Each individual inherits his phenotype from his parent with probability $\rho$ or from a random individual from the parental generations with probability $1-\rho$.
Therefore, the number of individuals with vertically transmitted phenotypes in the next generation is:
\begin{equation}
v \sim Bin(N, \rho),
\end{equation}
and the number of individuals with diagonally transmitted phenotypes is $N-v$.

Of the individuals with vertically transmitted phenotypes, the number of individuals with phenotype $A$ depends on the parent's fitness, the frequency of $A$, and on random sampling, and is 
\begin{equation}
a_V \sim Bin(v, \frac{w_A}{\bar{w}} x),
\end{equation}
where $x$ is the frequency of phenotype $A$ in the parent generation and $\bar{w}$ is the mean fitness in the parent generation
\begin{equation}
\bar{w} = x w_A + (1-x) w_B.
\end{equation}

Of the individuals with diagonally transmitted phenotypes, the number of individuals with phenotype $A$ depends only on the frequency of $A$ and random sampling, and is 
\begin{equation}
a_D \sim Bin(v, x).
\end{equation}

These $a_V$ and $a_D$ are inherently separate random events, as $a_V$ depends on events that determine the reproductive success of individuals in the parental generations, whereas $a_D$ does not.

Therefore, the number of individuals with phenotype $A$ in the next generation $X_{t+1}$ given their frequency in the current generation is $x = X_t/N$, is
\begin{equation}
(X_{t+1}|X_t=Nx) = a_V + a_D.
\end{equation}
However, $(X_{t+1}|X_t=Nx)$ is not a binomial, as \textit{the sum of two binomials with different success probabilities is not a binomial}, and is underdispersed compared to a binomial (Nedelman and Wallenius, 1986):
\begin{equation}
Var(X_{t+1}|X_t=Nx) \le Var(a_V) + Var(a_D),
\end{equation}
where $a$ is a binomial only if the equality is obtained, which occurs for $\rho=0$ and $1$ or for $w_A=w_B$.

Indeed, according to eq.~2 in Nedelman and Wallenius (1986), 
\begin{equation}
Var(X_{t+1}|X_t=Nx) = 
N \Big(\bar{p} (1-\bar{p}) - \rho\Big(\frac{w_A}{\bar{w}} x - \bar{p}\Big)^2 + (1-\rho)(\bar{p} - x)^2\Big),
\end{equation}
where
\begin{equation}
\bar{p} = \rho \frac{w_A}{\bar{w}} x + (1-\rho) x.
\end{equation}

Substituting $\frac{w_A}{\bar{w}} \to 1+\frac{s}{N}(1-x)$ (see eq.~61 in main text), we have, up to terms smaller than $1/N$,
\begin{align}
\bar{p} = 
	x(1 + \rho \frac{s}{N}(1-x))\\
\bar{p} (1-\bar{p} ) = 
	x(1-x)\Big(1 + \rho\frac{s}{N}(1-2x)\Big)\\
\frac{w_A}{\bar{w}} x - \bar{p} = 
	(1-x)\Big(1+\frac{s}{N}(1-\rho x)\Big) \\
\bar{p} - x = 
x (1-x) \rho \frac{s}{N} \\
	\Big(\frac{w_A}{\bar{w}} x - \bar{p}\Big)^2 = (1-x)^2(1 + 2\frac{s}{N}(1-\rho x)) \\
(\bar{p} - x)^2 = 0
\end{align}

So, the variance in the frequency of $A$ in the next generation $Var\Big(\frac{X_{t+1}}{N}\Big)$ given the frequency $x$ in the current generation $t$, up to terms of order smaller than $1/N$, is
\begin{equation}
\frac{1}{N^2}Var(X_{t+1}|X_t=Nx) = 
\frac{1}{N} \Big( x(1-x) - \rho (1-x)^2 \Big).
\end{equation}

Therefore, the diffusion approximation variance term $\sigma^2(x)$, is
\begin{equation}
\sigma^2(x) = x(1-x) - \rho (1-x)^2
\end{equation}
which is smaller than the value given in eq.~66 of the main text if $\rho>0$ and $0<x<1$.

\section*{References}

1.  Nedelman, J and Wallenius, T., 1986. Bernoulli trials, Poisson trials, surprising variances, and Jensen’s Inequality. The American Statistician, 40(4):286–289.

\end{document}

### 2017-09-17 ms/sum_binomials.tex
found a mistake... sigma^2 is just like in the ms

@@
 
-\date{\today}							% Activate to display a given date or no date
-
 \begin{document}
@@
 \section*{Finite population model}
+\textbf{\today}
 
@@
 \begin{equation}
-a_D \sim Bin(v, x).
+a_D \sim Bin(N-v, x).
 \end{equation}
@@
 
-Substituting $\frac{w_A}{\bar{w}} \to 1+\frac{s}{N}(1-x)$ (see eq.~61 in main text), we have, up to terms smaller than $1/N$,
+Substituting $\frac{w_A}{\bar{w}} \to 1+\frac{s}{N}(1-x)$ (see eq.~61 in main text), we have
 \begin{align}
-\bar{p} = 
-	x(1 + \rho \frac{s}{N}(1-x))\\
-\bar{p} (1-\bar{p} ) = 
-	x(1-x)\Big(1 + \rho\frac{s}{N}(1-2x)\Big)\\
-\frac{w_A}{\bar{w}} x - \bar{p} = 
-	(1-x)\Big(1+\frac{s}{N}(1-\rho x)\Big) \\
-\bar{p} - x = 
-x (1-x) \rho \frac{s}{N} \\
-	\Big(\frac{w_A}{\bar{w}} x - \bar{p}\Big)^2 = (1-x)^2(1 + 2\frac{s}{N}(1-\rho x)) \\
-(\bar{p} - x)^2 = 0
+Var(X_{t+1}|X_t=Nx) = 
+\frac{1}{N} x (1-x) (N^2 + N \rho s (1-2x) - \rho s^2 x (1-x)).
 \end{align}
 
-So, the variance in the frequency of $A$ in the next generation $Var\Big(\frac{X_{t+1}}{N}\Big)$ given the frequency $x$ in the current generation $t$, up to terms of order smaller than $1/N$, is
+So, the variance in the frequency of $A$ in the next generation $Var\Big(\frac{X_{t+1}}{N}\Big)$ given the frequency $x$ in the current generation $t$ is
+\begin{multline}
+Var\Big(\frac{X_{t+1}}{N} \Big| \frac{X_{t}}{N} = x\Big) = 
+\frac{1}{N^2}Var(X_{t+1}|X_t=Nx) = \\
+\frac{1}{N} x (1-x) \Big(1 + \frac{1}{N} \rho s (1-2x) - \rho \frac{s^2}{N^2} x (1-x)\Big).
+\end{multline}
+
+Therefore, the diffusion approximation variance term $\sigma^2(x)$, up to terms of the order $1/N$, is
 \begin{equation}
-\frac{1}{N^2}Var(X_{t+1}|X_t=Nx) = 
-\frac{1}{N} \Big( x(1-x) - \rho (1-x)^2 \Big).
+\sigma^2(x) = x(1-x),
 \end{equation}
-
-Therefore, the diffusion approximation variance term $\sigma^2(x)$, is
-\begin{equation}
-\sigma^2(x) = x(1-x) - \rho (1-x)^2
-\end{equation}
-which is smaller than the value given in eq.~66 of the main text if $\rho>0$ and $0<x<1$.
+which is the same as the value given in eq.~66 of the main text if $\rho>0$ and $0<x<1$.
 

### 2017-09-19 ms/ms.tex
update

@@
 \vfil
-\centerline{August 25, 2017, v.2}
+\centerline{September 15, 2017, v.1}
 
@@
 
-In some animals, transmission of microbes occasioned by the sharing or manipulation of food or other consumable resources during a social interaction.
+In some animals, transmission of microbes may occur during sharing or manipulation of food or other consumable resources during a social interaction.
 Although transmission of the microbiome in humans is likely to be mostly vertical (Rosenberg and Zilber-Rosenberg 2016), in other organisms there is multi-generational food sharing during which symbionts from the parental cohort may be transmitted obliquely to younger individuals (Theis et al.\ 2016).
@@
 
-Another perspective on the evolutionary consequences of fluctuating environments (and, as a result, fluctuating fitnesses) has derived from the phenomenon of phenotypic switching (Balaban et al.\ 2004; Kussell and Leibler 2005; Thattai and van Oudenaarden 2004; Salath\'e et al.\ 2009; Liberman et al.\ 2011; Gaal et al.\ 2010).
-In these studies, mutation caused the organism to switch phenotypes (usually treated as haploid genotypes), and the problem has usually been posed in terms of the optimal rate of mutation when the phenotypic fitnesses fluctuate over time.
+Another perspective on the evolutionary consequences of fluctuating environments (and, as a result, fluctuating fitnesses) derives from the phenomenon of phenotypic switching (Balaban et al.\ 2004; Kussell and Leibler 2005; Thattai and van Oudenaarden 2004; Salath\'e et al.\ 2009; Liberman et al.\ 2011; Gaal et al.\ 2010).
+In these studies, mutation causes the organism to switch phenotypes (usually treated as haploid genotypes), and the problem has usually been couched in terms of the optimal rate of mutation in models where the phenotypic fitnesses fluctuate over time.
 These models did not include social learning, and the evolution was regarded as a mode of bet-hedging against future environmental change.
@@
 In our formulation, both the parental phenotype and the distribution of phenotypes in the whole population contribute to an offspring's phenotype.
-Using conventional modifier theory (Feldman and Liberman 1986), we show that in a cyclic selection regime with cycles of periods 1 or 2, an allele reducing the rate of vertical transmission is expected to increase in frequency when rare and in so doing to increase the mean fitness of the population.
-However, for cycles of greater length, interesting non-monotonicities emerge, both in the uninvadable rate of vertical transmission, and the rate that maximizes the geometric time average of the population mean fitness, which we will refer to as the ``geometric mean fitness.'' We develop the models in very large populations with cyclic selection and with random fitnesses and also in the case where drift occurs via sampling from generation to generation in a finite population.
+Using conventional modifier theory (Feldman and Liberman 1986), we show that in a symmetric cyclic selection regime with cycles of periods 1 or 2, an allele reducing the rate of vertical transmission is expected to increase in frequency when rare and in so doing to increase the mean fitness of the population.
+However, for cycles of greater length or period asymmetry, interesting non-monotonicities emerge, both in the uninvadable rate of vertical transmission, and the rate that maximizes the geometric time average of the population mean fitness, which we will refer to as the ``geometric mean fitness.'' We develop the models in very large populations with cyclic selection and with random fitnesses and also in the case where drift occurs via sampling from generation to generation in a finite population.
 
@@
 
-\item{1.}As $k=l$ and the above result holds also when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is not stable.
-Thus neither phenotype can become extinct, and we have a {\sl protected polymorphism} (Prout 1968).
+\item{1.}Since $k=l$ and the above result holds also when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is not stable.
+Thus neither phenotype can be lost, and we have a {\sl protected polymorphism} (Prout 1968).
 
@@
 \noindent{\it Remark}.
-It should be pointed out that the local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations $A$ was favored $k$ times and $B$ was favored $l$ times, and not their order in the cycle.
+It should be pointed out that the local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations $A$ is favored $k$ times and $B$ is favored $l$ times, and not their order in the cycle.
 
@@
 For more general cyclic fitness regimes, the polynomial that gives the equilibria is of higher order, and it is conceivable that more than one stable polymorphism could exist for given values of $\rho$, $W$, and $w$.
-We investigated this numerically, and as shown in Supplemental Figure S2, the $A1B2$ regime also appears to allow a single stable polymorphism.
-For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplemental Figure S3 shows that only a single stable equilibrium appears to exist.
+We have been able to show that when neither fixation in $A$ or $B$ is stable, in the $AkBk$ case this cannot occur.
+In fact we have the following.
+
+\proclaim Result 5. In the $AkBk$ selection regimes, if the fixations in $A$ and $B$ are locally unstable, a single stable polymorphic equilibrium exists.\par
+
+The proof of Result 5 is in Supplementary Material SP1. 
+For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplemental Figure S3 illustrates the convergence to a single stable polymorphism.
+
+We have not been able to prove that for selection regimes $AkBl$ with $l\ne k$ there is a single stable polymorphic equilibrium when the two fixations are unstable.
+However, the numerical examples in Figure 1 (for $AkBl$) and in Figure 7 (for the special case $A1B2$) all exhibit a single stable polymorphic equilibrium when fixations in $A$ and $B$ are unstable.
+These numerical results suggest that for $W>w>0$ and $0<\rho<1$ the high order polynomial that gives the equilibria has only a single root that is a globally stable polymorphism.
 \medskip
@@
 
- \proclaim Result 5. Suppose $E\left[\log (1+\rho s_t)\right]>0$. Then $x^*=0$, the fixation of phenotype $B$, is not stochastically locally stable. In fact $P\left(\lim_{t\to\infty}x_t=0\right)=0$.
+ \proclaim Result 6. Suppose $E\left[\log (1+\rho s_t)\right]>0$. Then $x^*=0$, the fixation of phenotype $B$, is not stochastically locally stable. In fact $P\left(\lim_{t\to\infty}x_t=0\right)=0$.
 
-  \proclaim Result 6.
+  \proclaim Result 7.
 Suppose $E[\log(1+\rho s_t)]<0$.
@@
 
- The proofs of Results 5 and  6 can be found in Supplementary Materials SP1 and SP2.
+ The proofs of Results 6 and  7 can be found in Supplementary Materials SP2 and SP3.
 
@@
 $$\overline w =w_A(x_1+x_3) +w_B(x_2+x_4).\eqno(36)$$
+Note that under these assumptions the $M/m$ locus and the $A/B$ phenotypic dichotomy do not undergo anything analogous to recombination, which might be introduced if the $A/B$ were viewed as haploid genetic variants.
+
 Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} (e.g., Feldman and Liberman 1986; Altenberg et al.\ 2017) to invasion by allele $M$.
-
 Because we assume the initial existence of a stable polymorphic equilibrium that depends on $\rho$ (the rate determined by $m$), we cannot assume a {\sl constant environment}, which always leads to fixation of the favored type, independent of $\rho$.
@@
  Hence when $P=\rho$ one of the eigenvalues of $\cl$ is 1.
- \smallskip
 
@@
 
+ \smallskip
+
 We can prove the following result.
 
- \proclaim Result 7. $\cl$ has two positive eigenvalues and\hfil\break
+ \proclaim Result 8. $\cl$ has two positive eigenvalues and\hfil\break
   {\hglue.82truein}(i) when $P>\rho$ the two eigenvalues are less than $1$,\hfil\break
@@
   \noindent{\it Proof}.
-As $\cl$ is a positive matrix by the Perron-Frobenius theory, $\cl$ has a positive eigenvalue, and as $a_0>0$ and $a_2=1$ the product of the two eigenvalues of $\cl$ is positive.
+As $\cl$ is a positive matrix, by the Perron-Frobenius theory $\cl$ has a positive eigenvalue, and as $a_0>0$ and $a_2=1$ the product of the two eigenvalues of $\cl$ is positive.
 Thus in fact $\cl$ has two positive eigenvalues.
@@
 
+ We conclude that in the $A1B1$ selection regime an allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
+Thus in this case, evolution tends to reduce vertical transmission, and hence increase the rate of oblique transmission.
+The evolutionary dynamics of the reduction in $\rho$ under the $A1B1$ cycling regime are shown in Figure 5, which also illustrates the change in phenotype frequencies over time.
+In the case of identically distributed random fitnesses $w_A$ and $w_B$, Figure 6 shows  an example of the success of modifiers that reduce $\rho$.
+
+ Thus in the cycling fitness regime  $A1B1$  and the symmetric random fitness regime of Figure 6, there is a {\sl reduction principle} for the rate of vertical transmission (Feldman and Liberman 1986; Altenberg et al.\ 2017).
+  \medskip
+\centerline{[FIGURES 5, 6]}
+
+Mathematical analysis of modifiers of $\rho$ for more general $AkBl$ regimes is difficult, and we have made a numerical investigation of the evolutionarily stable value $\rho^*$ of the rate of vertical transmission for $k,l\ne 1$.
+For $A1B2$, $\rho^*=0$ still appears to be the stable state.
+However, as we increase $k$ and $l$, the stable value of $\rho$ changes, and becomes a function of the fitness difference between $A$ and $B$.
+Representative results are shown in Table 1, where it is clear that, even in the $AkBk$ case, increasing $k$ results in an increase of $\rho^*$.
+Figure 8 illustrates the results for the $AkBk$ case for $k$ up to 50 and shows that the value of $\rho^*$ depends on $k$ but also in a complex manner on $w/W$.
+For the values of $k$ and $l$ in Table 1, lower values of $W/w$ give a lower value of $\rho^*$, i.e., $\rho^*$ close to zero.
+However, as $k$ increases in the $AkBk$ case, values of $\rho^*$ increase as $W/w$ decreases.
+The modification process is clearly much more complex than the evolution of modifiers of recombination, mutation, or migration.
+
  It is natural to also ask how the population mean fitness is affected by modification of $\rho$. We have the following result.
 
-\proclaim Result 8. If $W>w$ and $0\le\rho\le 1$, then the mean fitness at the stable equilibrium in the $A1B1$ environment is a decreasing function of $\rho$.\par
+\proclaim Result 9. If $W>w$ and $0\le\rho\le 1$, then the mean fitness at the stable equilibrium in the $A1B1$ environment is a decreasing function of $\rho$.\par
 
@@
 
- We conclude that in the $A1B1$ selection regime an allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
-Thus in this case, evolution tends to reduce vertical transmission, and hence increase the rate of oblique transmission.
-The evolutionary dynamics of the reduction in $\rho$ under the $A1B1$ cycling regime are shown in Figure 5, which also illustrates the change in phenotype frequencies over time.
-In the case of identically distributed random fitnesses $w_A$ and $w_B$, Figure 6 shows  an example of the success of modifiers that reduce $\rho$.
+  Figure 2B illustrates the decrease with increasing $\rho$ of the geometric mean fitness (over time) at a polymorphic equilibrium  in the $A1B1$ regime.
+Figure 7 illustrates an example of the $A1B2$ selection regime: panel A shows the values of selection and $\rho$ that allow a polymorphic equlibrium of phenotypes $A$ and $B$, and Panel B shows that the mean fitness at this equilibrium is decreasing as $\rho$ increases.
 
- Figure 2B illustrates the decrease with increasing $\rho$ of the geometric mean fitness (over time) at a polymorphic equilibrium  in the $A1B1$ regime.
-Figure 7 illustrates an example of the $A1B2$ selection regime: panel A shows the values of selection and $\rho$ that allow polymorphism of phenotypes $A$ and $B$, and Panel B shows that the mean fitness at this equilibrium is decreasing as $\rho$ increases.
-
- Thus in  cycling fitness  $A1B1$ and  $A2B2$ regimes and the symmetric random fitness regime of Figure 6, there is a {\sl reduction principle} for the rate of vertical transmission (Feldman and Liberman 1986; Altenberg et al.\ 2017).
-For the $A2B2$ selection regime, numerical analysis (Figure 8) confirms the reduction principle for $\rho$.
-  \medskip
-\centerline{[FIGURES 5, 6,  7 HERE]}
-
- We investigated the dynamics of modifiers of $\rho$ in $AkBk$ regimes for $k\ge 3$.
-The reduction principle appears to fail for all $k\ge 3$.
-For each value of $k\ge 3$, there is a value of $\rho$ below which modifiers that increase $\rho$ invade and above which modifiers that reduce $\rho$ invade.
-For $k\le8$, the evolutionarily stable value of $\rho$, $\rho^*$  increases rapidly with $k$, but then the increase  becomes much slower.
-Note that for $k$ large enough, during the $Ak$ part of the cycle, phenotype $A$ becomes extremely close to fixation for even moderate values of $w$ relative to $W$.
-At this stage, any effect of the $M/m$ difference would be very difficult to detect computationally.
-The variation of $\rho^*$ with $k$ for selection regimes $AkBk$ is shown in Figure 8 for $k\le50$, $W=1$ with $w=0.1$, $0.5$, and $0.9$.
-
- The value of $\rho$ that maximizes the geometric mean fitness, denoted by $\hat\rho$, also varies with $k$ in the $AkBk$ regimes.
-It is somewhat surprising, given the modifier findings for $k\ge 3$, that $\hat\rho$ is actually zero for $k\le11$ and, depending on the fitnesses, is larger than zero for $k>11$.
+ The value of $\rho$ that maximizes the geometric mean fitness, denoted by $\hat\rho$,  varies with $k$ and with $W/w$ in the $AkBk$ regimes.
 This is shown in Figure 9, where for $W=1$ with $w=0.1$ and $0.5$, $\hat\rho$ is noticeably positive.
@@
 \medskip
-\centerline{[FIGURES 8, 9 HERE]}  
+\centerline{[FIGURES 7, 8, 9 HERE]}  
  \bigskip
@@
  for $j=0,1,2,\dots,N$.
- Thus the fluctuations in the numbers of phenotypes $A$ and $B$ in the  population of size $N$ are generated by the Wright-Fisher Markov chain process where, given that $X_t=Nx$, $X_{t+1}$ has a binomial distribution with parameters $(N,x')$.
+ Thus the fluctuations in the numbers of phenotypes $A$ and $B$ in the  population of size $N$ are generated by the Wright-Fisher Markov chain  where, given that $X_t=Nx$, $X_{t+1}$ has a binomial distribution with parameters $(N,x')$.
 
- This Markov chain process has two absorbing states, $X_t =N$ and $X_t=0$, corresponding to the two fixations in $A$ and $B$, respectively, and we are interested in the fixation probabilities and the time to fixation of these two absorbing states as functions of the initial frequency $x$ and also of $\rho$, $w_A$, and $w_B$.
+ This Markov chain  has two absorbing states, $X_t =N$ and $X_t=0$, corresponding to the two fixations in $A$ and $B$, respectively, and we are interested in the fixation probabilities and the time to fixation of these two absorbing states as functions of the initial frequency $x$ and also of $\rho$, $w_A$, and $w_B$.
 
@@
 
-  \proclaim Result 9. When $s>0$ so that the phenotype $A$ is favored, the fixation probability $u(x)$ is monotone increasing in $\rho$.\par
+  \proclaim Result 10. When $s>0$ so that the phenotype $A$ is favored, the fixation probability $u(x)$ is monotone increasing in $\rho$.\par
 
- The proof of Result 9 is in Supplementary Materials SP3.  
+ The proof of Result 10 is in Supplementary Materials SP4.  
   Figure 10 compares the fixation probability and time to fixation derived numerically from  simulating the Wright-Fisher Markov chain with the diffusion-derived values of $u(x)$ and $T(x)$.
@@
 
- \proclaim Result 10.
+ \proclaim Result 11.
 The mean $\mu(x)$ and variance $\sigma^2(x)$ of the change in the frequency of $A$ in one generation for the diffusion approximation in the case of a cycling environment $AkBl$, where $k+l=n$, are
@@
 
- The proof of Result 10, based on induction on $n$, is given in Supplemental Materials SP4.
+ The proof of Result 11, based on induction on $n$, is given in Supplemental Materials SP5.
 
@@
 In particular, with deterministically cycling symmetric fitness values (the $AkBl$ model), it is impossible for fixation in $A$ and $B$ to both be stable.
-If $k=l$, for example, neither fixation is stable, and from numerical iteration it appears that there is a simple stable polymorphic equilibrium (with phenotypes $A$ and $B$ present; see Result 3).
+If $k=l$, for example, neither fixation is stable, and there is a single stable polymorphic equilibrium (with phenotypes $A$ and $B$ present; see Result 5).
 In the $A1B1$ case this  polymorphism is globally stable.
 In the $AkBl$ case, bounds on $l/k$ that determine the instability of both fixations, and hence the protection of polymorphism are given by inequalities (20), which depend on both the fitness differences and the rate $\rho$ of vertical transmission.
+We conjecture that with $k\ne l$ there is a unique stable polymorphism when both fixations are unstable.
 
@@
  When the fitnesses $w_A$ and $w_B$ are treated as random variables, rather than varying cyclically, stochastic local stability is the appropriate analog to local stability in the case of cyclic fitness variation.
-While fixations in phenotypes $A$ and $B$ cannot both be stable in this case, both may be unstable and a polymorphic distribution may result.
+While fixations in phenotypes $A$ and $B$ cannot both be stochastically stable in this case, both may be unstable and a polymorphic distribution may result.
 The variance of this distribution is greater for larger values of $\rho$.
@@
 
- We have shown that the rate of vertical transmission tends to decrease when it is under the control of a genetic modifier.
-This is true in the deterministic case with cycling environments, and also in the random selection case, when the fitnesses of $A$ and $B$ are identically distributed and independent between generations.
-In light of these results, together with the finite population result that fixation probability is higher, and time to fixation faster with greater $\rho$, we can conclude that increased oblique transmission in our model leads to greater phenotypic diversity; i.e., it is more likely to result in  phenotypic polymorphism.
+ We have shown that in the $A1B1$ case, the rate of vertical transmission tends to decrease when it is under the control of a genetic modifier.
+This is also true in the random selection case, when the fitnesses of $A$ and $B$ are identically distributed and independent between generations.
+ %In light of these results, together with the finite population result that fixation probability is higher, and time to fixation faster with greater $\rho$, we can conclude that increased oblique transmission in our model leads to greater phenotypic diversity; i.e., it is more likely to result in  phenotypic polymorphism.
 This can be seen in Figure 1, where it is also clear that the existence and stability of a polymorphism is more sensitive to $(1-\rho)$ than to the fitness difference between $A$ and $B$.
@@
 
-\ref Mazel, D. Integrons: agents of bacterial evolution. {\it Nat.\ Rev.\ Microbiol.} {\bf 4}: 608--620.
+\ref Mazel, D. 2006. Integrons: agents of bacterial evolution. {\it Nat.\ Rev.\ Microbiol.} {\bf 4}: 608--620.
 
-\ref Milkman, R., and M. M. Bridges. Molecular evolution of the {\it Escherichia coli} chromosome. III. Clonal frames. {\it Genetics} {\bf 126} 505--517.
+\ref Milkman, R., and M. M. Bridges. 1990. Molecular evolution of the {\it Escherichia coli} chromosome. III. Clonal frames. {\it Genetics} {\bf 126} 505--517.
 
-\ref Redfield, R. J. Evolution of natural transformation: testing the DNA repair hypothesis in {\it Bacillus subtilis} and {\it Haemophilus influenzae}. {\it Genetics} {\bf 133}: 755--761.
+\ref Redfield, R. J. 1993. Evolution of natural transformation: testing the DNA repair hypothesis in {\it Bacillus subtilis} and {\it Haemophilus influenzae}. {\it Genetics} {\bf 133}: 755--761.
 
@@
 
- \noindent{\bf SP1. Proof of Result 5}
+ \noindent{\bf SP1. Proof of Uniqueness in Result 5}
  \medskip
 
- \noindent{\it Proof of Result 5}. Rewrite recursion (31) as
- $${x_t+1\over x_t} =(1+\rho s_t)\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right].\eqno(\hbox{\tenrm S}1)$$
- Then
- $$\log x_{t+1} -\log x_t =\log(1+\rho s_t) +\log\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right].\eqno(\hbox{\tenrm S}2)$$
- Summation yields
- $${1\over t}\left[\log x_t-\log x_0\right] ={1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n) +{1\over t}\sum_{n=0}^{t-1}\log\left[1-x_n{\rho s_n(1+s_n)\over (1+\rho s_n)(1+x_ns_n)}\right].\eqno(\hbox{\tenrm S}3)$$
-Let $\mu=E\left[\log(1+\rho s_t)\right]$. As $\{s_t\}_{t\ge 0}$ are independent and identically distributed random variables, the {\sl strong law of large numbers} applies and
-$$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)=\mu\eqno(\hbox{\tenrm S}4)$$
-almost surely.
+ Following (10) the transformation of the frequency $x$ of phenotype $A$ is
+ $$x'=F(x) =\underbrace{F_B\circ\cdots\circ F_B}_{k\ \hbox{\sevenrm times}}\;\circ\; \underbrace{F_A\circ\cdots\circ F_A}_{k\ \hbox{\sevenrm times}}(x).\eqno(\hbox{\tenrm S}1)$$
+ Using (28) and (29) we can write
+ $$\eqalignno{
+ &F'_A(x) =\rho{wW\over \bigl[\overline w(x)\bigr]^2} +(1-\rho),\quad \overline w(x)=(W-w)x +w&(\hbox{\tenrm S}2)\cr\noalign{\smallskip}
+ &F'_B(x) =\rho{wW\over \bigl[\widetilde w(x)\bigr]^2} +(1-\rho),\quad \widetilde w(x)=(w-W)x +W.&(\hbox{\tenrm S}3)\cr
+ }$$
+Since $F'_A(x) >0$, $F'_B(x)>0$ for  $0\le x\le 1$,  all the functions $F_A$, $F_B$, $F_A\circ\cdots\circ F_A$, $F_B\circ\cdots\circ F_B$, and $F$ are monotone increasing for  $0\le x\le 1$.
 
-Let $\zeta$ be such that ${1\over t}\sum_{n=0}^{t-1}\log[1+\rho s_n(\zeta)]=\mu$ and assume that $\lim_{t\to\infty}x_t(\zeta)=0$. As the random variables $\{s_t\}_{t\ge 0}$ are uniformally bounded,
-$$x_t(\zeta){\rho s_t(\zeta)[1+s_t(\zeta)]\over [1+\rho s_t(\zeta)][1+x_t(\zeta)s_t(\zeta)]}\;\;\harr{}{t\to\infty}\;\;0\eqno(\hbox{\tenrm S}5)$$
-and
-$$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log\left[1-x_n(\zeta){\rho x_n(\zeta)[1+s_n(\zeta)]\over [1+\rho s_n(\zeta)][1+x_t(\zeta)s_n(\zeta)]}\right]=0.\eqno(\hbox{\tenrm S}6)$$
-Thus (S3) implies that
-$$\lim_{t\to\infty}{1\over t}\left[\log x_t(\zeta) -\log x_0(\zeta)\right]=\mu.\eqno(\hbox{\tenrm S}7)$$
-If $\mu =E\left[\log(1+s_t)\right]>0$, then from (S7) we deduce that $\lim_{t\to\infty}x_t(\zeta)=\infty$, a contradiction.
-Therefore when $\mu>0$, $P\left(\lim_{t\to\infty}x_t=0\right)=0$, and fixation of $B$ ($x^*=0$) is stochastically locally unstable.
+From Result 2, the two fixations $x=0$ and $x=1$ are not stable because $F'(0)>1$ and $F'(1)>1$. Therefore
+$$\eqalignno{
+&F(x)-x>0\ \hbox{\tenrm for}\ x>0\ \hbox{\tenrm ``near''}\ x=0,&\hbox{\tenrm (S4)}\cr
+&F(x)-x<0\ \hbox{\tenrm for}\ x<1\ \hbox{\tenrm ``near''}\ x=1.&\hbox{\tenrm (S5)}\cr}$$
+Hence, as $F(x)-x$ is a continuous function of $x$ for $0\le x\le 1$, there exists (at least one) polymorphic equilibrium $x^*$ with $0<x^*<1$ such that $F(x^*)=x^*$.
 
- Thus by Result 5, for $x^*=0$ to be stochastically locally stable it is necessary that $E[\log(1+\rho s_t)]\le 0$. In fact, the strict inequality is sufficient.
+If there is more than one polymorphic equilibrium, and as there is a finite number of equilibria, let $x^*$ be the ``closest'' polymorphic equilibrium to $x=0$.
+Since $F(x)>x$ for $0<x<x^*$, $F(x)<x$ for $x>x^*$ (at least ``near'' $x^*$), and $F(x)$ is a monotone increasing function in $[0,1]$,  $x^*$ must be locally stable.
 
- Figure 3 presents a numerical example of the dynamics of recursion (31) with a specific random selection coefficient $s_t$.
+Let $\hat x=F(\hat x)$ with $0<\hat x<1$ be any polymorphic equilibrium, then from (S1) its evolution in the $k+k$ generations is 
+$$\vbox{\halign{
+\hfil$#$\hfil &&\hfil\ $#$\hfil\cr
+&A&&A&&\cdots&&A&&B&&B&&\cdots&&B&&A\cr
+\hat x=&\hat x_0&\to&\hat x_1&\to&\cdots&\to&x_{k-1}&\to&\hat y_0&\to&\hat y_1&\to&\cdots&\to&\hat y_{k-1}&\to&\hat x_0&=\hat x.\cr
+}}\eqno(\hbox{\tenrm S}6)$$
+Due to the symmetry between phenotypes $A$ and $B$ we have
+$$\hat y_t=1-\hat x_t,\qquad \overline w\bigl(\hat x_t\bigr) =\widetilde w\bigl(\hat y_t\bigr)\eqno(\hbox{\tenrm S}7)$$
+for all $t=0,1,2,\dots,k-1$.
+
+The polymorphic equilibrium $\hat x$ is locally stable if $F'(\hat x)<1$, or from (S1), (S2), and (S3), if
+$$\prod_{t=0}^{k-1}\left\{\rho{wW\over\bigl[\overline w(\hat x_t)\bigr]^2} +(1-\rho)\right\}\cdot\prod_{t=0}^{k-1} \left\{\rho{wW\over\bigl[\widetilde w(\hat y_t)\bigr]^2} +(1-\rho)\right\}<1.\eqno(\hbox{\tenrm S}8)$$
+Applying (S7), we conclude that $\hat x$ is locally stable if
+$$\prod_{t=0}^{k-1}\left\{\rho{wW\over\bigl[\overline w(\hat x_t)\bigr]^2} +(1-\rho)\right\}<1.\eqno(\hbox{\tenrm S}9)$$
+
+As $x^*$, the ``closest'' polymorhpic equilibrium to $x=0$, is stable, then (S9) implies that
+$$\prod_{t=0}^{k-1}\left\{\rho{wW\over\bigl[\overline w(x_t^*)\bigr]^2} +(1-\rho)\right\}\le 1,\eqno(\hbox{\tenrm S}10)$$
+ where $x_t^*$ for $t=0,1,\dots,k-1$ is defined as in (S6).
+
+ If $\hat x$ is any  polymorphic equilibrium other than $x^*$, then $\hat x>x^*$ or $\hat x_0>x_0^*$ by (S4). Since $F_A$ is a monotone increasing function and
+ $$\hat x_{t+1} =F_A\bigl(\hat x_t\bigr),\quad x_{t+1}^* =F_A\bigl(x_t^*\bigr)\quad t=0,1,\dots,k-1,\eqno(\hbox{\tenrm S}11)$$
+ then by induction we have $\hat x_t>x_t^*$ for all $t=0,1,2,\dots,k-1$.
+In addition, as $\overline w(x) =(W-w)x+w$ and $W>w$, we also have $\overline w(\hat x_t)>\overline w(x_t^*)$ for all $t=0,1,2,\dots,k-1$, and
+ $$\prod_{t=0}^{k-1}\left\{\rho{wW\over\bigl[\overline w(\hat x_t)\bigr]^2} +(1-\rho)\right\}<\prod_{t=0}^{k-1} \left\{\rho{wW\over\bigl[\overline w(x_t^*)\bigr]^2} +(1-\rho)\right\}\le 1.\eqno(\hbox{\tenrm S}12)$$
+Hence $\hat x$ is also locally stable.
+But it is impossible that {\sl all} polymorphic equilibria are stable unless there is only one stable polymorphic equilibrium.
+Therefore $x^*$ is the unique stable polymorphic equilibrium, and since $F(x)>x$ for $0<x<x^*$ with $F(x)<x$ for $x^*<x<1$, and $F(x)$ is monotone increasing in $[0,1]$, therefore $x^*$ is globally stable.
   \bigskip
@@
 
+ Rewrite recursion (31) as
+ $${x_t+1\over x_t} =(1+\rho s_t)\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right].\eqno(\hbox{\tenrm S}13)$$
+ Then
+ $$\log x_{t+1} -\log x_t =\log(1+\rho s_t) +\log\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right].\eqno(\hbox{\tenrm S}14)$$
+ Summation yields
+ $${1\over t}\left[\log x_t-\log x_0\right] ={1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n) +{1\over t}\sum_{n=0}^{t-1}\log\left[1-x_n{\rho s_n(1+s_n)\over (1+\rho s_n)(1+x_ns_n)}\right].\eqno(\hbox{\tenrm S}15)$$
+Let $\mu=E\left[\log(1+\rho s_t)\right]$. As $\{s_t\}_{t\ge 0}$ are independent and identically distributed random variables, the {\sl strong law of large numbers} applies and
+$$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)=\mu\eqno(\hbox{\tenrm S}16)$$
+almost surely.
+
+Let $\zeta$ be such that ${1\over t}\sum_{n=0}^{t-1}\log[1+\rho s_n(\zeta)]=\mu$ and assume that $\lim_{t\to\infty}x_t(\zeta)=0$. As the random variables $\{s_t\}_{t\ge 0}$ are uniformally bounded,
+$$x_t(\zeta){\rho s_t(\zeta)[1+s_t(\zeta)]\over [1+\rho s_t(\zeta)][1+x_t(\zeta)s_t(\zeta)]}\;\;\harr{}{t\to\infty}\;\;0\eqno(\hbox{\tenrm S}17)$$
+and
+$$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log\left[1-x_n(\zeta){\rho x_n(\zeta)[1+s_n(\zeta)]\over [1+\rho s_n(\zeta)][1+x_t(\zeta)s_n(\zeta)]}\right]=0.\eqno(\hbox{\tenrm S}18)$$
+Thus (S15) implies that
+$$\lim_{t\to\infty}{1\over t}\left[\log x_t(\zeta) -\log x_0(\zeta)\right]=\mu.\eqno(\hbox{\tenrm S}19)$$
+If $\mu =E\left[\log(1+s_t)\right]>0$, then from (S19) we deduce that $\lim_{t\to\infty}x_t(\zeta)=\infty$, a contradiction.
+Therefore when $\mu>0$, $P\left(\lim_{t\to\infty}x_t=0\right)=0$, and fixation of $B$ ($x^*=0$) is stochastically locally unstable.
+
+ Thus by Result 6, for $x^*=0$ to be stochastically locally stable it is necessary that $E[\log(1+\rho s_t)]\le 0$. In fact, the strict inequality is sufficient.
+
+ Figure 3 presents a numerical example of the dynamics of recursion (31) with a specific random selection coefficient $s_t$.
+\vfil\break
+%\bigskip
+%\bigskip
+
+ \noindent{\bf SP3. Proof of Result 7}
+ \medskip
+
 Let $\mu=E[\log(1+\rho s_t)]$. Then as $\{s_t\}_{t\ge 0}$ are independent and identically distributed random variables, the strong law of large number applies and almost surely
- $$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)=\mu<0.\eqno(\hbox{\tenrm S}8)$$
+ $$\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)=\mu<0.\eqno(\hbox{\tenrm S}20)$$
  Appealing to the Egoroff Theorem, for any $\var>0$ there exists $T$ such that 
- $$P\left({1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)<{\mu\over 2}\ \hbox{\tenrm for all}\ t\ge T\right)\ge 1-\var.\eqno(\hbox{\tenrm S}9)$$
+ $$P\left({1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)<{\mu\over 2}\ \hbox{\tenrm for all}\ t\ge T\right)\ge 1-\var.\eqno(\hbox{\tenrm S}21)$$
  As $0\le \rho\le 1$ and the $\{s_t\}_{t\ge 0}$ are uniformly bounded, we can find a $\delta'>0$ such that
- $$x_t<\delta'\Longrightarrow\left|\log\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right]\right|<-{\mu\over 4}.\eqno(\hbox{\tenrm S}10)$$
+ $$x_t<\delta'\Longrightarrow\left|\log\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right]\right|<-{\mu\over 4}.\eqno(\hbox{\tenrm S}22)$$
  Also, as $0\le x_t\le 1$ for all $t$,
- $$x_{t+1}=x_t{1+\rho s_t +x_t(1-\rho)s_t\over 1+x_ts_t}< Kx_t,\eqno(\hbox{\tenrm S}11)$$
+ $$x_{t+1}=x_t{1+\rho s_t +x_t(1-\rho)s_t\over 1+x_ts_t}< Kx_t,\eqno(\hbox{\tenrm S}23)$$
  where $K$ is independent of $t$. It follows that there exists a $\delta$ with $0<\delta<\delta'$ such that
- $$x_o<\delta\Longrightarrow x_t<\delta'\ \hbox{\tenrm for all}\ t=0,1,2,\dots,T-1.\eqno(\hbox{\tenrm S}12)$$
+ $$x_o<\delta\Longrightarrow x_t<\delta'\ \hbox{\tenrm for all}\ t=0,1,2,\dots,T-1.\eqno(\hbox{\tenrm S}24)$$
  Let $\xi$ be a realization of the evolutionary process such that
- $${1\over t}\sum_{n=0}^{t-1}\log[1+\rho s_n(\xi)]<{\mu\over 2}\ \hbox{\tenrm for all}\  t\ge T\eqno(\hbox{\tenrm S}13)$$
+ $${1\over t}\sum_{n=0}^{t-1}\log[1+\rho s_n(\xi)]<{\mu\over 2}\ \hbox{\tenrm for all}\  t\ge T\eqno(\hbox{\tenrm S}25)$$
  and assume $x_0<\delta$. Then
@@
  &={1\over T}\sum_{n=0}^{T-1}\log[1+\rho s_n(\xi)]+{1\over T}\sum_{n=0}^{T-1}\log[1-x_n(\xi){\rho s_n(\xi)[1+s_n(\xi)]\over [1+\rho s_n(\xi)][1+x_n(\xi)s_n(\xi)]}\cr\noalign{\smallskip}
- &<{\mu\over 2}- {\mu\over 4} ={\mu\over 4}<0, }\eqno(\hbox{\tenrm S}14)$$
+ &<{\mu\over 2}- {\mu\over 4} ={\mu\over 4}<0, }\eqno(\hbox{\tenrm S}26)$$
  and therefore $x_T(\xi) <x_0(\xi) <\delta'$. Invoking induction we get that for $t\ge T$
- $${1\over t}\log{x_t(\xi)\over x_0}\le {\mu\over 4},\eqno(\hbox{\tenrm S}15)$$
+ $${1\over t}\log{x_t(\xi)\over x_0}\le {\mu\over 4},\eqno(\hbox{\tenrm S}27)$$
  or for all $t\ge T$
- $$x_t(\xi)\le x_0\exp\left({\mu\over 4}t\right).\eqno(\hbox{\tenrm S}16)$$
+ $$x_t(\xi)\le x_0\exp\left({\mu\over 4}t\right).\eqno(\hbox{\tenrm S}28)$$
 
@@
 Therefore we have shown that for given $\var>0$ there is a $\delta>0$ such that if $0<x_0<\delta$, then $P\left(\lim_{t\to\infty}x_t =0\right)\ge 1-\var$, and therefore $x^*=0$, the fixation in $B$ is stochastically locally stable.
-The second statement of Result 5 follows from the convexity of the log function and Jensen's inequality.
-\bigskip
-\bigskip
-
- \noindent{\bf SP3. Proof of Result 9}
- \medskip
-
-Without loss of generality and for the ease of representation, we will show that for $t>0$, 
-  $$v (x;t) ={1-e^{-tx}\over 1-e^{-t}}\eqno(\hbox{\tenrm S}17)$$
-  is monotone increasing as a function of $t$. Observe that
-  $${\partial v\over\partial t} = {\left(1-e^{-t}\right)xe^{-tx} -\left(1-e^{-tx}\right)e^{-t}\over \left(1-e^{-t}\right)^2}.\eqno(\hbox{\tenrm S}18)$$
-
-   \noindent For the monotonicity we have to show that
-  $$f(x;t) =\left(1-e^{-t}\right)xe^{-tx} -\left(1 -e^{-tx}\right)e^{-t} \ge 0\eqno(\hbox{\tenrm S}19)$$ 
-  when $t>0$ and $0\le x\le 1$. Note that $f(0;t)=0$ and $f(1;t)=0$. Also
-  $${\partial f\over\partial x} =\left(1-e^{-t}\right)\left(e^{-tx}-txe^{-tx}\right) -te^{-tx}e^{-t},\eqno(\hbox{\tenrm S}20)$$
-  or
-  $${\partial f\over\partial x} =e^{-tx}\left[\left(1-e^{-t}\right)\left(1-tx\right) -te^{-t}\right] =e^{-tx}g(x;t),\eqno(\hbox{\tenrm S}21)$$
- say, where for fixed $t$, $g(x;t)$ is a linear function of $x$, which vanishes at $x_0 =(1-e^{-t} -te^{-t})/t(1-e^{-t})$.
-If $t>0$, $e^t >1+t$, so $1>e^{-t}(1+t)$ and $x_0>0$.
-Also if $t>0$, $e^{-t}>1-t$, and so $1-e^{-t} -te^{-t} <t(1-e^{-t})$ and $x_0<1$.
-Since $g(0,t) =1 -e^{-t} -te^{-t} >0$ and $g(1;t) =(1-e^{-t})(1-t) -te^{-t} <0$ for $t>0$, we deduce that ${\partial f\over\partial x}(x,t)>0$ for $ 0<x<x_0$ and ${\partial f\over\partial x}(x,t)<0$ for $x_0<x<1$ for all $t>0$.
-These facts, combined with $f(0,t) =f(1;t) =0$, prove that $f(x;t)\ge 0$ for $0\le x\le 1$ (in fact, $f(x;t)>0$ for $0<x<1$), and inequality (S19) is satisfied as desired.
+The second statement of Result 6 follows from the convexity of the log function and Jensen's inequality.
 \bigskip
@@
 
+Without loss of generality and for the ease of representation, we will show that for $t>0$, 
+  $$v (x;t) ={1-e^{-tx}\over 1-e^{-t}}\eqno(\hbox{\tenrm S}29)$$
+  is monotone increasing as a function of $t$. Observe that
+  $${\partial v\over\partial t} = {\left(1-e^{-t}\right)xe^{-tx} -\left(1-e^{-tx}\right)e^{-t}\over \left(1-e^{-t}\right)^2}.\eqno(\hbox{\tenrm S}30)$$
+
+   \noindent For the monotonicity we have to show that
+  $$f(x;t) =\left(1-e^{-t}\right)xe^{-tx} -\left(1 -e^{-tx}\right)e^{-t} \ge 0\eqno(\hbox{\tenrm S}31)$$ 
+  when $t>0$ and $0\le x\le 1$. Note that $f(0;t)=0$ and $f(1;t)=0$. Also
+  $${\partial f\over\partial x} =\left(1-e^{-t}\right)\left(e^{-tx}-txe^{-tx}\right) -te^{-tx}e^{-t},\eqno(\hbox{\tenrm S}32)$$
+  or
+  $${\partial f\over\partial x} =e^{-tx}\left[\left(1-e^{-t}\right)\left(1-tx\right) -te^{-t}\right] =e^{-tx}g(x;t),\eqno(\hbox{\tenrm S}33)$$
+ say, where for fixed $t$, $g(x;t)$ is a linear function of $x$, which vanishes at $x_0 =(1-e^{-t} -te^{-t})/t(1-e^{-t})$.
+If $t>0$, $e^t >1+t$, so $1>e^{-t}(1+t)$ and $x_0>0$.
+Also if $t>0$, $e^{-t}>1-t$, and so $1-e^{-t} -te^{-t} <t(1-e^{-t})$ and $x_0<1$.
+Since $g(0,t) =1 -e^{-t} -te^{-t} >0$ and $g(1;t) =(1-e^{-t})(1-t) -te^{-t} <0$ for $t>0$, we deduce that ${\partial f\over\partial x}(x,t)>0$ for $ 0<x<x_0$ and ${\partial f\over\partial x}(x,t)<0$ for $x_0<x<1$ for all $t>0$.
+These facts, combined with $f(0,t) =f(1;t) =0$, prove that $f(x;t)\ge 0$ for $0\le x\le 1$ (in fact, $f(x;t)>0$ for $0<x<1$), and inequality (S31) is satisfied as desired.
+\bigskip
+\bigskip
+
+ \noindent{\bf SP5. Proof of Result 11}
+ \medskip
+
  The proof is based on induction on $n$, where in order to prove (74), we show that if $X_t$ is the number of individuals with phenotype $A$ at stage $t$ of the cycle, and $x$ is the initial frequency of $A$, then
- $$E\left({X_t\over N} -x\right) \simeq {1\over N}\rho S_t x(1-x),\qquad V\left({X_t\over N}\right)\simeq {1\over N}tx(1-x),\eqno(\hbox{\tenrm S}22)$$
+ $$E\left({X_t\over N} -x\right) \simeq {1\over N}\rho S_t x(1-x),\qquad V\left({X_t\over N}\right)\simeq {1\over N}tx(1-x),\eqno(\hbox{\tenrm S}34)$$
  where $N$ is the size of the population.
-When $t=1$, (S22) coincides with the constant environment case.
-Assuming (S22), we go to $t+1$.
+When $t=1$, (S34) coincides with the constant environment case.
+Assuming (S34), we go to $t+1$.
 Now $X_{t+1}$ given $X_t=Ny$ has a binomial distribution with parameters $(N,y')$.
 Hence
- $$E\left({X_{t+1}\over N} -{X_t\over N} \mid X_t=Ny\right) =y'-y.\eqno(\hbox{\tenrm S}23)$$
+ $$E\left({X_{t+1}\over N} -{X_t\over N} \mid X_t=Ny\right) =y'-y.\eqno(\hbox{\tenrm S}35)$$
  Following (57), $y'-y\simeq (1/N)\rho s_{t+1}y(1-y)$, and so
- $$E\left({X_{t+1}\over N} -{X_t\over N}\mid X_t\right)\simeq {1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right).\eqno(\hbox{\tenrm S}24)$$
+ $$E\left({X_{t+1}\over N} -{X_t\over N}\mid X_t\right)\simeq {1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right).\eqno(\hbox{\tenrm S}36)$$
  Observe that
@@
  E\left[{X_t\over N}\left(1-{X_t\over N}\right)\right] &= E\left({X_t\over N}\right) -E\left[\left({X_t\over N}\right)^2\right]\cr\noalign{\smallskip}
-    &= E\left({X_t\over N}\right) -V\left({X_t\over N}\right) -\left[E\left({X_t\over N}\right)\right]^2.\cr}\eqno(\hbox{\tenrm S}25)$$
+    &= E\left({X_t\over N}\right) -V\left({X_t\over N}\right) -\left[E\left({X_t\over N}\right)\right]^2.\cr}\eqno(\hbox{\tenrm S}37)$$
  By the induction assumption, $V(X_t/N)\simeq (1/N)tx(1-x)$, and ignoring terms of order $1/N^2$ we have
  $$E\left({X_{t+1}\over N} -{X_t\over N}\right)\simeq {1\over N}\rho s_{t+1} E\left({X_t\over N}\right)\left[1-E\left({X_t\over N}\right)\right].
-\eqno(\hbox{\tenrm S}26)$$
-Applying (S22) we have
+\eqno(\hbox{\tenrm S}38)$$
+Applying (S34) we have
 $$\eqalign{
 E\left({X_t\over N}\right)  &\simeq x+{1\over N}\rho S_tx(1-x)\cr\noalign{\smallskip}
-1-E\left({X_t\over N}\right) &\simeq 1-x-{1\over N}\rho S_tx(1-x),} \eqno(\hbox{\tenrm S}27)$$
+1-E\left({X_t\over N}\right) &\simeq 1-x-{1\over N}\rho S_tx(1-x),} \eqno(\hbox{\tenrm S}39)$$
 and ignoring terms $O(1/N^2)$, we find
-$$E\left({X_{t+1}\over N} -{X_t\over N}\right) \simeq {1\over N}\rho s_{t+1}x(1-x).\eqno(\hbox{\tenrm S}28)$$
+$$E\left({X_{t+1}\over N} -{X_t\over N}\right) \simeq {1\over N}\rho s_{t+1}x(1-x).\eqno(\hbox{\tenrm S}40)$$
 Thus
@@
 E\left({X_{t+1}\over N} -x\right) &= E\left({X_{t+1}\over N} -{X_t\over N}\right) +E\left({X_t\over N} -x\right)\cr\noalign{\smallskip}
-&\simeq {1\over N}\rho s_{t+1}x(1-x) +{1\over N}\rho S_t x(1-x),}\eqno(\hbox{\tenrm S}29)$$
+&\simeq {1\over N}\rho s_{t+1}x(1-x) +{1\over N}\rho S_t x(1-x),}\eqno(\hbox{\tenrm S}41)$$
 and since $S_t +s_{t+1} =S_{t+1}$,
- $$E\left({X_{t+1}\over N} -x\right)\simeq	{1\over N}\rho S_{t+1}x(1-x)\eqno(\hbox{\tenrm S}30)$$
+ $$E\left({X_{t+1}\over N} -x\right)\simeq	{1\over N}\rho S_{t+1}x(1-x)\eqno(\hbox{\tenrm S}42)$$
  as desired.
@@
  We now compute $V(X_{t+1}/N)$ using the induction assumption and the formula
- $$V\left({X_{t+1}\over N}\right) =E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] +V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right],\eqno(\hbox{\tenrm S}31)$$
- where by (S22)
- $$E\left({X_{t+1}\over N}\mid X_t\right)\simeq {X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\eqno(\hbox{\tenrm S}32)$$
+ $$V\left({X_{t+1}\over N}\right) =E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] +V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right],\eqno(\hbox{\tenrm S}43)$$
+ where by (S34)
+ $$E\left({X_{t+1}\over N}\mid X_t\right)\simeq {X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\eqno(\hbox{\tenrm S}44)$$
  and
- $$V\left({X_{t+1}\over N}\mid X_t\right)\simeq{1\over N}{X_t\over N}\left(1-{X_t\over N}\right).\eqno(\hbox{\tenrm S}33)$$
+ $$V\left({X_{t+1}\over N}\mid X_t\right)\simeq{1\over N}{X_t\over N}\left(1-{X_t\over N}\right).\eqno(\hbox{\tenrm S}45)$$
  Here we used the fact that by (63) $y'(1-y')\simeq y(1-y)$. Now
- $$E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] \simeq {1\over N}E\left[{X_t\over N}\left(1-{X_t\over N}\right)\right]\simeq {1\over N}x(1-x),\eqno(\hbox{\tenrm S}34)$$
- where we use the same computations as led from (S25) to (S28).
- $$V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right] =V\left[{X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right].\eqno(\hbox{\tenrm S}35)$$
+ $$E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] \simeq {1\over N}E\left[{X_t\over N}\left(1-{X_t\over N}\right)\right]\simeq {1\over N}x(1-x),\eqno(\hbox{\tenrm S}46)$$
+ where we use the same computations as led from (S37) to (S40).
+ $$V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right] =V\left[{X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right].\eqno(\hbox{\tenrm S}47)$$
  Since $(X_t/N)\bigl[1-(X_t/N)\bigr]$ is a random variable taking values in [0,1],  its variance is less than 1/4 and
- $$V\left[{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right] \le {1\over 4N^2}\rho^2 s_{t+1}^2.\eqno(\hbox{\tenrm S}36)$$
+ $$V\left[{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right] \le {1\over 4N^2}\rho^2 s_{t+1}^2.\eqno(\hbox{\tenrm S}48)$$
  We ignore terms $O(1/N^2)$ so that the random variable $(1/N)\rho s_{t+1}(X_t/N)\bigl[1-(X_t/N)\bigr]$ is almost constant. As a result,
- $$V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right]\simeq V\left({X_t\over N}\right) \simeq {1\over N}tx(1-x),\eqno(\hbox{\tenrm S}37)$$
- by the induction assumption. Combining (S34) and (S37) gives
- $$V\left({X_{t+1}\over N}\right)\simeq {1\over N}x(1-x) +{1\over N}tx(1-x) ={1\over N}(t+1)x(1-x)\eqno(\hbox{\tenrm S}38)$$
+ $$V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right]\simeq V\left({X_t\over N}\right) \simeq {1\over N}tx(1-x),\eqno(\hbox{\tenrm S}49)$$
+ by the induction assumption. Combining (S46) and (S49) gives
+ $$V\left({X_{t+1}\over N}\right)\simeq {1\over N}x(1-x) +{1\over N}tx(1-x) ={1\over N}(t+1)x(1-x)\eqno(\hbox{\tenrm S}50)$$
  as expected.

### 2017-09-19 cluster/bethedging.sh
obsolete

@@ -1,14 +1,0 @@
-python src/simulation.py --Ne 100000 --n 500 --η1 0.0 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.0001 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.001 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.01 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.02 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.03 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.04 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.05 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.06 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.07 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.08 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.09 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
-python src/simulation.py --Ne 100000 --n 500 --η1 0.1 --ω0 2.0 --ω1 0.0  --π0 0.5  --env A --reps 10 --cpus 3
[…]

### 2017-09-19 src/get_bibtex.py

@@ -1,52 +1,0 @@
-import re
-
-from bibtexparser import load as load_bib
-from bibtexparser.bwriter import BibTexWriter
-from bibtexparser.bibdatabase import BibDatabase
-import click
-
-
-species_pattern = re.compile(
-	r'({\\textless}i{\\textgreater}\w.*?{\\textless}/i{\\textgreater})'
-)
-
-@click.command()
-@click.argument('keys_filename', type=click.Path(exists=True, file_okay=True, dir_okay=False, readable=True))
-@click.argument('bibtex_filename', type=click.Path(exists=True, file_okay=True, dir_okay=False, readable=True))
-@click.argument('output_filename', type=click.Path(file_okay=True, dir_okay=False, writable=True))
-@click.option('-v/-V', '--verbose/--no-verbose', default=False)
-def main(keys_filename, bibtex_filename, output_filename, verbose):
-	with open(keys_filename) as f:
-		citation_keys = (line.strip() for line in f.readlines())
-	if verbose:
-		print("Read {} keys from {}".format(len(citation_keys), citation_keys))
-
-	with open(bibtex_filename) as f:
-		main_bib = load_bib(f)
-	if verbose:
-		print("Read {} entries from {}".format(len(main_bib.entries), bibtex_filename))
-
-	out_bib = BibDatabase()
-	for key in citation_keys:
-		e = main_bib.entries_dict[key]		
-		title = e['title']
-		groups = species_pattern.findall(title)
-		for grp in groups:
-			s = grp.replace(
-				'{\\textless}i{\\textgreater}', ''
-			).replace(
-				'{\\textless}/i{\\textgreater}', ''
-			)
-			s = '\\textit{\\uppercase{' + s[0] + '}' + s[1:] + '}'
-			title = title.replace(grp, s)
-		e['title'] = title		
-		out_bib.entries.append(e)
-	if verbose:
-		print("Writing {} entries to {}".format(len(out_bib.entries), output_filename))
-	writer = BibTexWriter()
-	with open(output_filename, 'w') as f:
-	    f.write(writer.write(out_bib))
-
-
-if __name__ == '__main__':
-	main()

### 2017-09-19 src/get_citations.py

@@ -1,19 +1,0 @@
-import re
-
-import click
-
-@click.command()
-@click.argument('markdown_filename', type=click.Path(exists=True, file_okay=True, dir_okay=False, readable=True))
-@click.argument('keys_filename', type=click.Path(file_okay=True, dir_okay=False, writable=True))
-def main(markdown_filename, keys_filename):
-	pattern = re.compile('@([-\w]+\d{4}[a-z]?)')
-	with open(markdown_filename) as f:
-		groups = (pattern.findall(line) for line in f)
-		groups = sum((g for g in groups if g), [])	
-	groups = set(groups)
-	with open(keys_filename, 'wt') as f:
-		for g in groups:
-			print(g, file=f)
-
-if __name__ == '__main__':
-	main()

### 2017-09-19 src/plots.py

@@ -1,183 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-TODO
-
-Created on Sun Jan 08 10:43:42 2017
-
-@author: yoav@yoavram.com
-"""
-import numpy as np
-import matplotlib.pyplot as plt
-import seaborn as sns
-from simulation import simulation, parse_env, parse_π0
-
-
-def plot_π(π, ϵ=None, bands=8, label=None, ax=None):    
-    if ax is None:
-        fig, ax = plt.subplots()
-    for i in range(bands):
-        low, high = np.percentile(π, [0.5*(2**i), 100-0.5*(2**i)], 1)
-        ax.fill_between(range(π.shape[0]), low, high, color='b', alpha=0.1)
-    ax.plot(π.mean(axis=1), c='y', lw=3)
-    if ϵ is not None:
-        ax.axhline((ϵ == 0).mean(), color='k', ls='--')
-    ax.set(
-        ylim=(0, 1),
-        ylabel='$π$',
-        xlabel='t'
-    )
-    if label is not None:
-        ax.set_label(label)
-    ax.set_clip_on(False)
-    sns.despine()
-    return ax
-
-def plot_logNtN0(π, ω0, ω1, p0, ϵ, ax=None):
-    if ax is None:
-        fig, ax = plt.subplots()
-    
-    # estimated log(Nt/N0)
-    ω = np.array(
-        [
-            [ω0, ω1],
-            [ω1, ω0]
-        ],
-    dtype=float)
-    π_bar = π.mean(axis=1)
-    Δ = np.log(π_bar * ω[ϵ, 0] + (1 - π_bar) * ω[ϵ, 1]) # eq. 17
-    logNtN0 = Δ.cumsum() # eq. 18
-    ax.plot(logNtN0, 'b')
-    # max possible log(Nt/N0)
-    maxlogNtN0 = np.repeat(np.log(ω0), logNtN0.size).cumsum()
-    ax.plot(maxlogNtN0, 'y') 
-    # bet-hedging  log(Nt/N0), eq. 11
-    ax.plot(np.repeat(
-        p0 * np.log(p0 * ω0 + (1 - p0) * ω1) + (1 - p0) * np.log(p0 * ω1 + (1 - p0) * ω0),
-        logNtN0.size
-    ).cumsum(), 'k')
-
-    ax.set(
-        xlabel='t',
-        ylabel='$log(N_t/N_0)$',
[…]

### 2017-09-19 src/random_simulation.py

@@ -1,55 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-TODO
-
-Created on Thu Nov 17 16:24:45 2016
-
-@author: yoav@yoavram.com
-"""
-import click
-from ultrachronic import repeat
-import numpy as np
-from simulation import __version__
-from simulation import __name__ as prog_name
-from simulation import _main
-
-def min_l_for_fixation(k, η, W, w):
-    return k * (1 + (1 - η) *(W - w) / w)
-
-def max_l_for_no_fixation(k, η, W, w):
-    return k * (1 + ((1 - η) *(W - w) / w)/(1+η*(1 - η)*(W-w)**2/(W*w)))
-
-@click.command()
-@click.version_option(version=__version__, prog_name=prog_name)
-@click.option('--Ne', default=None, type=int, help="Population size")
-@click.option('--ω0', default=1.0, help="Fitness of phenotype 0 in environment 0")
-@click.option('--π0', default='0.5', type=str,
-              help="Initial density of phenotype 0; one of Nx, U, Cx, x where 0<x<1, N is normal (default), U is uniform, C is constant.")
-@click.option('--light', default=False, is_flag=True, help="Follow only last generation to save memory")
-@click.option('--output_folder', default='output', type=click.Path(),
-              help="Output folder to which results will be written, defaults to 'output'")
-def main(ne, ω0, π0, output_folder, light):
-    W = ω0
-    N = ne
-    η = round(10 ** np.random.uniform(-3, -1), 4)
-    w = round(10 ** np.random.uniform(-2, -1), 4)
-    k = int(np.random.randint(1, 40))
-
-    Ls = [
-        k,
-        np.floor(max_l_for_no_fixation(k, η, W, w)) // 2,
-        np.floor(max_l_for_no_fixation(k, η, W, w)),
-        np.ceil(min_l_for_fixation(k, η, W, w)),
-        2 * np.ceil(min_l_for_fixation(k, η, W, w)),
-    ]
-    Ls = map(int, Ls)
-
-    for l in Ls:
-        n = min(1000000, int(500 * (k + l)))
-        env = 'A{:d}B{:d}'.format(k, l)
-        print('env={}, n={}'.format(env, n))
-        _main(N, n, η, None, 0, None, W, w, π0, 0, env, output_folder, light)
-
-
-if __name__ == '__main__':
-    main()

### 2017-09-19 src/simulation.py

@@ -1,399 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Thu Nov 17 16:24:45 2016
-
-@author: yoav@yoavram.com
-"""
-import datetime
-import os
-import gzip
-import json
-import glob
-import re
-import warnings
-# filter seaborn nasty warning
-warnings.filterwarnings('ignore', 'The `IPython.html` package has been deprecated. You should import from `notebook` instead. `IPython.html.widgets` has moved to `ipywidgets`.')
-
-import click
-import numpy as np
-import pandas as pd
-from ultrachronic import repeat
-
-__version__ = '0.0.6'
-env_pattern = re.compile("A(?P<k>\d+)B(?P<l>\d+)")
-
-
-def simulation(N, n, η, ρ, ω0, ω1, π0, ϵ=None, light=False):
-    """Run a single simulation.
-    
-    Parameters
-    ----------
-    N : int
-        constant population size
-    n : int
-        number of generations
-    η : float
-        learning rate, 0 <= η <= 1
-    ρ : float
-        conformity effect, 0 <= ρ <= 1
-    ω0, ω1 : float
-        two fitness values for the two phenotyes in the two environments, ω > 0
-    π0 : function
-        function that given N, returns an initial value for π.
-    ϵ : numpy.ndarray
-        ϵ[t] is the environment at time t
-    light : bool
-        if True, simulation is memory-light, returning only π of last generation
-
-    Returns
-    -------
-    π : np.ndarray
-        π[t, i] is the probability for phenotype 0 in individual i at time t
-    """
-    ω = np.array(
-        [
-            [ω0, ω1],
-            [ω1, ω0]
-        ],
-    dtype=float)
-    
-    if ϵ is None:
-        ϵ = np.random.randint(0, 2, n)
-    # π[t, i] is the probability for phenotype 0 atin individual i at time t
-    if light:
-        π = np.zeros((2, N), dtype=float)
-    else:
-        π = np.zeros((n, N), dtype=float)
-    π[0, :] = π0(N)
-    π[0, π[0, :] < 0] = 0
-    π[0, π[0, :] > 1] = 1
-    π[1, :] = π[0, :] # for light simulations
-
-    for t in range(n - 1):
-        ϵ_ = ϵ[t]
-        t_ = t + 1
-        if light:
[…]

### 2017-09-19 src/test_simulation.py

@@ -1,56 +1,0 @@
-#!/usr/bin/env python
-# -*- coding: utf-8 -*-
-from unittest import TestCase, main
-
-import numpy as np
-import simulation
-
-
-class SimulationConstantEnvironment(TestCase):
-
-    def setUp(self):
-        self.ε = np.zeros(1000, dtype=int)
-        def isogenic(N):
-            x = np.ones(N) * 0.5
-            return x
-        self.π0 = isogenic
-
-    def test_default_pop(self):
-        π = simulation.simulation(1000, 200, 0.1, 0, 2.0, 0.2, self.π0, self.ε)
-
-        assert np.allclose(π[-1, :], 1), np.unique(π[-1, :])
-
-    def test_light(self):
-        π = simulation.simulation(1000, 200, 0.1, 0, 2.0, 0.2, self.π0, self.ε, light=True)
-        assert π.ndim == 1, π.shape
-        assert np.allclose(π, 1), np.unique(π)
-
-    def test_η_0(self):
-        # this won't work with isogenic - totally up to selection here
-        def uniform(N):
-            x = np.random.rand(N)
-            return x
-        π = simulation.simulation(1000, 1000, 0, 0, 2.0, 0.2, uniform, self.ε)
-
-        assert (π[-1, :] > 0.99).all(), np.unique(π[-1, :])
-
-
-class DeterministicConstantEnvironment(TestCase):
-
-    def setUp(self):
-        self.ε = np.zeros(1000, dtype=int)
-
-    def test_isogenic_pop_η_0_s_0(self):
-        π, f = simulation.deterministic(1000, 100, 0, 0, 1, 0, 0.5, self.ϵ)
-
-        assert np.allclose(π[-1], 0.5), np.unique(π[-1])
-
-    def test_isogenic_pop_η_0(self):
-        π, f = simulation.deterministic(1000, 200, 0.1, 0, 1, 0, 0.5, self.ϵ)
-
-        assert np.allclose(π[-1], 1), np.unique(π[-1])
-
-    def test_uniform_pop_η_0(self):
-        π, f = simulation.deterministic(1000, 200, 0.1, 0, 1, 0, lambda: simulation.uniform(10), self.ϵ)
-
-        assert np.allclose(π[-1], 1), np.unique(π[-1])

### 2017-09-19 ms/collect_figures.py
collect figures to a zip file

@@ -1,0 +1,37 @@
+# import tempfile
+import shutil
+import os.path
+from os.path import join as join_path
+from os.path import split as split_path
+import re
+from zipfile import ZipFile
+
+if __name__ == '__main__':
+	ms_dir = '../ms'
+	figures_dir = '../figures'
+	ms_figures_dir = join_path(ms_dir, 'figures')
+	zip_fname = '/Users/yoavram/Dropbox/Milpitas/figures.zip'
+	tex_files = ['figures.tex', 'figures_supp.tex']
+	pattern = re.compile(r'\\includegraphics\{(.*\.\w{3})\}')
+
+	if os.path.exists(ms_figures_dir):
+		shutil.rmtree(ms_figures_dir)
+		os.mkdir(ms_figures_dir)
+
+	figures = []
+	for fn in tex_files:
+		with open(join_path(ms_dir, fn)) as f:
+			matches = (pattern.match(line) for line in f)
+			matches = (m for m in matches if m is not None)
+			filenames = (m.groups()[0] for m in matches)
+			filenames = (split_path(fn)[-1] for fn in filenames)
+			filenames = (join_path(figures_dir, fn) for fn in filenames)
+			figures.extend(filenames)
+
+	with ZipFile(zip_fname, 'w') as z:
+		for fn in figures:
+			shutil.copy(fn, join_path(ms_figures_dir, split_path(fn)[-1]))
+			z.write(fn)
+
+	print("{} figures copied to {} and zipped to {}".format(
+		len(figures), ms_figures_dir, zip_fname))

### 2017-09-19 ms/figures.tex
figures tex

--- 2017-08-24 ms/figures.tex
@@
 \centering
-\includegraphics{../figures/env_A1B1.pdf}
-\caption{Frequency of phenotype \emph{A} after every two generations in
-selection regime \(A1B1\). The orange line is the finite population model
-(eqs. 52-53; average of 100 simulations). The blue line is the infinite population model
-(eq.\ 23), and the green line is the  solution of \(G(x)=0\) (see eq.\ 24).
-Parameters: \(W=1\); For the finite population model (orange lines), population size is \(N=10,000\) and initial frequency of \(A\) is \(x_0=0.5\).}\label{fig:env_A1B1}
-\end{figure}
-
-%Figure 3
-\begin{figure}
-\centering
 \includegraphics{../figures/A1B1_equilibrium.pdf}
@@
 
-%Figure 4
+%Figure 3
 \begin{figure}
@@
 The gray lines mark combinations of \emph{p} and \emph{s} for which \(E[\log{(1-\rho+\rho\frac{w_A}{w_B})}]=0\) and \(E[\log{(1-\rho+\rho\frac{w_B}{w_A})}]=0\).
-According to Result 6, between these lines fixation of either phenotype will be stochastically locally unstable, and indeed we have a stationary polymorphism between the lines, with phenotype frequencies neither 0 or 1.
+According to Result 6, between these lines fixation of either phenotype will be stochastically locally unstable, and we expect a stationary polymorphism between the lines.
 Parameters: initial frequency of \(A\) is \(x_0=1/10,000\) and the vertical transmission rate is \(\rho=0.1\).}\label{fig:stochastic_env_x_t}
@@
 
-%Figure 5
+%Figure 4
 \begin{figure}
@@
 
-%Figure 6
+%Figure 5
 \begin{figure}
@@
 
-%Figure 7
+%Figure 6
 \begin{figure}
@@
 \includegraphics[height=\dimexpr \textheight - 8\baselineskip\relax]{../figures/stoch_modifier_invasions.png}
-\caption{Consecutive fixation of modifiers that reduce the vertical
-transmission rate $\rho$ under symmetric randomly changing selection. The figure shows
-results of numerical simulations of evolution with two modifier alleles
-(eq.\ 33). When a modifier allele fixes
-(frequency\textgreater{}99.9\%), a new modifier allele is introduced
-with a vertical transmission rate one order of magnitude lower (vertical
-dashed lines). \textbf{(A)} The frequency of phenotype \emph{A} in the
-population over time. \textbf{(B)} The frequency of the invading
-modifier allele over time. Parameters: vertical transmission rate of the initial
-resident modifier allele, \(\rho_0 =0.1\); fitness values: $W=1$
-and $w=0.1$ with probability 0.5 and $w=0.1$ and $W=1$
-also with probability 0.5. The x-axis is on a log-scale, as each sequential invasion takes
-an order of magnitude longer to
-complete.}\label{fig:stoch_modifier_invasions}
+\caption{Consecutive fixation of modifiers that reduce the vertical transmission rate $\rho$ under symmetric randomly changing selection.
+The figure shows results of numerical simulations of evolution with two modifier alleles
+(eq.\ 33).
+When a modifier allele fixes (frequency\textgreater{}99.9\%),
+a new modifier allele is introduced with a vertical transmission rate one order of magnitude lower (vertical dashed lines).
+\textbf{(A)} The frequency of phenotype \emph{A} in the population over time.
+\textbf{(B)} The frequency of the invading modifier allele over time.
+Parameters: vertical transmission rate of the initial resident modifier allele is \(\rho_0 =0.1\) and the fitness values ratio is $w_A/w_B=10$ with probability 0.5 and $w_A/w_B=0.1$ also with probability 0.5.
+The x-axis is on a log-scale, as each sequential invasion takes an order of magnitude longer to complete.}\label{fig:stoch_modifier_invasions}
 \end{figure}
 
-%Figure 8
+%Figure 7
 \begin{figure}
@@
 
-%Figure 9
+%Figure 8
 \begin{figure}
@@
 
-%Figure 10
+%Figure 9
 \begin{figure}
@@
 
-%Figure 11
+%Figure 10
 \begin{figure}
@@
 
-%Figure 12
+%Figure 11
 \begin{figure}

### 2017-10-03 ms/stability_analysis.tex
first draft of computational stability analysis for modifiers

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{lineno}
\linenumbers
% math typesetting 
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}

%SetFonts

%SetFonts

\pagestyle{headings}
\markright{Yoav Ram, \today\hfill}

\begin{document}
Here I describe how I analyzed the stability of a modifier allele $m$ with vertical transmission rate $\rho$ to invasion by a modifier $M$ with a vertical transmission rate $P$, as described in the main text, eqs. 34-35, in environmental regime $AkBl$.
\\
The analysis is similar to that used in the main text to analyze stability in $A1B1$, but it uses computation because the cases where $k>1$ or $l>1$ cannot be analyzed using closed form expressions.
\\
The analysis includes the following steps for a fixed $W, w, k,$ and $l$:
\begin{enumerate}

\item Find the stable frequency vector $\vec{x^*}$ without invader $M$, i.e. $x^*_3=x^*_4=0$. This is done by minimizing the expression $|x_{k+l} - x_{0}|$ where $x_{t}$ is defined in eq. 9 of the main text. The minimization is done by iterating the recurrence until it converges, i.e. until the difference is smaller than machine precision (roughly $10^-8$ when subtracting similar small numbers).

\item Define $F_A(x)$ by eq. 35 of main text with $w_A=W$ and $w_B=w$ ($W>w$), and similarly $F_B(x)$ with $w_B=W$ and $w_A=w$.

\item Define, similar to eq. 37 of main text,
\begin{equation}
\vec{F} = \underbrace{F_A \circ \ldots \circ F_A}_{k \text{ times}} \circ
	\underbrace{F_B \circ \ldots \circ F_B}_{l \text{ times}}.
\label{eq:F}\end{equation}

\item Calculate the linear approximation $\vec{J}$ of $\vec{F}$  near $\vec{x^*}$, that is, the Jacobian matrix of $\vec{F}$ at $\vec{x^*}$:
\begin{equation}
\vec{J}_{ij} = \frac{\partial F_i}{\partial x_j}.
\label{eq:jacobian}\end{equation}

\item Define $\vec{L}=\vec{L_{ex}}$ as in eqs. 40-45 in the main text such that
\begin{equation}
\vec{J} = \begin{pmatrix}
\vec{L_{in}} & * \\
* & \vec{L_{ex}}
\end{pmatrix}.
\end{equation}

\item Calculate the eigenvalues $\lambda_1 > \lambda_2$ of $\vec{L}$; by the Perron-Frobenius theorem, the leading eigenvalue $\lambda_1$ is real and positive.

\item Finally, if $\lambda_1>1$ then $M$ can invade into $m$ and if $\lambda_1<1$, $m$ is stable to invasion by $M$.

\end{enumerate}

These steps give us a function $\lambda_1(W, w, \rho, P, k, l)$ for the eigenvalue that determines the stability of a modifier $m$ with rate $\rho$ to invasion of modifier $M$ with rate $P$ with fitness values $W, w$ in environmental regime $AkBl$ (see~\autoref{fig:AkBk_stable_modifier_w_0.1}).
\\

To find a vertical transmission rate $\rho^*$ that is stable to invasion,
we use the following algorithm (in the spirit of gradient descent, or rather, ascent):
\begin{algorithm} 
% https://en.wikibooks.org/wiki/LaTeX/Algorithms
\caption{Modifier gradient ascent algorithm}\label{algorithm}
\begin{algorithmic}[1]
\State $\rho_{0} = 0.5$
\State $t = 0$
\While {$\frac{|\rho_{t} - \rho_{t-1}|}{\rho_{t-1}} < \epsilon \;\text{and}\; \eta > \eta_0$}
	\State $P_{t} \gets \rho_{t} + \eta \frac{\partial \lambda_1(\rho_{t}, \rho_{t})}{\partial P}$
	\If {$\lambda_1(\rho_{t}, P_{t})$} 
	\State {$\rho_{t+1} \gets P_{t}$}
	\Else
	\State {$\eta \gets \eta / 2$}
	\EndIf
	\State $t \gets t+1$
\EndWhile
\State $\rho^* \gets \rho_t$
\end{algorithmic}
\end{algorithm}

where here $\lambda_1(\rho_, \rho)$ means $\lambda_1(W, w \rho, P, k, l)$ as $W, w, k$ and $l$ are all fixed, $\eta$ is a "learning rate" parameter, initialized to 0.005-0.05 (lower for larger $k$ and $l$), and $\eta_0$ and $\epsilon$ determine the stopping and convergence of the algorithm (balancing computational time and precision requirements, I use $\eta_0= \epsilon = 0.01$).
\\
The arrows in~\autoref{fig:AkBk_stable_modifier_w_0.1} show examples of resulting $\{\rho_{t}\}$ sequences.

\paragraph{Results}
Examining ~\autoref{fig:AkBk_stable_modifier_w_0.1}, this method seems to work well for $k=1$ and $10$, less so for $k=20$ and not at all for $k=40$.
This is probably because when the environment is constant for a long period of time, the frequencies are close to the boundaries (i.e. 0 and 1) and the gradient just after the environment changes is very sharp.
Therefore, the calculation of $\vec{J}$ is unstable.
Also, because most of the time the frequencies don't change by much, the elements of $\vec{L}$ are small, leading to unstable calculation of the leading eigenvalue.
\\

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\linewidth]{{../figures/AkBk_stable_modifier_w_0.1}.pdf}
  \caption{Stable oblique transmission in $A1kBk$ with selection. The figure shows $\lambda_1(W, w, \rho, P, k, l)$, the leading eigenvalue of the external stability matrix for combinations of $\rho$ and $P$, the vertical transmission rates of resident and invader modifiers, respectively.
  The arrows show an "evolutionary" path $\rho_{0} \to \rho_{1} \to \ldots \to \rho^*$: at each step, a modifier with rate $P_t$ invades the resident modifier with rate $\rho_t$ and becomes the resident modifier $\rho_{t+1}$, until consecutive invaders have similar rates (see Algorithm~\autoref{algorithm}).
  Here, $W=1$ and $w=0.1$}
  \label{fig:AkBk_stable_modifier_w_0.1}
\end{figure}

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\linewidth]{{../figures/AkBk_stable_modifier_w_0.5}.pdf}
  \caption{Stable oblique transmission in $A1kBk$ with intermediate selection.
  Same as~\autoref{fig:AkBk_stable_modifier_w_0.1} except here, $W=1$ and $w=0.5$}
  \label{fig:AkBk_stable_modifier_w_0.5}
\end{figure}

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\linewidth]{{../figures/AkBk_stable_modifier_w_0.9}.pdf}
  \caption{Stable oblique transmission in $A1kBk$ with intermediate selection.
  Same as~\autoref{fig:AkBk_stable_modifier_w_0.1} except here, $W=1$ and $w=0.9$}
  \label{fig:AkBk_stable_modifier_w_0.9}
\end{figure}

\paragraph{Note}
Crucially, the Jacobian $\vec{J}$ in~\autoref{eq:jacobian} can be calculated using \textit{automatic differentiation} with good precision from a function that iteratively calculates $\vec{F}$ according to~\autoref{eq:F}. Similarly, the partial derivative $\frac{\partial \lambda_1}{\partial P}$ in Algorithm~\autoref{algorithm} can be calculated from a function that calculates $\lambda_1$ using simple arithmetic operations, as the eigenvalues can be found from a quadratic formula.
\\
Note that \textit{automatic differentiation} does not mean \textit{symbolic} or \textit{numerical differentiation}, which can lead to inefficient or inaccurate estimation of $\vec{J}$ when $k$ is not very small. Rather, from Bartholomew-Biggs et al. (2000): 
\begin{quotation}\textit{
Automatic differentiation is a set of techniques for transforming a program that calculates numerical values of a function, into a program which calculates numerical values for derivatives of that function with about the same accuracy and efficiency as the function values themselves.
}\end{quotation}

\paragraph{References} ~\\
Bartholomew-Biggs, M., S. Brown, B. Christianson, and L. Dixon. 2000. Automatic differentiation of algorithms. J. Comput. Appl. Math. 124:171–190.

\end{document}

### 2017-10-04 ms/stability_analysis.tex
suggesting another optimization algorithm using a loss function

@@
 
+\paragraph{Finding $\rho^*$}
 To find a vertical transmission rate $\rho^*$ that is stable to invasion,
-we use the following algorithm (in the spirit of gradient descent, or rather, ascent):
+I suggest Algorithm \autoref{algorithm}.
+
 \begin{algorithm} 
@@
 
-where here $\lambda_1(\rho_, \rho)$ means $\lambda_1(W, w \rho, P, k, l)$ as $W, w, k$ and $l$ are all fixed, $\eta$ is a "learning rate" parameter, initialized to 0.005-0.05 (lower for larger $k$ and $l$), and $\eta_0$ and $\epsilon$ determine the stopping and convergence of the algorithm (balancing computational time and precision requirements, I use $\eta_0= \epsilon = 0.01$).
+Here $\lambda_1(\rho_, P)=\lambda_1(W, w, \rho, P, k, l)$ as $W, w, k$ and $l$ are all fixed, $\eta$ is a "learning rate" parameter, initialized to 0.005-0.05 (lower for larger $k$ and $l$), and $\eta_0$ and $\epsilon$ determine the stopping and convergence of the algorithm (balancing computational time and precision requirements, I use $\eta_0= \epsilon = 0.01$).
 \\
 The arrows in~\autoref{fig:AkBk_stable_modifier_w_0.1} show examples of resulting $\{\rho_{t}\}$ sequences.
+However, the results are sensitive to the choice of $\eta$.
+
+\paragraph{Another approach to find $\rho^*$}
+Note that
+\begin{align}
+\lambda_1(\rho^* \pm \epsilon, \rho^*) > 1 > \lambda_1(\rho^*, \rho^*), \\
+\lambda_1(\rho^*, \rho^* \pm \epsilon) < 1 < \lambda_1(\rho^*, \rho^*)
+\end{align}
+where the first inequality assumes that a modifier with $\rho^*$ can invade resident modifiers with slightly smaller or higher rates and the second inequality is the definition of a stable rate.
+These inequalities imply that at $(\rho^*,\rho^*)$
+\begin{align}
+\lambda_1 = 1, \\
+\frac{\partial \lambda_1}{\partial \rho} = \frac{\partial \lambda_1}{\partial P} = 0, \\
+\frac{\partial^2 \lambda_1}{\partial \rho^2} > 0, \\
+\frac{\partial^2 \lambda_1}{\partial P^2} < 0. 
+\end{align}
+
+We can use the following loss function
+\begin{equation}
+\mathbb{L}(z) =  \Big(\frac{\partial\lambda_1}{\partial \rho}(z, z)\Big)^2 + 
+		\Big(\frac{\partial\lambda_1}{\partial P}(z, z)\Big)^2
+\label{eq:loss} \end{equation}
+Minimizing $\mathbb{L}(z)$ will result in a point $(z,z)$ at which the gradients of $\lambda_1(\rho, P)$ are both zero, and therefore $(z,z)$ will be a stationary point of $\lambda_1$.
+As $\lambda_1(z, z)=1$ for any $z \in (0,1)$, the point $(z,z)$ cannot be a minimum or a maximum, and therefore it must be a saddle point (\textbf{is that right?}).
 
@@
     \includegraphics[width=\linewidth]{{../figures/AkBk_stable_modifier_w_0.1}.pdf}
-  \caption{Stable oblique transmission in $A1kBk$ with selection.
-The figure shows $\lambda_1(W, w, \rho, P, k, l)$, the leading eigenvalue of the external stability matrix for combinations of $\rho$ and $P$, the vertical transmission rates of resident and invader modifiers, respectively.
+  \caption{Stable oblique transmission in $AkBk$ with selection.
+The colors show $\lambda_1(W, w, \rho, P, k, l)$, the leading eigenvalue of the external stability matrix for combinations of $\rho$ (y-axis) and $P$ (x-axis), the vertical transmission rates of resident and invader modifiers, respectively.
+Purple shows values lower than 1, leading to stability of $\rho$; white equals to 1; green larger than 1, leading to invasion of $P$..
   The arrows show an "evolutionary" path $\rho_{0} \to \rho_{1} \to \ldots \to \rho^*$: at each step, a modifier with rate $P_t$ invades the resident modifier with rate $\rho_t$ and becomes the resident modifier $\rho_{t+1}$, until consecutive invaders have similar rates (see Algorithm~\autoref{algorithm}).
-  Here, $W=1$ and $w=0.1$}
+  Here, $W=1$ and $w=0.1$.}
   \label{fig:AkBk_stable_modifier_w_0.1}
@@
   \caption{Stable oblique transmission in $A1kBk$ with intermediate selection.
-  Same as~\autoref{fig:AkBk_stable_modifier_w_0.1} except here, $W=1$ and $w=0.5$}
+  Same as~\autoref{fig:AkBk_stable_modifier_w_0.1} except here, $W=1$ and $w=0.5$.}
   \label{fig:AkBk_stable_modifier_w_0.5}
@@
   \caption{Stable oblique transmission in $A1kBk$ with intermediate selection.
-  Same as~\autoref{fig:AkBk_stable_modifier_w_0.1} except here, $W=1$ and $w=0.9$}
+  Same as~\autoref{fig:AkBk_stable_modifier_w_0.1} except here, $W=1$ and $w=0.9$.}
   \label{fig:AkBk_stable_modifier_w_0.9}

### 2017-10-05 src/geom.py
merged geom2 and oblique

@@ -1,14 +1,120 @@
-from oblique import *
- 
+from itertools import cycle, count
+from uuid import uuid4 as uuid
+
+import numpy as np
+import numba
+import scipy.optimize
+
+
+@numba.jit()
+def recurrence(x, ρ, wA, wB, N=0, err=1e-14):
+    N = int(N)
+    wbar = x * wA + (1 - x) * wB
+    # same precision around x=0 and x=1
+    if np.all(x > 0.5): 
+        x = 1 - (ρ * (1 - x) * wB/wbar + (1 - ρ) * (1 - x))
+    else:
+        x = ρ * x * wA/wbar + (1 - ρ) * x
+    # move from edges to get out of non-stable equilibriums at x=0 and x=1
+    if hasattr(x, 'size') and x.size > 1:
+        x[x < err ] = err
+        x[x > 1 - err] = 1 - err
+    else:
+        if x < err:
+            x = err
+        elif x > 1 - err:
+            x = 1 - err
+    if N > 0:
+        x = np.random.binomial(N, x) / N
+    return x
+    
+
+def is_polymorphism(w, ρ, k, l):
+    # see eq 22
+    s = 1 - w
+    if ρ*s == 0:
+        return True
+    elif k == l:
+        return True
+    elif l > k:
+        return l/k < -np.log(1+ρ*s/(1-s))/np.log(1-ρ*s) 
+    else: # k > l
+        return k/l < -np.log(1+ρ*s/(1-s))/np.log(1-ρ*s) 
+
+
+def max_ρ_with_polymorphism(w, k, l, num_pts=1001):
+    for ρ_ in np.linspace(0, 1, num_pts)[::-1]:
+        if is_polymorphism(w, ρ_, k, l): 
+            return ρ_
+
+
+def geom_avg_wbar(x, w, k, l):
+    x = x[-k-l:]
+    wA = np.array([1]*k + [w]*l, dtype=float).reshape((k+l, 1))
+    wB = np.array([w]*k + [1]*l, dtype=float).reshape((k+l, 1))
+    wbars = x * wA + (1 - x) * wB
+    assert (wbars > 0).all(), wbars
+    res = scipy.stats.mstats.gmean(wbars)
+    assert (0 < res).all(), (res, k, l)
+    assert (res < 1).all(), (res, k, l)
+    return res
+
+
[…]

### 2017-10-05 src/geom2.py

@@ -1,33 +1,0 @@
-from uuid import uuid4 as uuid
-import numpy as np
-from oblique import max_ρ_with_polymorphism, geom_wbar_target, geom_avg_wbar, stable_cycle
-import scipy.optimize
-
-def optimal_ρ(w, k, l):    
-    max_ρ = max_ρ_with_polymorphism(w, k, l)
-    if np.isclose(max_ρ, 0):
-        return np.nan
-    res = scipy.optimize.minimize_scalar(
-        geom_wbar_target, args=(w, k, l), bounds=[0, max_ρ], method='bounded')
-    if not res.success: 
-        print('Minimization failed for w={}, k={}, l={}'.format(w, k, l))
-        return np.nan
-    return res.x
-
-if __name__ == '__main__':
-	ks = np.arange(1, 51, 1, dtype=int)
-	ws = np.array([0.1, 0.5, 0.9])
-
-	ρs = np.empty((ws.size, ks.size))
-	for i, w in enumerate(ws):
-		for j, k in enumerate(ks):
-			l = k
-			print("w={}, k={}, l={}".format(w, k, l))
-			ρs[i, j] = optimal_ρ(w, k, l)
-			wbar0 = geom_avg_wbar(stable_cycle(0, w, k, l), w, k, l)
-			wbarρ = geom_avg_wbar(stable_cycle(ρs[i, j], w, k, l), w, k, l)
-			if wbar0 > wbarρ:
-				ρs[i, j] = 0
-			print("ρ={}".format(ρs[i, j]))
-	fname = 'geom2_{}.npz'.format(uuid().hex)
-	np.savez_compressed(fname, ks=ks, ws=ws, ρs=ρs)

### 2017-10-05 src/geom2.py
calc fitness optimal rho

@@ -1,0 +1,28 @@
+from uuid import uuid4 as uuid
+import numpy as np
+from oblique import max_ρ_with_polymorphism, geom_wbar_target
+import scipy.optimize
+
+def optimal_ρ(w, k, l):    
+    max_ρ = max_ρ_with_polymorphism(w, k, l)
+    if np.isclose(max_ρ, 0):
+        return np.nan
+    res = scipy.optimize.minimize_scalar(
+        geom_wbar_target, args=(w, k, l), bounds=[0, max_ρ], method='bounded')
+    if not res.success: 
+        print('Minimization failed for w={}, k={}, l={}'.format(w, k, l))
+        return np.nan
+    return res.x
+
+if __name__ == '__main__':
+	ks = np.arange(1, 51, 1, dtype=int)
+	ws = np.array([0.1, 0.5, 0.9])
+
+	ρs = np.empty((ws.size, ks.size))
+	for i, w in enumerate(ws):
+		for j, k in enumerate(ks):
+			print("w={}, k={}, l={}".format(w, k, k))
+			ρs[i, j] = optimal_ρ(w, k, k)
+			print("ρ={}".format(ρs[i, j]))
+	fname = 'geom2_{}.npz'.format(uuid().hex)
+	np.savez_compressed(fname, ks=ks, ws=ws, ρs=ρs)

### 2017-10-05 src/stability_analysis.py
calculate stable rho (need to test)

@@ -1,0 +1,234 @@
+import sys
+from uuid import uuid4 as uuid
+from concurrent.futures import ProcessPoolExecutor as Executor
+from concurrent.futures import wait
+from functools import lru_cache
+import warnings
+
+import autograd.numpy as np
+from autograd import grad, jacobian
+from scipy import optimize as opt
+
+
+def f1(x, wA, wB, ρ, dx=0):    
+    wbar = x * wA + (1 - x) * wB
+    x = ρ * x * wA/wbar + (1 - ρ) * x
+    if dx > 0:
+        x += dx * np.sign(wB - wA)
+    return x
+df1 = grad(f1)
+
+def test_f1():
+    W, w = 1.0, 0.5
+    ρ = 0.5
+    x = 0.25
+    dx = 1e-4
+    assert np.isclose(
+        f1(x + dx, W, w, ρ, dx=1e-10), 
+        f1(x, W, w, ρ, dx=1e-10) + df1(x, W, w, ρ, dx=1e-10) * dx
+    )
+
+
+def F1(x, W, w, ρ, k, l, dx=0): 
+    for _ in range(k):
+        x = f1(x, W, w, ρ, dx=dx)
+    for _ in range(l):
+        x = f1(x, w, W, ρ, dx=dx)
+    return x
+dF1 = jacobian(F1)
+
+def test_F1():
+    W, w = 1.0, 0.5
+    ρ = 0.5
+    x = 0.25
+    dx = 1e-4
+    assert np.isclose(
+        F1(x + dx, W, w, ρ, 1, 1, dx=1e-10), 
+        F1(x, W, w, ρ, 1, 1, dx=1e-10) + dF1(x, W, w, ρ, 1, 1, dx=1e-10) * dx
+    )
+
+def stablex1(W, w, ρ, k, l, x0=0.5, dx=0):
+    x = x0
+    x_prev = -1
+    while not np.isclose(x, x_prev):
+        x, x_prev = F1(x, W, w, ρ, k, l, dx=dx), x
+#     assert np.isclose(F1(x, W, w, ρ, k, l, dx=dx), x)
+    return x
+
+lru_cache()
+def test_stablex1():
+    W, w = 1.0, 0.5
+    ρ = 0.5
+    k, l = 1, 1
+    x = stablex1(W, w, ρ, k, l, dx=1e-8)
+    assert np.isclose(x, F1(x, W, w, ρ, k, l, dx=1e-8))
+
+
[…]

### 2017-10-05 ms/stability_analysis.tex
fix wording

--- 2017-10-03 ms/stability_analysis.tex
@@
 \pagestyle{headings}
-\markright{Yoav Ram, \today\hfill}
+\markright{Yoav Ram, \today \; v2\hfill}
 
@@
 \item Calculate the eigenvalues $\lambda_1 > \lambda_2$ of $\vec{L}$; by the Perron-Frobenius theorem, the leading eigenvalue $\lambda_1$ is real and positive.
+Denote $\lambda_1(\rho, P)$ the resulting leading eigenvalue with resident rate $\rho$ and invader rate $P$.
+Note that
+\begin{equation}
+\lambda_1(\rho, \rho) = 1,
+\end{equation}
+for any $\rho \in (0,1)$.
 
-\item Finally, if $\lambda_1>1$ then $M$ can invade into $m$ and if $\lambda_1<1$, $m$ is stable to invasion by $M$.
+\item The evolutionary stable rate $\rho^*$ is defined as stable to invasion, that is, 
+\begin{equation}
+\lambda_1(\rho^*,\rho^* \pm dP) < 1 = \lambda_1(\rho^*,\rho^*)
+\end{equation}
+for a small enough $dP>0$.
+Therefore, 
+\begin{equation}
+\frac{\partial \lambda_1}{\partial P}\big(\rho^*,\rho^*\big) = 0. 
+\label{eq:dλdP=0} \end{equation}
+
+\item We use Brent's root-finding method (Brent, 1971) to find $\rho^*$ that satisfies \autoref{eq:dλdP=0}. If, due to numerical instability of the process, we have
+\begin{equation}
+\frac{\partial \lambda_1}{\partial P}\big(0,0\big) \cdot \frac{\partial \lambda_1}{\partial P}\big(1,1) > 0,
+\end{equation}
+i.e., the partial derivative sign is identical at $\rho=P=0$ and $\rho=P=1$, then we cannot use Brent's method.
+In these cases we assume that the partial derivative doesn't have a root in $(0,1)$ and we determine the stable rate $\rho^*$ by the rule
+\begin{equation}
+\rho^* = \begin{cases}
+0, \quad \text{if } \frac{\partial \lambda_1}{\partial P}\big(0,0\big) \le 0, \\
+1, \quad \text{if } \frac{\partial \lambda_1}{\partial P}\big(0,0\big) > 0.
+\end{cases}
+\end{equation}
 
@@
 
-These steps give us a function $\lambda_1(W, w, \rho, P, k, l)$ for the eigenvalue that determines the stability of a modifier $m$ with rate $\rho$ to invasion of modifier $M$ with rate $P$ with fitness values $W, w$ in environmental regime $AkBl$ (see~\autoref{fig:AkBk_stable_modifier_w_0.1}).
+\paragraph{Results}
+\autoref{fig:AkBk_stable_modifier_w_0.1}, \autoref{fig:AkBk_stable_modifier_w_0.5} and \autoref{fig:AkBk_stable_modifier_w_0.9}, show the leading eigenvalue $\lambda_1$ of the external stability matrix $\vec{L}$ for different choices of environmental cycles $AkBk$, resident rate $\rho$ and invader rate $P$.
+The arrows show a series of invading modifiers generated using Algorithm~\autoref{algorithm} (this algorithm cannot be efficiently used to determine the stable rate as it requires fine tuning of the meta-parameter $\eta$).
+
+The leading eigenvalue calculation is good for small $k$ but for large $k$ and especially for $w=0.1$ the calculation is unstable.
+This is probably because when the environment is constant for a long period of time, the frequencies are close to the boundaries (i.e.
+0 and 1) and the gradient just after the environment changes is very sharp.
+Therefore, the calculation of $\vec{J}$ is unstable.
+Also, because most of the time the frequencies don't change by much, the elements of $\vec{L}$ are small, leading to unstable calculation of the leading eigenvalue.
 \\
 
-To find a vertical transmission rate $\rho^*$ that is stable to invasion,
-we use the following algorithm (in the spirit of gradient descent, or rather, ascent):
 \begin{algorithm} 
@@
 \begin{algorithmic}[1]
-\State $\rho_{0} = 0.5$
-\State $t = 0$
+\State $\eta > 0$ is a step size
+\State $\rho_{0} = 0.5$ is a starting condition
+\State $t = 0$ is an iteration counter
 \While {$\frac{|\rho_{t} - \rho_{t-1}|}{\rho_{t-1}} < \epsilon \;\text{and}\; \eta > \eta_0$}
@@
 
-where here $\lambda_1(\rho_, \rho)$ means $\lambda_1(W, w \rho, P, k, l)$ as $W, w, k$ and $l$ are all fixed, $\eta$ is a "learning rate" parameter, initialized to 0.005-0.05 (lower for larger $k$ and $l$), and $\eta_0$ and $\epsilon$ determine the stopping and convergence of the algorithm (balancing computational time and precision requirements, I use $\eta_0= \epsilon = 0.01$).
-\\
-The arrows in~\autoref{fig:AkBk_stable_modifier_w_0.1} show examples of resulting $\{\rho_{t}\}$ sequences.
-
-\paragraph{Results}
-Examining ~\autoref{fig:AkBk_stable_modifier_w_0.1}, this method seems to work well for $k=1$ and $10$, less so for $k=20$ and not at all for $k=40$.
-This is probably because when the environment is constant for a long period of time, the frequencies are close to the boundaries (i.e.
-0 and 1) and the gradient just after the environment changes is very sharp.
-Therefore, the calculation of $\vec{J}$ is unstable.
-Also, because most of the time the frequencies don't change by much, the elements of $\vec{L}$ are small, leading to unstable calculation of the leading eigenvalue.
-\\
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=\linewidth]{../figures/AkBk_stable_optimal_rate.pdf}
+  \caption{Fitness "optimal" and evolutionary stable vertical transmission rate in $AkBk$ selection regime.
+\textbf{(A)} The vertical transmission rate $\hat{rho}$ that maximized the geometric average of the population mean fitness is zero, i.e.
+complete oblique transmission, when selection cycles quickly between favoring phenotype $A$ and $B$, and then abruptly transitions to ~0.2, followed by a slow decrease (see Fig.
+SX for details on the abrupt transition).
+  \textbf{(B)} The evolutionary stable rate $\rho^*$ which cannot be invaded by modifiers with either higher or lower vertical transmission rate $P$, rapidly increases from zero when selection cycles are short ($k=1$ or $2$) to roughly 1 when the environment cycles are longer.
+The stable rate was found by finding the root of $\frac{\partial \lambda_1}{\partial P}$, the leading eigenvalue of the external stability matrix w.r.t to the rate of the invading modifier (see Supplementary Information for details).
+The dashed lines shows $1-\frac{1}{k-1}$, which seems to fit the values for $w=0.5$ (Carja et al.
+2011).
+The values for $w=0.1$ (blue) could not be calculated for $k>21$ due to numerical instability when selection is strong and the duration between selection fluctuations is long.
+  In all cases, $W=1$}
+  \label{fig:AkBk_stable_modifier_w_0.1}
+\end{figure}
 
@@
     \includegraphics[width=\linewidth]{{../figures/AkBk_stable_modifier_w_0.1}.pdf}
-  \caption{Stable oblique transmission in $A1kBk$ with selection.
-The figure shows $\lambda_1(W, w, \rho, P, k, l)$, the leading eigenvalue of the external stability matrix for combinations of $\rho$ and $P$, the vertical transmission rates of resident and invader modifiers, respectively.
+  \caption{The leading eigenvalue of the external stability matrix in $AkBk$ with selection.
+The colors show $\lambda_1(W, w, \rho, P, k, l)$, the leading eigenvalue of the external stability matrix for combinations of $\rho$ (y-axis) and $P$ (x-axis), the vertical transmission rates of resident and invader modifiers, respectively.
+Purple shows values lower than 1, leading to stability of $\rho$; white equals to 1; green larger than 1, leading to invasion of $P$.
   The arrows show an "evolutionary" path $\rho_{0} \to \rho_{1} \to \ldots \to \rho^*$: at each step, a modifier with rate $P_t$ invades the resident modifier with rate $\rho_t$ and becomes the resident modifier $\rho_{t+1}$, until consecutive invaders have similar rates (see Algorithm~\autoref{algorithm}).
-  Here, $W=1$ and $w=0.1$}
+  Here, $W=1$ and $w=0.1$.}
   \label{fig:AkBk_stable_modifier_w_0.1}
@@
     \includegraphics[width=\linewidth]{{../figures/AkBk_stable_modifier_w_0.5}.pdf}
-  \caption{Stable oblique transmission in $A1kBk$ with intermediate selection.
-  Same as~\autoref{fig:AkBk_stable_modifier_w_0.1} except here, $W=1$ and $w=0.5$}
+  \caption{The leading eigenvalue of the external stability matrix in $AkBk$ with intermediate selection.
+  Same as~\autoref{fig:AkBk_stable_modifier_w_0.1} except here, $W=1$ and $w=0.5$.}
   \label{fig:AkBk_stable_modifier_w_0.5}
@@
     \includegraphics[width=\linewidth]{{../figures/AkBk_stable_modifier_w_0.9}.pdf}
-  \caption{Stable oblique transmission in $A1kBk$ with intermediate selection.
-  Same as~\autoref{fig:AkBk_stable_modifier_w_0.1} except here, $W=1$ and $w=0.9$}
+  \caption{The leading eigenvalue of the external stability matrix in $AkBk$ with intermediate selection.
+  Same as~\autoref{fig:AkBk_stable_modifier_w_0.1} except here, $W=1$ and $w=0.9$.}
   \label{fig:AkBk_stable_modifier_w_0.9}
@@
 \paragraph{References} ~\\
-Bartholomew-Biggs, M., S. Brown, B. Christianson, and L. Dixon. 2000. Automatic differentiation of algorithms. J. Comput. Appl. Math. 124:171–190.
+\begin{itemize}
+
+\item Bartholomew-Biggs, M., S. Brown, B. Christianson, and L. Dixon. 2000. Automatic differentiation of algorithms. J. Comput. Appl. Math. 124:171–190.
+
+\item Brent, R. P. 1971. An algorithm with guaranteed convergence for finding a zero of a function. Comput. J. 14:422–425.
+
+\end{itemize}
 

### 2017-10-05 src/geom.py
obsolete?

@@ -1,120 +1,0 @@
-from itertools import cycle, count
-from uuid import uuid4 as uuid
-
-import numpy as np
-import numba
-import scipy.optimize
-
-
-@numba.jit()
-def recurrence(x, ρ, wA, wB, N=0, err=1e-14):
-    N = int(N)
-    wbar = x * wA + (1 - x) * wB
-    # same precision around x=0 and x=1
-    if np.all(x > 0.5): 
-        x = 1 - (ρ * (1 - x) * wB/wbar + (1 - ρ) * (1 - x))
-    else:
-        x = ρ * x * wA/wbar + (1 - ρ) * x
-    # move from edges to get out of non-stable equilibriums at x=0 and x=1
-    if hasattr(x, 'size') and x.size > 1:
-        x[x < err ] = err
-        x[x > 1 - err] = 1 - err
-    else:
-        if x < err:
-            x = err
-        elif x > 1 - err:
-            x = 1 - err
-    if N > 0:
-        x = np.random.binomial(N, x) / N
-    return x
-    
-
-def is_polymorphism(w, ρ, k, l):
-    # see eq 22
-    s = 1 - w
-    if ρ*s == 0:
-        return True
-    elif k == l:
-        return True
-    elif l > k:
-        return l/k < -np.log(1+ρ*s/(1-s))/np.log(1-ρ*s) 
-    else: # k > l
-        return k/l < -np.log(1+ρ*s/(1-s))/np.log(1-ρ*s) 
-
-
-def max_ρ_with_polymorphism(w, k, l, num_pts=1001):
-    for ρ_ in np.linspace(0, 1, num_pts)[::-1]:
-        if is_polymorphism(w, ρ_, k, l): 
-            return ρ_
-
-
-def geom_avg_wbar(x, w, k, l):
-    x = x[-k-l:]
-    wA = np.array([1]*k + [w]*l, dtype=float).reshape((k+l, 1))
-    wB = np.array([w]*k + [1]*l, dtype=float).reshape((k+l, 1))
-    wbars = x * wA + (1 - x) * wB
-    assert (wbars > 0).all(), wbars
-    res = scipy.stats.mstats.gmean(wbars)
-    assert (0 < res).all(), (res, k, l)
-    assert (res < 1).all(), (res, k, l)
-    return res
-
-
-@numba.jit()
[…]

### 2017-10-05 src/stable.py

@@ -1,16 +1,0 @@
-from oblique import *
-from uuid import uuid4 as uuid
-import json
-import sys
-
-if __name__ == '__main__':    
-    w = float(sys.argv[1])
-    k = int(sys.argv[2])
-    l = k      
-    ρ = evol_stable(w=w, k=k, l=k, reps=1, PRINT=True)
-    fname = 'stable_{}.json'.format(uuid().hex)
-    with open(fname, 'wt') as f:
-    	json.dump(
-            dict(w=w, k=k, l=l, ρ=ρ),
-            f
-        )

### 2017-10-05 src/table.py

@@ -1,40 +1,0 @@
-from oblique import *
-from uuid import uuid4 as uuid
-import sys
-import json
-
-if __name__ == '__main__':
-    w = float(sys.argv[1])
-    k = int(sys.argv[2])
-    l = int(sys.argv[3])
-    
-    print("k={}, l={}, w={}".format(k, l, w))
-    # find optimal ρ, this time using minimization of geom_wbar_target
-    print("find optimal ρ")
-    max_ρ = max_ρ_with_polymorphism(w, k, l)
-    if round(max_ρ, 7) == 0:
-        opt_ρ = np.nan
-    else:
-        res = scipy.optimize.minimize_scalar(
-            geom_wbar_target, args=(w, k, l), bounds=[0, max_ρ], method='bounded')
-        if not res.success: 
-            print('Minimization failed for w={}, k={}, l={}'.format(w, k, l))
-            opt_ρ  = np.nan
-        else:
-            opt_ρ = res.x
-    print("optimal ρ: {:.2g}".format(opt_ρ))
-    # find an evolutionary stable ρ, if k!=l
-    if k != l:
-        print("find stable ρ")
-        stable_ρ = evol_stable(w=w, k=k, l=l, reps=1, PRINT=True)    
-        print("stable ρ: {:.2g}".format(stable_ρ))
-    else: # bring it from stable_xxx.json
-        stable_ρ = np.nan
-
-    fname = 'table_{}.json'.format(uuid().hex)
-    with open(fname, 'wt') as f:
-        json.dump(
-            dict(w=w, k=k, l=l, opt_ρ=opt_ρ, stable_ρ=stable_ρ),
-            f
-        )
-        

### 2017-10-17 ms/collect_figures.py
update pattern to match includegraphics options

@@ -12,7 +12,7 @@ if __name__ == '__main__':
 	ms_figures_dir = join_path(ms_dir, 'figures')
 	zip_fname = '/Users/yoavram/Dropbox/Milpitas/figures.zip'
 	tex_files = ['figures.tex', 'figures_supp.tex']
-	pattern = re.compile(r'\\includegraphics\{(.*\.\w{3})\}')
+	pattern = re.compile(r'\\includegraphics(?:\[.*\])?\{(.*\.\w{3})\}')
 
 	if os.path.exists(ms_figures_dir):
 		shutil.rmtree(ms_figures_dir)
@@ -30,6 +30,7 @@ if __name__ == '__main__':
 
 	with ZipFile(zip_fname, 'w') as z:
 		for fn in figures:
+			print(fn)
 			shutil.copy(fn, join_path(ms_figures_dir, split_path(fn)[-1]))
 			z.write(fn)
 

### 2017-10-17 ms/figures.tex
update by marc; merged stability analysis to SP6; w/W->w; revised Figure S4 into S6 with different axis

@@
 \includegraphics{../figures/A1B1_equilibrium.pdf}
-\caption{Properties of $A1B1$ environment.
-\textbf{(A)} Stable frequency of phenotype $A$ in the \emph{A1B1} selection regime and \textbf{(B)} stable mean fitness in selection regime \emph{A1B1} as functions of the vertical transmission rate \(\rho\) and the fitness ratio \(w/W\) between the favored and unfavored phenotypes.
-In all cases, \emph{W=1}.
-}\label{fig:AkBl_equilibrium}
+\caption{Properties of stability in \emph{A1B1} selection regime.
+\textbf{(A)} Stable frequency of phenotype $A$ and \textbf{(B)} stable mean fitness as functions of the vertical transmission rate \(\rho\) and the fitness of the disfavored phenotype \(w\).
+Black contour lines join \(\rho\) and \(w\) combinations that result in the same stable value.
+In all cases, fitness of the favored phenotype is \emph{W=1}.
+}\label{fig:A1B1_equilibrium}
 \end{figure}
@@
 \caption{Stochastic local stability. 
-The figure shows the frequency of phenotype \emph{A} after \(10^6\) generations in a very large population evolving in a stochastic environment (eq.\ 31).
-The fitness of phenotypes \emph{A} and \emph{B} are \(w_A=1+s_t\) and \(w_B=1\),
+The figure shows the frequency of phenotype \emph{A} after \(10^6\) generations in a very large population evolving in a stochastic environment (eq.\ 28).
+The fitnesses of phenotypes \emph{A} and \emph{B} are \(w_A=1+s_t\) and \(w_B=1\),
 where \(s_t\) is \emph{s} with probability \emph{p} and \emph{-s} with probability \(1-p\).
 The gray lines mark combinations of \emph{p} and \emph{s} for which \(E[\log{(1-\rho+\rho\frac{w_A}{w_B})}]=0\) and \(E[\log{(1-\rho+\rho\frac{w_B}{w_A})}]=0\).
-According to Result 6, between these lines fixation of either phenotype will be stochastically locally unstable, and we expect a stationary polymorphism between the lines.
-Parameters: initial frequency of \(A\) is \(x_0=1/10,000\) and the vertical transmission rate is \(\rho=0.1\).}\label{fig:stochastic_env_x_t}
+According to Result 6, between these lines fixation of either phenotype is not stochastically locally stable, and we expect a stationary polymorphism between the lines.
+Here, initial frequency of \(A\) is \(x_0=1/10,000\) and the vertical transmission rate is \(\rho=0.1\).}\label{fig:stochastic_env_x_t}
 \end{figure}
@@
 \centering
-\includegraphics{../figures/stoch_two_iid_rvs.png}
+\includegraphics{../figures/stoch_two_iid_rvs.pdf}
 \caption{Effect of vertical transmission rate $\rho$ on phenotype polymorphism in a randomly changing environment.
-Dynamics of the frequency of phenotype \(A\) over time starting at \(x_0=10^{-5}\) when the fitness of phenotypes \(A\) and \(B\) are identically and independently distributed random variables.
-As the vertical transmission rate \(\rho\) increases from 0.001 to 0.5 the frequency reaches a polymorphic distribution with \(E(x_t)\to0.5\) faster,
+Dynamics of the frequency of phenotype \(A\) over time starting at \(x_0=10^{-5}\) when the fitnesses of phenotypes \(A\) and \(B\) are identically and independently distributed random variables.
+As the vertical transmission rate \(\rho\) increases from 0.001 to 0.5, the frequency reaches a polymorphic distribution with \(E(x_t)\to0.5\) faster,
 but the variance also increases.
-The fitness of phenotypes \(A\) and \(B\), \(w_A\) and \(w_B\), are both exponential random variables with expected value 2.}\label{fig:stoch_two_iid_rvs}
+The fitnesses of phenotypes \(A\) and \(B\), \(w_A\) and \(w_B\), are both exponential random variables with expected value 2.}\label{fig:stoch_two_iid_rvs}
 \end{figure}
@@
 \caption{Consecutive fixation of modifiers that reduce the vertical
-transmission rate in environmental regime \emph{A1B1}. The figure shows
+transmission rate in selection regime \emph{A1B1}. The figure shows
 results of numerical simulations of evolution with two modifier alleles
-(eq.\ 35). When a modifier allele fixes
+(eq.\ 32). When a modifier allele fixes
 (frequency\textgreater{}99.9\%), a new modifier allele is introduced
@@
 the initial resident modifier allele, \(\rho_0 =0.1\); fitness values:
-$W=1$, $w=0.1$ (\textbf{A-C}), 0.5 (\textbf{D-F}), and 0.9
+$W=1$; $w=0.1$ (\textbf{A-C}), 0.5 (\textbf{D-F}), and 0.9
 (\textbf{G-I}). The x-axis is on a log-scale, as each sequential invasion
@@
 \centering
-\includegraphics[height=\dimexpr \textheight - 8\baselineskip\relax]{../figures/stoch_modifier_invasions.png}
+\includegraphics[height=\dimexpr \textheight - 8\baselineskip\relax]{../figures/stoch_modifier_invasions.pdf}
 \caption{Consecutive fixation of modifiers that reduce the vertical transmission rate $\rho$ under symmetric randomly changing selection.
 The figure shows results of numerical simulations of evolution with two modifier alleles
-(eq.\ 33).
+(eq.\ 32).
 When a modifier allele fixes (frequency\textgreater{}99.9\%),
@@
 \textbf{(B)} The frequency of the invading modifier allele over time.
-Parameters: vertical transmission rate of the initial resident modifier allele is \(\rho_0 =0.1\) and the fitness values ratio is $w_A/w_B=10$ with probability 0.5 and $w_A/w_B=0.1$ also with probability 0.5.
+Parameters: vertical transmission rate of the initial resident modifier allele is \(\rho_0 =0.1\) and the ratio of fitness values  is $w_A/w_B=10$ with probability 0.5 and $w_A/w_B=0.1$ also with probability 0.5.
 The x-axis is on a log-scale, as each sequential invasion takes an order of magnitude longer to complete.}\label{fig:stoch_modifier_invasions}
@@
 \includegraphics{../figures/A1B2_polymorphism.pdf}
-\caption{Stable frequency of phenotype \emph{A} and population mean fitness in selection regime
-$A1B2$ as a function of the vertical transmission rate \(\rho\) and the fitness ratio \(W/w\) between the favored and unfavored phenotypes.
-In the grey area \(B\) reaches fixation and the frequency of \(A\) is 0.
-\textbf{(A)} Stable frequency of phenotype $A$ at the end of each period of three generations.
-\textbf{(B)} Geometric average of the stable mean fitness over the three-generation period: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
+\caption{Stable frequency of phenotype \emph{A} and geometric mean fitness in selection regime
+$A1B2$ as a function of the vertical transmission rate \(\rho\) and the fitness of the disfavored phenotype \(w\).
+\textbf{(A)} Stable frequency of phenotype $A$ at the end of each three generations cycle.
+\textbf{(B)} Geometric average of the stable population mean fitness over the three generations cycle: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
+Gray contour lines join \(\rho\) and \(w\) combinations that result in the same stable value.
 In all cases, \emph{W=1}.}\label{fig:A1B2_polymorphism}
@@
 \centering
-\includegraphics{../figures/rho_evol_stability.pdf}
-\caption{Evolutionarily stable vertical transmission rate in $AkBk$ selection regime. 
-The evolutionary stable rate $\rho^*$ was determined as in Carja et al.\ (2014) using simulations of eq.\ 35: for each $k$ from 1 to 50, ten initial rates were randomly chosen between 0 and 1 for the resident modifier $m$; after the population evolved with only $m$ for 1,000 generations towards an equilibrium, invader modifier alleles $M$ were consecutively introduced at frequency $10^{-4}$; their value for $\rho$ was a product of the resident rate and a random number from an exponential distribution with a mean of 1; if an invader increased in frequency after $5,000 \times 2k$ generations, the invader was considered successful and became the resident modifier for the next invasion, and allowed to evolve for 1,000 generations to reach a new equilibrium before the next invasion; if the invader was unsuccessful, a new invader was drawn.
-After at least 500 invasion attempts and 50 unsuccessful ones, the evolutionary process stopped.
-The markers show the average for ten such processes for each $k$ value.
-95\% confidence intervals (calculated with bootstrap, 1,000 resamples) are too small to be seen.
-The dashed line shows the relation $\rho=1-1/(k-1)$ (Carja et al.
-2014), which is close to the simulation results for \(w=0.5\).
-In all cases, $W=1$.}\label{fig:rho_evol_stability}
+\includegraphics{../figures/AkBk_stable_optimal_rate.pdf}
+\caption{Fitness ``optimal'' and evolutionary stable vertical transmission rate in $AkBk$ selection regime. 
+\textbf{(A)} The vertical transmission rate $\hat{\rho}$ that maximized the geometric average of the population mean fitness is zero (complete oblique transmission) when selection cycles quickly between favoring phenotype $A$ and $B$, and then abruptly transitions to $\approx 0.2$, followed by a slow decrease (see Fig.
+S5 and S6 for details on the abrupt transition).
+\textbf{(B)} The evolutionary stable rate $\rho^*$, which cannot be invaded by modifiers with either higher or lower vertical transmission rate $P$, rapidly increases from zero when selection cycles are short ($k=1$ or $2$) to roughly 1 when selection cycles are longer.
+The stable rate was found by finding the root of $\frac{\partial \lambda_1}{\partial P}$, the derivative of the leading eigenvalue of the external stability matrix w.r.t to the rate of the invading modifier (see \emph{Supplemental Material SP6} for details).
+The dashed line shows $1-\frac{1}{k-1}$, which fit the values for $w=0.5$ (Carja et al.
+2011).
+The values for $w=0.1$ (blue) could not be calculated for $k > 19$ due to numerical instability when selection is strong and the duration between selection fluctuations is long.
+In all cases, $W=1$.}\label{fig:AkBk_stable_optimal_rate}
 \end{figure}
@@
 \centering
-\includegraphics{../figures/AkBk_stable_wbar_slice.pdf}
-\caption{Vertical transmission rates that maximize population mean fitness in the \emph{AkBk} selection regime.
-Populations evolved with different vertical transmission rates \(\rho\) and in different \emph{AkBk} selection regimes for \(1000 \times 2k\) generations, and the geometric average of the population mean fitness in the final \(2k\) generations was calculated.
-For each \(k\), the rate \(\hat{\rho}\) that maximizes the geometric average is shown.
-In all cases, $W=1$. See Fig. S3 for a broader view.} \label{AkBk_stable_wbar_slice}
+\includegraphics{../figures/fixation_prob_time.pdf}
+\caption{Fixation probability and time in a finite population.
+\textbf{(A)} Fixation probability \(u(x)\) of phenotype \(A\) (eq.\ 55), and \textbf{(B)}
+Expected time to fixation \(T(x)\) of phenotype \(A\) (eq.\ 56) conditioned on its
+fixation, starting with a single copy in a population of size \(N\). The
+figure compares two estimates: Wright-Fisher simulations (blue circles) and diffusion
+equation approximation (green solid line). Parameters: selection coefficient, \(s=w_A-w_B=0.1\),
+population size, \(N=10,000\).}\label{fixation_prob_time}
 \end{figure}
@@
 \centering
-\includegraphics{../figures/fixation_prob_time.pdf}
-\caption{Fixation probability and time in a finite population.
-\textbf{(A)} Fixation probability \(u(x)\) of phenotype \(A\) (eq.\ 70), and \textbf{(B)}
-Expected time to fixation \(T(x)\) of phenotype \(A\) (eq.\ 72), conditioned on its
-fixation, starting with a single copy in a population of size \(N\). The
-figure compares two estimates: simulations (blue circles) and diffusion
-equation approximation (green solid line). Parameters: Selection coefficient, \(s=w_A-w_B=0.1\);
-Population size, \(N=10,000\).}\label{fixation_prob_time}
-\end{figure}
-
-%Figure 11
-\begin{figure}
-\centering
 \includegraphics{../figures/lk_fix_prob.pdf}
@@
 Fixation probability of phenotype $A$ when starting with a single copy in a population of size $N$: $u(1/N) = (1-\exp(-2 \rho \frac{k-l}{k+l}(W-w))/(1-\exp(-2 N \rho \frac{k-l}{k+l}(W-w))$ (see eqs.
-115-116).
+59--60).
 \emph{k} and \emph{l} are the number of
 generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by
-selection. In all cases, $W = 1$, $w = 1-s$, Population size, \(N=10,000\).} \label{lk_fix_prob}
+selection. In all cases, fitness of the favored phenotype, $W = 1$; fitness of the unfavored phenotype is $w=1-s$ and the population size is \(N=10,000\).} \label{lk_fix_prob}
 \end{figure}

### 2017-10-17 ms/figures_supp.tex
update by marc; merged stability analysis to SP6; w/W->w; revised Figure S4 into S6 with different axis

--- 2017-08-24 ms/figures_supp.tex
@@
 \def\fps@figure{htbp}
+\@fpsep\textheight
 \makeatother
@@
 \centering
+\includegraphics{../figures/env_A1B1.pdf}
+\caption{Frequency of phenotype \emph{A} after every two generations in
+selection regime \(A1B1\). The orange line is the finite population model
+(eqs. 52-53; average of 100 simulations). The blue line is the infinite population model
+(eq.\ 21), and the green line is the  solution of \(Q(x)=0\) (eq.\ 22).
+In all cases, \(W=1\); for the finite population model (orange lines), population size is \(N=10,000\) and initial frequency of \(A\) is \(x_0=0.5\).}\label{fig:env_A1B1}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics{../figures/env_A1B2.pdf}
@@
 different initial frequency of phenotype \(A\) (0.01--0.99).
-See also Figure 1 and Figure S2. In all cases, $W=1$.}\label{fig:env_A1B2}
+See also Figures 1 and  7. In all cases, $W=1$.}\label{fig:env_A1B2}
 \end{figure}
@@
 \centering
-\includegraphics{../figures/AkBk_stable_wbar.pdf}
-\caption{Stable population mean fitness in selection regime \emph{AkBk} as a function of the vertical transmission rate \(\rho\) and the number \(k\) of generations in which phenotypes \emph{A} and \emph{B} are favored by selection, for different selection intensities.
-Colors represent the geometric average of the population mean fitness over \(2k\) generations after \(1,000 \times 2k\) generations of evolution.
-Blue markers show the maximum average mean fitness for each period \(k\).
-For example, with \(w=0.1\), \(\hat{\rho}=0\) maximizes the average fitness for \(k \le 11\), then \(\hat{\rho}\) increases to \(\hat{\rho} \approx 0.24\), and then continues to decrease as \(k\) increases, down to \(\hat{\rho} \approx 0.15\) for \(k=50\).
-Contour lines represent \(\rho\) and \(k\) combinations that produce the same average mean fitness. 
-In all cases, $W=1$.}\label{fig:AkBk_stable_wbar}
-\end{figure}
-
-\begin{figure}
-\centering
 \includegraphics{../figures/env_A3B10.pdf}
@@
 
+\begin{figure}
+\centering
+\includegraphics{../figures/AkBk_stable_wbar.pdf}
+\caption{Stable population mean fitness in selection regime \emph{AkBk} as a function of the vertical transmission rate \(\rho\) and the number \(k\) of generations in which phenotypes \emph{A} and \emph{B} are favored by selection, for different selection intensities: \textbf{(A)} \(w=0.1\), \textbf{(B)} \(w=0.5\), and \textbf{(C)} \(w=0.9\).
+Colors represent the geometric average of the stable population mean fitness over \(2k\) generations, calculated by iterating eq.
+10 until phenotype frequencies stabilized, and for at least \(1,000 \) generations.
+Blue markers show the maximum average mean fitness for each period \(k\).
+For example, with \(w=0.1\), \(\hat{\rho}=0\) maximizes the average fitness for \(k \le 11\), then \(\hat{\rho}\) increases to \(\hat{\rho} \approx 0.24\), and then continues to decrease as \(k\) increases, down to \(\hat{\rho} \approx 0.15\) for \(k=50\) (see also Fig.
+8A).
+Contour lines represent \(\rho\) and \(k\) combinations that produce the same average mean fitness. 
+In all cases, $W=1$.}\label{fig:AkBk_stable_wbar}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/AkBk_w0.5_geomwbar.pdf}
+\caption{The geometric average of the stable population mean fitness over the $2k$ generation period peaks at $\rho=0$ for $k \le 30$ (red, blue and green lines) and at $\rho \approx 0.23$ for $k=31$ and $32$ (purple and orange lines).
+See also Figs. 8A, S5B. The inset zooms out to show that the geometric mean fitness is strictly and significantly decreasing for $\rho>0.3$ (reaching $\approx 0.7$ for $\rho=1$).
+In all cases, $W=1$, $w=0.5$.}\label{fig:AkBk_w0.5_geomwbar}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/AkBk_stable_modifier_w_0.5.pdf}
+ \caption{The leading eigenvalue of the external stability matrix in $AkBk$ with selection.
+The colors show $\lambda_1(\rho, P)$, the leading eigenvalue of the external stability matrix for combinations of $\rho$ (y-axis) and $P$ (x-axis), the vertical transmission rates of resident and invader modifiers, respectively (see \emph{Supplemental Material SP6} for details).
+Purple shows values lower than 1, leading to stability of $\rho$; white equals to 1; green larger than 1, leading to invasion of $P$.
+  The arrows show an "evolutionary" path $\rho_{0} \to \rho_{1} \to \ldots \to \rho^*$: at each step, a modifier with rate $P_t$ invades the resident modifier with rate $\rho_t$ and becomes the resident modifier $\rho_{t+1}$, until consecutive invaders have similar rates.
+  Here, $W=1$ and $w=0.5$.}\label{fig:AkBk_stable_modifier_w_0.5}
+\end{figure}
+
 \end{document}

### 2017-10-17 ms/ms.tex
update by marc; merged stability analysis to SP6; w/W->w; revised Figure S4 into S6 with different axis

@@
 \def\cl{\hbox{\tenbf L}}
+\def\cj{\hbox{\tenbf J}}
 \def\ci{\hbox{\tenbf I}}
@@
 \vfil
-\centerline{September 15, 2017, v.1}
+\centerline{October 17, 2017}
 
@@
 
-\vfil\break
+\bigskip\bigskip
 
@@
  Therefore the frequency $x'$ of  phenotype $A$ after one generation is given by the recursion equation
- $$x'= \rho{w_Ax\over \overline w} +(1-\rho)x=     {w_A\over \overline w}x\bigl[(1-\rho)x +\rho\bigr] +{w_B\over \overline w}(1-x)\bigl[(1-\rho)x\bigr],\eqno(2)$$
+ $$x'= \rho{w_A\over \overline w}x +(1-\rho)x=     {w_A\over \overline w}x\bigl[(1-\rho)x +\rho\bigr] +{w_B\over \overline w}(1-x)\bigl[(1-\rho)x\bigr],\eqno(2)$$
  where $\overline w$ is the {\sl mean fitness}, namely
@@
 \noindent{\it Proof}.
-If we rewrite (4) as $x'=x\cdot F(x)$, it can be seen that $F(1)=1$, and for $\rho>0$ and $0<x<1$
-$$\eqalign{
-&F(x)>1\quad\hbox{\tenrm when}\quad w_A>w_B,\cr
-&F(x)<1\quad\hbox{\tenrm when}\quad w_A<w_B.}\eqno(5)$$
+If we rewrite (4) as $x'=x\cdot f(x)$, it can be seen that $f(1)=1$, and for $\rho>0$ and $0<x<1$
+$$\eqalign{
+&f(x)>1\quad\hbox{\tenrm when}\quad w_A>w_B,\cr
+&f(x)<1\quad\hbox{\tenrm when}\quad w_A<w_B.}\eqno(5)$$
 Hence, as $w_A>0$ and $w_B>0$, both fixations in $A$ or in $B$  ($x^*=1$ for fixation in $A$ and $x^*=0$ for fixation in $B$) are equilibrium points of (4).
@@
 $$\eqalign{
-&x^{t+1}>x^t\quad\hbox{\tenrm when}\quad w_A>w_B,\cr
-&x^{t+1}<x^t\quad\hbox{\tenrm when}\quad w_A<w_B.}\eqno(6)$$
-
-\noindent Thus, as $x^*=1$ or $x^*=0$ are the only equilibrium points, then
+&x_{t+1}>x_t\quad\hbox{\tenrm when}\quad w_A>w_B,\cr
+&x_{t+1}<x_t\quad\hbox{\tenrm when}\quad w_A<w_B,}\eqno(6)$$
+and since $x^*=1$ or $x^*=0$ are the only equilibrium points, we have
 $$\eqalign{
@@
 &\lim_{t\to\infty}x_t=0,\ \ \hbox{\tenrm for all}\ 0\le x_0< 1,\ \ \hbox{\tenrm when}\ w_A<w_B.}\eqno(7)$$
-Therefore fixation of the favored phenotype is globally stable, and ultimately all individuals have the favored phenotype.
-
-If $\rho=0$ or $w_A=w_B>0$, the recursion (4) reduces to $x'=x$; the phenotype frequencies do not change over time.
+Therefore fixation of the favored phenotype is globally stable.
+
+If $\rho=0$ or $w_A=w_B>0$, the phenotype frequencies do not change over time.
 With {\sl extreme selection},  for example $w_A=1$ and $w_B=0$ (or $w_B=1$ and $w_A=0$), recursion (4) reduces to
@@
 
-Next we consider the case where the environment is not constant but rather changes periodically, such that the favored phenotype changes after a fixed number of generations.
-Simple examples are $A1B1=ABABAB\dots$, in which the favored genotype switches every generation, or $A2B1=AABAABAAB\dots$, where every two generations in which selection favors $A$ are followed by a single generation in which selection favors $B$.
-
+Suppose the environment is not constant but rather changes periodically, such that the favored phenotype changes after a fixed number of generations.
+Simple examples are $A1B1=ABABAB\dots$, in which the favored phenotype switches every generation, or $A2B1=AABAABAAB\dots$, where every two generations in which selection favors $A$ are followed by a single generation in which selection favors $B$.
   In general, $AkBl$ denotes a selection regime in which the period is of $(k+l)$ generations with $k$ generations favoring phenotype $A$ followed by $l$ generations favoring $B$.
@@
 
-  \noindent If $x_t$ denotes the frequency of exhibiting the phenotype $A$ at generation $t$ starting with $x_0$ initially, then as we are interested in the values of $x_t$ for $t=n(k+l)$ with $n=0,1,\dots$ at the end of complete periods then in fact
+  \noindent If $x_t$ denotes the frequency of  the phenotype $A$ at generation $t$ starting with $x_0$ initially, then as we are interested in the values of $x_t$ for $t=n(k+l)$ with $n=0,1,\dots$ at the end of complete periods, we can write
   $$x_{(n+1)(k+l)}=F(x_{n(k+l)}),\quad n=0,1,2,\dots,\eqno(9)$$
@@
    $$F(x) \simeq\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^lx.\eqno(11)$$
-   Hence the local stability of $x^*=0$ is determined by the product $\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l$, and $x^*=0$  is locally stable if this product  is less than 1 and unstable if it is larger than 1.
+   Hence the local stability of $x^*=0$ is determined by the product $\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l$;  $x^*=0$  is locally stable if this product  is less than 1 and unstable if it is larger than 1.
 
@@
    Observe that
-   $$\eqalign{
-   \left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right) &= \left(1-\rho+\rho{W\over w}\right)\left(1-\rho+\rho{w\over W}\right)\cr
-   &= \left(1-\rho\right)^2+ \rho^2 +\rho(1-\rho){W^2+w^2\over Ww}.}\eqno(14)$$
-   Thus,
-   $$\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right) =1+\rho(1-\rho){\left(W-w\right)^2\over wW}>1.\eqno(15)$$
-   Since $0<\rho<1$ and $0<w<W$,  the $B$ fixation is unstable.
+   $$\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right) = 1+\rho(1-\rho){\left(W-w\right)^2\over wW}>1.\eqno(14)$$
+   Since $0<\rho<1$ and $0<w<W$,  fixation on $B$ is unstable.
   \bigskip
@@
 
-\item{1.}Since $k=l$ and the above result holds also when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is not stable.
-Thus neither phenotype can be lost, and we have a {\sl protected polymorphism} (Prout 1968).
+\item{1.}Since $k=l$ and the above result holds also when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is also not stable.
+Thus neither phenotype can be lost, and there is a {\sl protected polymorphism} (Prout 1968).
 
 \item{2.}Observe that for general $k$, $l$, the condition for local stability of fixation in $A$ is
-$$\bigl[f_A(0)\bigr]^l \bigl[f_B(0)\bigr]^k <1, \eqno(16)$$
+$$\bigl[f_A(0)\bigr]^l \bigl[f_B(0)\bigr]^k <1, \eqno(15)$$
 and that of $B$ is
-$$\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l <1. \eqno(17)$$
+$$\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l <1. \eqno(16)$$
 Therefore, following Result 2, 
-$$\bigl[f_A(0)\bigr]^{k+l}\bigl[f_B(0)\bigr]^{k+l}>1,\eqno(18)$$
+$$\bigl[f_A(0)\bigr]^{k+l}\bigl[f_B(0)\bigr]^{k+l}>1,\eqno(17)$$
 and it is impossible that both fixations are stable.
@@
 \noindent{\it Proof}.
-Let $a=1+\rho{W-w\over w}$ and $b=1+\rho{w-W\over W}$.
-Then since $0<\rho <1$ and $0<w<W$, we have $a>1$ and $0<b<1$.
+Let $a=1+\rho{W-w\over w}$ and $b=1+\rho{w-W\over W}$, and our assumption entails  $a>1$ and $0<b<1$.
 Following (11), fixation in $B$ is not stable if $a^kb^l>1$, and similarly  fixation in $A$ is unstable if $a^lb^k>1$.
 Therefore both fixations are unstable if
-$$a^kb^l>1\quad\hbox{\tenrm and}\quad a^lb^k>1,\eqno(19)$$
+$$a^kb^l>1\quad\hbox{\tenrm and}\quad a^lb^k>1,\eqno(18)$$
 or equivalently if
-$$k\log a +l\log b>0\quad\hbox{\tenrm and}\quad l\log a+k\log b>0.\eqno(20)$$
-Now (20) holds if and only if
-$$k\,{\log(1/b)\over\log a} < l < k\,{\log a\over\log(1/b)}.\eqno(21)$$ 
-Inequalities (21) are consistent if and only if $\log(1/b)<\log a$, i.e., $ab>1$. But by (15),
-$$ab=\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right)>1.\eqno(22)$$
-
-\noindent{\it Remark}.
-It should be pointed out that the local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations $A$ is favored $k$ times and $B$ is favored $l$ times, and not their order in the cycle.
-
-When neither fixation in $A$ or $B$ is stable, we have a case of protected polymorphism, and we expect to have one or more polymorphic equilibria.
-Figure 1 shows the relationship between $k,l$ and $\rho$ that supports polymorphism of $A$ and $B$, or fixation, for different values of $W$ and $w$.
+$$k\log a +l\log b>0\quad\hbox{\tenrm and}\quad l\log a+k\log b>0.\eqno(19)$$
+Now (19) holds if and only if
+$$k\,{\log(1/b)\over\log a} < l < k\,{\log a\over\log(1/b)}.\eqno(20)$$
+
+Inequalities (20) are consistent if and only if $\log(1/b)<\log a$, i.e., $ab>1$, which according to (14), is true.
+
+The local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations $A$ is favored $k$ times and $B$ is favored $l$ times, and not their order in the cycle.
+
+When neither fixation in $A$ or $B$ is stable, there is a protected polymorphism, and we expect to have one or more polymorphic equilibria.
+Figure 1 shows the relationship between $k,l$ and $\rho$ that gives polymorphism of $A$ and $B$, or fixation, for different values of $s=W-w$.
 
@@
 &F_A(x)=x{x(1-\rho)(W-w) +\rho W+(1-\rho)w\over x(W-w) +w},\cr\noalign{\smallskip}
-&F_B(y) = y{y(1-\rho)(w-W) +\rho w+(1-\rho)W\over y(w-W)+W}.}\eqno(23)$$
+&F_B(y) = y{y(1-\rho)(w-W) +\rho w+(1-\rho)W\over y(w-W)+W}.}\eqno(21)$$
 The equilibrium equation is $x=F_B\bigl(F_A(x)\bigr)$, which reduces to a fourth degree polynomial equation in $x$.
 Since the two fixations in $B$ and $A$ are equilibria corresponding to the two solutions $x=0$ and $x=1$, the other equilibria correspond to solutions of a quadratic equation $Q(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with
-$$\alpha_2=1,\quad \alpha_1={W+w\over (2-\rho)(W-w)}-1,\quad \alpha_0={-w\over (2-\rho)(W-w)}.\eqno(24)$$
+$$\alpha_2=1,\quad \alpha_1={W+w\over (2-\rho)(W-w)}-1,\quad \alpha_0={-w\over (2-\rho)(W-w)}.\eqno(22)$$
 As $0<\rho<1$ and $0<w<W$ we have
-$$Q(0) = {-w\over (2-\rho)(W-w)}<0\eqno(25)$$
-and
-$$Q(1) ={W\over (2-\rho)(W-w)} >0.\eqno(26)$$
+$$Q(0) = {-w\over (2-\rho)(W-w)}<0\quad\hbox{\tenrm and}\quad Q(1) ={W\over (2-\rho)(W-w)} >0.\eqno(23)$$
 Also, as $\alpha_2=1$ and $\alpha_0<0$, the quadratic equation $Q(x)=0$ has two real roots, one negative and one positive $x^*$ satisfying $0<x^*<1$.
@@
 Then
- $$H(0)=0,\quad H(x^*)=x^*,\quad H(1)=1.\eqno(27)$$
+ $$H(0)=0,\quad H(x^*)=x^*,\quad H(1)=1.\eqno(24)$$
  Also 
- $$F'_A(x) = {x^2(1-\rho)\left(W-w\right)^2 +2xw(1-\rho)(W-w) +w\left[\rho W +(1-\rho)w\right]\over \left[x(W-w) +w\right]^2}\eqno(28)$$
+ $$F'_A(x) = {x^2(1-\rho)\left(W-w\right)^2 +2xw(1-\rho)(W-w) +w\left[\rho W +(1-\rho)w\right]\over \left[x(W-w) +w\right]^2}\eqno(25)$$
  and
-  $$F'_B(x) = {x^2(1-\rho)\left(w-W\right)^2 +2xW(1-\rho)(w-W) +W\left[\rho w +(1-\rho)W\right]\over \left[x(w-W) +W\right]^2}.\eqno(29)$$
-  As $0<\rho<1$ and $0<w<W$, we have $F'_A(x)>0$ for  $0\le x\le 1$. Observe that the numerator of $F'_B(x)$ is linear in $\rho$; its value when $\rho=1$ is $wW>0$, and when $\rho=0$ it is
-  $$x^2\left(w-W\right)^2 +2xW(w-W) +W^2 =\left[x(w-W) +W\right]^2 >0.\eqno(30)$$
-  Hence $F'_{B}(x)>0$ for all $0\le x\le 1$, and $H'(x) =F'_B\bigl(F_A(x)\bigr)F'_A(x)$ is positive when $0\le x\le 1$. Thus $H(x)$ is a monotone increasing function of $x$ for $0\le x\le 1$.
-Thus, starting from any initial value $0<x_0 <1$ we have  $x_t\,\harr{}{t\to\infty}\,x^*$.
-Supplementary Figure S1 illustrates examples of how the frequency of $A$ changes over time in the $A1B1$ regime of cycling selection, while Figure 2A shows the equilibrium frequencies $x^*$ as a function of $\rho$, $W$, and $w$ in this regime.
+  $$F'_B(x) = {x^2(1-\rho)\left(w-W\right)^2 +2xW(1-\rho)(w-W) +W\left[\rho w +(1-\rho)W\right]\over \left[x(w-W) +W\right]^2}.\eqno(26)$$
+ From our assumptions on $\rho$, $w$, and $W$, we have $F'_A(x)>0$ for  $0\le x\le 1$.
+Observe that the numerator of $F'_B(x)$ is linear in $\rho$; its value when $\rho=1$ is $wW>0$, and when $\rho=0$ it is
+  $$x^2\left(w-W\right)^2 +2xW(w-W) +W^2 =\left[x(w-W) +W\right]^2 >0.\eqno(27)$$
+  Hence $F'_{B}(x)>0$ for all $0\le x\le 1$, and $H'(x) =F'_B\bigl(F_A(x)\bigr)F'_A(x)$ is positive when $0\le x\le 1$.
+Thus $H(x)$ is monotone increasing for $0\le x\le 1$; $H(x)>x$ for $0<x<x^*$, $H(x)<x$ for $x^*<x<1$.
+Starting from any initial value $0<x_0 <1$ we have  $x_t\,\harr{}{t\to\infty}\,x^*$.
+Supplementary Figure S1, panels A, C, and E, illustrate examples of how the frequency of $A$ changes over time in the $A1B1$ regime of cycling selection.
 
@@
 
-The proof of Result 5 is in Supplementary Material SP1. 
+The proof of Result 5 is in Supplementary Material SP1.  Figure 2A shows the stable equilibrium frequencies $x^*$ as a function of $\rho$, $W$, and $w$ in the $A1B1$ regime.
 For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplemental Figure S3 illustrates the convergence to a single stable polymorphism.
@@
 However, the numerical examples in Figure 1 (for $AkBl$) and in Figure 7 (for the special case $A1B2$) all exhibit a single stable polymorphic equilibrium when fixations in $A$ and $B$ are unstable.
-These numerical results suggest that for $W>w>0$ and $0<\rho<1$ the high order polynomial that gives the equilibria has only a single root that is a globally stable polymorphism.
+These numerical results suggest that for $W>w>0$ and $0<\rho<1$ the high order polynomial that gives the equilibria has only a single root corresponding to a globally stable polymorphism.
+Supplementary Figure S4 shows that this is the case for the $A3B10$ regime.
 \medskip
@@
  Corresponding to (4), with $w_A=1+s_t$, $w_B=1$ the recursion equation is
- $$x_{t+1}=x_t {1+\rho s_t +x_t(1-\rho)s_t\over 1+x_ts_t}\quad t=0,1,2,\dots\; .\eqno(31)$$
+ $$x_{t+1}=x_t {1+\rho s_t +x_t(1-\rho)s_t\over 1+x_ts_t}\quad t=0,1,2,\dots\; .\eqno(28)$$
 
@@
  A constant equilibrium state $x^*$ is said to be {\sl stochastically locally stable} if for any $\var>0$ there exists a $\delta>0$ such that $|x_0-x^*|<\delta$ implies
- $$P\left(\lim_{t\to\infty}x_t =x^*\right)\ge 1-\var.\eqno(32)$$
+ $$P\left(\lim_{t\to\infty}x_t =x^*\right)\ge 1-\var.\eqno(29)$$
  Thus stochastic local stability holds for $x^*$ provided for any initial $x_0$ sufficiently near $x^*$ the process $x_t$ converges to $x^*$ with high probability.
@@
  Using the general notation for the fitness parameters $w_A$ and $w_B$, the stochastic local stability of the $B$ fixation is determined by the sign of $E\left[\log\left(1-\rho+\rho{w_A\over w_B}\right)\right]$ and that of the $A$ fixation by the sign of $E\left[\log\left(1-\rho+\rho{w_B\over w_A}\right)\right]$.
-For example, if the first sign  is negative, fixation in $B$ is stochastically locally stable, and when it is positive, with probability one convergence to fixation in phenotype $B$ does not occur.
+For example, if the  sign of the first  is negative, fixation in $B$ is stochastically locally stable, and when it is positive, with probability one convergence to fixation in phenotype $B$ does not occur.
 It is also true  that if $E(w_A/w_B)\le 1$, then fixation of $B$ is stochastically locally stable.
-Following (14),
- $$\left(1-\rho +\rho{w_A\over w_B}\right)\left(1-\rho+\rho{w_B\over w_A}\right) >1.\eqno(33)$$
+Following (14), for all realizations of $w_A$ and $w_B$,
+ $$\log\left(1-\rho +\rho{w_A\over w_B}\right) +\log\left(1-\rho +\rho{w_B\over w_A}\right)>0.\eqno(30)$$
  Therefore, as in the case of periodically changing environments $AkBl$, it is impossible that both fixations are simultaneously stochastically locally stable.
 It is possible, however, that neither fixation is  stochastically locally stable, in which case we expect the population to converge to a polymorphic distribution.
-Figure 3 illustrates how the properties of $s_t$ in recursion (31) affect the frequency of phenotype $A$, and in particular the stochastic local stability of fixation in phenotype $B$.
-Figure 4 shows the dynamics of the frequency of $A$ in a case where $w_A$ and $w_B$ are identically distributed and independent; in this case the expectation of the stationary distribution is 1/2, and its variance increases as $\rho$ increases.
+Figure 3 illustrates how the properties of $s_t$ in recursion (28) affect the frequency of phenotype $A$, and in particular the stochastic local stability of fixation in phenotype $B$.
+Figure 4 shows the dynamics of the frequency of $A$ in a case where $w_A$ and $w_B$ are identically distributed and independent; in this case the expectation of the stationary distribution is ${1\over 2}$, and its variance increases as $\rho$ increases.
 \medskip
@@
 For example, is there an evolutionarily stable rate of oblique transmission?
-To answer this question we use a modifier model.
-
- Suppose that the vertical transmission rate is controlled by a genetic locus with two possible alleles $m$ and $M$.
+To answer this question we use a modifier model, for which we suppose that the vertical transmission rate is controlled by a genetic locus with two possible alleles $m$ and $M$.
 Let the vertical transmission rates determined by $m$ and $M$ be $\rho$ and $P$, respectively.
@@
  \hbox{\tenrm fitness}&w_A&w_B&w_A&w_B\cr\noalign{\smallskip}
- \hbox{\tenrm vertical transmission rate}&\rho&\rho&P&P\cr}\eqno(34)$$
+ \hbox{\tenrm vertical transmission rate}&\rho&\rho&P&P\cr}\eqno(31)$$
  Following the rationale leading to equation (2), the next generation pheno-genotype frequencies $x'_1$, $x'_2$, $x'_3$, $x'_4$ are
@@
 \overline wx'_4 &= w_Ax_3(1-P)(x_2+x_4) +w_Bx_4\bigl[(1-P)(x_2+x_4)+P\bigr], 
-}\eqno(35)$$
+}\eqno(32)$$
 with $\overline w$, the mean fitness, given by
-$$\overline w =w_A(x_1+x_3) +w_B(x_2+x_4).\eqno(36)$$
-Note that under these assumptions the $M/m$ locus and the $A/B$ phenotypic dichotomy do not undergo anything analogous to recombination, which might be introduced if the $A/B$ were viewed as haploid genetic variants.
+$$\overline w =w_A(x_1+x_3) +w_B(x_2+x_4).\eqno(33)$$
+Note that under these assumptions the $M/m$ locus and the $A/B$ phenotypic dichotomy do not undergo anything analogous to recombination, which might be introduced if $A/B$ phenotypes were viewed as haploid genetic variants.
 
 Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} (e.g., Feldman and Liberman 1986; Altenberg et al.\ 2017) to invasion by allele $M$.
-Because we assume the initial existence of a stable polymorphic equilibrium that depends on $\rho$ (the rate determined by $m$), we cannot assume a {\sl constant environment}, which always leads to fixation of the favored type, independent of $\rho$.
-We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$.
-Specifically, following (35) with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
-$$\underline x''=T_2(T_1\underline x),\eqno(37)$$
-where $\underline x'=T_1\underline x$ is given by (35) with $w_A=W$, $w_B=w$, and $\underline x''=T_2\underline x'$ is given by (35) with $w_A=w$, $w_B=W$.
+Constant environment always leads to fixation of the favored type, independent of $\rho$.
+We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$ (but see Supplemental Material SP6 for computational analysis of the general $AkBl$ case).
+Specifically, from (32) with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
+$$\underline x''=T_2(T_1\underline x),\eqno(34)$$
+where $\underline x'=T_1\underline x$ is given by (32) with $w_A=W$, $w_B=w$, and $\underline x''=T_2\underline x'$ is given by (32) with $w_A=w$, $w_B=W$.
 Here $\underline x$, $\underline x'$, $\underline x''$ are the frequency vectors.
 
-For the $A1B1$ case, when only the $m$ allele is present with associated rate $\rho$, with $0<\rho<1$ and $0<w<W$, a unique stable equilibrium $\underline x^* =(x_1^*,1-x_1^*,0,0)$ exists.
-$x_1^*$ is the only positive root of the quadratic equation $G(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with $\alpha_2,\alpha_1,\alpha_0$ specified in (24).
-Solving $G(x)=0$ gives
-$$x_1^* ={1\over 2} -{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}\over 2\cdot(2-\rho)(W-w)},\eqno(38)$$
+For the $A1B1$ case, when only the $m$ allele is present with associated rate $\rho$,  $0<\rho<1$, and $0<w<W$, a unique stable equilibrium $\underline x^* =(x_1^*,1-x_1^*,0,0)$ exists.
+$x_1^*$ is the only positive root of the quadratic equation $Q(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with $\alpha_2,\alpha_1,\alpha_0$ specified in equation (21).
+Solving $Q(x)=0$ gives
+$$x_1^* ={1\over 2} -{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}\over 2\cdot(2-\rho)(W-w)},\eqno(35)$$
  and it can be seen that
- $${\sqrt{Ww}-w\over W-w}<x_1^* <{1\over 2}.\eqno(39)$$
-
- The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ near $x^*$, which is derived from (35) and given by
+ $${\sqrt{Ww}-w\over W-w}<x_1^* <{1\over 2}.\eqno(36)$$
+
+ The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ near $x^*$, which is derived from (32) and given by
  $$\overline w^*\cl_1 =\left[\matrix{W\bigl[(1-P)x_1^* +P\bigr] & w(1-P)x_1^*\cr\noalign{\medskip}
- W(1-P)x_2^* & w\bigl[(1-P)x_2^* +P\bigr]\cr}\right]\eqno(40)$$
+ W(1-P)x_2^* & w\bigl[(1-P)x_2^* +P\bigr]\cr}\right]\eqno(37)$$
  and
  $$\overline w^{**}\cl_2 =\left[\matrix{w\bigl[(1-P)x_1^{**} +P\bigr] & W(1-P)x_1^{**}\cr\noalign{\medskip}
- w(1-P)x_2^{**} & W\bigl[(1-P)x_2^{**} +P\bigr]\cr}\right],\eqno(41)$$
+ w(1-P)x_2^{**} & W\bigl[(1-P)x_2^{**} +P\bigr]\cr}\right],\eqno(38)$$
 where $\underline x^{**} =T_1\underline x^*$, $x_1^{**}=x_2^*=1-x_1^*$, $x_2^{**}=x_1^*$,  and
-$$\overline w^* =Wx_1^* +wx_2^*,\qquad \overline w^{**} =wx_1^{**} +Wx_2^{**}.\eqno(42)$$
+$$\overline w^* =Wx_1^* +wx_2^*,\qquad \overline w^{**} =wx_1^{**} +Wx_2^{**}.\eqno(39)$$
 Due to the symmetry between the two phenotypes $A$ and $B$ in the $A1B1$ case, we have $x_1^{**} =x_2^*$ and $x_2^{**} =x_1^*$ so that $\overline w^{**} =\overline w^*$ and in fact
 $$\overline w^*\cl_2 = \left[\matrix{w\bigl[(1-P)x_2^* +P\bigr] & W(1-P)x_2^*\cr\noalign{\medskip}
- w(1-P)x_1^* & W\bigl[(1-P)x_1^* +P\bigr]\cr}\right].\eqno(43)$$
-  \smallskip
-
-   \noindent{\it Remark}. 
- Observe that as $\underline x^* =T_2(T_1\underline x^*)$ with $x_3^*=x_4^*=0$, from (40) and (41) with $P=\rho$ we have
- $$\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right]= \cl_2\cdot\cl_1\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right] =\cl\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right].\eqno(44)$$
+ w(1-P)x_1^* & W\bigl[(1-P)x_1^* +P\bigr]\cr}\right].\eqno(40)$$
+ Note that as $\underline x^* =T_2(T_1\underline x^*)$ with $x_3^*=x_4^*=0$, from (37) and (38) with $P=\rho$ we have
+ $$\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right]= \cl_2\cdot\cl_1\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right] =\cl\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right].\eqno(41)$$
  Hence when $P=\rho$ one of the eigenvalues of $\cl$ is 1.
 
- In general $\cl=\cl_2\cdot\cl_1$, and using (40) and (43) we get
+ In general $\cl=\cl_2\cdot\cl_1$, and using (37) and (40) we have
  $$\eqalign{
@@
  w(1-P)\bigl[1-x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr] & Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[W(1-P)x_2^*\bigr]^2
- }\right].}\eqno(45)$$
+ }\right].}\eqno(42)$$
  The external stability of $\underline x^*$ is determined by the eigenvalues of $\cl$, namely the roots of its characteristic polynomial $R(\lambda) =\det(\cl-\lambda\ci)$, with $\ci$ the $2\times 2$ identity matrix.
-From (45), $R(\lambda)=a_2\lambda^2 +a_1\lambda +a_0$, where
- $$a_0={P^2W^2w^2\over\left(\overline w^*\right)^4},\qquad a_1=-{2PWw +\left(1-P\right)^2\left[Wx_2^* +wx_1^*\right]^2\over \left(\overline w^*\right)^2},\qquad a_2=1.\eqno(46)$$
+From (42), $R(\lambda)=a_2\lambda^2 +a_1\lambda +a_0$, where
+ $$a_0={P^2W^2w^2\over\left(\overline w^*\right)^4},\qquad a_1=-{2PWw +\left(1-P\right)^2\left[Wx_2^* +wx_1^*\right]^2\over \left(\overline w^*\right)^2},\qquad a_2=1.\eqno(43)$$
 
@@
   \noindent{\it Proof}.
-As $\cl$ is a positive matrix, by the Perron-Frobenius theory $\cl$ has a positive eigenvalue, and as $a_0>0$ and $a_2=1$ the product of the two eigenvalues of $\cl$ is positive.
-Thus in fact $\cl$ has two positive eigenvalues.
-Let $R(1)=R(1;P)$, then from (46)
-  $$R(1;P) = {W^2w^2-\left(\overline w^*\widetilde w^*\right)^2\over\left(\overline w^*\right)^4}P^2 +2P{\left(\widetilde w^*\right)^2 -Ww\over \left(\overline w^*\right)^2} +{\left(\overline w^*\right)^2-\left(\widetilde w^*\right)^2\over\left(\overline w^*\right)^2},\eqno(47)$$
+As $\cl$ is a positive matrix, by the Perron-Frobenius theorem $\cl$ has a positive eigenvalue, and as $a_0>0$ and $a_2=1$ the product of the two eigenvalues of $\cl$ is positive.
+Thus $\cl$ has two positive eigenvalues.
+Let $R(1)=R(1;P)$, then from (43)
+  $$R(1;P) = {W^2w^2-\left(\overline w^*\widetilde w^*\right)^2\over\left(\overline w^*\right)^4}P^2 +2P{\left(\widetilde w^*\right)^2 -Ww\over \left(\overline w^*\right)^2} +{\left(\overline w^*\right)^2-\left(\widetilde w^*\right)^2\over\left(\overline w^*\right)^2},\eqno(44)$$
 where $\widetilde w^* =Wx_2^* +wx_1^*$.
 
-By (39) $\bigl(\sqrt{Ww}-w\bigr)/(W-w) <x_1^* <{1\over 2}$, from which it is easily seen that
-$$\sqrt{Ww}< \overline w^*<\widetilde w^* .\eqno(48)$$
+By (36) $\bigl(\sqrt{Ww}-w\bigr)/(W-w) <x_1^* <{1\over 2}$, from which it is easily seen that
+$$\sqrt{Ww}< \overline w^*<\widetilde w^* .\eqno(45)$$
 
  When $P=\rho$ one of the eigenvalues of $\cl$ is 1; hence $R(1;\rho)=0$.
-Another root of $R(1;P)=0$ is $\bigl[(\overline w^*)^2 +\overline w^*\widetilde w^*\bigr]/\bigl[Ww +\overline w^*\widetilde w^*\bigr]$, which by (48) is larger than 1.
-As $R(1;0)=\bigl[(\overline w^*)^2 -(\widetilde w^*)^2\bigr]/(\overline w^*)^2 <0$ by (48), we deduce that when $0<P<\rho$, $R(1;P)<0$, whereas when $\rho<P<1$, $R(1;P)>0$.
+Another root of $R(1;P)=0$ is $\bigl[(\overline w^*)^2 +\overline w^*\widetilde w^*\bigr]/\bigl[Ww +\overline w^*\widetilde w^*\bigr]$, which by (45) is larger than 1.
+As $R(1;0)=\bigl[(\overline w^*)^2 -(\widetilde w^*)^2\bigr]/(\overline w^*)^2 <0$ by (45), we deduce that when $0<P<\rho$, $R(1;P)<0$, whereas when $\rho<P<1$, $R(1;P)>0$.
  Hence, when $P<\rho$, $R(1)<0$, and since $a_2=1$, $R(+\infty)>0$, we conclude that $R(\lambda)=0$ has a positive root larger than 1 and the largest positive eigenvalue of $\cl$ is larger than 1.
@@
 As $R(\lambda)=0$ has two positive roots and as $a_2>0$, $R(\lambda)$ is convex, either the two positive roots are less than 1 or both larger than one.
-But the product of the two roots is $P^2W^2w^2/(\overline w^*)^2<1$ by (48); thus when $P>\rho$ the two positive eigenvalues of $\cl$ are less than 1.
+But the product of the two roots is $P^2W^2w^2/(\overline w^*)^2<1$ by (45); thus when $P>\rho$ the two positive eigenvalues of $\cl$ are less than 1.
 
  We conclude that in the $A1B1$ selection regime an allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
-Thus in this case, evolution tends to reduce vertical transmission, and hence increase the rate of oblique transmission.
+Thus in this case, evolution tends to reduce vertical transmission, and hence increase the rate of oblique transmission,  and there is a {\sl reduction principle} for the rate of vertical transmission (Feldman and Liberman 1986; Altenberg et al.\ 2017).
 The evolutionary dynamics of the reduction in $\rho$ under the $A1B1$ cycling regime are shown in Figure 5, which also illustrates the change in phenotype frequencies over time.
+
  In the case of identically distributed random fitnesses $w_A$ and $w_B$, Figure 6 shows  an example of the success of modifiers that reduce $\rho$.
-
- Thus in the cycling fitness regime  $A1B1$  and the symmetric random fitness regime of Figure 6, there is a {\sl reduction principle} for the rate of vertical transmission (Feldman and Liberman 1986; Altenberg et al.\ 2017).
+We have not, however, been able to prove that there is a reduction principle for this class of fluctuating fitnesses.
+
   \medskip
@@
 
-Mathematical analysis of modifiers of $\rho$ for more general $AkBl$ regimes is difficult, and we have made a numerical investigation of the evolutionarily stable value $\rho^*$ of the rate of vertical transmission for $k,l\ne 1$.
-For $A1B2$, $\rho^*=0$ still appears to be the stable state.
-However, as we increase $k$ and $l$, the stable value of $\rho$ changes, and becomes a function of the fitness difference between $A$ and $B$.
-Representative results are shown in Table 1, where it is clear that, even in the $AkBk$ case, increasing $k$ results in an increase of $\rho^*$.
-Figure 8 illustrates the results for the $AkBk$ case for $k$ up to 50 and shows that the value of $\rho^*$ depends on $k$ but also in a complex manner on $w/W$.
-For the values of $k$ and $l$ in Table 1, lower values of $W/w$ give a lower value of $\rho^*$, i.e., $\rho^*$ close to zero.
-However, as $k$ increases in the $AkBk$ case, values of $\rho^*$ increase as $W/w$ decreases.
-The modification process is clearly much more complex than the evolution of modifiers of recombination, mutation, or migration.
-
- It is natural to also ask how the population mean fitness is affected by modification of $\rho$. We have the following result.
+Values of $\rho^*$ for some $AkBl$ examples (see Supplemental Material SP6 for analysis details) are recorded in Table 1 for different values of $w$ relative to $W=1$.
+Interestingly, for the $A1B2$ regime, with $w=0.1$, the evolutionarily stable value of $\rho$ is substantially greater than zero, as it is in the $A3B10$ and $A5B30$ regimes.
+$AkBk$ results are plotted in Figure 8B.
+In the $A2B2$ regime, $\rho^* =0$ and there is reduction of vertical transmission for all selection values tested.
+However, for $AkBk$ regimes with $k>2$, we find $\rho^*\ne 0$, and depending on $w$, $\rho^*$ can be as high as 0.95.
+In Table 1, blank values for $\rho^*$ indicate that our method was numerically unstable  and that a precise value for $\rho^*$ could not be obtained.
+This is why, in Figure 8B, no $\rho^*$ points are shown for $AkBk$ with $k>19$.
+In Table 1, the word ``fixation'' indicates that fixation of $B$ occurs, at which point there can be no effect of modification of $\rho$; $\rho^*$ cannot be calculated in such cases.
+\bigskip\bigskip
+
+\noindent{\bf GEOMETRIC MEAN FITNESS AND RATE OF VERTICAL TRANSMISSION}
+\smallskip
+
+In fluctuating environments, the geometric mean fitness of genotypes has been shown to determine their evolutionary dynamics (Haldane and Jayakar  1963; Karlin and Lieberman 1974; Wakano et al.\ 2004).
+For the evolution of mutation rates that are controlled by genetic modifiers, the stable mutation rate and the mutation rate that maximizes the geometric mean fitness of the population appear to be the same when the period of environmental fluctuation is low enough (Liberman et al.\ 2011).
+We can ask the same question here: is the stable rate $\rho^*$ the same as the rate $\hat\rho$ that maximizes the equilibrium value of the geometric mean fitness in fluctuating environments.
+First consider the $A1B1$ selection regime.
+We have the following result.
 
@@
 
-\noindent{\it Proof.} In $A1B1$ the stable frequency of phenotype $A$ is by (38)
-$$x^* ={1\over 2} -{W+w-Z\over 2(2-\rho)(W-w)},\eqno(49)$$
+\noindent{\it Proof.} In $A1B1$ the stable frequency of phenotype $A$ is by (35)
+$$x^* ={1\over 2} -{W+w-Z\over 2(2-\rho)(W-w)},\eqno(46)$$
 where $Z =\sqrt{(1-\rho)^2(W-w)^2 +4Ww}>0$.
-The mean fitness at the stable equilibrium  is\hfil\break $\overline w^*\cdot\overline w^{**}$, and as $\overline w^* =\overline w^{**}$  by equation (42), this allows us to reduce the problem to properties of $\overline w^*$.
+The mean fitness at the stable equilibrium  is\hfil\break $\overline w^*\cdot\overline w^{**}$, and as $\overline w^* =\overline w^{**}$  by equation (39), this allows us to reduce the problem to properties of $\overline w^*$.
 Now since $W>w$,  $\overline w^*$ is an increasing linear function of $x^*$:
-$$\overline w^* =x^*W +(1-x^*)w = x^*(W-w) +w.\eqno(50)$$
-Thus $\overline w^*$ is decreasing in $\rho$ if $dx^*/d\rho$ is negative. Using (49),
-$${dx^*\over d\rho} =- {(1-\rho)(W-w)\over 2(2-\rho)Z} -{W+w-Z\over 2(2-\rho)(W-w)} = {x^*-{1\over 2}\over 2-\rho} -{(1-\rho)(W-w)\over 2(2-\rho)Z}.\eqno(51)$$
-From (39) $0<x^*\le{1\over 2}$ and therefore $dx^*/d\rho <0$, which completes the proof.
-
-  Figure 2B illustrates the decrease with increasing $\rho$ of the geometric mean fitness (over time) at a polymorphic equilibrium  in the $A1B1$ regime.
-Figure 7 illustrates an example of the $A1B2$ selection regime: panel A shows the values of selection and $\rho$ that allow a polymorphic equlibrium of phenotypes $A$ and $B$, and Panel B shows that the mean fitness at this equilibrium is decreasing as $\rho$ increases.
-
- The value of $\rho$ that maximizes the geometric mean fitness, denoted by $\hat\rho$,  varies with $k$ and with $W/w$ in the $AkBk$ regimes.
-This is shown in Figure 9, where for $W=1$ with $w=0.1$ and $0.5$, $\hat\rho$ is noticeably positive.
-For $W=1$, $w=0.9$, however, $\hat\rho=0$ maximizes the geometric mean fitness.
-More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Supplementary Figure S4.
-\medskip
-\centerline{[FIGURES 7, 8, 9 HERE]}  
+$$\overline w^* =x^*W +(1-x^*)w = x^*(W-w) +w.\eqno(47)$$
+Thus $\overline w^*$ is decreasing in $\rho$ if $dx^*/d\rho$ is negative. Using (46),
+$${dx^*\over d\rho} =- {(1-\rho)(W-w)\over 2(2-\rho)Z} -{W+w-Z\over 2(2-\rho)(W-w)} = {x^*-{1\over 2}\over 2-\rho} -{(1-\rho)(W-w)\over 2(2-\rho)Z}.\eqno(48)$$
+From (36) $0<x^*\le{1\over 2}$ and therefore $dx^*/d\rho <0$, which completes the proof.
+
+  Figure 2B illustrates the decrease kf the geometric mean fitness with increasing $\rho$ at a polymorphic equilibrium in the $A1B1$ regime.
+The values of $\hat\rho$ and $\rho^*$ are the same in $A1B1$ and $A2B2$ regimes, namely both are zero.
+Figure 7B shows the geometric mean fitness in the $A1B2$ regime, and we see that for small values of $w$, this mean fitness decreases as $\rho$ increases.
+At $w=0.1$, Table 1 shows that $\hat\rho$ is 0.00065, whereas $\rho^*=0.821$.
+In all $AkBk$ regimes that we tested with $w=0.9$, the value of $\hat\rho$ was also zero, substantially different from the values of $\rho^*$, as shown in Figures 8A and 8B.
+Also in Figure 8A, we see that with $w=0.1$, $\hat\rho$ changes from zero to positive in the $AkBk$ regimes with $k \ge 12$, while with $w=0.5$, the change occurs at $k=31$.
+ In Figure 8, with $w=0.1$, $\hat\rho$ is between 0.15 and 0.24 for $12\le k\le 50$, while with $w=0.5$, $\hat\rho$ exceeds 0.2 for $31\le k\le 50$.
+More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Table 1, Supplementary Figures S5 and S6, and Supplemental Material SP6.
+\medskip
+\centerline{[TABLE 1 AND FIGURES 7, 8 HERE]}  
  \bigskip
@@
 Also, let $x'$ represent the frequency of the phenotype $A$ in the infinite population model in the next generation, namely (see (2)),
- $$x'={w_Ax\over\overline w}\rho +(1-\rho)x.\eqno(52)$$
+ $$x'={w_Ax\over\overline w}\rho +(1-\rho)x.\eqno(49)$$
  Then, according to the Wright-Fisher model (Ewens 2004), $X_{t+1}$, the number of individuals of phenotype $A$ at generation $(t+1)$, is determined by the probability
- $$P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\matrix{N\cr j\cr}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},\eqno(53)$$
+ $$P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\matrix{N\cr j\cr}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},\eqno(50)$$
  for $j=0,1,2,\dots,N$.
@@
  This Markov chain  has two absorbing states, $X_t =N$ and $X_t=0$, corresponding to the two fixations in $A$ and $B$, respectively, and we are interested in the fixation probabilities and the time to fixation of these two absorbing states as functions of the initial frequency $x$ and also of $\rho$, $w_A$, and $w_B$.
-
- To these ends we use a {\sl diffusion approximation}  to the process $\{X_t\}$.
-Specifically, we compute the mean $\mu(x)$ and the variance $\sigma^2(x)$ of the change in {\sl one generation} in the frequency of phenotype $A$ given that at the beginning of the generation $X_t=Nx$.
-
- To compute $\mu(x)$, observe that by (52)   
+ To these ends we use a diffusion approximation  to the process $\{X_t\}$.
+Specifically, we compute the mean $\mu(x)$ and the variance $\sigma^2(x)$ of the change in one generation  in the frequency of phenotype $A$ given that at the beginning of the generation $X_t=Nx$.
+
+ To compute $\mu(x)$, observe that by (49)   
  $$\eqalignno{
  x' -x &={w_Ax\over\overline w}\rho +(1-\rho)x -x =\rho x\left[{w_A\over \overline w}-1\right]\cr
- &= \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},&(54)\cr}$$
+ &= \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},&(51)\cr}$$
  since $\overline w =w_A x +w_B(1-x)$.
@@
 That is, we assume
- $$w_A\simeq 1+{s_1\over N},\qquad w_B\simeq 1+{s_2\over N}.\eqno(55)$$
- Therefore, up to terms of order smaller than ${1\over N}$,
- $${w_A-w_B\over w_Ax +w_B(1-x)}\simeq {{s_1-s_2\over N}\over 1+ {s_1\over N}x +{s_2\over N}(1-x)} \simeq {s_1 -s_2\over N}.\eqno(56)$$
- Let $s=s_1-s_2$, then
- $$x'-x\simeq {1\over N}\rho sx(1-x)\eqno(57)$$
+ $$w_A -w_B ={s\over N}.\eqno(52)$$
+ Then
+ $$x'-x\simeq {1\over N}\rho sx(1-x)\eqno(53)$$
  up to terms of order small than ${1\over N}$. Since one generation corresponds to $\Delta t\simeq {1\over N}$, we  conclude that
- $$\mu(x) =\rho sx(1-x);\quad 0\le x\le 1.\eqno(58)$$
-
- We now compute $\sigma^2(x)$. From the binomial distribution of $X_{t+1}$ with parameters $N$ and $x'$, the variance of $X_{t+1} -X_t$ given $X_t =Nx$ is $Nx'(1-x')$. Following (52) we have
- $$\eqalignno{
- x'(1-x') &=x\left[{w_A\over \overline w}\rho +(1-\rho)\right](1-x)\left[{w_B\over\overline w}\rho +(1-p)\right]&(59)\cr
- &=x(1-x)\left[{w_Aw_B\over\left(\overline w\right)^2}\rho^2 +{w_A +w_B\over\overline w}\rho(1-\rho) +\left(1-\rho\right)^2\right].&(60)}$$
- Assuming again that $w_A\simeq 1+{s_1\over N}$, $w_B\simeq 1+{s_2\over N}$, up to terms of order smaller than ${1\over N}$ we have
- $${w_A\over\overline w}\simeq 1+{s_1-s_2\over N}(1-x),\qquad {w_B\over\overline w} \simeq 1 +{s_2-s_1\over N}x.\eqno(61)$$
- Consequently
- $$\eqalign{
- {w_A +w_B\over\overline w} &\simeq 2 +{s_1-s_2\over N}(1-2x)\cr\noalign{\smallskip}
- {w_A\cdot w_B\over\left(\overline w\right)^2} &\simeq 1+{s_1-s_2\over N} (1-2x).}
- \eqno(62)$$
- Thus (62) combined with (60) imply that, up to terms of order smaller than ${1\over N}$,
- $$x'(1-x') =x(1-x)\left[1+ \rho{s\over N}(1-2x)\right],\eqno(63)$$
- and the variance, $V$, of the frequency of the phenotype $A$ given that $X_t =Nx$ is
- $$\eqalign{
- V\left({X_{t+1}-X_t\over N}\mid X_t =Nx\right) &={1\over N^2}V\bigl(X_{t+1}\mid X_t=Nx\bigr)\cr \noalign{\smallskip}
- &={x'(1-x')\over N},}\eqno(64)$$
- and in one generation ($\Delta t\simeq{1\over N}$),
- $$\sigma^2(x)\cdot{1\over N} \simeq{1\over N} x(1-x)\left[1 +\rho{s\over N}(1-2x)\right],\eqno(65)$$
- which up to terms of order smaller than ${1\over N}$ gives
- $$\sigma^2(x) =x(1-x).\eqno(66)$$
-
- We  can compute now the fixation probabilities and the expected time to fixation.
-Let $u(x)$ be the probability of fixation in $A$ given that the initial frequency  of $A$ is $x$, and let $T(x)$ be the expected time until the population becomes fixed if the initial frequency of $A$ is $x$.
-Then (following, e.g., Karlin and Levikson 1974) we have
- $$u(x) ={\int_0^xs(\eta)d\eta\over\int_0^1s(\eta)d\eta},\ \ \hbox{\tenrm where}\ \ s(\eta)=\exp\left[-\int^\eta{2\mu(\xi)\over\sigma^2(\xi)}d\xi\right],\eqno(67)$$
- and
- $$T(x)=\int_0^1G(x,\xi)d\xi,\eqno(68)$$
- where
- $$G(x,\xi)=\left\{{2u(x)\bigl[u(1) -u(\xi)\bigr]{C\over s(\xi)\sigma^2(\xi)},\quad 0\le x\le\xi\le 1 \atop
- 2\bigl[u(1) -u(x)\bigr]u(\xi){C\over s(\xi)\sigma^2(\xi)},\quad 0\le\xi\le x\le 1}\right.\eqno(69)$$
- and $C=\int_0^1s(\eta)d\eta$.
-
- In our case, $\mu(\xi)=\rho s\xi(1-\xi)$ and $\sigma^2(\xi)=\xi(1-\xi)$. Therefore $2\mu(\xi)/
- \sigma^2(\xi) =2\rho s$, $s(\eta) =e^{-2\rho s\eta}$, and it can be easily seen that
- $$u(x) ={1-e^{-2\rho sx}\over 1-e^{-2\rho s}}.\eqno(70)$$
- Now
- $$C=\int_0^1s(\eta)d\eta =\int_0^1e^{-2\rho s\eta}d\eta  ={1-e^{-2\rho s}\over 2\rho s},\eqno(71)$$
-Therefore  $T(x)$ is given by
- $$T(x)={1-u(x)\over\rho s}\int_0^x{e^{2\rho s\xi}-1\over \xi(1-\xi)}d\xi +{u(x)\over\rho s}\int_x^1{1-e^{-2\rho s(1-\xi)}\over \xi(1-\xi)}d\xi,\eqno(72)$$
- where $u(x)$ is given in (70), and in generations, $T(x)$ is multiplied by $N$.
-  Unfortunately the integrals in (72) cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ (Karlin and Levikson 1974), and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.
+ $$\mu(x) =\rho sx(1-x);\quad 0\le x\le 1.\eqno(54)$$
+ In the same way, we can compute $\sigma^2(x)$, which is $x(1-x)$. These values of $\mu(x)$ and $\sigma^2(x)$ allow us to compute $u(x)$, namely
+  $$u(x) ={1-e^{-2\rho sx}\over 1-e^{-2\rho s}}.\eqno(55)$$
+ The expected time to fixation in $A$ starting from a initial frequency of $x$ is is given by
+ $$T(x)={1-u(x)\over\rho s}\int_0^x{e^{2\rho s\xi}-1\over \xi(1-\xi)}d\xi +{u(x)\over\rho s}\int_x^1{1-e^{-2\rho s(1-\xi)}\over \xi(1-\xi)}d\xi,\eqno(56)$$
+ where $u(x)$ is given in (55), and in generations, $T(x)$ is multiplied by $N$.
+  Unfortunately the integrals in (56) cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ (Ewens 2004, p.\ 160), and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.
 
@@
 
- The proof of Result 10 is in Supplementary Materials SP4.  
-  Figure 10 compares the fixation probability and time to fixation derived numerically from  simulating the Wright-Fisher Markov chain with the diffusion-derived values of $u(x)$ and $T(x)$.
+ The proof of Result 10 is in Supplementary Material SP4.  
+  Figure 9 compares the fixation probability and time to fixation derived numerically from  simulating the Wright-Fisher Markov chain with the diffusion-derived values of $u(x)$ and $T(x)$.
 The fit is seen to be very good.
@@
 
-  \centerline{[FIGURE 10 HERE]}
+  \centerline{[FIGURE 9 HERE]}
 
@@
 Also let
- $${1\over N}s_t \simeq w_A^t -w_B^t,\qquad S_t =\sum_{i=1}^t s_i,\qquad t=1,2,\dots,n.\eqno(73)$$
+ $${1\over N}s_t \simeq w_A^t -w_B^t,\qquad S_t =\sum_{i=1}^t s_i,\qquad t=1,2,\dots,n.\eqno(57)$$
  Following Karlin and Levikson (1974) we have the following result.
@@
  $$\eqalign{
- \mu(x) &= \rho S_n x(1-x)\cr \sigma^2(x) &= nx(1-x).}\eqno(74)$$
-
+ \mu(x) &= \rho S_n x(1-x)\cr \sigma^2(x) &= nx(1-x).}\eqno(58)$$
  The proof of Result 11, based on induction on $n$, is given in Supplemental Materials SP5.
 
- Using the moments in eqs.\ (74), the fixation probability $u(x)$ and the expected time $T(x)$ to fixation can be computed. Comparing to (70) and (72), with $s$  replaced by $S_n/n$, we find
- $$u(x) ={1-e^{-2\rho{S_n\over n}x}\over 1-e^{-2\rho{S_n\over n}}},\eqno(75)$$
+ Using the moments in eqs.\ (58), the fixation probability $u(x)$ and the expected time $T(x)$ to fixation from an initial frequency of $x$ can be computed where $s$ is replaced by $s_n/n$. We find
+ $$u(x) ={1-e^{-2\rho{S_n\over n}x}\over 1-e^{-2\rho{S_n\over n}}},\eqno(59)$$
  and $T(x)$ can be computed similarly.
 
- In the special case of the $AkBl$ cycling environment, we write $n=k+l$, and if $w_A=W$, $w_B=w$ for $k$ generations, and $w_A=w$, $w_B=W$ for $l$ generations, we have
- $$S_n=S_{k+l} =(k-l)(W-w).\eqno(76).$$
- Figure 11 illustrates how $(k-l)$, $\rho$, and $(W-w)$, which enter the formula for $u(x)$ in (75), interact to affect fixation probabilities.
-
- \medskip
-\centerline{[FIGURE 11 HERE]}  
+ In the  case of the $AkBl$ cycling environment, we write $n=k+l$, and if $w_A=W$, $w_B=w$ for $k$ generations, and $w_A=w$, $w_B=W$ for $l$ generations, we have
+ $$S_n=S_{k+l} =N(k-l)(W-w).\eqno(60)$$
+ Figure 10 illustrates how $(k-l)$, $\rho$, and $(W-w)$, which enter the formula for $u(x)$ in (59), interact to affect fixation probabilities.
+
+ \medskip
+\centerline{[FIGURE 10 HERE]}  
   \bigskip
@@
  \centerline{\bf TABLE 1}
- \smallskip
-
- $$\vbox{\halign{\hfil #&&\qquad\hfil#\hfil\cr
- &&&&evolutionarily&fitness\cr
- &&&&stable&optimal\cr
- &$k$&$l$&$w$&$\rho^*$&$\hat\rho$\cr
- 0&3&10&0.1&0.5286296&0.01010101\cr
- 1&5&30&0.1&0.24641176&0.01010101\cr
- 2&3&10&0.5&0.38577372&0.02020202\cr
- 3&5&30&0.5&0.46443287&0.01010101\cr
- 4&3&10&0.9&0.9681569&0.05050505\cr
- 5&5&30&0.9&0.64959375&0.02020202\cr
-6&12&12&0.1&0.84855774&0.24242424\cr
-7&30&30&0.5&0.96077135&0\cr
-8&1&1&0.1&0&0\cr
-9&1&2&0.1&0&0.01010101\cr
-10&2&2&0.1&0.00015691&0\cr
-11&1&1&0.5&0&0\cr
-12&1&2&0.5&0&0.22222222\cr
-13&2&2&0.5&5.68E-05&0\cr
-14&1&1&0.9&0&0\cr
-15&1&2&0.9&0&0.36363636\cr
-16&2&2&0.9&5.06E-05&0\cr}}$$
+ \centerline{Values of $\rho^*$ (stable $\rho$)$^{\hbox{\sevenrm (a)}}$ and $\hat\rho$ (``optimal'' $\rho$)$^{\hbox{\sevenrm (b)}}$}
+ %\smallskip
+ \vskip-.25truein
+ $$\vbox{\halign{#\hfil&&\qquad#\hfil\cr
+ %&&&&evolutionarily&fitness\cr
+ %&&&&stable&optimal\cr
+ \noalign{\hrule\vskip 8pt}
+ $k$&$l$&$w^\dagger$&\omit\hfill$\hat\rho$\hfill&\omit\hfill$\rho^*$\hfill\cr
+ \noalign{\vskip 8pt\hrule\vskip 8pt}
+ 1&1&0.1&0.0000000&0.0000000\cr
+ 1&1&0.5&0.0000000&0.0000000\cr
+ 1&1&0.9&0.0000000&0.0000000\cr
+ 1&2&0.1&0.00065&0.821\cr
+ 1&2&0.5&fixation&fixation\cr
+ 1&2&0.9&fixation&fixation\cr
+ 2&2&0.1&0.0000000&0.0000000\cr
+ 2&2&0.5&0.0000000&0.0000000\cr
+ 2&2&0.9&0.0000000&0.0000000\cr
+ 3&10&0.1&0.00031&0.619\cr
+ 3&10&0.5&fixation&fixation\cr
+ 3&10&0.9&fixation&fixation\cr
+ 5&30&0.1&0.00027&0.6503\cr
+ 5&30&0.5&fixation&fixation\cr
+ 5&30&0.9&fixation&fixation\cr
+ 12&12&0.1&0.24347&0.84924\cr
+ 12&12&0.5&0.0000000&0.91209\cr
+ 12&12&0.9&0.0000000&0.95686\cr
+ 20&20&0.1&0.223925\cr
+ 20&20&0.5&0.0000000&0.94643\cr
+ 20&20&0.9&0.0000000&0.98304\cr
+ 30&30&0.1&0.1932801\cr
+ 30&30&0.5&0.0000000&0.96331\cr
+ 30&30&0.9&0.0000000&0.0000000\cr
+ 50&50&0.1&0.15419\cr
+ 50&50&0.5&0.22107&0.9768\cr
+ 50&50&0.9&0.0000000&0.99581\cr
+\noalign{\vskip 8pt\hrule}
+\noalign{\medskip}
+\multispan4$^\dagger$\ \hbox{\tenrm Note:}\ $W=1$\hfill\cr
+}}$$
+\medskip
+\vbox{\hsize6.5truein
+(a): $\rho^*$ is the uninvadable value of the vertical transmission rate.
+
+(b): $\hat\rho$ is the geometric mean value of $\rho$ at the stable equilibrium of the $AkBl$ cycle.}
  \vfil\break
@@
 
-From Result 2, the two fixations $x=0$ and $x=1$ are not stable because $F'(0)>1$ and $F'(1)>1$. Therefore
+From Result 2, the two fixations $x=0$ and $x=1$ are not stable because $F'(0)>1$ and $F'(1)>1$ (see eq.\ 15). Therefore
 $$\eqalignno{
@@
  as expected.
+\bigskip
+\bigskip
+
+ \noindent{\bf SP6. Calculation of stable vertical transmission rate in $AkBk$}
+ \medskip
+
+Here we describe the analysis of the stability of a modifier allele $m$ with vertical transmission rate $\rho$ to invasion by a modifier $M$ with a vertical transmission rate $P$, as described in eq.
+32, in environmental regime $AkBl$.
+
+The analysis is similar to that used in Result 8 to analyze stability in $A1B1$, but it is numerical because the cases where $k>1$ or $l>1$ require solving polynomials of degree $>6$  in order to obtain closed form expressions.
+
+The analysis includes the following steps for fixed $W, w, k,$ and $l$.
+First, we find the stable frequency of phenotype $A$ with a single modifier $x^*$.
+This is done by minimizing the expression $|x_{k+l} - x_{0}|$ where $x_{t}$ is defined in eq.
+9.
+The minimization is done by iterating the recurrence (9) until it converges, i.e.
+until the difference $|x_{k+l} - x_{0}|$ is smaller than available machine precision (roughly $10^{-8}$ when subtracting similar small numbers).
+Next, we set the frequency vector with two modifiers to $\underline x^*=(x^*, 1-x^*, 0, 0)$, that is, to the stable frequencies in the absence of modifier $M$.
+
+Now we define $F_A(\underline x)$ by eq. 32 with $w_A=W$ and $w_B=w$ ($W>w$), and similarly $F_B(\underline x)$ with $w_B=W$ and $w_A=w$.
+Also, we define, similar to eq. 34, $F(\underline x)$ by composition
+$$F=\underbrace{F_B\circ \cdots\circ F_B}_{l\ \hbox{\sevenrm times}} \circ\underbrace{F_A\circ \cdots\circ F_A}_{k\ \hbox{\sevenrm times}}.\eqno(\hbox{\tenrm S}51)$$
+
+To obtain a linear approximation of $F(\underline x)$ "near" $\underline x^*$, we calculate the Jacobian matrix of $F(\underline x)$ at $\underline x = \underline x^*$,
+$$
+\cj_{ij} = \cj(\underline x^*)_{ij} = {\partial F(\underline x^*)_i \over \partial x_j},
+\eqno(\hbox{\tenrm S}52)$$
+and we define the 2-by-2 external stability matrix $\cl=\cl_{ex}$ as in eqs. 37-38 such that (note that the upper-right block is $\underline 0$ because $x^*_3=x^*_4=0$)
+$$
+\cj = \left[\matrix{\cl_{in} & \underline 0\cr\noalign{\medskip}
+ * & \cl_{ex}\cr}\right].
+\eqno(\hbox{\tenrm S}53)$$
+
+We calculate the eigenvalues $\lambda_1 > \lambda_2$ of $\cl$ using the quadratic formula as the characteristic polynomial of $\cl$ has degree 2.
+By the Perron-Frobenius theorem, the leading eigenvalue $\lambda_1$ is real and positive.
+Denote by $\lambda_1(\rho, P)$ the resulting leading eigenvalue with resident rate $\rho$ and invader rate $P$.
+Note that for any $\rho \in (0,1)$
+$$
+\lambda_1(\rho, \rho) = 1.
+\eqno(\hbox{\tenrm S}54)$$
+The evolutionary stable rate $\rho^*$ is defined to be stable to invasion, that is, for a small enough value $\partial P>0$ we have
+$$
+\lambda_1(\rho^*,\rho^* \pm \partial P) < 1 = \lambda_1(\rho^*,\rho^*),
+\eqno(\hbox{\tenrm S}55)$$
+where the equality is given by eq. S54.
+Therefore, 
+$$
+{\partial \lambda_1 \over \partial P} \big(\rho^*,\rho^*\big) = 0. 
+\eqno(\hbox{\tenrm S}56)$$
+
+We use Brent's root-finding method (Brent, 1971) to find $\rho^*$ that satisfies eq. S56.
+If, due to numerical instability of the described numerical process, we have
+$$
+\partial {\lambda_1 \over \partial P} \big(0,0\big) \cdot {\partial \lambda_1 \over \partial P} \big(1,1) > 0,
+\eqno(\hbox{\tenrm S}57)$$
+i.e., the partial derivative sign is identical at $\rho=P=0$ and $\rho=P=1$, then we cannot use Brent's method.
+In these cases we assume that the partial derivative doesn't have a root in $(0,1)$ and we determine the stable rate $\rho^*$ by the rule
+$$
+\rho^* =\left\{\matrix{
+0 &\hbox{\tenrm if}\ {\partial \lambda_1 \over \partial P} \big(0,0\big) \le 0\cr 
+\noalign{\smallskip}
+1 & \hbox{\tenrm if}\ {\partial \lambda_1 \over \partial P} \big(0,0\big) > 0\cr}\right.\eqno(\hbox{\tenrm S}58)$$
+
+Supplementary Figure S7 shows the leading eigenvalue $\lambda_1$ of the external stability matrix $\cl$ for different choices of environmental cycles $AkBk$, resident rate $\rho$ and invader rate $P$ when $W=1$ and $w=0.5$.
+The arrows show a series of invading modifiers.
+
+The described numerical analysis is fine for small $k$, but for large $k$ and especially for $w=0.1$ the calculation is unstable.
+This is probably because when the environment is constant for a long period of time, most of the time the frequencies $x_i$ are close to the boundaries (i.e. 0 and 1).
+
+Crucially, the Jacobian $\cj$ in eq.
+S52 is calculated using {\sl automatic differentiation} from a function that iteratively calculates $F(\underline x^*)$ according to eq.
+S51.
+Similarly, the partial derivative ${\partial \lambda_1 \over \partial P}$ in eq.
+S56 is calculated from a function that calculates $\lambda_1$ using simple arithmetic operations.
+Note that {\sl automatic differentiation} does not mean {\sl symbolic} or {\sl numerical differentiation}, which can lead to inefficient or inaccurate estimation of $\cj$ when $k$ is not very small.
+Rather, from Bartholomew-Biggs et al.
+(2000):
+{\sl "Automatic differentiation is a set of techniques for transforming a program that calculates numerical values of a function, into a program which calculates numerical values for derivatives of that function with about the same accuracy and efficiency as the function values themselves."}
 

### 2017-11-03 poster/poster.tex
first draft

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% a0poster Landscape Poster
% LaTeX Template
% Version 1.0 (22/06/13)
%
% The a0poster class was created by:
% Gerlinde Kettl and Matthias Weiser (tex@kettl.de)
% 
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[a0,landscape]{a0poster}

\usepackage{multicol} % This is so we can have multiple columns of text side-by-side
\columnsep=100pt % This is the amount of white space between the columns in the poster
\columnseprule=3pt % This is the thickness of the black line between the columns in the poster

\usepackage[svgnames]{xcolor} % Specify colors by their 'svgnames', for a full list of all colors available see here: http://www.latextemplates.com/svgnames-colors

\usepackage{times} % Use the times font
%\usepackage{palatino} % Uncomment to use the Palatino font

\usepackage{graphicx} % Required for including images
\graphicspath{{figures/}} % Location of the graphics files
\usepackage{booktabs} % Top and bottom rules for table
\usepackage[font=small,labelfont=bf]{caption} % Required for specifying captions to tables and figures
\usepackage{amsfonts, amsmath, amsthm, amssymb} % For math fonts, symbols and environments
\usepackage{wrapfig} % Allows wrapping text around tables and figures

\begin{document}

%----------------------------------------------------------------------------------------
%	POSTER HEADER 
%----------------------------------------------------------------------------------------

% The header is divided into three boxes:
% The first is 55% wide and houses the title, subtitle, names and university/organization
% The second is 25% wide and houses contact information
% The third is 19% wide and houses a logo for your university/organization or a photo of you
% The widths of these boxes can be easily edited to accommodate your content as you see fit

\begin{minipage}[b]{1\linewidth}
\veryHuge \color{NavyBlue} \textbf{Vertical and Oblique Transmission under Fluctuating Selection} \color{Black}\\ % Title
%\veryHuge \color{NavyBlue} \textbf{\qquad\qquad\qquad\qquad under Fluctuating Selection}\\[1cm] % Subtitle
\huge \textbf{Yoav Ram$^1,a$, Uri Liberman$^{2}$, and Marcus W. Feldman$^1$}\\ % Author(s)
\huge $^1$Department of Biology, Stanford University\\ % University/organization
\huge $^{2}$School of Mathematical Sciences, Tel-Aviv University\\ % University/organization
\huge $^{a}$yoav@yoavram.com\\ % Email
\end{minipage}
%
%\begin{minipage}[b]{0.25\linewidth}
%\color{DarkSlateGray}\Large \textbf{Contact Information:}\\
%Phone: +1 (650) 440-9625\\ % Phone number
%Email: \texttt{yoav@yoavram.com}\\ % Email address
%Twitter: \texttt{@yoavram}\\ % Twitter handle
%\end{minipage}
%%
%\begin{minipage}[b]{0.19\linewidth}
%\includegraphics[width=20cm]{cehg_logo} % Logo or a photo of you, adjust its dimensions here
%\end{minipage}

\vspace{1cm} % A bit of extra whitespace between the header and poster content

%----------------------------------------------------------------------------------------

\begin{multicols}{4} % This is how many columns your poster will be broken into, a poster with many figures may benefit from less columns whereas a text-heavy poster benefits from more

%----------------------------------------------------------------------------------------
%	ABSTRACT
%----------------------------------------------------------------------------------------

%\color{Navy} % Navy color for the abstract
%
%\begin{abstract}
%
%Sed fringilla tempus hendrerit. Vestibulum ante ipsum primis in faucibus orci luctus et ultrices posuere cubilia Curae; Etiam ut elit sit amet metus lobortis consequat sit amet in libero. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Phasellus vel sem magna. Nunc at convallis urna. isus ante. Pellentesque condimentum dui. Etiam sagittis purus non tellus tempor volutpat. Donec et dui non massa tristique adipiscing. Quisque vestibulum eros eu. Phasellus imperdiet, tortor vitae congue bibendum, felis enim sagittis lorem, et volutpat ante orci sagittis mi. Morbi rutrum laoreet semper. Morbi accumsan enim nec tortor consectetur non commodo nisi sollicitudin. Proin sollicitudin. Pellentesque eget orci eros. Fusce ultricies, tellus et pellentesque fringilla, ante massa luctus libero, quis tristique purus urna nec nibh.
%
%\end{abstract}

%----------------------------------------------------------------------------------------
%	INTRODUCTION
%----------------------------------------------------------------------------------------

\color{Navy} % SaddleBrown color for the introduction

\section*{Introduction}

Oblique transmission occurs when a trait is passed to offspring from non-parental adult of the parental generation.

Examples include social learning in humans, birds, and mammals, pathogen andsymbiont transmission, and HGT in microbes.

%----------------------------------------------------------------------------------------
%	MODEL
%----------------------------------------------------------------------------------------

\color{DarkSlateGray} % DarkSlateGray color for the rest of the content

\section*{Model}

We model an infinite population with two phenotype -- $A$ and $B$ -- with frequencies $x$ and $1-x$ and fitnesses $w_A$ and $w_B$.

Offspring inherit its phenotype from its parent with probability $\rho$ or from a random adult with probability $1-\rho$.

The frequency of $A$ in the next generation $x'$ is given by
\begin{equation}
x' = \rho \frac{w_A}{\bar{w}}x + (1-\rho) x
\end{equation}

%----------------------------------------------------------------------------------------
%	RESULTS
%----------------------------------------------------------------------------------------

\section*{Results: Constant environment}

When the environment is constant, the fitness parameters $w_A$ and $w_B$ do not change between generations, and we have the following result.

\paragraph*{Result 1.} \textit{If $0<\rho\le 1$ and both $w_A$ and $w_B$ are positive with $w_A\ne w_B$, then fixation in the phenotype $A\ (B)$ is globally stable when $w_A>w_B$ $(w_A<w_B)$.}

\section*{Results: Periodically changing environment}

Consider an environmental regime $AkBl$ which favors phenotype $A$ for $k$ generations and then phenotypes $B$ for $l$ generations.
Let $W$ be the fitness of the favored phenotype and $w$ be that of the other phenotype where $0<w<W$.

\paragraph*{Result 2.}\textit{Fixation of both $A$ and $B$ is unstable if
\begin{equation}
-{\log(b)\over\log a} < {l \over k} < -{\log a\over\log(b)}.
\end{equation} 
Specifically, if $k=l$ then there is a protected polymorphism.}

\begin{center}\vspace{1cm}
\includegraphics[width=0.8\linewidth]{lk_phase_plane_F}
\captionof{figure}{Oblique transmission maintains polymorphism.}
\end{center}\vspace{1cm}

\paragraph*{Result 3.}\textit{In the case $A1B1$ with there exists a unique stable polymorphism:
\begin{equation}
x^* ={1\over 2} -{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}\over 2\cdot(2-\rho)(W-w)}.
\end{equation}
}

\begin{center}\vspace{1cm}
\includegraphics[width=0.8\linewidth]{A1B1_equilibrium_B}
\captionof{figure}{In $A1B1$ selection regime, the stable population mean fitness decrease as $\rho$ increases.
}
\end{center}\vspace{1cm}

%\begin{center}\vspace{1cm}
%\begin{tabular}{l l l l}
%\toprule
%\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2} \\
%\midrule
%Treatment 1 & 0.0003262 & 0.562 \\
%Treatment 2 & 0.0015681 & 0.910 \\
%Treatment 3 & 0.0009271 & 0.296 \\
%\bottomrule
%\end{tabular}
%\captionof{table}{\color{Green} Table caption}
%\end{center}\vspace{1cm}

\section*{Modifier model}

Consider a modifier locus with alleles $m/M$ that determines the vertical transmission rates $\rho/P$:

\begin{equation}
\begin{matrix}
 \text{ pheno-genotype}&mA&mB&MA&MB\\
 \text{ frequency}&x_1&x_2&x_3&x_4\\
 \text{ fitness}&w_A&w_B&w_A&w_B\\
 \text{ vertical transmission rate}&\rho&\rho&P&P\\
\end{matrix}
\end{equation}

\section*{Results: $A1B1$ selection regime}

\paragraph*{Result 4.}\textit{In $A1B1$ selection regime, a modifier allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
}

\begin{center}\vspace{1cm}
\includegraphics[width=0.8\linewidth]{A1B1_modifier_invasions}
\captionof{figure}{In $A1B1$ selection regime, modifier alleles that reduce the vertical transmission rate successfully invade and fix in the population.
}
\end{center}\vspace{1cm}

\section*{Results: $AkBl$ selection regime}

\begin{center}\vspace{1cm}
\includegraphics[width=0.8\linewidth]{AkBk_stable_optimal_rate}
\captionof{figure}{The evolutionary stable vertical transmission rate $\rho^*$ and the transmission rate $\hat{\rho}$ that maximized the geometric average of the population mean fitness over $k+l$ generations are only the same for $k=l=1$ and $k=l=2$.
}
\end{center}\vspace{1cm}

\section*{Conclusions}

\begin{itemize}
\item Oblique transmission maintains polymorphism in constant and periodically changing environments.
\item Oblique transmission is favored by selection in rapidly changing environments.
\item Vertical transmission is favored by selection in slowly changing and constant environments.
\end{itemize}

%\color{DarkSlateGray} % Set the color back to DarkSlateGray for the rest of the content

%----------------------------------------------------------------------------------------
%	REFERENCES
%----------------------------------------------------------------------------------------

\nocite{*} % Print all references regardless of whether they were cited in the poster or not
\bibliographystyle{plain} % Plain referencing style
\bibliography{sample} % Use the example bibliography file sample.bib

%----------------------------------------------------------------------------------------
%	ACKNOWLEDGEMENTS
%----------------------------------------------------------------------------------------

\section*{Acknowledgements}

Partial support by the Stanford Center for Computational, Evolutionary and Human Genomics

%----------------------------------------------------------------------------------------

\end{multicols}
\end{document}

### 2017-11-05 poster/poster.tex
finished

@@
 \columnseprule=3pt % This is the thickness of the black line between the columns in the poster
+%\raggedcolumns
 
@@
 \begin{minipage}[b]{1\linewidth}
-\veryHuge \color{NavyBlue} \textbf{Vertical and Oblique Transmission under Fluctuating Selection} \color{Black}\\ % Title
-%\veryHuge \color{NavyBlue} \textbf{\qquad\qquad\qquad\qquad under Fluctuating Selection}\\[1cm] % Subtitle
-\huge \textbf{Yoav Ram$^1,a$, Uri Liberman$^{2}$, and Marcus W. Feldman$^1$}\\ % Author(s)
-\huge $^1$Department of Biology, Stanford University\\ % University/organization
-\huge $^{2}$School of Mathematical Sciences, Tel-Aviv University\\ % University/organization
-\huge $^{a}$yoav@yoavram.com\\ % Email
+
+\center \veryHuge \color{NavyBlue} \textbf{Vertical and Oblique Transmission under Fluctuating Selection} \color{Black}\\ % Title
+
+\vspace{1cm}
+
+\huge \textbf{Yoav Ram$^{1,a}$, Uri Liberman$^{2}$, and Marcus W. Feldman$^1$}\\ % Author(s)
+
+\vspace{1cm}
+
+\huge $^1$ Department of Biology, Stanford University\\ % University/organization
+\huge $^{2}$ School of Mathematical Sciences, Tel-Aviv University\\ % University/organization
+%\huge $^{a}$yoav@yoavram.com\\ % Email
+
+\vspace{1cm}
+
+\Large $^{a}$ Postdoctoral fellow of the Stanford Center for Computational, Evolutionary and Human Genomics
+
 \end{minipage}
@@
 %%
-%\begin{minipage}[b]{0.19\linewidth}
-%\includegraphics[width=20cm]{cehg_logo} % Logo or a photo of you, adjust its dimensions here
+%\begin{minipage}[b]{0.2\linewidth}
+%\includegraphics[width=0.5\linewidth]{avatar_150x225} % Logo or a photo of you, adjust its dimensions here
 %\end{minipage}
 
-\vspace{1cm} % A bit of extra whitespace between the header and poster content
+\vspace{4cm} % A bit of extra whitespace between the header and poster content
 
@@
 
-%----------------------------------------------------------------------------------------
 % INTRODUCTION
-%----------------------------------------------------------------------------------------
 
-\color{Navy} % SaddleBrown color for the introduction
+\color{DarkSlateGray}
 
@@
 
-%----------------------------------------------------------------------------------------
 % MODEL
-%----------------------------------------------------------------------------------------
-
-\color{DarkSlateGray} % DarkSlateGray color for the rest of the content
 
@@
 
-%----------------------------------------------------------------------------------------
-%	RESULTS
-%----------------------------------------------------------------------------------------
-
-\section*{Results: Constant environment}
-
-When the environment is constant, the fitness parameters $w_A$ and $w_B$ do not change between generations, and we have the following result.
-
-\paragraph*{Result 1.} \textit{If $0<\rho\le 1$ and both $w_A$ and $w_B$ are positive with $w_A\ne w_B$, then fixation in the phenotype $A\ (B)$ is globally stable when $w_A>w_B$ $(w_A<w_B)$.}
-
-\section*{Results: Periodically changing environment}
-
-Consider an environmental regime $AkBl$ which favors phenotype $A$ for $k$ generations and then phenotypes $B$ for $l$ generations.
+We consider an environmental regime $AkBl$ which favors phenotype $A$ for $k$ generations and then phenotypes $B$ for $l$ generations.
 Let $W$ be the fitness of the favored phenotype and $w$ be that of the other phenotype where $0<w<W$.
 
-\paragraph*{Result 2.}\textit{Fixation of both $A$ and $B$ is unstable if
+%\section*{Constant environment}
+%
+%When the environment is constant, the fitness parameters $w_A$ and $w_B$ do not change between generations, and we have the following result.
+%
+%\paragraph*{Result 1.} \textit{If $0<\rho\le 1$ and both $w_A$ and $w_B$ are positive with $w_A\ne w_B$, then fixation in the phenotype $A\ (B)$ is globally stable when $w_A>w_B$ $(w_A<w_B)$.}
+
+\section*{Phenotype dynamics: $AkBl$}
+
+\paragraph*{Result 1.}\textit{Fixation of both $A$ and $B$ is unstable if
 \begin{equation}
--{\log(b)\over\log a} < {l \over k} < -{\log a\over\log(b)}.
+\frac{\log a}{\log(b)} < \frac{l}{k} < \frac{\log(b)}{\log a}.
 \end{equation} 
@@
 
-\paragraph*{Result 3.}\textit{In the case $A1B1$ with there exists a unique stable polymorphism:
+\columnbreak
+
+\section*{Phenotype dynamics: $A1B1$}
+
+\paragraph*{Result 2.}\textit{In the case $A1B1$ with there exists a unique stable polymorphism:
 \begin{equation}
-x^* ={1\over 2} -{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}\over 2\cdot(2-\rho)(W-w)}.
+x^* =\frac{1}{2} -\frac{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}}{2\cdot(2-\rho)(W-w)}.
 \end{equation}
@@
 
-%\begin{center}\vspace{1cm}
-%\begin{tabular}{l l l l}
-%\toprule
-%\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2} \\
-%\midrule
-%Treatment 1 & 0.0003262 & 0.562 \\
-%Treatment 2 & 0.0015681 & 0.910 \\
-%Treatment 3 & 0.0009271 & 0.296 \\
-%\bottomrule
-%\end{tabular}
-%\captionof{table}{\color{Green} Table caption}
-%\end{center}\vspace{1cm}
-
 \section*{Modifier model}
@@
 Consider a modifier locus with alleles $m/M$ that determines the vertical transmission rates $\rho/P$:
-
 \begin{equation}
@@
 
-\section*{Results: $A1B1$ selection regime}
+The next generation pheno-genotype frequencies are
+\begin{equation}
+\begin{aligned}
+\hat{w}x'_1 &= w_A x_1\bigl[(1-\rho)(x_1+x_3) +\rho\bigr] +w_B x_2(1-\rho)(x_1+x_3)\\
+\hat{w}x'_2 &= w_A x_1(1-\rho)(x_2+x_4) +w_B x_2\bigl[(1-\rho)(x_2+x_4)+\rho\bigr]\\ 
+\hat{w}x'_3 &= w_A x_3\bigl[(1-P)(x_1+x_3) +P\bigr] +w_B x_4(1-P)(x_1+x_3)\\
+\hat{w}x'_4 &= w_A x_3(1-P)(x_2+x_4) +w_B x_4\bigl[(1-P)(x_2+x_4)+P\bigr], 
+\end{aligned}
+\end{equation}
 
-\paragraph*{Result 4.}\textit{In $A1B1$ selection regime, a modifier allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
+\paragraph{Evolutionary stable vertical transmission rate.}
+We start with a single modifier $m$ with rate $\rho$ at equilibrium between phenotypes $A$ and $B$, and then introduce modifier $M$ with rate $P$ at a low frequency.
+If $m$ is stable to invasion of every $M$ with any rate $P$, then we declare $\rho$ to be the stable rate $\rho^*$.
+
+\columnbreak
+
+\section*{Modifier dynamics: $A1B1$}
+
+\paragraph*{Result 3.}\textit{In $A1B1$ selection regime, a modifier allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
 }
+
+Similar dynamics were observed in environments which randomly flip between $A$ and $B$ every generation.
 
@@
 
-\section*{Results: $AkBl$ selection regime}
+\columnbreak
+
+\section*{Modifier dynamics: $AkBl$}
+
+In the general case there is no reduction of the transmission rate, nor is there maximization of the population mean fitness.
 
@@
 \end{center}\vspace{1cm}
+
+For $k \ne l$, there is a polymorphism between $A$ and $B$ for $w=0.1$ (but not $w=0.5$ or $0.9$), in which case the stable and optimal rates given in the table.
+
+\begin{center}\vspace{1cm}
+\begin{tabular}{p{2.5cm} p{2.5cm} p{4.5cm} p{4.5cm}}
+\toprule
+\textbf{$k$} & \textbf{$l$ } & \textbf{$\hat{\rho}$} & \textbf{$\rho^*$} \\
+\midrule
+1 & 2 & 0.00065 & 0.821 \\
+3 & 10 & 0.00031 & 0.619 \\
+5 & 30 & 0.00027 & 0.6503 \\
+\bottomrule
+\end{tabular}
+\captionof{table}{Values of $\rho^*$ (stable rate) and $\hat\rho$ (fitness optimal rate) in $AkBl$ selection regime.}
+\end{center}\vspace{1cm}
+
+% Conclusions
 
@@
 
-%----------------------------------------------------------------------------------------
 % REFERENCES
-%----------------------------------------------------------------------------------------
 
-\nocite{*} % Print all references regardless of whether they were cited in the poster or not
-\bibliographystyle{plain} % Plain referencing style
-\bibliography{sample} % Use the example bibliography file sample.bib
-
-%----------------------------------------------------------------------------------------
-%	ACKNOWLEDGEMENTS
-%----------------------------------------------------------------------------------------
-
-\section*{Acknowledgements}
-
-Partial support by the Stanford Center for Computational, Evolutionary and Human Genomics
+%\nocite{*} % Print all references regardless of whether they were cited in the poster or not
+%\bibliographystyle{plain} % Plain referencing style
+%\bibliography{sample} % Use the example bibliography file sample.bib
 

### 2017-11-07 poster/poster.tex
last minute formatting, additions, phrasing, typos

@@
 %\raggedcolumns
+\usepackage{fontawesome}
 
@@
 
-\huge \textbf{Yoav Ram$^{1,a}$, Uri Liberman$^{2}$, and Marcus W. Feldman$^1$}\\ % Author(s)
+\huge \textbf{Yoav Ram$^{1}$, Uri Liberman$^{2}$, and Marcus W. Feldman$^1$}\\ % Author(s)
 
@@
 
-\vspace{1cm}
+%\vspace{1cm}
 
-\Large $^{a}$ Postdoctoral fellow of the Stanford Center for Computational, Evolutionary and Human Genomics
+\Large YR is a postdoctoral fellow of the Stanford Center for Computational, Evolutionary and Human Genomics\\ 
+\faicon{envelope-o}\; yoav@yoavram.com \quad
+\faicon{phone}\; 650.440.9625 \quad
+\faicon{globe}\; www.yoavram.com
 
@@
 
-\vspace{4cm} % A bit of extra whitespace between the header and poster content
+\vspace{2cm} % A bit of extra whitespace between the header and poster content
 
@@
 
-Examples include social learning in humans, birds, and mammals, pathogen andsymbiont transmission, and HGT in microbes.
+Examples include social learning in humans, birds, and mammals, pathogen and symbiont transmission, and HGT in microbes.
 
@@
 
-We model an infinite population with two phenotype -- $A$ and $B$ -- with frequencies $x$ and $1-x$ and fitnesses $w_A$ and $w_B$.
+We model an infinite population with two phenotype, $A$ and $B$, with frequencies $x$ and $1-x$.
+An offspring inherits its phenotype from its parent with probability $\rho$ or from a random adult with probability $1-\rho$.
 
-Offspring inherit its phenotype from its parent with probability $\rho$ or from a random adult with probability $1-\rho$.
+We consider an environmental regime $AkBl$ which favors phenotype $A$ for $k$ generations and then phenotype $B$ for $l$ generations.
+Let $W$ be the fitness of the favored phenotype and $w$ be that of the other phenotype where $0<w<W$.
 
-The frequency of $A$ in the next generation $x'$ is given by
+The frequency of $A$ in the next generation $x'$, when $A$ is \mbox{favored} over $B$, is given by
 \begin{equation}
-x' = \rho \frac{w_A}{\bar{w}}x + (1-\rho) x
+x' = \rho \frac{W}{\bar{w}}x + (1-\rho) x
 \end{equation}
-
-We consider an environmental regime $AkBl$ which favors phenotype $A$ for $k$ generations and then phenotypes $B$ for $l$ generations.
-Let $W$ be the fitness of the favored phenotype and $w$ be that of the other phenotype where $0<w<W$.
+where $\bar{w}=xW + (1-x)w$ is the population mean fitness.
 
@@
 
-\section*{Phenotype dynamics: $AkBl$}
+\section*{Phenotype dynamics: \emph{AkBl}}
 
@@
 \begin{equation}
-\frac{\log a}{\log(b)} < \frac{l}{k} < \frac{\log(b)}{\log a}.
+\frac{\log(1+\rho \frac{W-w}{w})}{\log(1+\rho \frac{w-W}{W})} < 
+\frac{l}{k} <
+\frac{\log(1+\rho \frac{w-W}{W})}{\log(1+\rho \frac{W-w}{w})}.
 \end{equation} 
@@
 \includegraphics[width=0.8\linewidth]{lk_phase_plane_F}
-\captionof{figure}{Oblique transmission maintains polymorphism.}
+\captionof{figure}{\textbf{Oblique transmission maintains polymorphism.}}
 \end{center}\vspace{1cm}
@@
 
-\section*{Phenotype dynamics: $A1B1$}
+\section*{Phenotype dynamics: \emph{A1B1}}
 
-\paragraph*{Result 2.}\textit{In the case $A1B1$ with there exists a unique stable polymorphism:
+\paragraph*{Result 2.}\textit{In selection regime $A1B1$ there exists a unique stable polymorphism:
 \begin{equation}
-x^* =\frac{1}{2} -\frac{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}}{2\cdot(2-\rho)(W-w)}.
+x^* =\frac{1}{2} -\frac{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}}{2\cdot(2-\rho)(W-w)},
 \end{equation}
+where $x^*$ is the stable frequency of phenotype $A$ after the end of each two-generation cycle.
 }
+
+Since $\bar{w}^*$ is a decreasing function function of $x^*$, the stable population mean fitness $\bar{w}^*$ is a decreasing function of $\rho$.
 
@@
 \includegraphics[width=0.8\linewidth]{A1B1_equilibrium_B}
-\captionof{figure}{In $A1B1$ selection regime, the stable population mean fitness decrease as $\rho$ increases.
+\captionof{figure}{\textbf{Mean fitness in A1B1 selection regime}.
+The stable population mean fitness $\bar{w}^*$ decreases as the vertical transmission rate $\rho$ increases.
 }
@@
 
-Consider a modifier locus with alleles $m/M$ that determines the vertical transmission rates $\rho/P$:
+Consider a modifier locus with alleles $m/M$ that determine the vertical transmission rates $\rho/P$:
 \begin{equation}
@@
  \text{ frequency}&x_1&x_2&x_3&x_4\\
- \text{ fitness}&w_A&w_B&w_A&w_B\\
+ \text{ fitness when $A$ is favored}&W&w&W&w\\
+ \text{ fitness when $B$ is favored}&w&W&w&W\\
  \text{ vertical transmission rate}&\rho&\rho&P&P\\
@@
 
-The next generation pheno-genotype frequencies are
+The next generation pheno-genotype frequencies when $A$ is favored are
 \begin{equation}
 \begin{aligned}
-\hat{w}x'_1 &= w_A x_1\bigl[(1-\rho)(x_1+x_3) +\rho\bigr] +w_B x_2(1-\rho)(x_1+x_3)\\
-\hat{w}x'_2 &= w_A x_1(1-\rho)(x_2+x_4) +w_B x_2\bigl[(1-\rho)(x_2+x_4)+\rho\bigr]\\ 
-\hat{w}x'_3 &= w_A x_3\bigl[(1-P)(x_1+x_3) +P\bigr] +w_B x_4(1-P)(x_1+x_3)\\
-\hat{w}x'_4 &= w_A x_3(1-P)(x_2+x_4) +w_B x_4\bigl[(1-P)(x_2+x_4)+P\bigr], 
+\hat{w}x'_1 &= W x_1\bigl[(1-\rho)(x_1+x_3) +\rho\bigr] + w x_2(1-\rho)(x_1+x_3)\\
+\hat{w}x'_2 &= W x_1(1-\rho)(x_2+x_4) + w x_2\bigl[(1-\rho)(x_2+x_4)+\rho\bigr]\\ 
+\hat{w}x'_3 &= W x_3\bigl[(1-P)(x_1+x_3) +P\bigr] + w x_4(1-P)(x_1+x_3)\\
+\hat{w}x'_4 &= W x_3(1-P)(x_2+x_4) + w x_4\bigl[(1-P)(x_2+x_4)+P\bigr], 
 \end{aligned}
@@
 \paragraph{Evolutionary stable vertical transmission rate.}
-We start with a single modifier $m$ with rate $\rho$ at equilibrium between phenotypes $A$ and $B$, and then introduce modifier $M$ with rate $P$ at a low frequency.
-If $m$ is stable to invasion of every $M$ with any rate $P$, then we declare $\rho$ to be the stable rate $\rho^*$.
+We start with modifier allele $m$ with rate $\rho$ at equilibrium between phenotypes $A$ and $B$, and then introduce modifier allele $M$ with rate $P$ at a low frequency.
+If $m$ is stable to invasion of any $M$ with any rate $P$, then we declare $\rho$ to be the stable rate $\rho^*$.
 
@@
 
-\section*{Modifier dynamics: $A1B1$}
+\section*{Modifier dynamics: \emph{A1B1}}
 
-\paragraph*{Result 3.}\textit{In $A1B1$ selection regime, a modifier allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
+\paragraph*{Result 3.}\textit{In selection regime $A1B1$, a modifier allele $m$ with associated vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
 }
 
-Similar dynamics were observed in environments which randomly flip between $A$ and $B$ every generation.
+That is, there is a \emph{reduction principle} for the vertical transmission rate in selection regime $A1B1$.
+
+Similar dynamics were observed in environments which randomly flip between favoring $A$ and $B$ every generation.
 
@@
 \includegraphics[width=0.8\linewidth]{A1B1_modifier_invasions}
-\captionof{figure}{In $A1B1$ selection regime, modifier alleles that reduce the vertical transmission rate successfully invade and fix in the population.
+\captionof{figure}{\textbf{Modifier invasion in A1B1 selection regime}.
+Modifier alleles that reduce the vertical transmission rate successfully invade and fix in the population.
+Dashed lines show invasion times. Initial resident allele with $\rho=0.1$.
 }
@@
 
-\section*{Modifier dynamics: $AkBl$}
+\section*{Modifier dynamics: \emph{AkBk}}
 
@@
 \includegraphics[width=0.8\linewidth]{AkBk_stable_optimal_rate}
-\captionof{figure}{The evolutionary stable vertical transmission rate $\rho^*$ and the transmission rate $\hat{\rho}$ that maximized the geometric average of the population mean fitness over $k+l$ generations are only the same for $k=l=1$ and $k=l=2$.
+\captionof{figure}{\textbf{Stability and optimality in selection regime $AkBk$.} \textbf{(A)} The transmission rate $\hat{\rho}$ that maximizes the geometric average of the population mean fitness over $k+l$ generations, and \textbf{(B)} the evolutionary stable vertical transmission rate $\rho^*$.
+$\rho^*$ and $\hat{\rho}$ are equal to 0 and to each other only when $k=l=1$ and $k=l=2$.
 }
@@
 
-For $k \ne l$, there is a polymorphism between $A$ and $B$ for $w=0.1$ (but not $w=0.5$ or $0.9$), in which case the stable and optimal rates given in the table.
+\section*{Modifier dynamics: \emph{AkBl}}
+
+For $k \ne l$, there is a polymorphism between $A$ and $B$ when $w=0.1$ (but not when $w=0.5$ or $0.9$), in which case the stable and optimal rates are given below.
 
@@
 \end{tabular}
-\captionof{table}{Values of $\rho^*$ (stable rate) and $\hat\rho$ (fitness optimal rate) in $AkBl$ selection regime.}
-\end{center}\vspace{1cm}
+\captionof{table}{\textbf{Stability and optimality in selection regime AkBl.}
+Values of evolutionary stable vertical transmission rate $\rho^*$ and the transmission rate $\hat\rho$ that maximizes geometric mean fitness.}
+\label{table}
+\end{center}%\vspace{1cm}
 

### 2017-12-04 ms/ms.tex
version submitted to pnas

@@
 \vfil
-\centerline{October 17, 2017}
+\centerline{October 25, 2017, v.\ 1}
 
 \break
+
+\noindent{\bf ABSTRACT}
+\smallskip
+
+Vertical and oblique cultural transmission of a dichotomous phenotype is studied under constant, periodic cycling, and randomly fluctuating selection.
+Conditions are derived for the existence of a stable polymorphism in a periodically cycling selection regime.
+Under such a selection regime, the fate of a genetic modifier of the rate of vertical transmission depends on the length of the cycle and the strength of selection.
+In general, the evolutionarily stable rate of vertical transmission differs markedly from the rate that maximizes the geometric mean fitness of the population.
+The evolution of rules of transmission has dramatically different dynamics than the more frequently studied modifiers of recombination, mutation, or migration.
+\bigskip
+\bigskip
+\bigskip
+
+\noindent{\bf SIGNIFICANCE}
+\smallskip
+
+Evolutionary dynamics of phenotypes in populations depend on how the traits are transmitted across generations, and how the environments that cause selection on the traits fluctuate over time.
+We show that under periodically fluctuating selection, a gene that increases the rate of vertical transmission is disfavored when the periods are short, but approaches an intermediate stable rate for longer periods.
+This stable rate differs markedly from the rate that maximizes the geometric mean fitness.
+The evolution of learning rules thus differs qualitatively from the evolution of genetically modified rules of genetic transmission.
+\vfil\break
+
 \noindent{\bf INTRODUCTION}
@@
 
-Another perspective on the evolutionary consequences of fluctuating environments (and, as a result, fluctuating fitnesses) derives from the phenomenon of phenotypic switching (Balaban et al.\ 2004; Kussell and Leibler 2005; Thattai and van Oudenaarden 2004; Salath\'e et al.\ 2009; Liberman et al.\ 2011; Gaal et al.\ 2010).
+Another perspective on the evolutionary consequences of fluctuating environments (and, as a result, fluctuating selection) derives from the phenomenon of phenotypic switching (Balaban et al.\ 2004; Kussell and Leibler 2005; Thattai and van Oudenaarden 2004; Salath\'e et al.\ 2009; Liberman et al.\ 2011; Gaal et al.\ 2010).
 In these studies, mutation causes the organism to switch phenotypes (usually treated as haploid genotypes), and the problem has usually been couched in terms of the optimal rate of mutation in models where the phenotypic fitnesses fluctuate over time.
@@
 
-In a recent analysis of evolution in fluctuating environments, Xue and Leibler (2016) allowed an organism to absorb information about the distribution of possible environments by learning the phenotypes of members of its parental lineage from previous generations.
+In a recent analysis of evolution under fluctuating selection, Xue and Leibler (2016) allowed an organism to absorb information about the distribution of possible environments by learning the phenotypes of members of its parental lineage from previous generations.
 They describe this as ``positive feedback that enhances the probability of the offspring to express the same phenotype as the parent.'' In this formulation there was ``reinforcement of the parent phenotype'' in an offspring, such as might occur through epigenetic inheritance.
 Although their analysis was not couched in terms of oblique and vertical transmission, as defined by Cavalli-Sforza and Feldman (1981), we have been stimulated by their analysis to develop a model in which oblique transmission, at a rate dependent on the trait frequency in the parental generation, occurs in addition to classical vertical transmission.
-We then ask how fluctuations in fitnesses interact with the rate of oblique transmission to affect evolutionary dynamics and how the rate of oblique transmission itself might evolve.
+We then ask how fluctuations in selection interact with the rate of oblique transmission to affect evolutionary dynamics and how the rate of oblique transmission itself might evolve.
 
@@
 
-\item{1.}Since $k=l$ and the above result holds also when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is also not stable.
+\item{1.}Since $k=l$ and the above result also holds  when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is also not stable.
 Thus neither phenotype can be lost, and there is a {\sl protected polymorphism} (Prout 1968).
 
-\item{2.}Observe that for general $k$, $l$, the condition for local stability of fixation in $A$ is
+\item{2.}For general $k$, $l$, the condition for local stability of fixation in $A$ is
 $$\bigl[f_A(0)\bigr]^l \bigl[f_B(0)\bigr]^k <1, \eqno(15)$$
@@
 The local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations $A$ is favored $k$ times and $B$ is favored $l$ times, and not their order in the cycle.
-
 When neither fixation in $A$ or $B$ is stable, there is a protected polymorphism, and we expect to have one or more polymorphic equilibria.
-Figure 1 shows the relationship between $k,l$ and $\rho$ that gives polymorphism of $A$ and $B$, or fixation, for different values of $s=W-w$.
-
-\medskip
-\centerline{[FIGURE 1 HERE]}
+Supplementary Figure S1 illustrates the relationship between $k,l$ and $\rho$ that gives polymorphism of $A$ and $B$, or fixation, for different values of $s=W-w$.
+
 For the simple case of $A1B1$ periodically changing environment we have the following.
@@
 Starting from any initial value $0<x_0 <1$ we have  $x_t\,\harr{}{t\to\infty}\,x^*$.
-Supplementary Figure S1, panels A, C, and E, illustrate examples of how the frequency of $A$ changes over time in the $A1B1$ regime of cycling selection.
+Supplementary Figure S2, panels A, C, and E, illustrate examples of how the frequency of $A$ changes over time in the $A1B1$ regime of cycling selection.
 
@@
 
-The proof of Result 5 is in Supplementary Material SP1.  Figure 2A shows the stable equilibrium frequencies $x^*$ as a function of $\rho$, $W$, and $w$ in the $A1B1$ regime.
-For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplemental Figure S3 illustrates the convergence to a single stable polymorphism.
+The proof of Result 5 is in Supplementary Material SP1.  Supplementary Figure S3A shows the stable equilibrium frequencies $x^*$ as a function of $\rho$, $W$, and $w$ in the $A1B1$ regime.
+For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplementary Figure S4 illustrates the convergence to a single stable polymorphism.
 
 We have not been able to prove that for selection regimes $AkBl$ with $l\ne k$ there is a single stable polymorphic equilibrium when the two fixations are unstable.
-However, the numerical examples in Figure 1 (for $AkBl$) and in Figure 7 (for the special case $A1B2$) all exhibit a single stable polymorphic equilibrium when fixations in $A$ and $B$ are unstable.
+However, the numerical examples in Supplementary Figure S1 (for $AkBl$) and in Figure 1 and Supplementary Figure S5 for the special case $A1B2$ all exhibit a single stable polymorphic equilibrium when fixations in $A$ and $B$ are unstable.
 These numerical results suggest that for $W>w>0$ and $0<\rho<1$ the high order polynomial that gives the equilibria has only a single root corresponding to a globally stable polymorphism.
-Supplementary Figure S4 shows that this is the case for the $A3B10$ regime.
-\medskip
-\centerline{[FIGURE 2 HERE]}  
+Supplementary Figure S6 shows that this is the case for the $A3B10$ regime.
+\medskip
+\centerline{[FIGURE 1 HERE]}  
  \bigskip
@@
 It is possible, however, that neither fixation is  stochastically locally stable, in which case we expect the population to converge to a polymorphic distribution.
-Figure 3 illustrates how the properties of $s_t$ in recursion (28) affect the frequency of phenotype $A$, and in particular the stochastic local stability of fixation in phenotype $B$.
-Figure 4 shows the dynamics of the frequency of $A$ in a case where $w_A$ and $w_B$ are identically distributed and independent; in this case the expectation of the stationary distribution is ${1\over 2}$, and its variance increases as $\rho$ increases.
-\medskip
-\centerline{[FIGURES 3, 4 HERE]}  
+Figure 2 illustrates how the properties of $s_t$ in recursion (28) affect the frequency of phenotype $A$, and in particular the stochastic local stability of fixation in phenotype $B$.
+Figure 3 shows the dynamics of the frequency of $A$ in a case where $w_A$ and $w_B$ are identically distributed and independent; in this case the expectation of the stationary distribution is ${1\over 2}$, and its variance increases as $\rho$ increases.
+\medskip
+\centerline{[FIGURES 2, 3 HERE]}  
  \bigskip
@@
 Constant environment always leads to fixation of the favored type, independent of $\rho$.
-We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$ (but see Supplemental Material SP6 for computational analysis of the general $AkBl$ case).
+We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$ (but see Supplementary Material SP6 for computational analysis of the general $AkBl$ case).
 Specifically, from (32) with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
@@
 For the $A1B1$ case, when only the $m$ allele is present with associated rate $\rho$,  $0<\rho<1$, and $0<w<W$, a unique stable equilibrium $\underline x^* =(x_1^*,1-x_1^*,0,0)$ exists.
-$x_1^*$ is the only positive root of the quadratic equation $Q(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with $\alpha_2,\alpha_1,\alpha_0$ specified in equation (21).
+$x_1^*$ is the only positive root of the quadratic equation $Q(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with $\alpha_2,\alpha_1,\alpha_0$ specified in equation (22).
 Solving $Q(x)=0$ gives
@@
 
-We can prove the following result.
+We then have the following result, which is proved in Supplementary Material SP4.
 
@@
 
-  \noindent{\it Proof}.
-As $\cl$ is a positive matrix, by the Perron-Frobenius theorem $\cl$ has a positive eigenvalue, and as $a_0>0$ and $a_2=1$ the product of the two eigenvalues of $\cl$ is positive.
-Thus $\cl$ has two positive eigenvalues.
-Let $R(1)=R(1;P)$, then from (43)
-  $$R(1;P) = {W^2w^2-\left(\overline w^*\widetilde w^*\right)^2\over\left(\overline w^*\right)^4}P^2 +2P{\left(\widetilde w^*\right)^2 -Ww\over \left(\overline w^*\right)^2} +{\left(\overline w^*\right)^2-\left(\widetilde w^*\right)^2\over\left(\overline w^*\right)^2},\eqno(44)$$
-where $\widetilde w^* =Wx_2^* +wx_1^*$.
-
-By (36) $\bigl(\sqrt{Ww}-w\bigr)/(W-w) <x_1^* <{1\over 2}$, from which it is easily seen that
-$$\sqrt{Ww}< \overline w^*<\widetilde w^* .\eqno(45)$$
-
- When $P=\rho$ one of the eigenvalues of $\cl$ is 1; hence $R(1;\rho)=0$.
-Another root of $R(1;P)=0$ is $\bigl[(\overline w^*)^2 +\overline w^*\widetilde w^*\bigr]/\bigl[Ww +\overline w^*\widetilde w^*\bigr]$, which by (45) is larger than 1.
-As $R(1;0)=\bigl[(\overline w^*)^2 -(\widetilde w^*)^2\bigr]/(\overline w^*)^2 <0$ by (45), we deduce that when $0<P<\rho$, $R(1;P)<0$, whereas when $\rho<P<1$, $R(1;P)>0$.
- Hence, when $P<\rho$, $R(1)<0$, and since $a_2=1$, $R(+\infty)>0$, we conclude that $R(\lambda)=0$ has a positive root larger than 1 and the largest positive eigenvalue of $\cl$ is larger than 1.
-
- When $P>\rho$, we have $R(1)>0$ and also $R(0)=a_0>0$.
-As $R(\lambda)=0$ has two positive roots and as $a_2>0$, $R(\lambda)$ is convex, either the two positive roots are less than 1 or both larger than one.
-But the product of the two roots is $P^2W^2w^2/(\overline w^*)^2<1$ by (45); thus when $P>\rho$ the two positive eigenvalues of $\cl$ are less than 1.
-
  We conclude that in the $A1B1$ selection regime an allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
 Thus in this case, evolution tends to reduce vertical transmission, and hence increase the rate of oblique transmission,  and there is a {\sl reduction principle} for the rate of vertical transmission (Feldman and Liberman 1986; Altenberg et al.\ 2017).
-The evolutionary dynamics of the reduction in $\rho$ under the $A1B1$ cycling regime are shown in Figure 5, which also illustrates the change in phenotype frequencies over time.
-
- In the case of identically distributed random fitnesses $w_A$ and $w_B$, Figure 6 shows  an example of the success of modifiers that reduce $\rho$.
+The evolutionary dynamics of the reduction in $\rho$ under the $A1B1$ cycling regime are shown in Figure 4, which also illustrates the change in phenotype frequencies over time.
+
+ In the case of identically distributed random fitnesses $w_A$ and $w_B$, Figure 5 shows  an example of the success of modifiers that reduce $\rho$.
 We have not, however, been able to prove that there is a reduction principle for this class of fluctuating fitnesses.
@@
   \medskip
-\centerline{[FIGURES 5, 6]}
-
-Values of $\rho^*$ for some $AkBl$ examples (see Supplemental Material SP6 for analysis details) are recorded in Table 1 for different values of $w$ relative to $W=1$.
+\centerline{[FIGURES 4, 5]}
+
+Values of $\rho^*$ for some $AkBl$ examples (see Supplementary Material SP7 and  Figure S10 for analytical details) are recorded in Table 1 for different values of $w$ relative to $W=1$.
 Interestingly, for the $A1B2$ regime, with $w=0.1$, the evolutionarily stable value of $\rho$ is substantially greater than zero, as it is in the $A3B10$ and $A5B30$ regimes.
-$AkBk$ results are plotted in Figure 8B.
+$AkBk$ results are plotted in Figure 6B.
 In the $A2B2$ regime, $\rho^* =0$ and there is reduction of vertical transmission for all selection values tested.
@@
 In Table 1, blank values for $\rho^*$ indicate that our method was numerically unstable  and that a precise value for $\rho^*$ could not be obtained.
-This is why, in Figure 8B, no $\rho^*$ points are shown for $AkBk$ with $k>19$.
+This is why, in Figure 6B, no $\rho^*$ points are shown for $AkBk$ with $k>19$.
 In Table 1, the word ``fixation'' indicates that fixation of $B$ occurs, at which point there can be no effect of modification of $\rho$; $\rho^*$ cannot be calculated in such cases.
@@
 
-In fluctuating environments, the geometric mean fitness of genotypes has been shown to determine their evolutionary dynamics (Haldane and Jayakar  1963; Karlin and Lieberman 1974; Wakano et al.\ 2004).
+Under fluctuating selection, the geometric mean fitness of genotypes has been shown to determine their evolutionary dynamics (Haldane and Jayakar  1963; Karlin and Lieberman 1974; Wakano et al.\ 2004).
 For the evolution of mutation rates that are controlled by genetic modifiers, the stable mutation rate and the mutation rate that maximizes the geometric mean fitness of the population appear to be the same when the period of environmental fluctuation is low enough (Liberman et al.\ 2011).
-We can ask the same question here: is the stable rate $\rho^*$ the same as the rate $\hat\rho$ that maximizes the equilibrium value of the geometric mean fitness in fluctuating environments.
+We can ask the same question here: is the stable rate $\rho^*$ the same as the rate $\hat\rho$ that maximizes the equilibrium value of the geometric mean fitness under fluctuating selection.
 First consider the $A1B1$ selection regime.
@@
 \noindent{\it Proof.} In $A1B1$ the stable frequency of phenotype $A$ is by (35)
-$$x^* ={1\over 2} -{W+w-Z\over 2(2-\rho)(W-w)},\eqno(46)$$
+$$x^* ={1\over 2} -{W+w-Z\over 2(2-\rho)(W-w)},\eqno(44)$$
 where $Z =\sqrt{(1-\rho)^2(W-w)^2 +4Ww}>0$.
@@
 Now since $W>w$,  $\overline w^*$ is an increasing linear function of $x^*$:
-$$\overline w^* =x^*W +(1-x^*)w = x^*(W-w) +w.\eqno(47)$$
-Thus $\overline w^*$ is decreasing in $\rho$ if $dx^*/d\rho$ is negative. Using (46),
-$${dx^*\over d\rho} =- {(1-\rho)(W-w)\over 2(2-\rho)Z} -{W+w-Z\over 2(2-\rho)(W-w)} = {x^*-{1\over 2}\over 2-\rho} -{(1-\rho)(W-w)\over 2(2-\rho)Z}.\eqno(48)$$
+$$\overline w^* =x^*W +(1-x^*)w = x^*(W-w) +w.\eqno(45)$$
+Thus $\overline w^*$ is decreasing in $\rho$ if $dx^*/d\rho$ is negative. Using (44),
+$${dx^*\over d\rho} =- {(1-\rho)(W-w)\over 2(2-\rho)Z} -{W+w-Z\over 2(2-\rho)(W-w)} = {x^*-{1\over 2}\over 2-\rho} -{(1-\rho)(W-w)\over 2(2-\rho)Z}.\eqno(46)$$
 From (36) $0<x^*\le{1\over 2}$ and therefore $dx^*/d\rho <0$, which completes the proof.
 
-  Figure 2B illustrates the decrease kf the geometric mean fitness with increasing $\rho$ at a polymorphic equilibrium in the $A1B1$ regime.
+  Figure 4 illustrates the decrease over time of the geometric mean fitness with increasing $\rho$ at a polymorphic equilibrium in the $A1B1$ regime.
 The values of $\hat\rho$ and $\rho^*$ are the same in $A1B1$ and $A2B2$ regimes, namely both are zero.
-Figure 7B shows the geometric mean fitness in the $A1B2$ regime, and we see that for small values of $w$, this mean fitness decreases as $\rho$ increases.
+Figure 1B shows the geometric mean fitness in the $A1B2$ regime, and we see that for small values of $w$, this mean fitness decreases as $\rho$ increases.
 At $w=0.1$, Table 1 shows that $\hat\rho$ is 0.00065, whereas $\rho^*=0.821$.
-In all $AkBk$ regimes that we tested with $w=0.9$, the value of $\hat\rho$ was also zero, substantially different from the values of $\rho^*$, as shown in Figures 8A and 8B.
-Also in Figure 8A, we see that with $w=0.1$, $\hat\rho$ changes from zero to positive in the $AkBk$ regimes with $k \ge 12$, while with $w=0.5$, the change occurs at $k=31$.
- In Figure 8, with $w=0.1$, $\hat\rho$ is between 0.15 and 0.24 for $12\le k\le 50$, while with $w=0.5$, $\hat\rho$ exceeds 0.2 for $31\le k\le 50$.
-More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Table 1, Supplementary Figures S5 and S6, and Supplemental Material SP6.
-\medskip
-\centerline{[TABLE 1 AND FIGURES 7, 8 HERE]}  
+In all $AkBk$ regimes that we tested with $w=0.9$, the value of $\hat\rho$ was also zero, substantially different from the values of $\rho^*$, as shown in Figures 6A and 6B.
+Also in Figure 6A, we see that with $w=0.1$, $\hat\rho$ changes from zero to positive in the $AkBk$ regimes with $k \ge 12$, while with $w=0.5$, the change occurs at $k=31$.
+ In Figure 6, with $w=0.1$, $\hat\rho$ is between 0.15 and 0.24 for $12\le k\le 50$, while with $w=0.5$, $\hat\rho$ exceeds 0.2 for $31\le k\le 50$.
+More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Table 1, Supplementary Figures S7 and S8, and Supplemental Material SP7.
+\medskip
+\centerline{[TABLE 1 AND FIGURE 6 HERE]}  
  \bigskip
@@
 Also, let $x'$ represent the frequency of the phenotype $A$ in the infinite population model in the next generation, namely (see (2)),
- $$x'={w_Ax\over\overline w}\rho +(1-\rho)x.\eqno(49)$$
+ $$x'={w_Ax\over\overline w}\rho +(1-\rho)x.\eqno(47)$$
  Then, according to the Wright-Fisher model (Ewens 2004), $X_{t+1}$, the number of individuals of phenotype $A$ at generation $(t+1)$, is determined by the probability
- $$P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\matrix{N\cr j\cr}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},\eqno(50)$$
+ $$P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\matrix{N\cr j\cr}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},\eqno(48)$$
  for $j=0,1,2,\dots,N$.
@@
 
- To compute $\mu(x)$, observe that by (49)   
+ To compute $\mu(x)$, observe that by (47)   
  $$\eqalignno{
  x' -x &={w_Ax\over\overline w}\rho +(1-\rho)x -x =\rho x\left[{w_A\over \overline w}-1\right]\cr
- &= \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},&(51)\cr}$$
+ &= \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},&(49)\cr}$$
  since $\overline w =w_A x +w_B(1-x)$.
@@
 That is, we assume
- $$w_A -w_B ={s\over N}.\eqno(52)$$
+ $$w_A -w_B ={s\over N}.\eqno(50)$$
  Then
- $$x'-x\simeq {1\over N}\rho sx(1-x)\eqno(53)$$
+ $$x'-x\simeq {1\over N}\rho sx(1-x)\eqno(51)$$
  up to terms of order small than ${1\over N}$. Since one generation corresponds to $\Delta t\simeq {1\over N}$, we  conclude that
- $$\mu(x) =\rho sx(1-x);\quad 0\le x\le 1.\eqno(54)$$
+ $$\mu(x) =\rho sx(1-x);\quad 0\le x\le 1.\eqno(52)$$
  In the same way, we can compute $\sigma^2(x)$, which is $x(1-x)$. These values of $\mu(x)$ and $\sigma^2(x)$ allow us to compute $u(x)$, namely
-  $$u(x) ={1-e^{-2\rho sx}\over 1-e^{-2\rho s}}.\eqno(55)$$
+  $$u(x) ={1-e^{-2\rho sx}\over 1-e^{-2\rho s}}.\eqno(53)$$
  The expected time to fixation in $A$ starting from a initial frequency of $x$ is is given by
- $$T(x)={1-u(x)\over\rho s}\int_0^x{e^{2\rho s\xi}-1\over \xi(1-\xi)}d\xi +{u(x)\over\rho s}\int_x^1{1-e^{-2\rho s(1-\xi)}\over \xi(1-\xi)}d\xi,\eqno(56)$$
- where $u(x)$ is given in (55), and in generations, $T(x)$ is multiplied by $N$.
-  Unfortunately the integrals in (56) cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ (Ewens 2004, p.\ 160), and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.
+ $$T(x)={1-u(x)\over\rho s}\int_0^x{e^{2\rho s\xi}-1\over \xi(1-\xi)}d\xi +{u(x)\over\rho s}\int_x^1{1-e^{-2\rho s(1-\xi)}\over \xi(1-\xi)}d\xi,\eqno(54)$$
+ where $u(x)$ is given in (53), and in generations, $T(x)$ is multiplied by $N$.
+  Unfortunately the integrals in (54) cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ (Ewens 2004, p.\ 160), and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.
 
@@
 
- The proof of Result 10 is in Supplementary Material SP4.  
-  Figure 9 compares the fixation probability and time to fixation derived numerically from  simulating the Wright-Fisher Markov chain with the diffusion-derived values of $u(x)$ and $T(x)$.
+ The proof of Result 10 is in Supplementary Material SP5.  
+  Supplementary Figure S9 compares the fixation probability and time to fixation derived numerically from  simulating the Wright-Fisher Markov chain with the diffusion-derived values of $u(x)$ and $T(x)$.
 The fit is seen to be very good.
-Note that when $N$ is large, the Wright-Fisher model exhibits persistent fluctuation around the deterministic expectation, as shown by the orange diagram in Supplemental Figure S1.
-  \medskip
-
-  \centerline{[FIGURE 9 HERE]}
+Note that when $N$ is large, the Wright-Fisher model exhibits persistent fluctuation around the deterministic expectation, as shown by the orange diagram in Supplemental Figure S2.
 
@@
 Also let
- $${1\over N}s_t \simeq w_A^t -w_B^t,\qquad S_t =\sum_{i=1}^t s_i,\qquad t=1,2,\dots,n.\eqno(57)$$
+ $${1\over N}s_t \simeq w_A^t -w_B^t,\qquad S_t =\sum_{i=1}^t s_i,\qquad t=1,2,\dots,n.\eqno(55)$$
  Following Karlin and Levikson (1974) we have the following result.
@@
  $$\eqalign{
- \mu(x) &= \rho S_n x(1-x)\cr \sigma^2(x) &= nx(1-x).}\eqno(58)$$
- The proof of Result 11, based on induction on $n$, is given in Supplemental Materials SP5.
-
- Using the moments in eqs.\ (58), the fixation probability $u(x)$ and the expected time $T(x)$ to fixation from an initial frequency of $x$ can be computed where $s$ is replaced by $s_n/n$. We find
- $$u(x) ={1-e^{-2\rho{S_n\over n}x}\over 1-e^{-2\rho{S_n\over n}}},\eqno(59)$$
+ \mu(x) &= \rho S_n x(1-x)\cr \sigma^2(x) &= nx(1-x).}\eqno(56)$$
+
+\noindent The proof of Result 11, based on induction on $n$, is given in Supplemental Materials SP6.
+
+ Using the moments in eqs.\ (56), the fixation probability $u(x)$ and the expected time $T(x)$ to fixation from an initial frequency of $x$ can be computed where $s$ is replaced by $s_n/n$. We find
+ $$u(x) ={1-e^{-2\rho{S_n\over n}x}\over 1-e^{-2\rho{S_n\over n}}},\eqno(57)$$
  and $T(x)$ can be computed similarly.
@@
  In the  case of the $AkBl$ cycling environment, we write $n=k+l$, and if $w_A=W$, $w_B=w$ for $k$ generations, and $w_A=w$, $w_B=W$ for $l$ generations, we have
- $$S_n=S_{k+l} =N(k-l)(W-w).\eqno(60)$$
- Figure 10 illustrates how $(k-l)$, $\rho$, and $(W-w)$, which enter the formula for $u(x)$ in (59), interact to affect fixation probabilities.
-
- \medskip
-\centerline{[FIGURE 10 HERE]}  
+ $$S_n=S_{k+l} =N(k-l)(W-w).\eqno(58)$$
+ Figure 7 shows an example of how $(k-l)$, $\rho$, and $(W-w)$, which enter the formula for $u(x)$ in (57), interact to affect fixation probabilities.
+More examples are illustrated in Supplementary Figure S11.
+
+ \medskip
+\centerline{[FIGURE 7 HERE]}  
   \bigskip
@@
  We have shown that in the $A1B1$ case, the rate of vertical transmission tends to decrease when it is under the control of a genetic modifier.
-This is also true in the random selection case, when the fitnesses of $A$ and $B$ are identically distributed and independent between generations.
- %In light of these results, together with the finite population result that fixation probability is higher, and time to fixation faster with greater $\rho$, we can conclude that increased oblique transmission in our model leads to greater phenotypic diversity; i.e., it is more likely to result in  phenotypic polymorphism.
-This can be seen in Figure 1, where it is also clear that the existence and stability of a polymorphism is more sensitive to $(1-\rho)$ than to the fitness difference between $A$ and $B$.
-
- The mean fitness of the population evaluated at a polymorphism of phenotypes $A$ and $B$ is shown to be an increasing function of the level of oblique transmission in the $A1B1$ (Figure 2B, Figures 5C, 5F, 5I) and $A1B2$ (Figure 7B) environments.
-We can compare this result with the asymptotic growth rate (AGR) of Xue and Leibler (2016), whose parameter  $\eta$ is the rate at which an individual learns from its parental lineage.
-For $k$ small enough, the $AkBk$ environment shows the AGR decreasing with $\eta$, but larger values of $k$ entail that the AGR has a maximum for an intermediate value of $\eta$.
-[FINISH WRITEUP]
+From numerical iteration, it appears that this is also true in the random selection case, when the fitnesses of $A$ and $B$ are identically distributed and independent between generations.
+However, for $AkBl$ selection regimes more complicated than $A1B1$, evolution of a modifier of vertical transmission is not straightforward.
+While reduction of $\rho$ occurs in the $A2B2$ regime, the uninvadable value $\rho^*$ is not zero for all the fitness values explored in $AkBk$ regimes with $k>2$ (Figure 6B, Table 1).
+In fact, $\rho^*$ increases sharply as $k$ increases beyond $k=2$.
+This is an unusual scenario for genetic modifiers, although it must be noted that a modifier of $\rho$ is not neutral; it affects primary selection, while neutral  modifiers of recombination, mutation, and migration affect induced or secondary selection.
+
+ The dependence of the modifier dynamics on the strength of selection (that is $w$ when $W=1$) is complicated by the approach of the system to fixation.
+When the phenotype frequencies become exceedingly small, dependence of the dynamics of the modifier of $\rho$ becomes extremely difficult to detect due to numerical instability; this is especially true for larger values of $k$ in $AkBk$ regimes when $w$ is small (Table 1, Figure 6B with $k\ge 19$).
+
+ Figure 6A (see also Table 1) shows that the value $\hat\rho$ that maximizes the geometric mean fitness is the same as the evolutionarily stable value $\rho^*$ in the $A1B1$ and $A2B2$ selection regimes.
+For $AkBk$ regimes with $k>2$, our numerical analysis shows the $\hat\rho$ depends strongly on the strength of selection; i.e., the value of $w$ relative to $W=1$.
+For $AkBl$ regimes with $w=0.1$, the difference between $\hat\rho$ and $\rho^*$ is seen even with the $A1B2$ environment.
+For $AkBk$ regimes and $w=0.9$, we find $\hat\rho=0$, while $\rho^*$ is close to $0.9$.
+For larger values of $k$, $\hat\rho$ lies between 0.15 and 0.25, while $\rho^*$ remains above 0.8 and can reach 0.99 for very large $k$.
+Comparing Figure 6A with the asymptotic growth rate of Xue and Leibler (2016), whose parameter $\eta$ is the rate at which an individual learns from its parental lineage, there is a similarity for our curves for $w=0.1$ and $0.5$ with their curve in the $AkBk$ environment.
+They show the AGR decreasing with $\eta$ in the $AkBk$ regime for small $k$, but larger values of $k$ entail that the AGR has a maximum for an intermediate value of $\eta$.
+
+ Although the models of Xue and Leibler (2016) and that analyzed here both incorporate parental and non-parental transmission, they do so in qualitatively different ways.
+The model treated in the present paper is squarely in the tradition of gene-culture coevolutionary theory, together with modifier theory from population genetics.
+The different findings from the two classes of models are interesting and suggest that further exploration of the overlaps and discrepancies between the two approaches would be worthwhile.
 
@@
 
- \noindent{\bf SP4. Proof of Result 10}
+ \noindent{\bf SP4. Proof of Result 8}
+ \medskip
+
+As $\cl$ is a positive matrix, by the Perron-Frobenius theorem $\cl$ has a positive eigenvalue, and as $a_0>0$ and $a_2=1$ the product of the two eigenvalues of $\cl$ is positive.
+Thus $\cl$ has two positive eigenvalues.
+Let $R(1)=R(1;P)$, then from (43)
+  $$R(1;P) = {W^2w^2-\left(\overline w^*\widetilde w^*\right)^2\over\left(\overline w^*\right)^4}P^2 +2P{\left(\widetilde w^*\right)^2 -Ww\over \left(\overline w^*\right)^2} +{\left(\overline w^*\right)^2-\left(\widetilde w^*\right)^2\over\left(\overline w^*\right)^2},\eqno(\hbox{\tenrm S}29)$$
+where $\widetilde w^* =Wx_2^* +wx_1^*$.
+
+By (36) $\bigl(\sqrt{Ww}-w\bigr)/(W-w) <x_1^* <{1\over 2}$, from which it is easily seen that
+$$\sqrt{Ww}< \overline w^*<\widetilde w^* .\eqno(\hbox{\tenrm S}30)$$
+
+ When $P=\rho$ one of the eigenvalues of $\cl$ is 1; hence $R(1;\rho)=0$.
+Another root of $R(1;P)=0$ is $\bigl[(\overline w^*)^2 +\overline w^*\widetilde w^*\bigr]/\bigl[Ww +\overline w^*\widetilde w^*\bigr]$, which by (S30) is larger than 1.
+As $R(1;0)=\bigl[(\overline w^*)^2 -(\widetilde w^*)^2\bigr]/(\overline w^*)^2 <0$ by (S30), we deduce that when $0<P<\rho$, $R(1;P)<0$, whereas when $\rho<P<1$, $R(1;P)>0$.
+ Hence, when $P<\rho$, $R(1)<0$, and since $a_2=1$, $R(+\infty)>0$, we conclude that $R(\lambda)=0$ has a positive root larger than 1 and the largest positive eigenvalue of $\cl$ is larger than 1.
+
+ When $P>\rho$, we have $R(1)>0$ and also $R(0)=a_0>0$.
+As $R(\lambda)=0$ has two positive roots and as $a_2>0$, $R(\lambda)$ is convex, either the two positive roots are less than 1 or both larger than one.
+But the product of the two roots is $P^2W^2w^2/(\overline w^*)^2<1$ by (S30); thus when $P>\rho$ the two positive eigenvalues of $\cl$ are less than 1.
+%\bigskip
+%\bigskip
+\vfil
+\break
+
+ \noindent{\bf SP5. Proof of Result 10}
  \medskip
@@
 Without loss of generality and for the ease of representation, we will show that for $t>0$, 
-  $$v (x;t) ={1-e^{-tx}\over 1-e^{-t}}\eqno(\hbox{\tenrm S}29)$$
+  $$v (x;t) ={1-e^{-tx}\over 1-e^{-t}}\eqno(\hbox{\tenrm S}31)$$
   is monotone increasing as a function of $t$. Observe that
-  $${\partial v\over\partial t} = {\left(1-e^{-t}\right)xe^{-tx} -\left(1-e^{-tx}\right)e^{-t}\over \left(1-e^{-t}\right)^2}.\eqno(\hbox{\tenrm S}30)$$
+  $${\partial v\over\partial t} = {\left(1-e^{-t}\right)xe^{-tx} -\left(1-e^{-tx}\right)e^{-t}\over \left(1-e^{-t}\right)^2}.\eqno(\hbox{\tenrm S}32)$$
 
    \noindent For the monotonicity we have to show that
-  $$f(x;t) =\left(1-e^{-t}\right)xe^{-tx} -\left(1 -e^{-tx}\right)e^{-t} \ge 0\eqno(\hbox{\tenrm S}31)$$ 
+  $$f(x;t) =\left(1-e^{-t}\right)xe^{-tx} -\left(1 -e^{-tx}\right)e^{-t} \ge 0\eqno(\hbox{\tenrm S}33)$$ 
   when $t>0$ and $0\le x\le 1$. Note that $f(0;t)=0$ and $f(1;t)=0$. Also
-  $${\partial f\over\partial x} =\left(1-e^{-t}\right)\left(e^{-tx}-txe^{-tx}\right) -te^{-tx}e^{-t},\eqno(\hbox{\tenrm S}32)$$
+  $${\partial f\over\partial x} =\left(1-e^{-t}\right)\left(e^{-tx}-txe^{-tx}\right) -te^{-tx}e^{-t},\eqno(\hbox{\tenrm S}34)$$
   or
-  $${\partial f\over\partial x} =e^{-tx}\left[\left(1-e^{-t}\right)\left(1-tx\right) -te^{-t}\right] =e^{-tx}g(x;t),\eqno(\hbox{\tenrm S}33)$$
+  $${\partial f\over\partial x} =e^{-tx}\left[\left(1-e^{-t}\right)\left(1-tx\right) -te^{-t}\right] =e^{-tx}g(x;t),\eqno(\hbox{\tenrm S}35)$$
  say, where for fixed $t$, $g(x;t)$ is a linear function of $x$, which vanishes at $x_0 =(1-e^{-t} -te^{-t})/t(1-e^{-t})$.
@@
 Since $g(0,t) =1 -e^{-t} -te^{-t} >0$ and $g(1;t) =(1-e^{-t})(1-t) -te^{-t} <0$ for $t>0$, we deduce that ${\partial f\over\partial x}(x,t)>0$ for $ 0<x<x_0$ and ${\partial f\over\partial x}(x,t)<0$ for $x_0<x<1$ for all $t>0$.
-These facts, combined with $f(0,t) =f(1;t) =0$, prove that $f(x;t)\ge 0$ for $0\le x\le 1$ (in fact, $f(x;t)>0$ for $0<x<1$), and inequality (S31) is satisfied as desired.
-\bigskip
-\bigskip
-
- \noindent{\bf SP5. Proof of Result 11}
- \medskip
-
- The proof is based on induction on $n$, where in order to prove (74), we show that if $X_t$ is the number of individuals with phenotype $A$ at stage $t$ of the cycle, and $x$ is the initial frequency of $A$, then
- $$E\left({X_t\over N} -x\right) \simeq {1\over N}\rho S_t x(1-x),\qquad V\left({X_t\over N}\right)\simeq {1\over N}tx(1-x),\eqno(\hbox{\tenrm S}34)$$
+These facts, combined with $f(0,t) =f(1;t) =0$, prove that $f(x;t)\ge 0$ for $0\le x\le 1$ (in fact, $f(x;t)>0$ for $0<x<1$), and inequality (S33) is satisfied as desired.
+\bigskip
+\bigskip
+
+ \noindent{\bf SP6. Proof of Result 11}
+ \medskip
+
+ The proof is based on induction on $n$, where in order to prove (72), we show that if $X_t$ is the number of individuals with phenotype $A$ at stage $t$ of the cycle, and $x$ is the initial frequency of $A$, then
+ $$E\left({X_t\over N} -x\right) \simeq {1\over N}\rho S_t x(1-x),\qquad V\left({X_t\over N}\right)\simeq {1\over N}tx(1-x),\eqno(\hbox{\tenrm S}36)$$
  where $N$ is the size of the population.
-When $t=1$, (S34) coincides with the constant environment case.
-Assuming (S34), we go to $t+1$.
+When $t=1$, (S36) coincides with the constant environment case.
+Assuming (S36), we go to $t+1$.
 Now $X_{t+1}$ given $X_t=Ny$ has a binomial distribution with parameters $(N,y')$.
 Hence
- $$E\left({X_{t+1}\over N} -{X_t\over N} \mid X_t=Ny\right) =y'-y.\eqno(\hbox{\tenrm S}35)$$
- Following (57), $y'-y\simeq (1/N)\rho s_{t+1}y(1-y)$, and so
- $$E\left({X_{t+1}\over N} -{X_t\over N}\mid X_t\right)\simeq {1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right).\eqno(\hbox{\tenrm S}36)$$
+ $$E\left({X_{t+1}\over N} -{X_t\over N} \mid X_t=Ny\right) =y'-y.\eqno(\hbox{\tenrm S}37)$$
+ Following Ewens (2004, chapter 5), $y'-y\simeq (1/N)\rho s_{t+1}y(1-y)$, and so
+ $$E\left({X_{t+1}\over N} -{X_t\over N}\mid X_t\right)\simeq {1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right).\eqno(\hbox{\tenrm S}38)$$
  Observe that
@@
  E\left[{X_t\over N}\left(1-{X_t\over N}\right)\right] &= E\left({X_t\over N}\right) -E\left[\left({X_t\over N}\right)^2\right]\cr\noalign{\smallskip}
-    &= E\left({X_t\over N}\right) -V\left({X_t\over N}\right) -\left[E\left({X_t\over N}\right)\right]^2.\cr}\eqno(\hbox{\tenrm S}37)$$
+    &= E\left({X_t\over N}\right) -V\left({X_t\over N}\right) -\left[E\left({X_t\over N}\right)\right]^2.\cr}\eqno(\hbox{\tenrm S}39)$$
  By the induction assumption, $V(X_t/N)\simeq (1/N)tx(1-x)$, and ignoring terms of order $1/N^2$ we have
  $$E\left({X_{t+1}\over N} -{X_t\over N}\right)\simeq {1\over N}\rho s_{t+1} E\left({X_t\over N}\right)\left[1-E\left({X_t\over N}\right)\right].
-\eqno(\hbox{\tenrm S}38)$$
-Applying (S34) we have
+\eqno(\hbox{\tenrm S}40)$$
+Applying (S36) we have
 $$\eqalign{
 E\left({X_t\over N}\right)  &\simeq x+{1\over N}\rho S_tx(1-x)\cr\noalign{\smallskip}
-1-E\left({X_t\over N}\right) &\simeq 1-x-{1\over N}\rho S_tx(1-x),} \eqno(\hbox{\tenrm S}39)$$
+1-E\left({X_t\over N}\right) &\simeq 1-x-{1\over N}\rho S_tx(1-x),} \eqno(\hbox{\tenrm S}41)$$
 and ignoring terms $O(1/N^2)$, we find
-$$E\left({X_{t+1}\over N} -{X_t\over N}\right) \simeq {1\over N}\rho s_{t+1}x(1-x).\eqno(\hbox{\tenrm S}40)$$
+$$E\left({X_{t+1}\over N} -{X_t\over N}\right) \simeq {1\over N}\rho s_{t+1}x(1-x).\eqno(\hbox{\tenrm S}42)$$
 Thus
@@
 E\left({X_{t+1}\over N} -x\right) &= E\left({X_{t+1}\over N} -{X_t\over N}\right) +E\left({X_t\over N} -x\right)\cr\noalign{\smallskip}
-&\simeq {1\over N}\rho s_{t+1}x(1-x) +{1\over N}\rho S_t x(1-x),}\eqno(\hbox{\tenrm S}41)$$
+&\simeq {1\over N}\rho s_{t+1}x(1-x) +{1\over N}\rho S_t x(1-x),}\eqno(\hbox{\tenrm S}43)$$
 and since $S_t +s_{t+1} =S_{t+1}$,
- $$E\left({X_{t+1}\over N} -x\right)\simeq	{1\over N}\rho S_{t+1}x(1-x)\eqno(\hbox{\tenrm S}42)$$
+ $$E\left({X_{t+1}\over N} -x\right)\simeq	{1\over N}\rho S_{t+1}x(1-x)\eqno(\hbox{\tenrm S}44)$$
  as desired.
@@
  We now compute $V(X_{t+1}/N)$ using the induction assumption and the formula
- $$V\left({X_{t+1}\over N}\right) =E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] +V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right],\eqno(\hbox{\tenrm S}43)$$
- where by (S34)
- $$E\left({X_{t+1}\over N}\mid X_t\right)\simeq {X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\eqno(\hbox{\tenrm S}44)$$
+ $$V\left({X_{t+1}\over N}\right) =E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] +V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right],\eqno(\hbox{\tenrm S}45)$$
+ where by (S36)
+ $$E\left({X_{t+1}\over N}\mid X_t\right)\simeq {X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\eqno(\hbox{\tenrm S}46)$$
  and
- $$V\left({X_{t+1}\over N}\mid X_t\right)\simeq{1\over N}{X_t\over N}\left(1-{X_t\over N}\right).\eqno(\hbox{\tenrm S}45)$$
- Here we used the fact that by (63) $y'(1-y')\simeq y(1-y)$. Now
- $$E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] \simeq {1\over N}E\left[{X_t\over N}\left(1-{X_t\over N}\right)\right]\simeq {1\over N}x(1-x),\eqno(\hbox{\tenrm S}46)$$
- where we use the same computations as led from (S37) to (S40).
- $$V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right] =V\left[{X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right].\eqno(\hbox{\tenrm S}47)$$
+ $$V\left({X_{t+1}\over N}\mid X_t\right)\simeq{1\over N}{X_t\over N}\left(1-{X_t\over N}\right).\eqno(\hbox{\tenrm S}47)$$
+ Here we used the fact that  $y'(1-y')\simeq y(1-y)$. Now
+ $$E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] \simeq {1\over N}E\left[{X_t\over N}\left(1-{X_t\over N}\right)\right]\simeq {1\over N}x(1-x),\eqno(\hbox{\tenrm S}48)$$
+ where we use the same computations as led from (S39) to (S42).
+ $$V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right] =V\left[{X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right].\eqno(\hbox{\tenrm S}49)$$
  Since $(X_t/N)\bigl[1-(X_t/N)\bigr]$ is a random variable taking values in [0,1],  its variance is less than 1/4 and
- $$V\left[{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right] \le {1\over 4N^2}\rho^2 s_{t+1}^2.\eqno(\hbox{\tenrm S}48)$$
+ $$V\left[{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right] \le {1\over 4N^2}\rho^2 s_{t+1}^2.\eqno(\hbox{\tenrm S}50)$$
  We ignore terms $O(1/N^2)$ so that the random variable $(1/N)\rho s_{t+1}(X_t/N)\bigl[1-(X_t/N)\bigr]$ is almost constant. As a result,
- $$V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right]\simeq V\left({X_t\over N}\right) \simeq {1\over N}tx(1-x),\eqno(\hbox{\tenrm S}49)$$
- by the induction assumption. Combining (S46) and (S49) gives
- $$V\left({X_{t+1}\over N}\right)\simeq {1\over N}x(1-x) +{1\over N}tx(1-x) ={1\over N}(t+1)x(1-x)\eqno(\hbox{\tenrm S}50)$$
+ $$V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right]\simeq V\left({X_t\over N}\right) \simeq {1\over N}tx(1-x),\eqno(\hbox{\tenrm S}51)$$
+ by the induction assumption. Combining (S48) and (S51) gives
+ $$V\left({X_{t+1}\over N}\right)\simeq {1\over N}x(1-x) +{1\over N}tx(1-x) ={1\over N}(t+1)x(1-x)\eqno(\hbox{\tenrm S}52)$$
  as expected.
-\bigskip
-\bigskip
-
- \noindent{\bf SP6. Calculation of stable vertical transmission rate in $AkBk$}
- \medskip
-
-Here we describe the analysis of the stability of a modifier allele $m$ with vertical transmission rate $\rho$ to invasion by a modifier $M$ with a vertical transmission rate $P$, as described in eq.
-32, in environmental regime $AkBl$.
-
+%\bigskip
+%\bigskip
+\vfil\break
+
+ \noindent{\bf SP7. Calculation of stable vertical transmission rate in $AkBk$}
+ \medskip
+
+Here we describe the analysis of the stability of a modifier allele $m$ with vertical transmission rate $\rho$ to invasion by a modifier $M$ with a vertical transmission rate $P$, as described in eq.\ (32), in environmental regime $AkBl$.
 The analysis is similar to that used in Result 8 to analyze stability in $A1B1$, but it is numerical because the cases where $k>1$ or $l>1$ require solving polynomials of degree $>6$  in order to obtain closed form expressions.
@@
 First, we find the stable frequency of phenotype $A$ with a single modifier $x^*$.
-This is done by minimizing the expression $|x_{k+l} - x_{0}|$ where $x_{t}$ is defined in eq.
-9.
+This is done by minimizing the expression $|x_{k+l} - x_{0}|$ where $x_{t}$ is defined in eq.\ 9.
 The minimization is done by iterating the recurrence (9) until it converges, i.e.
@@
 
-Now we define $F_A(\underline x)$ by eq. 32 with $w_A=W$ and $w_B=w$ ($W>w$), and similarly $F_B(\underline x)$ with $w_B=W$ and $w_A=w$.
-Also, we define, similar to eq. 34, $F(\underline x)$ by composition
-$$F=\underbrace{F_B\circ \cdots\circ F_B}_{l\ \hbox{\sevenrm times}} \circ\underbrace{F_A\circ \cdots\circ F_A}_{k\ \hbox{\sevenrm times}}.\eqno(\hbox{\tenrm S}51)$$
+Now we define $F_A(\underline x)$ by eqs.\ (32) with $w_A=W$ and $w_B=w$ ($W>w$), and similarly $F_B(\underline x)$ with $w_B=W$ and $w_A=w$.
+Also, we define, similar to eq.\ (34), $F(\underline x)$ by composition
+$$F=\underbrace{F_B\circ \cdots\circ F_B}_{l\ \hbox{\sevenrm times}} \circ\underbrace{F_A\circ \cdots\circ F_A}_{k\ \hbox{\sevenrm times}}.\eqno(\hbox{\tenrm S}53)$$
 
@@
 \cj_{ij} = \cj(\underline x^*)_{ij} = {\partial F(\underline x^*)_i \over \partial x_j},
-\eqno(\hbox{\tenrm S}52)$$
-and we define the 2-by-2 external stability matrix $\cl=\cl_{ex}$ as in eqs. 37-38 such that (note that the upper-right block is $\underline 0$ because $x^*_3=x^*_4=0$)
+\eqno(\hbox{\tenrm S}54)$$
+and the $2\times 2$ external stability matrix $\cl=\cl_{ex}$ is as in eqs.\ (37-38)  (note that the upper-right block is $\underline 0$ because $x^*_3=x^*_4=0$)
 $$
@@
  * & \cl_{ex}\cr}\right].
-\eqno(\hbox{\tenrm S}53)$$
+\eqno(\hbox{\tenrm S}55)$$
 
@@
 \lambda_1(\rho, \rho) = 1.
-\eqno(\hbox{\tenrm S}54)$$
-The evolutionary stable rate $\rho^*$ is defined to be stable to invasion, that is, for a small enough value $\partial P>0$ we have
+\eqno(\hbox{\tenrm S}56)$$
+The evolutionary stable rate $\rho^*$ is defined to be stable to invasion; that is, for a small enough value $\partial P>0$ we have
 $$
 \lambda_1(\rho^*,\rho^* \pm \partial P) < 1 = \lambda_1(\rho^*,\rho^*),
-\eqno(\hbox{\tenrm S}55)$$
-where the equality is given by eq. S54.
+\eqno(\hbox{\tenrm S}57)$$
+where the equality is given by eq.\ (S56).
 Therefore, 
@@
 {\partial \lambda_1 \over \partial P} \big(\rho^*,\rho^*\big) = 0. 
-\eqno(\hbox{\tenrm S}56)$$
-
-We use Brent's root-finding method (Brent, 1971) to find $\rho^*$ that satisfies eq. S56.
+\eqno(\hbox{\tenrm S}58)$$
+
+We use Brent's root-finding method (Brent, 1971) to find $\rho^*$ that satisfies eq.\ (S58).
 If, due to numerical instability of the described numerical process, we have
@@
 \partial {\lambda_1 \over \partial P} \big(0,0\big) \cdot {\partial \lambda_1 \over \partial P} \big(1,1) > 0,
-\eqno(\hbox{\tenrm S}57)$$
+\eqno(\hbox{\tenrm S}59)$$
 i.e., the partial derivative sign is identical at $\rho=P=0$ and $\rho=P=1$, then we cannot use Brent's method.
@@
 \noalign{\smallskip}
-1 & \hbox{\tenrm if}\ {\partial \lambda_1 \over \partial P} \big(0,0\big) > 0\cr}\right.\eqno(\hbox{\tenrm S}58)$$
-
-Supplementary Figure S7 shows the leading eigenvalue $\lambda_1$ of the external stability matrix $\cl$ for different choices of environmental cycles $AkBk$, resident rate $\rho$ and invader rate $P$ when $W=1$ and $w=0.5$.
+1 & \hbox{\tenrm if}\ {\partial \lambda_1 \over \partial P} \big(0,0\big) > 0.\cr}\right.\eqno(\hbox{\tenrm S}60)$$
+
+Supplementary Figure S10 shows the leading eigenvalue $\lambda_1$ of the external stability matrix $\cl$ for different choices of environmental cycles $AkBk$, resident rate $\rho$ and invader rate $P$ when $W=1$ and $w=0.5$.
 The arrows show a series of invading modifiers.
 
-The described numerical analysis is fine for small $k$, but for large $k$ and especially for $w=0.1$ the calculation is unstable.
-This is probably because when the environment is constant for a long period of time, most of the time the frequencies $x_i$ are close to the boundaries (i.e. 0 and 1).
-
-Crucially, the Jacobian $\cj$ in eq.
-S52 is calculated using {\sl automatic differentiation} from a function that iteratively calculates $F(\underline x^*)$ according to eq.
-S51.
-Similarly, the partial derivative ${\partial \lambda_1 \over \partial P}$ in eq.
-S56 is calculated from a function that calculates $\lambda_1$ using simple arithmetic operations.
+The  numerical analysis above is fine for small $k$, but for large $k$ and especially for $w=0.1$ the calculation is unstable.
+This is probably because when the environment is constant for a long period of time, most of the time the frequencies $x_i$ are very close to the boundaries (i.e., 0 and 1).
+
+Crucially, the Jacobian $\cj$ in eq.\ (S54) is calculated using {\sl automatic differentiation} from a function that iteratively calculates $F(\underline x^*)$ according to eq.\ (S53).
+Similarly, the partial derivative ${\partial \lambda_1 \over \partial P}$ in eq.\ (S58) is calculated from a function that calculates $\lambda_1$ using simple arithmetic operations.
 Note that {\sl automatic differentiation} does not mean {\sl symbolic} or {\sl numerical differentiation}, which can lead to inefficient or inaccurate estimation of $\cj$ when $k$ is not very small.
@@
 (2000):
-{\sl "Automatic differentiation is a set of techniques for transforming a program that calculates numerical values of a function, into a program which calculates numerical values for derivatives of that function with about the same accuracy and efficiency as the function values themselves."}
+{\sl ``Automatic differentiation is a set of techniques for transforming a program that calculates numerical values of a function, into a program which calculates numerical values for derivatives of that function with about the same accuracy and efficiency as the function values themselves.''}
+\bigskip
+\bigskip
+
+\noindent {\bf References to Supplemental Materials}
+
+\ref Bartholomew-Biggs, M., S. Brown, B. Christianson, and L. Dixon. 2000. Automatic differentiation of algorithms. {\it J. Comput. Appl. Math.} {\bf 124}: 171--190.
+
+\ref Brent, R. P. 1971. An algorithm with guaranteed convergence for finding a zero of a function. {\it Comput. J.} {\bf 14}: 422--425.
+
+\ref Ewens, W. 2004. {\it Mathematical Population Genetics}, 2nd Edition. New York: Springer.
 

### 2017-12-11 ms/PNAS/Supplementary.tex
pnas branch, middle of revision

--- 2017-12-04 ms/ms.tex
@@
 \def\ci{\hbox{\tenbf I}}
-
-\font\tit=cmbx10 at 14pt
-\ \vskip1.1in
-\centerline{\tit Vertical and Oblique Transmission under}\smallskip
-\centerline{\tit Fluctuating Selection}
-\vskip.3in
-
-\centerline{Yoav Ram$^1$, Uri Liberman$^{2}$, and Marcus W. Feldman$^1$}
-
-\vskip.5truein
-
-%\centerline{$*$ Corresponding author}
-\bigskip
-\centerline{$^1$Department of Biology}
-%\vskip-7pt
-\centerline{Stanford University}
-%\vskip-7pt
-\centerline{Stanford, CA 94305-5020}
-\bigskip
-
-\centerline{$^2$School of Mathematical Sciences}
-%\vskip-7pt
-\centerline{Tel Aviv University}
-%\vskip-7pt
-\centerline{Tel Aviv, Israel 69978}
-%\vskip-7pt
-%\centerline{uril@tauex.tau.ac.il}
-\vfil
-\centerline{October 25, 2017, v.\ 1}
-
-\break
-
-\noindent{\bf ABSTRACT}
-\smallskip
-
-Vertical and oblique cultural transmission of a dichotomous phenotype is studied under constant, periodic cycling, and randomly fluctuating selection.
-Conditions are derived for the existence of a stable polymorphism in a periodically cycling selection regime.
-Under such a selection regime, the fate of a genetic modifier of the rate of vertical transmission depends on the length of the cycle and the strength of selection.
-In general, the evolutionarily stable rate of vertical transmission differs markedly from the rate that maximizes the geometric mean fitness of the population.
-The evolution of rules of transmission has dramatically different dynamics than the more frequently studied modifiers of recombination, mutation, or migration.
-\bigskip
-\bigskip
-\bigskip
-
-\noindent{\bf SIGNIFICANCE}
-\smallskip
-
-Evolutionary dynamics of phenotypes in populations depend on how the traits are transmitted across generations, and how the environments that cause selection on the traits fluctuate over time.
-We show that under periodically fluctuating selection, a gene that increases the rate of vertical transmission is disfavored when the periods are short, but approaches an intermediate stable rate for longer periods.
-This stable rate differs markedly from the rate that maximizes the geometric mean fitness.
-The evolution of learning rules thus differs qualitatively from the evolution of genetically modified rules of genetic transmission.
-\vfil\break
-
-\noindent{\bf INTRODUCTION}
-\smallskip
-
-Cavalli-Sforza and Feldman (1981) distinguished two forms of non-parental phenotypic transmission in the context of cultural evolution.
-Horizontal transmission  occurs when a trait is passed between members of the same generation and is analogous to transmission of an infectious agent.
-Oblique transmission to offspring is from non-parental members of the parental generation.
-Evolution under either of these is expected to be more rapid than under purely vertical, i.e., parent-to-offspring, transmission (Lycett and Gowlett 2008; Bergstrom and Dugatkin 2012, Ch.
-19.4).
-
-Oblique transmission occurs via some mechanism of social learning, which may include imitation or active teaching.
-There has been an interesting debate over the past thirty years concerning the conditions under which social learning would have an advantage over individual learning or vertical (including genetic) transmission.
-This debate is usually couched in terms of the mode and tempo of environmental fluctuations that would affect fitness and hence evolution (Rogers 1988; Boyd and Richerson 1988, 1995; Feldman et al.
-1996; Wakano et al.
-2004: Aoki et al.
-2005; Wakano and Aoki 2007; Aoki and Feldman 2014).
-Mathematical analyses of models of competition between individual and social learning have generally shown that social learning has an advantage when the environment does not fluctuate too frequently.
-On the other hand, when the environmental changes are very frequent, individual learning is favored, while innate (genetic) determination of the trait does best when periods between environmental change are long on average.
-
-In some situations, oblique transmission of biological material is possible.
-In bacteria, phenotypes might be determined by heritable mobile genetic elements such as phages (Zinder and Lederberg 1952), plasmids (Lederberg and Tatum 1946), integrons (Mazel 2006), and transposons (Salyers et al.\ 2004).
-Similarly, some phenotypes are determined by genes that are commonly converted by uptake of foreign DNA, i.e., transformation (Milkman and Bridges 1990).
-In these cases, inheritance of a phenotype may combine vertical transmission from the parent cell, and oblique transmission from other cells, even if the latter did not originally evolve for that purpose (Redfield 1993).
-
-In some animals, transmission of microbes may occur during sharing or manipulation of food or other consumable resources during a social interaction.
-Although transmission of the microbiome in humans is likely to be mostly vertical (Rosenberg and Zilber-Rosenberg 2016), in other organisms there is multi-generational food sharing during which symbionts from the parental cohort may be transmitted obliquely to younger individuals (Theis et al.\ 2016).
-In such cases, fluctuations in the resource type or availability may have fitness effects that depend on features of the transmitted microbiome.
-This ecological perspective on community transmission is stressed by van Opstal and Bordenstein (2015), who emphasize the ``need to consider the relative roles of vertical and horizontal transmission of microbial communities.''
-
-Another perspective on the evolutionary consequences of fluctuating environments (and, as a result, fluctuating selection) derives from the phenomenon of phenotypic switching (Balaban et al.\ 2004; Kussell and Leibler 2005; Thattai and van Oudenaarden 2004; Salath\'e et al.\ 2009; Liberman et al.\ 2011; Gaal et al.\ 2010).
-In these studies, mutation causes the organism to switch phenotypes (usually treated as haploid genotypes), and the problem has usually been couched in terms of the optimal rate of mutation in models where the phenotypic fitnesses fluctuate over time.
-These models did not include social learning, and the evolution was regarded as a mode of bet-hedging against future environmental change.
-Optimal (that is, evolutionarily stable) mutation rates depend on many features of the fluctuations, for example,  degree of fitness symmetry, strength of selection, and variance in the period of fluctuation (Salath\'e et al.\ 2009).
-
-In a recent analysis of evolution under fluctuating selection, Xue and Leibler (2016) allowed an organism to absorb information about the distribution of possible environments by learning the phenotypes of members of its parental lineage from previous generations.
-They describe this as ``positive feedback that enhances the probability of the offspring to express the same phenotype as the parent.'' In this formulation there was ``reinforcement of the parent phenotype'' in an offspring, such as might occur through epigenetic inheritance.
-Although their analysis was not couched in terms of oblique and vertical transmission, as defined by Cavalli-Sforza and Feldman (1981), we have been stimulated by their analysis to develop a model in which oblique transmission, at a rate dependent on the trait frequency in the parental generation, occurs in addition to classical vertical transmission.
-We then ask how fluctuations in selection interact with the rate of oblique transmission to affect evolutionary dynamics and how the rate of oblique transmission itself might evolve.
-
-In our formulation, both the parental phenotype and the distribution of phenotypes in the whole population contribute to an offspring's phenotype.
-Using conventional modifier theory (Feldman and Liberman 1986), we show that in a symmetric cyclic selection regime with cycles of periods 1 or 2, an allele reducing the rate of vertical transmission is expected to increase in frequency when rare and in so doing to increase the mean fitness of the population.
-However, for cycles of greater length or period asymmetry, interesting non-monotonicities emerge, both in the uninvadable rate of vertical transmission, and the rate that maximizes the geometric time average of the population mean fitness, which we will refer to as the ``geometric mean fitness.'' We develop the models in very large populations with cyclic selection and with random fitnesses and also in the case where drift occurs via sampling from generation to generation in a finite population.
-
-\bigskip\bigskip
-
-\noindent{\bf THE MODEL}
-\smallskip
-
-Consider an infinite population whose members are characterized by their phenotype $\phi$, which can be of two types $\phi=A$ or $\phi=B$, with associated frequencies $x$ and $(1-x)$, respectively.
-We follow the evolution of $x$ over discrete non-overlapping generations.
-In each generation individuals are subject to  selection where the fitnesses of $A$ and $B$ are $w_A$ and $w_B$, respectively.
-
-An offspring inherits its phenotype from its parent via {\sl vertical transmission} with probability $\rho$ and from a random individual in the parental population via {\sl oblique transmission} with probability $(1-\rho)$.
-Therefore, given that the parent phenotype is $\phi$ and assuming {\sl uni-parental inheritance} (Zefferman 2016), the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is
-$$P(\phi' =A|\phi) =\left\{\matrix{(1-\rho)x+\rho &\hbox{\tenrm if}\ \phi=A\cr 
-\noalign{\smallskip}
- (1-\rho)x & \hbox{\tenrm if}\ \phi =B\cr}\right.\eqno(1)$$
- where $x=P(\phi=A)$ in the parent's generation before selection.
-
- Therefore the frequency $x'$ of  phenotype $A$ after one generation is given by the recursion equation
- $$x'= \rho{w_A\over \overline w}x +(1-\rho)x=     {w_A\over \overline w}x\bigl[(1-\rho)x +\rho\bigr] +{w_B\over \overline w}(1-x)\bigl[(1-\rho)x\bigr],\eqno(2)$$
- where $\overline w$ is the {\sl mean fitness}, namely
- $$\overline w=w_A x+w_B(1-x).\eqno(3)$$
- Equation (2) can be rewritten as
- $$x' =x\left[1+\rho(1-x){w_A-w_B\over \overline w}\right]    =x\cdot{x(1-\rho)(w_A-w_B) +\rho w_A +(1-\rho)w_B\over x(w_A-w_B) +w_B}.\eqno(4)$$
-
- In what follows we explore the evolution of the recursion (4), namely the equilibria and their stability properties, in the cases of {\sl constant environments} and {\sl changing environments}.
- \bigskip
- \bigskip
-
-\noindent{\bf CONSTANT ENVIRONMENT}
-\smallskip
-
-When the environment is constant, the fitness parameters $w_A$ and $w_B$ do not change between generations, and we have the following result.
-
-\proclaim Result 1. If $0<\rho\le 1$ and both $w_A$ and $w_B$ are positive with $w_A\ne w_B$, then fixation in the phenotype $A\ (B)$ is globally stable when $w_A>w_B$ $(w_A<w_B)$.\par
-
-\noindent{\it Proof}.
-If we rewrite (4) as $x'=x\cdot f(x)$, it can be seen that $f(1)=1$, and for $\rho>0$ and $0<x<1$
-$$\eqalign{
-&f(x)>1\quad\hbox{\tenrm when}\quad w_A>w_B,\cr
-&f(x)<1\quad\hbox{\tenrm when}\quad w_A<w_B.}\eqno(5)$$
-Hence, as $w_A>0$ and $w_B>0$, both fixations in $A$ or in $B$  ($x^*=1$ for fixation in $A$ and $x^*=0$ for fixation in $B$) are equilibrium points of (4).
-Moreover, if $x_t$ is the value of $x$ at the $t$-th generation $(t=0,1,2,\dots)$, from (4)  and (5) we have for any $0<x_0<1$ and all $t=0,1,2,\dots$,
-$$\eqalign{
-&x_{t+1}>x_t\quad\hbox{\tenrm when}\quad w_A>w_B,\cr
-&x_{t+1}<x_t\quad\hbox{\tenrm when}\quad w_A<w_B,}\eqno(6)$$
-and since $x^*=1$ or $x^*=0$ are the only equilibrium points, we have
-$$\eqalign{
-&\lim_{t\to\infty}x_t=1,\ \ \hbox{\tenrm for all}\ 0<x_0\le 1,\ \ \hbox{\tenrm when}\ w_A>w_B,\cr
-&\lim_{t\to\infty}x_t=0,\ \ \hbox{\tenrm for all}\ 0\le x_0< 1,\ \ \hbox{\tenrm when}\ w_A<w_B.}\eqno(7)$$
-Therefore fixation of the favored phenotype is globally stable.
-
-If $\rho=0$ or $w_A=w_B>0$, the phenotype frequencies do not change over time.
-With {\sl extreme selection},  for example $w_A=1$ and $w_B=0$ (or $w_B=1$ and $w_A=0$), recursion (4) reduces to
-$x'=(1-\rho)x+\rho$, and therefore 
-$x_t=1-(1-\rho)^t(1-x_0)$.
-Hence, for $\rho>0$,  $x_t\,\harr{}{t\to\infty}\,1$ for all initial values $0\le x_0\le 1$, and  the favored phenotype fixes.
-\bigskip
-\bigskip
-
-\noindent{\bf PERIODICALLY CHANGING ENVIRONMENT}
-\smallskip
-
-Suppose the environment is not constant but rather changes periodically, such that the favored phenotype changes after a fixed number of generations.
-Simple examples are $A1B1=ABABAB\dots$, in which the favored phenotype switches every generation, or $A2B1=AABAABAAB\dots$, where every two generations in which selection favors $A$ are followed by a single generation in which selection favors $B$.
-  In general, $AkBl$ denotes a selection regime in which the period is of $(k+l)$ generations with $k$ generations favoring phenotype $A$ followed by $l$ generations favoring $B$.
-
- Let $W$ be the fitness of the favored phenotype and $w$ be that of the other phenotype where $0<w<W$.
-Rewrite equation (4) as $x'=F_A(x)=xf_A(x)$ when $A$ is favored and $x'=F_B(x)=xf_B(x)$ when $B$ is favored.
-Then
-  $$\eqalign{
-  &f_A(x) = {x(1-\rho)(W-w) +\rho W +(1-\rho)w\over x(W-w) +w} = 1+\rho(1-x){W-w\over Wx +w(1-x)},\cr\noalign{\smallskip}
-  &f_B(x) = {x(1-\rho)(w-W) +\rho w +(1-\rho)W\over x(w-W) +W} = 1+\rho(1-x){w-W\over wx+W(1-x)}.}\eqno(8)$$
-
-  \noindent If $x_t$ denotes the frequency of  the phenotype $A$ at generation $t$ starting with $x_0$ initially, then as we are interested in the values of $x_t$ for $t=n(k+l)$ with $n=0,1,\dots$ at the end of complete periods, we can write
-  $$x_{(n+1)(k+l)}=F(x_{n(k+l)}),\quad n=0,1,2,\dots,\eqno(9)$$
-  where $F$ is the composed function
-  $$F=\underbrace{F_B\circ F_B \circ\cdots\circ F_B}_{l\ \hbox{\sevenrm times}} \circ\underbrace{F_A\circ F_A \circ\cdots\circ F_A}_{k\ \hbox{\sevenrm times}}.\eqno(10)$$
-  Clearly, since $F_A(0) =F_B(0) =0$ and $F_A(1) =F_B(1) =1$, both fixations in $A$ or in $B$ are equilibrium points.
-An interesting question is when these fixations are locally stable.
-We concentrate on $x^*=0$, the fixation of the phenotype $B$.
-As $x'=F_A(x) =xf_A(x)$ for $k$ generations and $x'=F_B(x) =xf_B(x)$ for $l$ generations, the linear approximation of $F(x)$ ``near'' $x=0$ is
-   $$F(x) \simeq\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^lx.\eqno(11)$$
-   Hence the local stability of $x^*=0$ is determined by the product $\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l$;  $x^*=0$  is locally stable if this product  is less than 1 and unstable if it is larger than 1.
-
-   From (8) we have
-   $$\eqalign{
-   &f_A(0) = 1+\rho{W-w\over w},\cr\noalign{\smallskip}
-   &f_B(0) = 1+\rho{w-W\over W}.}\eqno(12)$$
-   We start with the case $k=l$.
-
-   \proclaim Result 2. If $k=l$ and $0<w<W$ with $0<\rho<1$, fixation of $B$ is unstable.\par
-
-   \noindent {\it Proof.} The local stability of $x^*=0$, the fixation of $B$, is determined by the product
-   $$\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^k=\left[\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right)\right]^k.\eqno(13)$$
-   Observe that
-   $$\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right) = 1+\rho(1-\rho){\left(W-w\right)^2\over wW}>1.\eqno(14)$$
-   Since $0<\rho<1$ and $0<w<W$,  fixation on $B$ is unstable.
-  \bigskip
-
-\noindent{\it Conclusions}
-
-\item{1.}Since $k=l$ and the above result also holds  when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is also not stable.
-Thus neither phenotype can be lost, and there is a {\sl protected polymorphism} (Prout 1968).
-
-\item{2.}For general $k$, $l$, the condition for local stability of fixation in $A$ is
-$$\bigl[f_A(0)\bigr]^l \bigl[f_B(0)\bigr]^k <1, \eqno(15)$$
-and that of $B$ is
-$$\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l <1. \eqno(16)$$
-Therefore, following Result 2, 
-$$\bigl[f_A(0)\bigr]^{k+l}\bigl[f_B(0)\bigr]^{k+l}>1,\eqno(17)$$
-and it is impossible that both fixations are stable.
-Further, since by (12)  $f_A(0)>1$ and $0<f_B(0)<1$ when $0<w<W$, by choosing $k$ and $l$ appropriately, fixation on $A$ or fixation on $B$ (but not both) can be stable.
-In addition, we can have both fixations unstable giving the following result.
-
-\proclaim Result 3. With $0<\rho<1$ and $0<w<W$ in the case of $AkBl$ periodically changing environments, both fixations may be unstable, producing a protected polymorphism.\par
-
-\noindent{\it Proof}.
-Let $a=1+\rho{W-w\over w}$ and $b=1+\rho{w-W\over W}$, and our assumption entails  $a>1$ and $0<b<1$.
-Following (11), fixation in $B$ is not stable if $a^kb^l>1$, and similarly  fixation in $A$ is unstable if $a^lb^k>1$.
-Therefore both fixations are unstable if
-$$a^kb^l>1\quad\hbox{\tenrm and}\quad a^lb^k>1,\eqno(18)$$
-or equivalently if
-$$k\log a +l\log b>0\quad\hbox{\tenrm and}\quad l\log a+k\log b>0.\eqno(19)$$
-Now (19) holds if and only if
-$$k\,{\log(1/b)\over\log a} < l < k\,{\log a\over\log(1/b)}.\eqno(20)$$
-
-Inequalities (20) are consistent if and only if $\log(1/b)<\log a$, i.e., $ab>1$, which according to (14), is true.
-
-The local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations $A$ is favored $k$ times and $B$ is favored $l$ times, and not their order in the cycle.
-When neither fixation in $A$ or $B$ is stable, there is a protected polymorphism, and we expect to have one or more polymorphic equilibria.
-Supplementary Figure S1 illustrates the relationship between $k,l$ and $\rho$ that gives polymorphism of $A$ and $B$, or fixation, for different values of $s=W-w$.
-
-For the simple case of $A1B1$ periodically changing environment we have the following.
-
-\proclaim Result 4. In the case $A1B1$ with $0<\rho<1$ and $0<w<W$ the two fixations are unstable and there exists a unique stable polymorphism.
-
-\noindent{\it Proof}. Let $x$ be the initial frequency of $A$ and $x'$ its frequency  after one cycle of the $A1B1$ selection. Then $x' =F_B\bigl(F_A(x)\bigr)$ where by (8)
-$$\eqalign{
-&F_A(x)=x{x(1-\rho)(W-w) +\rho W+(1-\rho)w\over x(W-w) +w},\cr\noalign{\smallskip}
-&F_B(y) = y{y(1-\rho)(w-W) +\rho w+(1-\rho)W\over y(w-W)+W}.}\eqno(21)$$
-The equilibrium equation is $x=F_B\bigl(F_A(x)\bigr)$, which reduces to a fourth degree polynomial equation in $x$.
-Since the two fixations in $B$ and $A$ are equilibria corresponding to the two solutions $x=0$ and $x=1$, the other equilibria correspond to solutions of a quadratic equation $Q(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with
-$$\alpha_2=1,\quad \alpha_1={W+w\over (2-\rho)(W-w)}-1,\quad \alpha_0={-w\over (2-\rho)(W-w)}.\eqno(22)$$
-As $0<\rho<1$ and $0<w<W$ we have
-$$Q(0) = {-w\over (2-\rho)(W-w)}<0\quad\hbox{\tenrm and}\quad Q(1) ={W\over (2-\rho)(W-w)} >0.\eqno(23)$$
-Also, as $\alpha_2=1$ and $\alpha_0<0$, the quadratic equation $Q(x)=0$ has two real roots, one negative and one positive $x^*$ satisfying $0<x^*<1$.
-The latter determines a unique polymorphism.
-Let $H(x) =F_B\bigl(F_A(x)\bigr)$.
-Then
- $$H(0)=0,\quad H(x^*)=x^*,\quad H(1)=1.\eqno(24)$$
- Also 
- $$F'_A(x) = {x^2(1-\rho)\left(W-w\right)^2 +2xw(1-\rho)(W-w) +w\left[\rho W +(1-\rho)w\right]\over \left[x(W-w) +w\right]^2}\eqno(25)$$
- and
-  $$F'_B(x) = {x^2(1-\rho)\left(w-W\right)^2 +2xW(1-\rho)(w-W) +W\left[\rho w +(1-\rho)W\right]\over \left[x(w-W) +W\right]^2}.\eqno(26)$$
- From our assumptions on $\rho$, $w$, and $W$, we have $F'_A(x)>0$ for  $0\le x\le 1$.
-Observe that the numerator of $F'_B(x)$ is linear in $\rho$; its value when $\rho=1$ is $wW>0$, and when $\rho=0$ it is
-  $$x^2\left(w-W\right)^2 +2xW(w-W) +W^2 =\left[x(w-W) +W\right]^2 >0.\eqno(27)$$
-  Hence $F'_{B}(x)>0$ for all $0\le x\le 1$, and $H'(x) =F'_B\bigl(F_A(x)\bigr)F'_A(x)$ is positive when $0\le x\le 1$.
-Thus $H(x)$ is monotone increasing for $0\le x\le 1$; $H(x)>x$ for $0<x<x^*$, $H(x)<x$ for $x^*<x<1$.
-Starting from any initial value $0<x_0 <1$ we have  $x_t\,\harr{}{t\to\infty}\,x^*$.
-Supplementary Figure S2, panels A, C, and E, illustrate examples of how the frequency of $A$ changes over time in the $A1B1$ regime of cycling selection.
-
-For more general cyclic fitness regimes, the polynomial that gives the equilibria is of higher order, and it is conceivable that more than one stable polymorphism could exist for given values of $\rho$, $W$, and $w$.
-We have been able to show that when neither fixation in $A$ or $B$ is stable, in the $AkBk$ case this cannot occur.
-In fact we have the following.
-
-\proclaim Result 5. In the $AkBk$ selection regimes, if the fixations in $A$ and $B$ are locally unstable, a single stable polymorphic equilibrium exists.\par
-
-The proof of Result 5 is in Supplementary Material SP1.  Supplementary Figure S3A shows the stable equilibrium frequencies $x^*$ as a function of $\rho$, $W$, and $w$ in the $A1B1$ regime.
-For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplementary Figure S4 illustrates the convergence to a single stable polymorphism.
-
-We have not been able to prove that for selection regimes $AkBl$ with $l\ne k$ there is a single stable polymorphic equilibrium when the two fixations are unstable.
-However, the numerical examples in Supplementary Figure S1 (for $AkBl$) and in Figure 1 and Supplementary Figure S5 for the special case $A1B2$ all exhibit a single stable polymorphic equilibrium when fixations in $A$ and $B$ are unstable.
-These numerical results suggest that for $W>w>0$ and $0<\rho<1$ the high order polynomial that gives the equilibria has only a single root corresponding to a globally stable polymorphism.
-Supplementary Figure S6 shows that this is the case for the $A3B10$ regime.
-\medskip
-\centerline{[FIGURE 1 HERE]}  
- \bigskip
-
- \noindent{\bf RANDOMLY CHANGING ENVIRONMENT}
- \smallskip
-
- We now consider the  case where the environment changes according to a stochastic process.
-Without loss of generality, assume that the fitness parameters at generation $t$ ($t=0,1,2,\dots$) are $1+s_t$ for phenotype $A$ and $1$ for phenotype $B$, where the random variables $s_t$ for $t=0,1,2,\dots$ are independent and identically distributed.
-Also we assume that there are positive constants $C$ and $D$ such that $P(-1+C<s_t <D)=1$.
-
- Corresponding to (4), with $w_A=1+s_t$, $w_B=1$ the recursion equation is
- $$x_{t+1}=x_t {1+\rho s_t +x_t(1-\rho)s_t\over 1+x_ts_t}\quad t=0,1,2,\dots\; .\eqno(28)$$
-
-\noindent As $\{x_t\}$ for $t=0,1,2,\dots$ is a sequence of random variables, the notion of stability of the two fixation states needs clarification.
-Following Karlin and Lieberman (1974) and Karlin and Liberman (1975) we make the following definition.
-
- \noindent{\it Definition}: ``stochastic local stability''.
- A constant equilibrium state $x^*$ is said to be {\sl stochastically locally stable} if for any $\var>0$ there exists a $\delta>0$ such that $|x_0-x^*|<\delta$ implies
- $$P\left(\lim_{t\to\infty}x_t =x^*\right)\ge 1-\var.\eqno(29)$$
- Thus stochastic local stability holds for $x^*$ provided for any initial $x_0$ sufficiently near $x^*$ the process $x_t$ converges to $x^*$ with high probability.
-
- In our case there are two constant equilibria $x^*=0$ and $x^*=1$ corresponding to fixation in $B$ and $A$, respectively.
-We can characterize the stochastic local stability of these fixations as follows.
-
- \proclaim Result 6. Suppose $E\left[\log (1+\rho s_t)\right]>0$. Then $x^*=0$, the fixation of phenotype $B$, is not stochastically locally stable. In fact $P\left(\lim_{t\to\infty}x_t=0\right)=0$.
-
-  \proclaim Result 7.
-Suppose $E[\log(1+\rho s_t)]<0$.
-Then $x^*=0$, the fixation of phenotype $B$, is stochastically locally stable.
-In particular, if $E(s_t)\le 0$, $x^*=0$ is stochastically locally stable.
-
- The proofs of Results 6 and  7 can be found in Supplementary Materials SP2 and SP3.
-
- Using the general notation for the fitness parameters $w_A$ and $w_B$, the stochastic local stability of the $B$ fixation is determined by the sign of $E\left[\log\left(1-\rho+\rho{w_A\over w_B}\right)\right]$ and that of the $A$ fixation by the sign of $E\left[\log\left(1-\rho+\rho{w_B\over w_A}\right)\right]$.
-For example, if the  sign of the first  is negative, fixation in $B$ is stochastically locally stable, and when it is positive, with probability one convergence to fixation in phenotype $B$ does not occur.
-It is also true  that if $E(w_A/w_B)\le 1$, then fixation of $B$ is stochastically locally stable.
-Following (14), for all realizations of $w_A$ and $w_B$,
- $$\log\left(1-\rho +\rho{w_A\over w_B}\right) +\log\left(1-\rho +\rho{w_B\over w_A}\right)>0.\eqno(30)$$
- Therefore, as in the case of periodically changing environments $AkBl$, it is impossible that both fixations are simultaneously stochastically locally stable.
-It is possible, however, that neither fixation is  stochastically locally stable, in which case we expect the population to converge to a polymorphic distribution.
-Figure 2 illustrates how the properties of $s_t$ in recursion (28) affect the frequency of phenotype $A$, and in particular the stochastic local stability of fixation in phenotype $B$.
-Figure 3 shows the dynamics of the frequency of $A$ in a case where $w_A$ and $w_B$ are identically distributed and independent; in this case the expectation of the stationary distribution is ${1\over 2}$, and its variance increases as $\rho$ increases.
-\medskip
-\centerline{[FIGURES 2, 3 HERE]}  
- \bigskip
-
- \noindent{\bf EVOLUTIONARY STABILITY OF OBLIQUE TRANSMISSION}
- \smallskip
-
- An interesting question concerns the evolution of oblique transmission itself.
-For example, is there an evolutionarily stable rate of oblique transmission?
-To answer this question we use a modifier model, for which we suppose that the vertical transmission rate is controlled by a genetic locus with two possible alleles $m$ and $M$.
-Let the vertical transmission rates determined by $m$ and $M$ be $\rho$ and $P$, respectively.
-Thus there are four pheno-genotypes $mA$, $mB$, $MA$, $MB$ whose frequencies at a given generation are denoted by $x_1$, $x_2$, $x_3$, $x_4$, respectively.
-As the fitnesses are determined by the two phenotypes $A$ and $B$, and the modifier locus is selectively neutral, we have the following table.
- $$\matrix{
- \hbox{\tenrm pheno-genotype}&mA&mB&MA&MB\cr\noalign{\smallskip}
- \hbox{\tenrm frequency}&x_1&x_2&x_3&x_4\cr\noalign{\smallskip}
- \hbox{\tenrm fitness}&w_A&w_B&w_A&w_B\cr\noalign{\smallskip}
- \hbox{\tenrm vertical transmission rate}&\rho&\rho&P&P\cr}\eqno(31)$$
- Following the rationale leading to equation (2), the next generation pheno-genotype frequencies $x'_1$, $x'_2$, $x'_3$, $x'_4$ are
- $$\eqalign{
- \overline wx'_1 &= w_Ax_1\bigl[(1-\rho)(x_1+x_3) +\rho\bigr] +w_Bx_2(1-\rho)(x_1+x_3)\cr
- \overline wx'_2 &= w_Ax_1(1-\rho)(x_2+x_4) +w_Bx_2\bigl[(1-\rho)(x_2+x_4)+\rho\bigr]\cr 
-  \overline wx'_3 &= w_Ax_3\bigl[(1-P)(x_1+x_3) +P\bigr] +w_Bx_4(1-P)(x_1+x_3)\cr
-\overline wx'_4 &= w_Ax_3(1-P)(x_2+x_4) +w_Bx_4\bigl[(1-P)(x_2+x_4)+P\bigr], 
-}\eqno(32)$$
-with $\overline w$, the mean fitness, given by
-$$\overline w =w_A(x_1+x_3) +w_B(x_2+x_4).\eqno(33)$$
-Note that under these assumptions the $M/m$ locus and the $A/B$ phenotypic dichotomy do not undergo anything analogous to recombination, which might be introduced if $A/B$ phenotypes were viewed as haploid genetic variants.
-
-Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} (e.g., Feldman and Liberman 1986; Altenberg et al.\ 2017) to invasion by allele $M$.
-Constant environment always leads to fixation of the favored type, independent of $\rho$.
-We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$ (but see Supplementary Material SP6 for computational analysis of the general $AkBl$ case).
-Specifically, from (32) with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
-$$\underline x''=T_2(T_1\underline x),\eqno(34)$$
-where $\underline x'=T_1\underline x$ is given by (32) with $w_A=W$, $w_B=w$, and $\underline x''=T_2\underline x'$ is given by (32) with $w_A=w$, $w_B=W$.
-Here $\underline x$, $\underline x'$, $\underline x''$ are the frequency vectors.
-
-For the $A1B1$ case, when only the $m$ allele is present with associated rate $\rho$,  $0<\rho<1$, and $0<w<W$, a unique stable equilibrium $\underline x^* =(x_1^*,1-x_1^*,0,0)$ exists.
-$x_1^*$ is the only positive root of the quadratic equation $Q(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with $\alpha_2,\alpha_1,\alpha_0$ specified in equation (22).
-Solving $Q(x)=0$ gives
-$$x_1^* ={1\over 2} -{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}\over 2\cdot(2-\rho)(W-w)},\eqno(35)$$
- and it can be seen that
- $${\sqrt{Ww}-w\over W-w}<x_1^* <{1\over 2}.\eqno(36)$$
-
- The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ near $x^*$, which is derived from (32) and given by
- $$\overline w^*\cl_1 =\left[\matrix{W\bigl[(1-P)x_1^* +P\bigr] & w(1-P)x_1^*\cr\noalign{\medskip}
- W(1-P)x_2^* & w\bigl[(1-P)x_2^* +P\bigr]\cr}\right]\eqno(37)$$
- and
- $$\overline w^{**}\cl_2 =\left[\matrix{w\bigl[(1-P)x_1^{**} +P\bigr] & W(1-P)x_1^{**}\cr\noalign{\medskip}
- w(1-P)x_2^{**} & W\bigl[(1-P)x_2^{**} +P\bigr]\cr}\right],\eqno(38)$$
-where $\underline x^{**} =T_1\underline x^*$, $x_1^{**}=x_2^*=1-x_1^*$, $x_2^{**}=x_1^*$,  and
-$$\overline w^* =Wx_1^* +wx_2^*,\qquad \overline w^{**} =wx_1^{**} +Wx_2^{**}.\eqno(39)$$
-Due to the symmetry between the two phenotypes $A$ and $B$ in the $A1B1$ case, we have $x_1^{**} =x_2^*$ and $x_2^{**} =x_1^*$ so that $\overline w^{**} =\overline w^*$ and in fact
-$$\overline w^*\cl_2 = \left[\matrix{w\bigl[(1-P)x_2^* +P\bigr] & W(1-P)x_2^*\cr\noalign{\medskip}
- w(1-P)x_1^* & W\bigl[(1-P)x_1^* +P\bigr]\cr}\right].\eqno(40)$$
- Note that as $\underline x^* =T_2(T_1\underline x^*)$ with $x_3^*=x_4^*=0$, from (37) and (38) with $P=\rho$ we have
- $$\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right]= \cl_2\cdot\cl_1\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right] =\cl\left[\matrix{x_1^*\cr\noalign{\medskip} x_2^*\cr}\right].\eqno(41)$$
- Hence when $P=\rho$ one of the eigenvalues of $\cl$ is 1.
-
- In general $\cl=\cl_2\cdot\cl_1$, and using (37) and (40) we have
- $$\eqalign{
- &\left(\overline w^*\right)^2\cl =\cr\noalign{\smallskip}
- &\left[\matrix{Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[w(1-P)x_1^*\bigr]^2 & W(1-P)\bigl[P+x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr]\cr
- \noalign{\medskip}
- w(1-P)\bigl[1-x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr] & Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[W(1-P)x_2^*\bigr]^2
- }\right].}\eqno(42)$$
- The external stability of $\underline x^*$ is determined by the eigenvalues of $\cl$, namely the roots of its characteristic polynomial $R(\lambda) =\det(\cl-\lambda\ci)$, with $\ci$ the $2\times 2$ identity matrix.
-From (42), $R(\lambda)=a_2\lambda^2 +a_1\lambda +a_0$, where
- $$a_0={P^2W^2w^2\over\left(\overline w^*\right)^4},\qquad a_1=-{2PWw +\left(1-P\right)^2\left[Wx_2^* +wx_1^*\right]^2\over \left(\overline w^*\right)^2},\qquad a_2=1.\eqno(43)$$
-
- \smallskip
-
-We then have the following result, which is proved in Supplementary Material SP4.
-
- \proclaim Result 8. $\cl$ has two positive eigenvalues and\hfil\break
-  {\hglue.82truein}(i) when $P>\rho$ the two eigenvalues are less than $1$,\hfil\break
-  {\hglue.79truein}(ii) when $P<\rho$ the largest eigenvalue is larger than $1$,\hfil\break
-  {\hglue.76truein}(iii) when $P=\rho$, the largest eigenvalue is $1$.\par
-
- We conclude that in the $A1B1$ selection regime an allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$.
-Thus in this case, evolution tends to reduce vertical transmission, and hence increase the rate of oblique transmission,  and there is a {\sl reduction principle} for the rate of vertical transmission (Feldman and Liberman 1986; Altenberg et al.\ 2017).
-The evolutionary dynamics of the reduction in $\rho$ under the $A1B1$ cycling regime are shown in Figure 4, which also illustrates the change in phenotype frequencies over time.
-
- In the case of identically distributed random fitnesses $w_A$ and $w_B$, Figure 5 shows  an example of the success of modifiers that reduce $\rho$.
-We have not, however, been able to prove that there is a reduction principle for this class of fluctuating fitnesses.
-
-  \medskip
-\centerline{[FIGURES 4, 5]}
-
-Values of $\rho^*$ for some $AkBl$ examples (see Supplementary Material SP7 and  Figure S10 for analytical details) are recorded in Table 1 for different values of $w$ relative to $W=1$.
-Interestingly, for the $A1B2$ regime, with $w=0.1$, the evolutionarily stable value of $\rho$ is substantially greater than zero, as it is in the $A3B10$ and $A5B30$ regimes.
-$AkBk$ results are plotted in Figure 6B.
-In the $A2B2$ regime, $\rho^* =0$ and there is reduction of vertical transmission for all selection values tested.
-However, for $AkBk$ regimes with $k>2$, we find $\rho^*\ne 0$, and depending on $w$, $\rho^*$ can be as high as 0.95.
-In Table 1, blank values for $\rho^*$ indicate that our method was numerically unstable  and that a precise value for $\rho^*$ could not be obtained.
-This is why, in Figure 6B, no $\rho^*$ points are shown for $AkBk$ with $k>19$.
-In Table 1, the word ``fixation'' indicates that fixation of $B$ occurs, at which point there can be no effect of modification of $\rho$; $\rho^*$ cannot be calculated in such cases.
-\bigskip\bigskip
-
-\noindent{\bf GEOMETRIC MEAN FITNESS AND RATE OF VERTICAL TRANSMISSION}
-\smallskip
-
-Under fluctuating selection, the geometric mean fitness of genotypes has been shown to determine their evolutionary dynamics (Haldane and Jayakar  1963; Karlin and Lieberman 1974; Wakano et al.\ 2004).
-For the evolution of mutation rates that are controlled by genetic modifiers, the stable mutation rate and the mutation rate that maximizes the geometric mean fitness of the population appear to be the same when the period of environmental fluctuation is low enough (Liberman et al.\ 2011).
-We can ask the same question here: is the stable rate $\rho^*$ the same as the rate $\hat\rho$ that maximizes the equilibrium value of the geometric mean fitness under fluctuating selection.
-First consider the $A1B1$ selection regime.
-We have the following result.
-
-\proclaim Result 9. If $W>w$ and $0\le\rho\le 1$, then the mean fitness at the stable equilibrium in the $A1B1$ environment is a decreasing function of $\rho$.\par
-
-\noindent{\it Proof.} In $A1B1$ the stable frequency of phenotype $A$ is by (35)
-$$x^* ={1\over 2} -{W+w-Z\over 2(2-\rho)(W-w)},\eqno(44)$$
-where $Z =\sqrt{(1-\rho)^2(W-w)^2 +4Ww}>0$.
-The mean fitness at the stable equilibrium  is\hfil\break $\overline w^*\cdot\overline w^{**}$, and as $\overline w^* =\overline w^{**}$  by equation (39), this allows us to reduce the problem to properties of $\overline w^*$.
-Now since $W>w$,  $\overline w^*$ is an increasing linear function of $x^*$:
-$$\overline w^* =x^*W +(1-x^*)w = x^*(W-w) +w.\eqno(45)$$
-Thus $\overline w^*$ is decreasing in $\rho$ if $dx^*/d\rho$ is negative. Using (44),
-$${dx^*\over d\rho} =- {(1-\rho)(W-w)\over 2(2-\rho)Z} -{W+w-Z\over 2(2-\rho)(W-w)} = {x^*-{1\over 2}\over 2-\rho} -{(1-\rho)(W-w)\over 2(2-\rho)Z}.\eqno(46)$$
-From (36) $0<x^*\le{1\over 2}$ and therefore $dx^*/d\rho <0$, which completes the proof.
-
-  Figure 4 illustrates the decrease over time of the geometric mean fitness with increasing $\rho$ at a polymorphic equilibrium in the $A1B1$ regime.
-The values of $\hat\rho$ and $\rho^*$ are the same in $A1B1$ and $A2B2$ regimes, namely both are zero.
-Figure 1B shows the geometric mean fitness in the $A1B2$ regime, and we see that for small values of $w$, this mean fitness decreases as $\rho$ increases.
-At $w=0.1$, Table 1 shows that $\hat\rho$ is 0.00065, whereas $\rho^*=0.821$.
-In all $AkBk$ regimes that we tested with $w=0.9$, the value of $\hat\rho$ was also zero, substantially different from the values of $\rho^*$, as shown in Figures 6A and 6B.
-Also in Figure 6A, we see that with $w=0.1$, $\hat\rho$ changes from zero to positive in the $AkBk$ regimes with $k \ge 12$, while with $w=0.5$, the change occurs at $k=31$.
- In Figure 6, with $w=0.1$, $\hat\rho$ is between 0.15 and 0.24 for $12\le k\le 50$, while with $w=0.5$, $\hat\rho$ exceeds 0.2 for $31\le k\le 50$.
-More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Table 1, Supplementary Figures S7 and S8, and Supplemental Material SP7.
-\medskip
-\centerline{[TABLE 1 AND FIGURE 6 HERE]}  
- \bigskip
-
- \noindent{\bf FINITE POPULATION SIZE}
- \smallskip
-
- In order to include the effect of random drift due to finite population in the above deterministic model, we use the Wright-Fisher model.
-Let $X_t$ denote the number of individuals with phenotype $A$ in a population of fixed size $N$ at the $t$-th generation, and suppose $X_t=Nx$.
-Also, let $x'$ represent the frequency of the phenotype $A$ in the infinite population model in the next generation, namely (see (2)),
- $$x'={w_Ax\over\overline w}\rho +(1-\rho)x.\eqno(47)$$
- Then, according to the Wright-Fisher model (Ewens 2004), $X_{t+1}$, the number of individuals of phenotype $A$ at generation $(t+1)$, is determined by the probability
- $$P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\matrix{N\cr j\cr}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},\eqno(48)$$
- for $j=0,1,2,\dots,N$.
- Thus the fluctuations in the numbers of phenotypes $A$ and $B$ in the  population of size $N$ are generated by the Wright-Fisher Markov chain  where, given that $X_t=Nx$, $X_{t+1}$ has a binomial distribution with parameters $(N,x')$.
-
- This Markov chain  has two absorbing states, $X_t =N$ and $X_t=0$, corresponding to the two fixations in $A$ and $B$, respectively, and we are interested in the fixation probabilities and the time to fixation of these two absorbing states as functions of the initial frequency $x$ and also of $\rho$, $w_A$, and $w_B$.
- To these ends we use a diffusion approximation  to the process $\{X_t\}$.
-Specifically, we compute the mean $\mu(x)$ and the variance $\sigma^2(x)$ of the change in one generation  in the frequency of phenotype $A$ given that at the beginning of the generation $X_t=Nx$.
-
- To compute $\mu(x)$, observe that by (47)   
- $$\eqalignno{
- x' -x &={w_Ax\over\overline w}\rho +(1-\rho)x -x =\rho x\left[{w_A\over \overline w}-1\right]\cr
- &= \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},&(49)\cr}$$
- since $\overline w =w_A x +w_B(1-x)$.
- For the diffusion approximation, it is essential that the differential selection does not have a large effect per individual in each time period  $\Delta t$ $\bigl(\Delta t\simeq{1\over N}\bigr)$.
-That is, we assume
- $$w_A -w_B ={s\over N}.\eqno(50)$$
- Then
- $$x'-x\simeq {1\over N}\rho sx(1-x)\eqno(51)$$
- up to terms of order small than ${1\over N}$. Since one generation corresponds to $\Delta t\simeq {1\over N}$, we  conclude that
- $$\mu(x) =\rho sx(1-x);\quad 0\le x\le 1.\eqno(52)$$
- In the same way, we can compute $\sigma^2(x)$, which is $x(1-x)$. These values of $\mu(x)$ and $\sigma^2(x)$ allow us to compute $u(x)$, namely
-  $$u(x) ={1-e^{-2\rho sx}\over 1-e^{-2\rho s}}.\eqno(53)$$
- The expected time to fixation in $A$ starting from a initial frequency of $x$ is is given by
- $$T(x)={1-u(x)\over\rho s}\int_0^x{e^{2\rho s\xi}-1\over \xi(1-\xi)}d\xi +{u(x)\over\rho s}\int_x^1{1-e^{-2\rho s(1-\xi)}\over \xi(1-\xi)}d\xi,\eqno(54)$$
- where $u(x)$ is given in (53), and in generations, $T(x)$ is multiplied by $N$.
-  Unfortunately the integrals in (54) cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ (Ewens 2004, p.\ 160), and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.
-
-  It is important to note that the fixation probability $u(x)$ is a monotone increasing function of $\rho$ when $s>0$. In fact we have the following result.
-
-  \proclaim Result 10. When $s>0$ so that the phenotype $A$ is favored, the fixation probability $u(x)$ is monotone increasing in $\rho$.\par
-
- The proof of Result 10 is in Supplementary Material SP5.  
-  Supplementary Figure S9 compares the fixation probability and time to fixation derived numerically from  simulating the Wright-Fisher Markov chain with the diffusion-derived values of $u(x)$ and $T(x)$.
-The fit is seen to be very good.
-Note that when $N$ is large, the Wright-Fisher model exhibits persistent fluctuation around the deterministic expectation, as shown by the orange diagram in Supplemental Figure S2.
-
- We can also develop a diffusion approximation for the case of a cycling environment.
-Suppose that selection changes in cycles of length $n$ such that within the cycle the fitness parameters are $w_A^t$, $w_B^t$ for $t=1,2,\dots,n$.
-Also let
- $${1\over N}s_t \simeq w_A^t -w_B^t,\qquad S_t =\sum_{i=1}^t s_i,\qquad t=1,2,\dots,n.\eqno(55)$$
- Following Karlin and Levikson (1974) we have the following result.
-
- \proclaim Result 11.
-The mean $\mu(x)$ and variance $\sigma^2(x)$ of the change in the frequency of $A$ in one generation for the diffusion approximation in the case of a cycling environment $AkBl$, where $k+l=n$, are
- $$\eqalign{
- \mu(x) &= \rho S_n x(1-x)\cr \sigma^2(x) &= nx(1-x).}\eqno(56)$$
-
-\noindent The proof of Result 11, based on induction on $n$, is given in Supplemental Materials SP6.
-
- Using the moments in eqs.\ (56), the fixation probability $u(x)$ and the expected time $T(x)$ to fixation from an initial frequency of $x$ can be computed where $s$ is replaced by $s_n/n$. We find
- $$u(x) ={1-e^{-2\rho{S_n\over n}x}\over 1-e^{-2\rho{S_n\over n}}},\eqno(57)$$
- and $T(x)$ can be computed similarly.
-
- In the  case of the $AkBl$ cycling environment, we write $n=k+l$, and if $w_A=W$, $w_B=w$ for $k$ generations, and $w_A=w$, $w_B=W$ for $l$ generations, we have
- $$S_n=S_{k+l} =N(k-l)(W-w).\eqno(58)$$
- Figure 7 shows an example of how $(k-l)$, $\rho$, and $(W-w)$, which enter the formula for $u(x)$ in (57), interact to affect fixation probabilities.
-More examples are illustrated in Supplementary Figure S11.
-
- \medskip
-\centerline{[FIGURE 7 HERE]}  
-  \bigskip
-
- \noindent{\bf DISCUSSION}
- \smallskip
-
-Non-chromosomal modes of phenotypic transmission are receiving increasing attention (Whiten et al.
-2017; Jaenisch and Bird 2003; Allis and Jenuwein 2016), especially with respect to their potential role in adaptation and maintenance of diversity (Rivoire and Leibler 2014).
-Here we have focused on a dichotomous phenotype transmitted through a combination of parental and non-parental transmission.
-In addition to the roles that these transmission modes play in the dynamics of phenotypic diversity in large and small populations, we have also investigated a genetic model for the evolution of the transmission mode itself.
-
-Our model differs markedly from that of Xue and Leibler (2016), who took the individual phenotypic distribution (i.e., the probability that an individual develops one of a set of phenotypes) to be the inherited trait.
-In our model, the transmitted trait is the phenotype itself.
-Thus, with two phenotypic states $A$ and $B$, we track the frequency $x$ of $A$, whereas Xue and Leibler focus on the dynamics of the per-individual probability $\pi_A$ of learning the phenotype $A$.
-One interpretation of our model is as a mean-value approximation to the model of Xue and Leibler, where $x$, the state in our model, is the average of the population distribution of individual phenotype probabilities.
-
-In a constant environment, the higher the vertical transmission rate $\rho$, the faster is the approach to fixation of the favored phenotype: $A$ if $w_A>w_B$ or $B$ if $w_B>w_A$.
-Here $1-\rho$, the oblique transmission rate, represents the added chance that an offspring becomes $A$ by learning from the parent's population after learning from its parents who have undergone selection (eq.\ 2).
-This simple phenotypic model does not allow a polymorphism to be achieved in a constant environment, but with more oblique transmission, approach to fixation is retarded.
-
- With fluctuating environments, the dynamics of the phenotype frequencies are, in general, much more complicated.
-In particular, with deterministically cycling symmetric fitness values (the $AkBl$ model), it is impossible for fixation in $A$ and $B$ to both be stable.
-If $k=l$, for example, neither fixation is stable, and there is a single stable polymorphic equilibrium (with phenotypes $A$ and $B$ present; see Result 5).
-In the $A1B1$ case this  polymorphism is globally stable.
-In the $AkBl$ case, bounds on $l/k$ that determine the instability of both fixations, and hence the protection of polymorphism are given by inequalities (20), which depend on both the fitness differences and the rate $\rho$ of vertical transmission.
-We conjecture that with $k\ne l$ there is a unique stable polymorphism when both fixations are unstable.
-
- In deterministic one-locus, two-allele diploid population genetic models with cycling fitness regimes, Haldane and Jayakar (1963) first showed the relevance of the geometric mean of genotypic fitnesses (cf.\ inequalities (16) and (17)) for the maintenance (or loss) of polymorphism.
-However, with equal homozygote fitnesses, which alternated in strength as a two-generation cycle (cf.\ $A1B1$), Karlin and Liberman (1975) extended the Haldane-Jayakar results and found conditions under which both allelic fixations and polymorphic equilibrium could all be stable, with the evolution depending on initial allele frequencies, as well as the homozygote fitness differences between alternate generations.
-Our haploid model appears not to produce such dependence on the initial conditions.
-
- When the fitnesses $w_A$ and $w_B$ are treated as random variables, rather than varying cyclically, stochastic local stability is the appropriate analog to local stability in the case of cyclic fitness variation.
-While fixations in phenotypes $A$ and $B$ cannot both be stochastically stable in this case, both may be unstable and a polymorphic distribution may result.
-The variance of this distribution is greater for larger values of $\rho$.
-This is because the  stochastic local stability conditions involve $E\bigl\{\log[1-\rho+\rho(w_A/w_B)]\bigr\}$ and the effect of the variance of $(w_A/w_B)$ will clearly increase as $\rho$ increases.
-In the finite population case, a greater level of vertical transmission makes selection more effective, increasing the probability $u(x)$ of fixation and reducing the expected time to fixation.
-
- We have shown that in the $A1B1$ case, the rate of vertical transmission tends to decrease when it is under the control of a genetic modifier.
-From numerical iteration, it appears that this is also true in the random selection case, when the fitnesses of $A$ and $B$ are identically distributed and independent between generations.
-However, for $AkBl$ selection regimes more complicated than $A1B1$, evolution of a modifier of vertical transmission is not straightforward.
-While reduction of $\rho$ occurs in the $A2B2$ regime, the uninvadable value $\rho^*$ is not zero for all the fitness values explored in $AkBk$ regimes with $k>2$ (Figure 6B, Table 1).
-In fact, $\rho^*$ increases sharply as $k$ increases beyond $k=2$.
-This is an unusual scenario for genetic modifiers, although it must be noted that a modifier of $\rho$ is not neutral; it affects primary selection, while neutral  modifiers of recombination, mutation, and migration affect induced or secondary selection.
-
- The dependence of the modifier dynamics on the strength of selection (that is $w$ when $W=1$) is complicated by the approach of the system to fixation.
-When the phenotype frequencies become exceedingly small, dependence of the dynamics of the modifier of $\rho$ becomes extremely difficult to detect due to numerical instability; this is especially true for larger values of $k$ in $AkBk$ regimes when $w$ is small (Table 1, Figure 6B with $k\ge 19$).
-
- Figure 6A (see also Table 1) shows that the value $\hat\rho$ that maximizes the geometric mean fitness is the same as the evolutionarily stable value $\rho^*$ in the $A1B1$ and $A2B2$ selection regimes.
-For $AkBk$ regimes with $k>2$, our numerical analysis shows the $\hat\rho$ depends strongly on the strength of selection; i.e., the value of $w$ relative to $W=1$.
-For $AkBl$ regimes with $w=0.1$, the difference between $\hat\rho$ and $\rho^*$ is seen even with the $A1B2$ environment.
-For $AkBk$ regimes and $w=0.9$, we find $\hat\rho=0$, while $\rho^*$ is close to $0.9$.
-For larger values of $k$, $\hat\rho$ lies between 0.15 and 0.25, while $\rho^*$ remains above 0.8 and can reach 0.99 for very large $k$.
-Comparing Figure 6A with the asymptotic growth rate of Xue and Leibler (2016), whose parameter $\eta$ is the rate at which an individual learns from its parental lineage, there is a similarity for our curves for $w=0.1$ and $0.5$ with their curve in the $AkBk$ environment.
-They show the AGR decreasing with $\eta$ in the $AkBk$ regime for small $k$, but larger values of $k$ entail that the AGR has a maximum for an intermediate value of $\eta$.
-
- Although the models of Xue and Leibler (2016) and that analyzed here both incorporate parental and non-parental transmission, they do so in qualitatively different ways.
-The model treated in the present paper is squarely in the tradition of gene-culture coevolutionary theory, together with modifier theory from population genetics.
-The different findings from the two classes of models are interesting and suggest that further exploration of the overlaps and discrepancies between the two approaches would be worthwhile.
-
-\vfil
- \break
-
- \noindent{\bf References}
-
-\ref Allis, C. D., and T. Jenuwein. 2016. The molecular hallmarks of epigenetic control. {\it Nat.\ Rev.\ Genet.} {\bf 17}: 487--500.
-
-\ref Altenberg, L.,  U.
-Liberman, and M.
-W.
-Feldman.
-2017.
-A unified reduction principle for the evolution of mutation, migration, and recombination.
-{\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 114}: E2392--E2400.
-
-\ref Aoki, K., and M. W. Feldman. 2014. Evolution of learning strategies in temporally and spatially variable environments: a review of theory. {\it Theor.\ Pop.\ Biol.} {\bf 91}: 3--19.
-
-\ref Aoki, K., J.Y. Wakano, and M. W. Feldman. 2005. The emergence of social learning in
-a temporally changing environment: a theoretical model. {\it
-Curr.\ Anthropol.} {\bf 46}: 334--340.
-
-\ref Balaban, N. Q., J. Merrin, R. Chait, L. Kowalik, and S. Leibler. 2004. Bacterial persistence as a phenotypic switch. {\it Science} {\bf 305}: 1622--1625.
-
-\ref Bergstrom, C. T., and L. A. Dugatkin. 2012. {\it Evolution}. New York: W. W. Norton \& Co., Inc.
-
-\ref Boyd, R., and P.
-J.
-Richerson.
-1988.
-An evolutionary model of social learning: the effect of spatial and temporal variation.
-Pp.
-29--48 {\it in} Zentall, T., and B.G.
-Galef (eds.) {\it Social Learning}.
-Hillsdale, NJ: Lawrence Erlbaum Associates, Publishers.
-
-\ref Boyd, R., and P. J. Richerson. 1995. Why does culture increase human adaptability? {\it Ethol.\ Sociobiol.} {\bf 16}: 125--143.
-
-\ref Carja, O., U.
-Liberman, and  M.
-W.
-Feldman.
-2014.
-Evolution in changing environments: modifiers of mutation, recombination, and migration.
-{\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 111}: 17935--17940.
-
-\ref Cavalli-Sforza, L. L. and M. W. Feldman.  1981. {\it Cultural
-Transmission and Evolution:  A Quantitative Approach}.  Princeton, N.J.: Princeton University
-Press.
-
-\ref Ewens, W. 2004. {\it Mathematical Population Genetics}, 2nd Edition. New York: Springer.
-
-\ref Feldman, M. W. and U. Liberman.  1986. An evolutionary reduction
-principle for genetic modifiers.  {\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 83}:
-4824--4827.
-
-\ref Feldman, M. W., K. Aoki, and J. Kumm. 1996. Individual
-versus social learning:  Evolutionary analysis in a fluctuating environment.   {\it
-Anthropological Science} {\bf 104}: 209--231.
-
-\ref Ga\'al, B., J. W. Pitchford, and A. J. Wood. 2010. Exact results for the evolution of stochastic switching in variable asymmetric environments. {\it Genetics} {\bf 184}: 1113--1119.
-
-\ref Haldane, J. B. S., and S. D. Jayakar. 1963. Polymorphism due to selection of varying direction. {\it J. Genet.} {\bf 58}: 237--242.
-
-\ref Jaenisch, R., and A. Bird. 2003. Epigenetic regulation of gene expression: how the genome integrates intrinsic and environmental signals. {\it Nat.\ Genet.} {\bf 33}: 245--254.
-
-\ref Karlin, S., and B. Levikson. 1974. Temporal fluctuation in selection intensities: case of small population size. {\it Theor.\ Popul.\ Biol.} {\bf 6}: 383--412.
-
-\ref Karlin, S., and U. Lieberman. 1974. Random temporal variation in selection intensities: case of large population size. {Theor.\ Popul. Biol.} {\bf 6}: 355--382.
-
-\ref Karlin, S., and U. Liberman. 1975. Random temporal variation in selection intensities: one-locus two-allele model. {\it J. Math.\ Biol.} {\bf 2}: 1--17.
-
-\ref Lederberg, J., and E. L. Tatum. 1946. Gene recombination in {\it Escherichia coli}. {\it Nature} {\bf 158}: 558.
-
-\ref Liberman, U., J. Van Cleve, and M. W. Feldman. 2011. On the evolution of mutation in changing environments: recombination and phenotypic switching. {\it Genetics}  {\bf 187}: 837--851.
-
-\ref Lycett, S. J., and J. A. J. Gowlett. 2008. On questions surrounding the Acheulean `tradition'. {\it World Archaeol.} {\bf 40}: 295--315.
-
-\ref Mazel, D. 2006. Integrons: agents of bacterial evolution. {\it Nat.\ Rev.\ Microbiol.} {\bf 4}: 608--620.
-
-\ref Milkman, R., and M. M. Bridges. 1990. Molecular evolution of the {\it Escherichia coli} chromosome. III. Clonal frames. {\it Genetics} {\bf 126} 505--517.
-
-\ref Redfield, R. J. 1993. Evolution of natural transformation: testing the DNA repair hypothesis in {\it Bacillus subtilis} and {\it Haemophilus influenzae}. {\it Genetics} {\bf 133}: 755--761.
-
-\ref Rivoire, O., and S. Leibler. 2014. A model for the generation and transmission of variations in evolution. {\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 111}: E1940--E1949.
-
-\ref Rogers, A. R. 1988. Does biology constrain culture? {\it Am.\ Anthropol.} {\bf 90}: 819--831.
-
-\ref Rosenberg, E., and I. Zilber-Rosenberg. 2016. Microbes drive evolution of animals and plants: the hologenome concept. {\it mBio} {\bf 7}: e01395-15.
-
-\ref Salath\'e, M., J. Van Cleve, and M. W. Feldman. 2009. Evolution of stochastic switching rates in asymmetric fitness landscapes. {\it Genetics} {\bf 182}: 1159--1164.
-
-\ref Salyers, A.
-A., G.
-Whittle, and N.
-B.
-Shoemaker.
-2004.
-Conjugal and mobilizable transposons.
-Pp./ 125--143 {\it in} Miller, R., and M.
-Day (eds.) {\it Microbial Evolution}.
-Washington, D.C.: ASM Press.
-
-\ref Thattai, M., and A. van Oudenaarden. 2004. Stochastic gene expression in fluctuating environments. {\it Genetics} {\bf 167}: 523--530.
-
-\ref Theis, K.
-R., N.
-M.
-Dheilly, J.
-L.
-Klassen, R.
-M.
-Brucker, J.
-F.
-Baines, T.
-C.
-G.
-Bosch, J.
-F.
-Cryan, S.
-F.
-Gilbert, C.
-J.
-Goodnight, E.
-A.
-Lloyd, J.
-Sapp, P.
-Vandenkoornhuyse, I.
-Zilber-Rosenberg, E.
-Rosenberg, and S.
-R.
-Bordenstein.
-2016.
-Getting the hologenome concept right: an eco-evolutionary framework for hosts and their microbiomes.
-{\it mSystems} {\bf 1}: e00028-16.
-
-\ref van Opstal, E. J., and S. R. Bordenstein. 2015. Rethinking heritability of the microbiome. {\it Science} {\bf 349}: 1172--1173.
-
-\ref Wakano, J.
-Y., and K.
-Aoki.
-2007.
-A mixed strategy model for the emergence and intensification of social learning in a periodically changing natural environment.
-{\it Theor.\ Popul.\ Biol.} {\bf 70}: 486--497.
-
-\ref Wakano, J. Y., K. Aoki, and M. W. Feldman. 2004. Evolution of social learning: a
-mathematical analysis. {\it Theor.\ Pop.\ Biol.} {\bf 66}: 249--258.
-
-\ref Whiten, A., F. Ayala, M. W. Feldman, and K. N. Laland. 2017. The extension of biology through culture. {\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 114}: 7775--7781.
-
-\ref Xue, B.-K., and S. Leibler. 2016. Evolutionary learning of adaptation to varying environments through a transgenerational feedback. {\it Proc.\ Natl.\ Acad.\ Sci.\ USA} {\bf 113}: 11266--11271.
-
-\ref Zinder, N. D., and J. Lederberg. 1952.  Genetic exchange in {\it Salmonella}. {\it J. Bacteriol.} {\bf 64}: 679--699.
- \vfil
- \break
- \centerline{\bf TABLE 1}
- \centerline{Values of $\rho^*$ (stable $\rho$)$^{\hbox{\sevenrm (a)}}$ and $\hat\rho$ (``optimal'' $\rho$)$^{\hbox{\sevenrm (b)}}$}
- %\smallskip
- \vskip-.25truein
- $$\vbox{\halign{#\hfil&&\qquad#\hfil\cr
- %&&&&evolutionarily&fitness\cr
- %&&&&stable&optimal\cr
- \noalign{\hrule\vskip 8pt}
- $k$&$l$&$w^\dagger$&\omit\hfill$\hat\rho$\hfill&\omit\hfill$\rho^*$\hfill\cr
- \noalign{\vskip 8pt\hrule\vskip 8pt}
- 1&1&0.1&0.0000000&0.0000000\cr
- 1&1&0.5&0.0000000&0.0000000\cr
- 1&1&0.9&0.0000000&0.0000000\cr
- 1&2&0.1&0.00065&0.821\cr
- 1&2&0.5&fixation&fixation\cr
- 1&2&0.9&fixation&fixation\cr
- 2&2&0.1&0.0000000&0.0000000\cr
- 2&2&0.5&0.0000000&0.0000000\cr
- 2&2&0.9&0.0000000&0.0000000\cr
- 3&10&0.1&0.00031&0.619\cr
- 3&10&0.5&fixation&fixation\cr
- 3&10&0.9&fixation&fixation\cr
- 5&30&0.1&0.00027&0.6503\cr
- 5&30&0.5&fixation&fixation\cr
- 5&30&0.9&fixation&fixation\cr
- 12&12&0.1&0.24347&0.84924\cr
- 12&12&0.5&0.0000000&0.91209\cr
- 12&12&0.9&0.0000000&0.95686\cr
- 20&20&0.1&0.223925\cr
- 20&20&0.5&0.0000000&0.94643\cr
- 20&20&0.9&0.0000000&0.98304\cr
- 30&30&0.1&0.1932801\cr
- 30&30&0.5&0.0000000&0.96331\cr
- 30&30&0.9&0.0000000&0.0000000\cr
- 50&50&0.1&0.15419\cr
- 50&50&0.5&0.22107&0.9768\cr
- 50&50&0.9&0.0000000&0.99581\cr
-\noalign{\vskip 8pt\hrule}
-\noalign{\medskip}
-\multispan4$^\dagger$\ \hbox{\tenrm Note:}\ $W=1$\hfill\cr
-}}$$
-\medskip
-\vbox{\hsize6.5truein
-(a): $\rho^*$ is the uninvadable value of the vertical transmission rate.
-
-(b): $\hat\rho$ is the geometric mean value of $\rho$ at the stable equilibrium of the $AkBl$ cycle.}
- \vfil\break
 

### 2017-12-11 ms/PNAS/Supplementary_Figures.tex
pnas branch, middle of revision

--- 2017-10-17 ms/figures_supp.tex
@@
 
+% 1
 \begin{figure}
 \centering
-\includegraphics{../figures/env_A1B1.pdf}
+\includegraphics{../../figures/lk_phase_plane.pdf}
+\caption{Ratios of selection periods \(\frac{k}{l}\) that lead to
+fixation of phenotype \emph{A} (red) or polymorphism of phenotypes
+\emph{A} and \emph{B} (blue). \emph{k} and \emph{l} are the number of
+generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
+In all cases, \emph{W} = 1, \emph{w} = \(1-s\).}\label{fig:lk_phase_plane}
+\end{figure}
+
+% 2
+\begin{figure}
+\centering
+\includegraphics{../../figures/env_A1B1.pdf}
 \caption{Frequency of phenotype \emph{A} after every two generations in
@@
 
+% 3
 \begin{figure}
 \centering
-\includegraphics{../figures/env_A1B2.pdf}
+\includegraphics{../../figures/A1B1_equilibrium.pdf}
+\caption{Properties of stability in \emph{A1B1} selection regime.
+\textbf{(A)} Stable frequency of phenotype $A$ and \textbf{(B)} stable mean fitness as functions of the vertical transmission rate \(\rho\) and the fitness of the disfavored phenotype \(w\).
+Black contour lines join \(\rho\) and \(w\) combinations that result in the same stable value.
+In all cases, fitness of the favored phenotype is \emph{W=1}.
+}\label{fig:A1B1_equilibrium}
+\end{figure}
+
+% 4
+\begin{figure}
+\centering
+\includegraphics{../../figures/AkBk_x0.pdf}
+\caption{Convergence of the frequency of phenotype \emph{A} to a stable polymorphism in selection regime \emph{AkBk}.
+Comparison of dynamics starting with different initial frequencies of phenotype \(A\) (0.01--0.99), and different $k$, $\rho$ and $w$ values.
+The lines show the $x$ frequency of phenotype \emph{A} at the end of each period, after every $2k$ generations.
+In all cases, $W=1$.}\label{fig:AkBk_x0}
+\end{figure}
+
+% 5
+\begin{figure}
+\centering
+\includegraphics{../../figures/env_A1B2.pdf}
 \caption{Frequency of phenotype \emph{A} after every three generations in
@@
 
+% 6
 \begin{figure}
 \centering
-\includegraphics{../figures/AkBk_x0.pdf}
-\caption{Convergence of the frequency of phenotype \emph{A} to a stable polymorphism in selection regime \emph{AkBk}.
-Comparison of dynamics starting with different initial frequencies of phenotype \(A\) (0.01--0.99), and different $k$, $\rho$ and $w$ values.
-The lines show the $x$ frequency of phenotype \emph{A} at the end of each period, after every $2k$ generations.
-In all cases, $W=1$.}\label{fig:AkBk_x0}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/env_A3B10.pdf}
+\includegraphics{../../figures/env_A3B10.pdf}
 \caption{Frequency of phenotype \emph{A} after every thirteen generations in
@@
 
+% 7
 \begin{figure}
 \centering
-\includegraphics{../figures/AkBk_stable_wbar.pdf}
+\includegraphics{../../figures/AkBk_stable_wbar.pdf}
 \caption{Stable population mean fitness in selection regime \emph{AkBk} as a function of the vertical transmission rate \(\rho\) and the number \(k\) of generations in which phenotypes \emph{A} and \emph{B} are favored by selection, for different selection intensities: \textbf{(A)} \(w=0.1\), \textbf{(B)} \(w=0.5\), and \textbf{(C)} \(w=0.9\).
@@
 
+% 8
 \begin{figure}
 \centering
-\includegraphics{../figures/AkBk_w0.5_geomwbar.pdf}
+\includegraphics{../../figures/AkBk_w0.5_geomwbar.pdf}
 \caption{The geometric average of the stable population mean fitness over the $2k$ generation period peaks at $\rho=0$ for $k \le 30$ (red, blue and green lines) and at $\rho \approx 0.23$ for $k=31$ and $32$ (purple and orange lines).
@@
 
+% 9
 \begin{figure}
 \centering
-\includegraphics{../figures/AkBk_stable_modifier_w_0.5.pdf}
- \caption{The leading eigenvalue of the external stability matrix in $AkBk$ with selection.
-The colors show $\lambda_1(\rho, P)$, the leading eigenvalue of the external stability matrix for combinations of $\rho$ (y-axis) and $P$ (x-axis), the vertical transmission rates of resident and invader modifiers, respectively (see \emph{Supplemental Material SP6} for details).
-Purple shows values lower than 1, leading to stability of $\rho$; white equals to 1; green larger than 1, leading to invasion of $P$.
-  The arrows show an "evolutionary" path $\rho_{0} \to \rho_{1} \to \ldots \to \rho^*$: at each step, a modifier with rate $P_t$ invades the resident modifier with rate $\rho_t$ and becomes the resident modifier $\rho_{t+1}$, until consecutive invaders have similar rates.
-  Here, $W=1$ and $w=0.5$.}\label{fig:AkBk_stable_modifier_w_0.5}
+\includegraphics{../../figures/fixation_prob_time.pdf}
+\caption{Fixation probability and time in a finite population.
+\textbf{(A)} Fixation probability \(u(x)\) of phenotype \(A\) (eq.\ 55), and \textbf{(B)}
+Expected time to fixation \(T(x)\) of phenotype \(A\) (eq.\ 56) conditioned on its
+fixation, starting with a single copy in a population of size \(N\). The
+figure compares two estimates: Wright-Fisher simulations (blue circles) and diffusion
+equation approximation (green solid line). Parameters: selection coefficient, \(s=w_A-w_B=0.1\),
+population size, \(N=10,000\).}\label{fixation_prob_time}
+\end{figure}
+
+% 10
+\begin{figure}
+\centering
+\includegraphics{../../figures/AkBl_stable_modifier_w_0.1.pdf}
+ \caption{Evolutionarily stable vertical transmission rate in $AkBl$ selection regime.
+ The figure shows $\frac{\partial \lambda_1}{\partial P}$ the sensitivity of the leading eigenvalue of the external stability matrix $\mathbf{L}$ to changes in $P$ the vertical transmission rate of the invader allele as a function of $\rho$ the vertical transmission rate of the resident allele (see \emph{Section 2} in main text and \emph{Supplemental Material SP6} for details).
+ The shaded area marks $\rho$ values for which phenotype $B$ fixes and there is no polymorphism (Eq.~(20) in main text).
+ Without polymorphism, selection does not affect the transmission rate, so any rate in the shaded area is neutrally stable.
+ In panels A, B, D, G, J, and M, $\frac{\partial \lambda_1}{\partial P} < 0$ at the vicinity of $\rho=0$ and therefore the stable rate is $\rho^*=0$.
+ In panels B, C, E, F, H, I, K, and L, the stable rate $\rho^*$ can be identified as the $\rho$ value at which $\frac{\partial \lambda_1}{\partial P}$ changes from positive to negative.
+ In panels N and O, $\frac{\partial \lambda_1}{\partial P} > 0$ for all $\rho$ values that protect polymorphism, so there are only neutrally stable rates (in the shaded area).
+   Here, $W=1$ and $w=0.1$.}\label{fig:AkBl_stable_modifier_w_0.1}
+\end{figure}
+
+% 11
+\begin{figure}
+\centering
+\includegraphics{../../figures/lk_fix_prob.pdf}
+\caption{Fixation in a finite population with different ratios of selection periods \(\frac{k}{l}\).
+Fixation probability of phenotype $A$ when starting with a single copy in a population of size $N$: $u(1/N) = (1-\exp(-2 \rho \frac{k-l}{k+l}(W-w))/(1-\exp(-2 N \rho \frac{k-l}{k+l}(W-w))$ (see eqs.
+59--60).
+\emph{k} and \emph{l} are the number of
+generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by
+selection. In all cases, fitness of the favored phenotype, $W = 1$; fitness of the unfavored phenotype is $w=1-s$ and the population size is \(N=10,000\).} \label{lk_fix_prob}
+\end{figure}
+
+% 12
+\begin{figure}
+\centering
+\includegraphics{../../figures/A1B1_modifier_invasions.pdf}
+\caption{Consecutive fixation of modifiers that reduce the vertical
+transmission rate in selection regime \emph{A1B1}. The figure shows
+results of numerical simulations of evolution with two modifier alleles
+(Eq. (32) in main text).
+When a modifier allele fixes (frequency\textgreater{}99.9\%), a new modifier allele is introduced with a vertical transmission rate one order of magnitude lower (vertical
+dashed lines). \textbf{(A,D,G)} The frequency of phenotype \emph{A} in
+the population over time. \textbf{(B,E,H)} The frequency of the invading
+modifier allele over time. \textbf{(C,F,I)} The population geometric mean
+fitness over time; insets zoom in to show that the mean fitness 
+decreases slightly with each invasion. 
+Invading alleles are introduced at frequency 0.01\%; whenever their frequency drops below 0.01\% they are re-introduced.
+Parameters: vertical transmission rate of
+the initial resident modifier allele, \(\rho_0 =0.1\); fitness values:
+$W=1$; $w=0.1$ (\textbf{A-C}), 0.5 (\textbf{D-F}), and 0.9
+(\textbf{G-I}). The x-axis is on a log-scale, as each sequential invasion
+takes an order of magnitude longer to complete.
+Panels D-F are the same as in Figure 4 in the main text.
+}\label{fig:A1B1_modifier_invasions}
 \end{figure}

### 2017-12-11 ms/PNAS/revision.tex
pnas branch, middle of revision

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} = Template for a one-column mathematics article
% {pnasinvited} = Template for a PNAS invited submission

\title{Evolution of vertical and oblique transmission under fluctuating selection}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a]{Yoav Ram}
\author[b]{Uri Liberman} 
\author[a,1]{Marcus W. Feldman}

\affil[a]{Department of Biology, Stanford University, Stanford, CA 94305-5020}
\affil[b]{School of Mathematical Sciences, Tel Aviv University}
%\affil[c]{Affiliation Three}

% Please give the surname of the lead author for the running footer
\leadauthor{Ram}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Evolutionary dynamics of phenotypes in populations depend on how the traits are transmitted across generations, and how the environments that cause selection on the traits fluctuate over time. We show that under periodically fluctuating selection, a gene that increases the rate of vertical transmission is disfavored when the periods are short, but approaches an intermediate stable rate for longer periods. This stable rate differs markedly from the rate that maximizes the geometric mean fitness. The evolution of learning rules thus differs qualitatively from the evolution of genetically modified rules of genetic transmission.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Y.R., U.L., and M.W.F. all contributed to all sections of the paper.}
\authordeclaration{The authors declare no conflict of interest.}
%\equalauthors{\textsuperscript{1}A.O.(Author One) and A.T. (Author Two) contributed equally to this work (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: mfeldman@stanford.edu}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{periodic selection $|$ phenotypic polymorphism $|$ modifier theory $|$ fitness optimum}

\begin{abstract}
Conditions for the evolution and maintenance of social learning, in competition with individual learning, under fluctuating selection, is a well-studied problem in the theory of cultural evolution.
Here we study competition between vertical and oblique cultural transmission of a dichotomous phenotype under constant, periodically cycling, and randomly fluctuating selection.
Conditions are derived for the existence of a stable polymorphism in a periodically cycling selection regime.
Under such a selection regime, the fate of a genetic modifier of the rate of vertical transmission depends on the length of the cycle and the strength of selection.
In general, the evolutionarily stable rate of vertical transmission differs markedly from the rate that maximizes the geometric mean fitness of the population.
The evolution of rules of transmission has dramatically different dynamics from the more frequently studied modifiers of recombination, mutation, or migration.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

%\newcommand{\harr#1#2}{\smash{\mathop{\hbox to .2in{\rightarrowfill}}\limits^{\scriptstyle#1}_{\scriptstyle#2}}}

\newcommand{\var}{\varepsilon}

%\newcommand{\ref}{\par\noindent\hangindent.35truein}

\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{\mathbf{I}}

\newcommand{\ab}[1]{\begin{align*}#1\end{align*}}
\newcommand{\an}[1]{\begin{align}#1\end{align}}

%\newcommand{\ss}{\scriptstyle}

\begin{document}

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
%\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{C}avalli-Sforza and Feldman \cite{cavalli1981cultural} distinguished two forms of non-parental phenotypic transmission in the context of cultural evolution. Horizontal transmission  occurs when a trait is passed between members of the same generation and is analogous to transmission of an infectious agent. Oblique transmission to offspring is from non-parental members of the parental generation. Evolution under either of these is expected to be more rapid than under purely vertical, i.e., parent-to-offspring, transmission \cite{lycett2008questions,bergstrom2012evolution}.

Oblique transmission occurs via some mechanism of social learning, which may include imitation or active teaching. There has been an interesting debate over the past thirty years concerning the conditions under which social learning would have an advantage over individual learning or vertical (including genetic) transmission. This debate is usually couched in terms of the mode and tempo of environmental fluctuations that would affect fitness and hence evolution \cite{rogers1988does,boyd1988evolutionary,boyd1995does,feldman1996individual,wakano2004evolution,aoki2005emergence,wakano2006mixed,aoki2014evolution}. 
Mathematical analyses of models of competition between individual and social learning have generally shown that social learning has an advantage when the environment does not fluctuate too frequently. On the other hand, when the environmental changes are very frequent, individual learning is favored, while innate (genetic) determination of the trait does best when periods between environmental change are long on average.

In some situations, oblique transmission of biological material is possible.
In bacteria, phenotypes might be determined by heritable mobile genetic elements such as phages \cite{zinder1952genetic}, plasmids \cite{lederberg1946gene}, integrons \cite{mazel2006integrons}, and transposons \cite{salyers2004conjugative}.
Similarly, some phenotypes are determined by genes that are commonly converted by uptake of foreign DNA, i.e., transformation \cite{milkman1990molecular}.
In these cases, inheritance of a phenotype may combine vertical transmission from the parent cell, and oblique transmission from other cells, even if the mechanism for oblique transmission evolved to facilitate some other function \cite{redfield1993evolution}.

In some animals, transmission of microbes may occur during sharing or manipulation of food or other consumable resources during a social interaction. Although transmission of the microbiome in humans is likely to be mostly vertical \cite{rosenberg2016microbes}, in other organisms there is multi-generational food sharing during which symbionts from the parental cohort may be transmitted obliquely to younger individuals \cite{theis2016getting}. In such cases, fluctuations in the resource type or availability may have fitness effects that depend on features of the transmitted microbiome. This ecological perspective on community transmission is stressed by van Opstal and Bordenstein \cite{van2015rethinking} who emphasize the ``need to consider the relative roles of vertical and horizontal transmission of microbial communities.''

Another perspective on the evolutionary consequences of fluctuating environments (and, as a result, fluctuating selection) derives from the phenomenon of phenotypic switching \cite{balaban2004bacterial,kussell2005phenotypic,thattai2004stochastic,salathe2009evolution,liberman2011evolution,gaal2010exact}
In these studies, mutation causes the organism to switch phenotypes (usually treated as haploid genotypes), and the problem has usually been couched in terms of the optimal rate of mutation in models where the phenotypic fitnesses fluctuate over time. These models did not include social learning, and the evolution was regarded as a mode of bet-hedging against future environmental change. Optimal (that is, evolutionarily stable) mutation rates depend on many features of the fluctuations, for example,  degree of fitness symmetry, strength of selection, and variance in the period of fluctuation \cite{salathe2009evolution}.

In a recent analysis of evolution under fluctuating selection, Xue and Leibler \cite{xue2016evolutionary} allowed an organism to absorb information about the distribution of possible environments by learning the phenotypes of members of its parental lineage from previous generations. They describe this as ``positive feedback that enhances the probability of the offspring to express the same phenotype as the parent.'' In this formulation there was ``reinforcement of the parent phenotype'' in an offspring, such as might occur through epigenetic inheritance. Although their analysis was not couched in terms of oblique and vertical transmission, as defined by Cavalli-Sforza and Feldman \cite{cavalli1981cultural}, we have been stimulated by their analysis to develop a model in which oblique transmission, at a rate dependent on the trait frequency in the parental generation, occurs in addition to classical vertical transmission. We then ask how fluctuations in selection interact with the rate of oblique transmission to affect evolutionary dynamics and how the rate of oblique transmission itself might evolve.

In our formulation, both the parental phenotype and the distribution of phenotypes in the whole population contribute to an offspring's phenotype. Using conventional modifier theory \cite{feldman1986evolutionary}, we show that in a symmetric cyclic selection regime with cycles of periods 1 or 2, an allele reducing the rate of vertical transmission is expected to increase in frequency when rare and in so doing to increase the mean fitness of the population. However, for cycles of greater length or period asymmetry, interesting non-monotonicities emerge, both in the uninvadable rate of vertical transmission, and the rate that maximizes the geometric time average of the population mean fitness, which we will refer to as the ``geometric mean fitness.'' We develop the models in very large populations with cyclic selection and with random fitnesses and also in the case where drift occurs via sampling from generation to generation in a finite population.

\section{The Model}

Consider an infinite population whose members are characterized by their phenotype $\phi$, which can be of two types $\phi=A$ or $\phi=B$, with associated frequencies $x$ and $(1-x)$, respectively. We follow the evolution of $x$ over discrete non-overlapping generations. In each generation individuals are subject to  selection where the fitnesses of $A$ and $B$ are $w_A$ and $w_B$, respectively.

An offspring inherits its phenotype from its parent via {\sl vertical transmission} with probability $\rho$ and from a random individual in the parental population via {\sl oblique transmission} with probability $(1-\rho)$. Therefore, given that the parent phenotype is $\phi$ and assuming {\sl uni-parental inheritance} \cite{zefferman2016mothers}, the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is
\begin{equation}
P(\phi' =A|\phi) =\left\{\begin{array}{cc}(1-\rho)x+\rho &\mbox{if}\ \phi=A\\ 
 (1-\rho)x & \mbox{if}\ \phi =B\end{array}\right.
 \end{equation}
 where $x=P(\phi=A)$ in the parent's generation before selection.

 Therefore the frequency $x'$ of  phenotype $A$ after one generation is given by the recursion equation
 \begin{equation}
 \an{
 x'&=\rho{w_A\over \overline w}x +(1-\rho)x\\
 &=     {w_A\over \overline w}x\bigl[(1-\rho)x +\rho\bigr] +{w_B\over \overline w}(1-x)\bigl[(1-\rho)x\bigr],}
 \label{2}\end{equation}
 where $\overline w$ is the {\sl mean fitness}, namely
 \begin{equation}\overline w=w_A x+w_B(1-x).\end{equation}
 \eqref{2} can be rewritten as
 \begin{equation}
 \an{
 x' &=& x\left[1+\rho(1-x){w_A-w_B\over \overline w}\right]\\
     &=&x\cdot{x(1-\rho)(w_A-w_B) +\rho w_A +(1-\rho)w_B\over x(w_A-w_B) +w_B}.}
    \label{4} \end{equation}

 In what follows we explore the evolution of the recursion \eqref{4}, namely the equilibria and their stability properties, in the cases of {\sl constant environments} and {\sl changing environments}.

 \subsection{Constant Environment}

When the environment is constant, the fitness parameters $w_A$ and $w_B$ do not change between generations, and we have the following result.
\medskip

\noindent {\bf Result 1.} {\sl If $0<\rho\le 1$ and both $w_A$ and $w_B$ are positive with $w_A\ne w_B$, then fixation in the phenotype $A\ (B)$ is globally stable when $w_A>w_B$ $(w_A<w_B)$.}\par
\medskip

\noindent{\it Proof}.
If we rewrite \eqref{4} as $x'=x\cdot f(x)$, it can be seen that $f(1)=1$, and for $\rho>0$ and $0<x<1$
\begin{equation}
\begin{array}{ccc}
f(x)>1&\hbox{\tenrm when}& w_A>w_B,\\
f(x)<1&\hbox{\tenrm when}& w_A<w_B.\end{array}
\label{5}\end{equation}
Hence, as $w_A>0$ and $w_B>0$, both fixations in $A$ or in $B$  ($x^*=1$ for fixation in $A$ and $x^*=0$ for fixation in $B$) are equilibrium points of \eqref{4}. Moreover, if $x_t$ is the value of $x$ at the $t$-th generation $(t=0,1,2,\dots)$, from \eqref{4}  and \eqref{5} we have for any $0<x_0<1$ and all $t=0,1,2,\dots$,
\begin{equation}
\begin{array}{ccc}
x_{t+1}>x_t&\hbox{\tenrm when}&w_A>w_B,\cr
x_{t+1}<x_t&\hbox{\tenrm when}&w_A<w_B,\end{array}
\end{equation}
and since $x^*=1$ or $x^*=0$ are the only equilibrium points, we have
\begin{equation}
\begin{array}{ccc}
\lim_{t\to\infty}x_t=1,&\hbox{\tenrm for all}\ 0<x_0\le 1,&\hbox{\tenrm when}\ w_A>w_B,\cr
\lim_{t\to\infty}x_t=0,& \hbox{\tenrm for all}\ 0\le x_0< 1,&\hbox{\tenrm when}\ w_A<w_B.}\end{array}
\end{equation}
Therefore fixation of the favored phenotype is globally stable.

If $\rho=0$ or $w_A=w_B>0$, the phenotype frequencies do not change over time.
With {\sl extreme selection},  for example $w_A=1$ and $w_B=0$ (or $w_B=1$ and $w_A=0$),  \eqref{4} reduces to
$x'=(1-\rho)x+\rho$, and therefore 
$x_t=1-(1-\rho)^t(1-x_0)$.
Hence, for $\rho>0$,  $x_t\to 1$ as $t\to\infty$ for all initial values $0\le x_0\le 1$, and  the favored phenotype fixes.

\subsection{Periodically Changing Environment}

 Suppose the environment is not constant but rather changes periodically, such that the favored phenotype changes after a fixed number of generations. Simple examples are $A1B1=ABABAB\dots$, in which the favored phenotype switches every generation, or $A2B1=AABAABAAB\dots$, where every two generations in which selection favors $A$ are followed by a single generation in which selection favors $B$.
  In general, $AkBl$ denotes a selection regime in which the period is of $(k+l)$ generations with $k$ generations favoring phenotype $A$ followed by $l$ generations favoring $B$.

Let $W$ be the fitness of the favored phenotype and $w$ be that of the other phenotype where $0<w<W$. Rewrite \eqref{4} as $x'=F_A(x)=xf_A(x)$ when $A$ is favored and $x'=F_B(x)=xf_B(x)$ when $B$ is favored. Then
 \begin{equation}
 \an{
f_A(x) &= {x(1-\rho)(W-w) +\rho W +(1-\rho)w\over x(W-w) +w}\\
&= 1+\rho(1-x){W-w\over Wx +w(1-x)},\\
f_B(x) &= {x(1-\rho)(w-W) +\rho w +(1-\rho)W\over x(w-W) +W} \\
&= 1+\rho(1-x){w-W\over wx+W(1-x)}.
}\label{8}\end{equation}
  \noindent If $x_t$ denotes the frequency of  the phenotype $A$ at generation $t$ starting with $x_0$ initially, then as we are interested in the values of $x_t$ for $t=n(k+l)$ with $n=0,1,\dots$ at the end of complete periods, we can write
 \begin{equation}x_{(n+1)(k+l)}=F(x_{n(k+l)}),\quad n=0,1,2,\dots,\label{9}\end{equation}
  where $F$ is the composed function
\begin{equation}F=\underbrace{F_B\circ F_B \circ\cdots\circ F_B}_{l\ \hbox{\sevenrm times}} \circ\underbrace{F_A\circ F_A \circ\cdots\circ F_A}_{k\ \hbox{\sevenrm times}}.\label{10}\end{equation}
  Clearly, since $F_A(0) =F_B(0) =0$ and $F_A(1) =F_B(1) =1$, both fixations in $A$ or in $B$ are equilibrium points. An interesting question is when these fixations are locally stable. We concentrate on $x^*=0$, the fixation of the phenotype $B$. As $x'=F_A(x) =xf_A(x)$ for $k$ generations and $x'=F_B(x) =xf_B(x)$ for $l$ generations, the linear approximation of $F(x)$ ``near'' $x=0$ is
 \begin{equation}F(x) \simeq\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^lx.\label{11}\end{equation}
   Hence the local stability of $x^*=0$ is determined by the product $\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l$;  $x^*=0$  is locally stable if this product  is less than 1 and unstable if it is larger than 1.

   From \eqref{8} we have
   \begin{equation}
   f_A(0) = 1+\rho{W-w\over w},\qquad
   f_B(0) = 1+\rho{w-W\over W}.\label{12}\end{equation}
   We start with the case $k=l$.
   \medskip

   \noindent {\bf Result 2}. {\sl If $k=l$ and $0<w<W$ with $0<\rho<1$, fixation of $B$ is unstable.}\par
   \medskip

   \noindent {\it Proof.} The local stability of $x^*=0$, the fixation of $B$, is determined by the product
   \begin{equation}
   \bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^k=\left[\left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right)\right]^k.\label{13}\end{equation}
   Observe that
   \begin{equation}
   \left(1+\rho{W-w\over w}\right)\left(1+\rho{w-W\over W}\right) = 1+\rho(1-\rho){\left(W-w\right)^2\over wW}>1.\label{14}\end{equation}
   Since $0<\rho<1$ and $0<w<W$,  fixation on $B$ is unstable.
  \bigskip

\noindent{\it Conclusions}
\smallskip

\noindent 1.\ \ Since $k=l$ and the above result also holds  when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is also not stable. Thus neither phenotype can be lost, and there is a {\sl protected polymorphism} \cite{prout1968sufficient}.
\smallskip

\noindent 2.\ \ For general $k$, $l$, the condition for local stability of fixation in $A$ is
\begin{equation}\bigl[f_A(0)\bigr]^l \bigl[f_B(0)\bigr]^k <1, \label{15}\end{equation}
and that of $B$ is
\begin{equation}\bigl[f_A(0)\bigr]^k \bigl[f_B(0)\bigr]^l <1. \label{16}\end{equation}
Therefore, following Result 2, 
\begin{equation}\bigl[f_A(0)\bigr]^{k+l}\bigl[f_B(0)\bigr]^{k+l}>1,\label{17}\end{equation}
and it is impossible that both fixations are stable. Further, since by \eqref{12}  $f_A(0)>1$ and $0<f_B(0)<1$ when $0<w<W$, by choosing $k$ and $l$ appropriately, fixation on $A$ or fixation on $B$ (but not both) can be stable. In addition, we can have both fixations unstable giving the following result.
\medskip

\noindent {\bf Result 3}. {\sl With $0<\rho<1$ and $0<w<W$ in the case of $AkBl$ periodically changing environments, both fixations may be unstable, producing a protected polymorphism.}\par
\medskip

\noindent{\it Proof}. Let $a=1+\rho{W-w\over w}$ and $b=1+\rho{w-W\over W}$, and our assumption entails  $a>1$ and $0<b<1$. Following \eqref{11}, fixation in $B$ is not stable if $a^kb^l>1$, and similarly  fixation in $A$ is unstable if $a^lb^k>1$. Therefore both fixations are unstable if
\begin{equation}a^kb^l>1\quad\hbox{\tenrm and}\quad a^lb^k>1,\label{18}\end{equation}
or equivalently if
\begin{equation}k\log a +l\log b>0\quad\hbox{\tenrm and}\quad l\log a+k\log b>0.\label{19}\end{equation}
Now \eqref{19} holds if and only if
\begin{equation}k\,{\log(1/b)\over\log a} < l < k\,{\log a\over\log(1/b)}.\label{20}\end{equation}

The inequalities of \eqref{20} are consistent if and only if $\log(1/b)<\log a$, i.e., $ab>1$, which according to \eqref{14}, is true.

The linear approximation of $F(x)$ near $x^*=0$ (\eqref{11}) does not depend on the order in which phenotypes $A$ and $B$ are favored within a cycle of $k+l$ generations.
Therefore, the local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations, $A$ is favored $k$ times and $B$ is favored $l$ times, and not their order in the cycle. 
When neither fixation in $A$ or $B$ is stable, there is a protected polymorphism, and we expect to have one or more polymorphic equilibria. Supplementary Figure S1 illustrates the relationship between $k,l$ and $\rho$ that gives polymorphism of $A$ and $B$, or fixation, for different values of $s=W-w$.

For the simple case of $A1B1$ periodically changing environment we have the following.
\medskip

\noindent {\bf Result 4}. {\sl In the case $A1B1$ with $0<\rho<1$ and $0<w<W$ the two fixations are unstable and there exists a unique stable polymorphism.}\par
\medskip

\noindent{\it Proof}. Let $x$ be the initial frequency of $A$ and $x'$ its frequency  after one cycle of the $A1B1$ selection. Then $x' =F_B\bigl(F_A(x)\bigr)$ where by \eqref{8}
\begin{equation}
\an{
F_A(x)&=x{x(1-\rho)(W-w) +\rho W+(1-\rho)w\over x(W-w) +w},\\
F_B(y) &= y{y(1-\rho)(w-W) +\rho w+(1-\rho)W\over y(w-W)+W}.}
\label{21}\end{equation}
The equilibrium equation is $x=F_B\bigl(F_A(x)\bigr)$, which reduces to a fourth degree polynomial equation in $x$. Since the two fixations in $B$ and $A$ are equilibria corresponding to the two solutions $x=0$ and $x=1$, the other equilibria correspond to solutions of a quadratic equation $Q(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with
\begin{equation}\alpha_2=1,\  \alpha_1={W+w\over (2-\rho)(W-w)}-1,\  \alpha_0={-w\over (2-\rho)(W-w)}.\label{22}\end{equation}
As $0<\rho<1$ and $0<w<W$ we have
\begin{equation}
\an{
Q(0) &= {-w\over (2-\rho)(W-w)}<0\\
\hbox{\tenrm and}\ \ Q(1) &={W\over (2-\rho)(W-w)} >0.}\label{23}\end{equation}
Also, as $\alpha_2=1$ and $\alpha_0<0$, the quadratic equation $Q(x)=0$ has two real roots, one negative and one positive $x^*$ satisfying $0<x^*<1$. The latter determines a unique polymorphism. Let $H(x) =F_B\bigl(F_A(x)\bigr)$. Then 
 \begin{equation}H(0)=0,\quad H(x^*)=x^*,\quad H(1)=1.\label{24}\end{equation}
 Also 
\begin{equation}\begin{array}{ccc}
F'_A(x) &=& {x^2(1-\rho)\left(W-w\right)^2 +2xw(1-\rho)(W-w) +w\left[\rho W +(1-\rho)w\right]\over \left[x(W-w) +w\right]^2}\\
 \end{array}\label{25}\end{equation}
 and
\begin{equation}\begin{array}{ccc}
F'_B(x) &=  &{x^2(1-\rho)\left(w-W\right)^2 +2xW(1-\rho)(w-W) +W\left[\rho w +(1-\rho)W\right]\over \left[x(w-W) +W\right]^2}.\\
\end{array}\label{26}\end{equation}
 From our assumptions on $\rho$, $w$, and $W$, we have $F'_A(x)>0$ for  $0\le x\le 1$. Observe that the numerator of $F'_B(x)$ is linear in $\rho$; its value when $\rho=1$ is $wW>0$, and when $\rho=0$ it is
 \begin{equation}x^2\left(w-W\right)^2 +2xW(w-W) +W^2 =\left[x(w-W) +W\right]^2 >0.\label{27}\end{equation}
  Hence $F'_{B}(x)>0$ for all $0\le x\le 1$, and $H'(x) =F'_B\bigl(F_A(x)\bigr)F'_A(x)$ is positive when $0\le x\le 1$. Thus $H(x)$ is monotone increasing for $0\le x\le 1$; $H(x)>x$ for $0<x<x^*$, $H(x)<x$ for $x^*<x<1$.
Starting from any initial value $0<x_0 <1$ we have  $x_t\to x^*$ as $t\to\infty$. Supplementary Figure S2, panels A, C, and E, illustrate examples of how the frequency of $A$ changes over time in the $A1B1$ regime of cycling selection.

For more general cyclic fitness regimes, the polynomial that gives the equilibria is of higher order, and it is conceivable that more than one stable polymorphism could exist for given values of $\rho$, $W$, and $w$. We have been able to show that when neither fixation in $A$ or $B$ is stable, in the $AkBk$ case this cannot occur. In fact we have the following.
\medskip

\noindent {\bf Result 5}. {\sl In the $AkBk$ selection regimes, if the fixations in $A$ and $B$ are locally unstable, a single stable polymorphic equilibrium exists.}\par
\medskip

The proof of Result 5 is in Supplementary Material SP1.  Supplementary Figure S3A shows the stable equilibrium frequencies $x^*$ as a function of $\rho$, $W$, and $w$ in the $A1B1$ regime.
For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplementary Figure S4 illustrates the convergence to a single stable polymorphism.

% Figure 1
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/A1B2_polymorphism_vert.pdf}
\caption{Stable frequency of phenotype \emph{A} and geometric mean fitness in selection regime
$A1B2$ as a function of the vertical transmission rate \(\rho\) and the fitness of the disfavored phenotype \(w\).
\textbf{(A)} Stable frequency of phenotype $A$ at the end of each three generations cycle.
\textbf{(B)} Geometric average of the stable population mean fitness over the three generations cycle: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
Gray contour lines join \(\rho\) and \(w\) combinations that result in the same stable value.
In all cases, \emph{W=1}.}\label{fig:A1B2_polymorphism}
\end{figure}

We have not been able to prove that for selection regimes $AkBl$ with $l\ne k$ there is a single stable polymorphic equilibrium when the two fixations are unstable. However, the numerical examples in Supplementary Figure S1 (for $AkBl$) and in Figure 1 and Supplementary Figure S5 for the special case $A1B2$ all exhibit a single stable polymorphic equilibrium when fixations in $A$ and $B$ are unstable. These numerical results suggest that for $W>w>0$ and $0<\rho<1$ the high order polynomial that gives the equilibria has only a single root corresponding to a globally stable polymorphism. Supplementary Figure S6 shows that this is the case for the $A3B10$ regime.

\subsection{Randomly Changing Environment}

 We now consider the  case where the environment changes according to a stochastic process. Without loss of generality, assume that the fitness parameters at generation $t$ ($t=0,1,2,\dots$) are $1+s_t$ for phenotype $A$ and $1$ for phenotype $B$, where the random variables $s_t$ for $t=0,1,2,\dots$ are independent and identically distributed. Also we assume that there are positive constants $C$ and $D$ such that $P(-1+C<s_t <D)=1$.

 Corresponding to \eqref{4}, with $w_A=1+s_t$, $w_B=1$ the recursion equation is
 \begin{equation}
 x_{t+1}=x_t {1+\rho s_t +x_t(1-\rho)s_t\over 1+x_ts_t}\quad t=0,1,2,\dots\; .\label{28}\end{equation}

\noindent As $\{x_t\}$ for $t=0,1,2,\dots$ is a sequence of random variables, the notion of stability of the two fixation states needs clarification. Following Karlin and Lieberman \cite{karlin1974random} and Karlin and Liberman \cite{karlin1975random} we make the following definition.
 \medskip

 \noindent{\it Definition}: ``stochastic local stability''.
 A constant equilibrium state $x^*$ is said to be {\sl stochastically locally stable} if for any $\var>0$ there exists a $\delta>0$ such that $|x_0-x^*|<\delta$ implies
 \begin{equation}
 P\left(\lim_{t\to\infty}x_t =x^*\right)\ge 1-\var.\label{29}\end{equation}
 Thus stochastic local stability holds for $x^*$ provided for any initial $x_0$ sufficiently near $x^*$ the process $x_t$ converges to $x^*$ with high probability.

 In our case there are two constant equilibria $x^*=0$ and $x^*=1$ corresponding to fixation in $B$ and $A$, respectively. We can characterize the stochastic local stability of these fixations as follows.
 \medskip

 \noindent {\bf Result 6}. {\sl Suppose $E\left[\log (1+\rho s_t)\right]>0$. Then $x^*=0$, the fixation of phenotype $B$, is not stochastically locally stable. In fact $P\left(\lim_{t\to\infty}x_t=0\right)=0$.}
 \medskip

  \noindent {\bf Result 7}. {\sl Suppose $E[\log(1+\rho s_t)]<0$. Then $x^*=0$, the fixation of phenotype $B$, is stochastically locally stable. In particular, if $E(s_t)\le 0$, $x^*=0$ is stochastically locally stable.}
\smallskip

 The proofs of Results 6 and  7 can be found in Supplementary Materials SP2 and SP3.

 Using the general notation for the fitness parameters $w_A$ and $w_B$, the stochastic local stability of the $B$ fixation is determined by the sign of $E\left[\log\left(1-\rho+\rho{w_A\over w_B}\right)\right]$ and that of the $A$ fixation by the sign of $E\left[\log\left(1-\rho+\rho{w_B\over w_A}\right)\right]$. For example, if the  sign of the first  is negative, fixation in $B$ is stochastically locally stable, and when it is positive, with probability one convergence to fixation in phenotype $B$ does not occur. It is also true  that if $E(w_A/w_B)\le 1$, then fixation of $B$ is stochastically locally stable. Following \eqref{14}, for all realizations of $w_A$ and $w_B$,
 \begin{equation}
 \log\left(1-\rho +\rho{w_A\over w_B}\right) +\log\left(1-\rho +\rho{w_B\over w_A}\right)>0.\label{30}\end{equation}
 Therefore, as in the case of periodically changing environments $AkBl$, it is impossible that both fixations are simultaneously stochastically locally stable.  It is possible, however, that neither fixation is  stochastically locally stable, in which case we expect the population to converge to a polymorphic distribution. Figure 2 illustrates how the properties of $s_t$ in \eqref{28} affect the frequency of phenotype $A$, and in particular the stochastic local stability of fixation in phenotype $B$. Figure 3 shows the dynamics of the frequency of $A$ in a case where $w_A$ and $w_B$ are identically distributed and independent; in this case the expectation of the stationary distribution is ${1\over 2}$, and its variance increases as $\rho$ increases.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/stochastic_env_x_t.pdf}
\caption{Stochastic local stability. 
The figure shows the frequency of phenotype \emph{A} after \(10^6\) generations in a very large population evolving in a stochastic environment (eq.\ 28).
The fitnesses of phenotypes \emph{A} and \emph{B} are \(w_A=1+s_t\) and \(w_B=1\),
where \(s_t\) is \emph{s} with probability \emph{p} and \emph{-s} with probability \(1-p\).
The gray lines mark combinations of \emph{p} and \emph{s} for which \(E[\log{(1-\rho+\rho\frac{w_A}{w_B})}]=0\) and \(E[\log{(1-\rho+\rho\frac{w_B}{w_A})}]=0\).
According to Result 6, between these lines fixation of either phenotype is not stochastically locally stable, and we expect a stationary polymorphism between the lines.
Here, initial frequency of \(A\) is \(x_0=1/10,000\) and the vertical transmission rate is \(\rho=0.1\).}\label{fig:stochastic_env_x_t}
\end{figure}

%Figure 3
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/stoch_two_iid_rvs.pdf}
\caption{Effect of vertical transmission rate $\rho$ on phenotype polymorphism in a randomly changing environment. Dynamics of the frequency of phenotype \(A\) over time starting at \(x_0=10^{-5}\) when the fitnesses of phenotypes \(A\) and \(B\) are identically and independently distributed random variables.
As the vertical transmission rate \(\rho\) increases from 0.001 to 0.5, the frequency reaches a polymorphic distribution with \(E(x_t)\to0.5\) faster,
but the variance also increases.
The fitnesses of phenotypes \(A\) and \(B\), \(w_A\) and \(w_B\), are both exponential random variables with expected value 2.}\label{fig:stoch_two_iid_rvs}
\end{figure}

\section{Evolutionary Stability of Oblique Transmission}

An interesting question concerns the evolution of oblique transmission itself. For example, is there an evolutionarily stable rate of oblique transmission? To answer this question we use a modifier model, for which we suppose that the vertical transmission rate is controlled by a genetic locus with two possible alleles $m$ and $M$. Let the vertical transmission rates determined by $m$ and $M$ be $\rho$ and $P$, respectively. Thus there are four pheno-genotypes $mA$, $mB$, $MA$, $MB$ whose frequencies at a given generation are denoted by $x_1$, $x_2$, $x_3$, $x_4$, respectively. As the fitnesses are determined by the two phenotypes $A$ and $B$, and the modifier locus is selectively neutral, we have the following table.
 \begin{equation}
 \begin{array}{ccccc}
 \hbox{\tenrm pheno-genotype}&mA&mB&MA&MB\\
 \hbox{\tenrm frequency}&x_1&x_2&x_3&x_4\\
 \hbox{\tenrm fitness}&w_A&w_B&w_A&w_B\\
 \hbox{\tenrm vertical transmission rate}&\rho&\rho&P&P
 \end{array}\label{31}\end{equation}
Following the rationale leading to equation [2], the next generation pheno-genotype frequencies $x'_1$, $x'_2$, $x'_3$, $x'_4$ are
\begin{equation}
\begin{array}{l}
\overline wx'_1 = w_Ax_1\bigl[(1-\rho)(x_1+x_3) +\rho\bigr] +w_Bx_2(1-\rho)(x_1+x_3)\\
 \overline wx'_2 = w_Ax_1(1-\rho)(x_2+x_4) +w_Bx_2\bigl[(1-\rho)(x_2+x_4)+\rho\bigr]\\ 
  \overline wx'_3 = w_Ax_3\bigl[(1-P)(x_1+x_3) +P\bigr] +w_Bx_4(1-P)(x_1+x_3)\\
\overline wx'_4 = w_Ax_3(1-P)(x_2+x_4) +w_Bx_4\bigl[(1-P)(x_2+x_4)+P\bigr], 
\end{array}\label{32}\end{equation}
with $\overline w$, the mean fitness, given by
\begin{equation}
\overline w =w_A(x_1+x_3) +w_B(x_2+x_4).
\label{33}\end{equation}
Note that under these assumptions the $M/m$ locus and the $A/B$ phenotypic dichotomy do not undergo anything analogous to recombination, which might be introduced if $A/B$ phenotypes were viewed as haploid genetic variants.

Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} \cite{feldman1986evolutionary,altenberg2017unified} to invasion by allele $M$.
A constant environment always leads to fixation of the favored type, independent of $\rho$. We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$ (but see Supplementary Material SP6 for computational analysis of the general $AkBl$ case). Specifically, from \eqref{32} with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
\begin{equation}
\underline x''=T_2(T_1\underline x),
\label{34}\end{equation}
where the non-linear transformation $\underline x'=T_1\underline x$ is given by \eqref{32} with $w_A=W$, $w_B=w$, and the non-linear transformation $\underline x''=T_2\underline x'$ is given by \eqref{32} with $w_A=w$, $w_B=W$. Here $\underline x$, $\underline x'$, $\underline x''$ are the frequency vectors.

For the $A1B1$ case, when only the $m$ allele is present with associated rate $\rho$,  $0<\rho<1$, and $0<w<W$, a unique stable equilibrium $\underline x^* =(x_1^*,1-x_1^*,0,0)$ exists. $x_1^*$ is the only positive root of the quadratic equation $Q(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with $\alpha_2,\alpha_1,\alpha_0$ specified in \eqref{22}. Solving $Q(x)=0$ gives
\begin{equation}
x_1^* ={1\over 2} -{W+w-\sqrt{\left(1-\rho\right)^2\left(W-w\right)^2 +4Ww}\over 2\cdot(2-\rho)(W-w)},
\label{35}\end{equation}
 and it can be seen that
 \begin{equation}
 {\sqrt{Ww}-w\over W-w}<x_1^* <{1\over 2}.
 \label{36}\end{equation}

 The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ near $x^*$, which is derived from \eqref{32} and given by
 \begin{equation}
 \overline w^*\cl_1 =\left[\begin{array}{cc}{W\bigl[(1-P)x_1^* +P\bigr] & w(1-P)x_1^*\\
 W(1-P)x_2^* & w\bigl[(1-P)x_2^* +P\bigr]\end{array}\right]
 \label{37}\end{equation}
 and
 \begin{equation}
 \overline w^{**}\cl_2 =\left[\begin{array}{cc}w\bigl[(1-P)x_1^{**} +P\bigr] & W(1-P)x_1^{**}\\
 w(1-P)x_2^{**} & W\bigl[(1-P)x_2^{**} +P\bigr]\end{array}\right],
 \label{38}\end{equation}
where $\underline x^{**} =T_1\underline x^*$, $x_1^{**}=x_2^*=1-x_1^*$, $x_2^{**}=x_1^*$,  and
\begin{equation}
\overline w^* =Wx_1^* +wx_2^*,\qquad \overline w^{**} =wx_1^{**} +Wx_2^{**}.\label{39}\end{equation}
Due to the symmetry between the two phenotypes $A$ and $B$ in the $A1B1$ case, we have $x_1^{**} =x_2^*$ and $x_2^{**} =x_1^*$ so that $\overline w^{**} =\overline w^*$ and in fact
\begin{equation}
\overline w^*\cl_2 = \left[\begin{array}{cc}w\bigl[(1-P)x_2^* +P\bigr] & W(1-P)x_2^*\\
 w(1-P)x_1^* & W\bigl[(1-P)x_1^* +P\bigr]\end{array}\right].
 \label{40}\end{equation}
 Note that as $\underline x^* =T_2(T_1\underline x^*)$ with $x_3^*=x_4^*=0$, from \eqref{37} and \eqref{38} with $P=\rho$ we have
 \begin{equation}\left[\begin{array}{c}x_1^*\\ x_2^*\end{array}\right]= \cl_2\cdot\cl_1\left[\begin{array}{c}x_1^*\cr x_2^*\end{array}\right] =\cl\left[\begin{array}{c}x_1^*\\ x_2^*\end{array}\right].
 \label{41}\end{equation}
 Hence when $P=\rho$ one of the eigenvalues of $\cl$ is 1.

 In general $\cl=\cl_2\cdot\cl_1$, and using \eqref{37} and \eqref{40} we have
  \begin{equation}
  \an{
 \left(\overline w^*\right)^2\cl_{11} &=Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[w(1-P)x_1^*\bigr]^2 \\
 \left(\overline w^*\right)^2\cl_{12} &=W(1-P)\bigl[P+x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr]\\
 \left(\overline w^*\right)^2\cl_{21} &=w(1-P)\bigl[1-x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr]\\
 \left(\overline w^*\right)^2\cl_{22} &= Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[W(1-P)x_2^*\bigr]^2
 }
 \label{42}\end{equation}
 The external stability of $\underline x^*$ is determined by the eigenvalues of $\cl$, namely the roots of its characteristic polynomial $R(\lambda) =\det(\cl-\lambda\ci)$, with $\ci$ the $2\times 2$ identity matrix. From \eqref{42}, $R(\lambda)=a_2\lambda^2 +a_1\lambda +a_0$, where
 \begin{equation}
 \an{
 &a_0={P^2W^2w^2\over\left(\overline w^*\right)^4},\\
 & a_1=-{2PWw +\left(1-P\right)^2\left[Wx_2^* +wx_1^*\right]^2\over \left(\overline w^*\right)^2},\quad a_2=1.}
 \label{43}\end{equation}

We then have the following result, which is proved in Supplementary Material SP4.
 \medskip

 \noindent{\bf Result 8}. {\sl $\cl$ has two positive eigenvalues and\hfil\break
  {\hglue.22truein}(i) when $P>\rho$ the two eigenvalues are less than $1$,\hfil\break
  {\hglue.19truein}(ii) when $P<\rho$ the largest eigenvalue is larger than $1$,\hfil\break
  {\hglue.16truein}(iii) when $P=\rho$, the largest eigenvalue is $1$.}\par
  \medskip

 We conclude that in the $A1B1$ selection regime an allele $m$ producing vertical transmission rate $\rho$ is stable to the introduction of a modifier allele $M$ with associated rate $P$ if $P>\rho$, and it is unstable if $P<\rho$. Thus in this case, evolution tends to reduce vertical transmission, and hence increase the rate of oblique transmission,  and there is a {\sl reduction principle} for the rate of vertical transmission \cite{feldman1986evolutionary,altenberg2017unified}. The evolutionary dynamics of the reduction in $\rho$ under the $A1B1$ cycling regime are shown in Figure 4, which also illustrates the change in phenotype frequencies over time.

 %Figure 4
\begin{figure*}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/{A1B1_modifier_invasions_w0.5}.pdf}
\caption{Consecutive fixation of modifiers that reduce the vertical transmission rate in selection regime \emph{A1B1}.
The figure shows results of numerical simulations of evolution with two modifier alleles (\eqref{32}).
When a modifier allele fixes (frequency\textgreater{}99.9\%), a new modifier allele is introduced with a vertical transmission rate one order of magnitude lower (vertical dashed lines). \textbf{(A)} The frequency of phenotype \emph{A} in the population over time. \textbf{(B)} The frequency of the invading modifier allele over time. \textbf{(C)} The population geometric mean fitness over time; insets zoom in to show that the mean fitness increases slightly with each invasion.
Invading alleles are introduced at frequency 0.01\%; whenever their frequency drops below 0.01\% they are re-introduced.
Parameters: vertical transmission rate of the initial resident modifier allele, \(\rho_0 =0.1\); fitness values: $W=1$; $w=0.5$. The x-axis is on a log-scale, as each sequential invasion takes an order of magnitude longer to complete.
See Figure S12 for $w=0.1$ and $0.9$.}
\label{fig:A1B1_modifier_invasions}
\end{figure*}

%Figure 5
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/stoch_modifier_invasions.pdf}
\caption{Consecutive fixation of modifiers that reduce the vertical transmission rate $\rho$ under symmetric randomly changing selection.
The figure shows results of numerical simulations of evolution with two modifier alleles
(\eqref{32}).
When a modifier allele fixes (frequency\textgreater{}99.9\%),
a new modifier allele is introduced with a vertical transmission rate one order of magnitude lower (vertical dashed lines).
\textbf{(A)} The frequency of phenotype \emph{A} in the population over time.
\textbf{(B)} The frequency of the invading modifier allele over time.
Invading alleles are introduced at frequency 0.01\%; whenever their frequency drops below 0.01\% they are re-introduced.
Parameters: vertical transmission rate of the initial resident modifier allele is \(\rho_0 =0.1\) and the ratio of fitness values  is $w_A/w_B=10$ with probability 0.5 and $w_A/w_B=0.1$ also with probability 0.5.
The x-axis is on a log-scale, as each sequential invasion takes an order of magnitude longer to complete.}\label{fig:stoch_modifier_invasions}
\end{figure}

 In the case of identically distributed random fitnesses $w_A$ and $w_B$, Figure 5 shows  an example of the success of modifiers that reduce $\rho$. We have not, however, been able to prove that there is a reduction principle for this class of fluctuating fitnesses.

Values of the evolutionarily stable vertical transmission rate, $\rho^*$, for some $AkBl$ examples (see Supplementary Material SP7 and  Figure S10 for analytical details) are recorded in Table 1 for different values of $w$ relative to $W=1$.
Interestingly, for the $A1B2$ regime, with $w=0.1$, the evolutionarily stable value of $\rho$ is substantially greater than zero, as it is in the $A3B10$ and $A5B30$ regimes.
$AkBk$ results are plotted in Figure 6B.
In the $A2B2$ regime, $\rho^* =0$ and there is reduction of vertical transmission for all selection values tested.
However, for $AkBk$ regimes with $k>2$, we find $\rho^*\ne 0$, and depending on $w$, $\rho^*$ can be as high as 0.95.
In Table 1, blank values for $\rho^*$ indicate that our method was numerically unstable  and that a precise value for $\rho^*$ could not be obtained.
This is why, in Figure 6B, no $\rho^*$ points are shown for $AkBk$ with $k>19$.
In Table 1, the word ``fixation'' indicates that fixation of $B$ occurs, at which point there can be no effect of modification of $\rho$; $\rho^*$ cannot be calculated in such cases.

\section{Geometric Mean Fitness and Rate of Vertical Transmission}

Under fluctuating selection, the geometric mean fitness of genotypes has been shown to determine their evolutionary dynamics \cite{haldane1963polymorphism,karlin1974random,wakano2004evolution}. For the evolution of mutation rates that are controlled by genetic modifiers, the stable mutation rate and the mutation rate that maximizes the geometric mean fitness of the population appear to be the same when the period of environmental fluctuation is low enough \cite{liberman2011evolution}. We can ask the same question here: is the stable rate $\rho^*$ the same as the rate $\hat\rho$ that maximizes the equilibrium value of the geometric mean fitness under fluctuating selection?
First consider the $A1B1$ selection regime. We have the following result.
\medskip

\noindent{\bf Result 9}. {\sl If $W>w$ and $0\le\rho\le 1$, then the mean fitness at the stable equilibrium in the $A1B1$ environment is a decreasing function of $\rho$.}\par
\medskip

\noindent{\it Proof.} In $A1B1$ the stable frequency of phenotype $A$ is by \eqref{35}
\begin{equation}
x^* ={1\over 2} -{W+w-Z\over 2(2-\rho)(W-w)},
\label{44}\end{equation}
where $Z =\sqrt{(1-\rho)^2(W-w)^2 +4Ww}>0$.   The mean fitness at the stable equilibrium  is $\overline w^*\cdot\overline w^{**}$, and as $\overline w^* =\overline w^{**}$  by  \eqref{39}, this allows us to reduce the problem to properties of $\overline w^*$.  Now since $W>w$,  $\overline w^*$ is an increasing linear function of $x^*$:
\begin{equation}
\overline w^* =x^*W +(1-x^*)w = x^*(W-w) +w.
\label{45}\end{equation}
Thus $\overline w^*$ is decreasing in $\rho$ if $dx^*/d\rho$ is negative. Using \eqref{44},
\begin{equation}
\an{
{dx^*\over d\rho} &=- {(1-\rho)(W-w)\over 2(2-\rho)Z} -{W+w-Z\over 2(2-\rho)(W-w)}\\
& = {x^*-{1\over 2}\over 2-\rho} -{(1-\rho)(W-w)\over 2(2-\rho)Z}.}
\label{46}\end{equation}
From \eqref{36} $0<x^*\le{1\over 2}$ and therefore $dx^*/d\rho <0$, which completes the proof.

Figure 4 illustrates the increase over time of the geometric mean fitness with decreasing $\rho$ at a polymorphic equilibrium in the $A1B1$ regime.
The values of $\hat\rho$ and $\rho^*$ are the same in $A1B1$ and $A2B2$ regimes, namely both are zero.
Figure 1B shows the geometric mean fitness in the $A1B2$ regime, and we see that for small values of $w$, this mean fitness increases as $\rho$ decreases.
At $w=0.1$, Table 1 shows that $\hat\rho$ is 0.00065, whereas $\rho^*=0.821$.
In all $AkBk$ regimes that we tested with $w=0.9$, the value of $\hat\rho$ was also zero, substantially different from the values of $\rho^*$, as shown in Figures 6A and 6B.
Also in Figure 6A, we see that with $w=0.1$, $\hat\rho$ changes from zero to positive in the $AkBk$ regimes with $k \ge 12$, while with $w=0.5$, the change occurs at $k=31$.
In Figure 6, with $w=0.1$, $\hat\rho$ is between 0.15 and 0.24 for $12\le k\le 50$, while with $w=0.5$, $\hat\rho$ exceeds 0.2 for $31\le k\le 50$.
More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Table 1, Supplementary Figures S7 and S8, and Supplemental Material SP7.

\begin{table}[h]
\centering
\caption{Values of $\rho^*$ (stable $\rho$)$^{\hbox{\sevenrm (a)}}$ and $\hat\rho$ (``optimal'' $\rho$)$^{\hbox{\sevenrm (b)}}$}
\begin{tabular}{rrcll}
$k$ & $l$ & $w^{\dagger}$ & $\rho^*$ & $\hat\rho$ \\
\midrule
 1	&	1	&	0.1		&	0.000000	&	0.000000	\cr
 1	&	1	&	0.5		&	0.000000	&	0.000000	\cr
 1	&	1	&	0.9		&	0.000000	&	0.000000	\cr
 1	&	2	&	0.1		&	0.000000	&	0.00065		\cr
 1	&	2	&	0.5		&	fixation	&	fixation	\cr
 1	&	2	&	0.9		&	fixation	&	fixation	\cr
 2	&	2	&	0.1		&	0.000000	&	0.000000	\cr
 2	&	2	&	0.5		&	0.000000	&	0.000000	\cr
 2	&	2	&	0.9		&	0.000000	&	0.000000	\cr
 3	&	10	&	0.1		&	>0.4265		&	0.00031		\cr
 3	&	10	&	0.5		&	fixation	&	fixation	\cr
 3	&	10	&	0.9		&	fixation	&	fixation	\cr
 5	&	30	&	0.1		&	>0.1489		&	0.00027     \cr
 5	&	30	&	0.5		&	fixation	&	fixation	\cr
 5	&	30	&	0.9		&	fixation	&	fixation	\cr
 12	&	12	&	0.1	    &	0.84924		&	0.24347		\cr
 12	&	12	&	0.5		&	0.91209		&	0.000000	\cr
 12	&	12	&	0.9		&	0.95686		&	0.000000	\cr
 20	&	20	&	0.1	    &				&	0.223925	\cr
 20	&	20	&	0.5		&	0.94643		&	0.000000	\cr
 20	&	20	&	0.9		&	0.98304		&	0.000000	\cr
 30	&	30	&	0.1	    &				&	0.193280	\cr
 30	&	30	&	0.5		&	0.96331		&	0.000000	\cr
 30	&	30	&	0.9		&	0.99136		&	0.000000	\cr
 50	&	50	&	0.1	    &				&	0.15419		\cr
 50	&	50	&	0.5		&	0.9768		&	0.22107		\cr
 50	&	50	&	0.9		&	0.99581		&	0.000000	\cr
\bottomrule
\end{tabular}

\addtabletext{\noindent $^\dagger$Note: $W=1$.\hfill\par
\noindent (a): $\rho^*$ is the uninvadable value of the vertical transmission rate.\hfill\par
(b): $\hat\rho$ is the geometric mean value of $\rho$ at the stable equilibrium of the $AkBl$ cycle.\hfill}
\end{table}

%Figure 6
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/AkBk_stable_optimal_rate.pdf}
\caption{Fitness ``optimal'' and evolutionary stable vertical transmission rate in $AkBk$ selection regime. 
\textbf{(A)} The vertical transmission rate $\hat{\rho}$ that maximized the geometric average of the population mean fitness is zero (complete oblique transmission) when selection cycles quickly between favoring phenotype $A$ and $B$, and then abruptly transitions to $\approx 0.2$, followed by a slow decrease (see Fig. S5 and S6 for details on the abrupt transition).
\textbf{(B)} The evolutionary stable rate $\rho^*$, which cannot be invaded by modifiers with either higher or lower vertical transmission rate $P$, rapidly increases from zero when selection cycles are short ($k=1$ or $2$) to roughly 1 when selection cycles are longer.
The dashed line shows $1-\frac{1}{k-1}$, which fit the values for $w=0.5$ (Carja et al. 2011). The values for $w=0.1$ (blue) could not be calculated for $k > 19$ due to numerical instability when selection is strong and the duration between selection fluctuations is long.
In all cases, $W=1$. See \emph{Supplemental Material SP6} for details on how we calculated the stable rate.}\label{fig:AkBk_stable_optimal_rate}
\end{figure}

\section{Finite Population Size}

 In order to include the effect of random drift due to finite population in the above deterministic model, we use the Wright-Fisher model. Let $X_t$ denote the number of individuals with phenotype $A$ in a population of fixed size $N$ at the $t$-th generation, and suppose $X_t=Nx$. Also, let $x'$ represent the frequency of the phenotype $A$ in the infinite population model in the next generation, namely (see \eqref{2}),
 \begin{equation}
 x' = \rho {w_A \over \overline w} x +(1-\rho)x.
 \label{47}\end{equation}
 Then, according to the Wright-Fisher model \cite{ewens2004mathematical}, $X_{t+1}$, the number of individuals of phenotype $A$ at generation $(t+1)$, is determined by the probability
 \begin{equation}
 P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\matrix{N\cr j\cr}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},
 \label{48}\end{equation}
 for $j=0,1,2,\dots,N$.
 Thus the fluctuations in the numbers of phenotypes $A$ and $B$ in the  population of size $N$ are generated by the Wright-Fisher Markov chain  where, given that $X_t=Nx$, $X_{t+1}$ has a binomial distribution with parameters $(N,x')$.

 This Markov chain  has two absorbing states, $X_t =N$ and $X_t=0$, corresponding to the two fixations in $A$ and $B$, respectively, and we are interested in the fixation probabilities and the time to fixation of these two absorbing states as functions of the initial frequency $x$ and also of $\rho$, $w_A$, and $w_B$.
 To these ends we use a diffusion approximation to the process $\{X_t\}$, which
% Specifically, we compute the mean $\mu(x)$ and the variance $\sigma^2(x)$ of the change in one generation  in the frequency of phenotype $A$ given that at the beginning of the generation $X_t=Nx$.
% 
% To compute $\mu(x)$, observe that by \eqref{47}   
% \begin{equation}
% \an{
% x' -x &={w_Ax\over\overline w}\rho +(1-\rho)x -x =\rho x\left[{w_A\over \overline w}-1\right]\cr
% &= \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},}
% \label{49}\end{equation}
% since $\overline w =w_A x +w_B(1-x)$.
% For the diffusion approximation, it is essential that the differential selection does not have a large effect per individual in each time period  $\Delta t$ $\bigl(\Delta t\simeq{1\over N}\bigr)$. That is, we assume
% \begin{equation}
% w_A -w_B ={s\over N}.
% \label{50}\end{equation}
% Then
% \begin{equation}
% x'-x\simeq {1\over N}\rho sx(1-x)
% \label{51}\end{equation}
% up to terms of order small than ${1\over N}$. Since one generation corresponds to $\Delta t\simeq {1\over N}$, we  conclude that
% \begin{equation}
% \mu(x) =\rho sx(1-x);\quad 0\le x\le 1.
% \label{52}\end{equation}
% In the same way, we can compute $\sigma^2(x)$, which is $x(1-x)$. These values of $\mu(x)$ and $\sigma^2(x)$ 
 allows us to compute $u(x)$, the probability that phenotype $A$ goes to fixation when it's initial frequency is $x$, namely
  \begin{equation}
  u(x) ={1-e^{-2\rho sx}\over 1-e^{-2\rho s}}.
 \label{53} \end{equation}
 The expected time to fixation in $A$ starting from a initial frequency of $x$ is given by
 \begin{equation}
 T(x)={1-u(x)\over\rho s}\int_0^x{e^{2\rho s\xi}-1\over \xi(1-\xi)}d\xi +{u(x)\over\rho s}\int_x^1{1-e^{-2\rho s(1-\xi)}\over \xi(1-\xi)}d\xi,
 \label{54}\end{equation}
 where $u(x)$ is given in \eqref{53}, and in generations, $T(x)$ is multiplied by $N$ (the derivation is in Supplementary Material SPX).
  Unfortunately the integrals in \eqref{54} cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ as Ewens points out on p.\ 160 \cite{ewens2004mathematical}, and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.

  It is important to note that the fixation probability $u(x)$ is a monotone increasing function of $\rho$ when $s>0$. In fact we have the following result.
 \medskip

\noindent{\bf Result 10}. {\sl When $s>0$ so that the phenotype $A$ is favored, the fixation probability $u(x)$ is monotone increasing in $\rho$.}\par
 \medskip

 The proof of Result 10 is in Supplementary Material SP5.  
  Supplementary Figure S9 compares the fixation probability and time to fixation derived numerically from  simulating the Wright-Fisher Markov chain with the diffusion-derived values of $u(x)$ and $T(x)$. The fit is seen to be very good. Note that when $N$ is large, the Wright-Fisher model exhibits persistent fluctuation around the deterministic expectation, as shown by the orange traces in Supplemental Figure S2.

 We can also develop a diffusion approximation for the case of a cycling environment. Suppose that selection changes in cycles of length $n$ such that within the cycle the fitness parameters are $w_A^t$, $w_B^t$ for $t=1,2,\dots,n$. Also let
 \begin{equation}
 {1\over N}s_t \simeq w_A^t -w_B^t,\qquad S_t =\sum_{i=1}^t s_i,\qquad t=1,2,\dots,n.
 \label{55}\end{equation}
 Following Karlin and Levikson \cite{karlin1974temporal} we have the following result.
 \medskip

 \noindent{\bf Result 11}. {\sl The mean $\mu(x)$ and variance $\sigma^2(x)$ of the change in the frequency of $A$ in one generation for the diffusion approximation in the case of a cycling environment $AkBl$, where $k+l=n$, are
 \begin{equation}
 \an{
 \mu(x) &= \rho S_n x(1-x)\cr \sigma^2(x) &= nx(1-x).}
 \label{56}\end{equation}}

\noindent The proof of Result 11, based on induction on $n$, is given in Supplemental Materials SP6.

 Using the moments in \eqref{56}, the fixation probability $u(x)$ and the expected time $T(x)$ to fixation from an initial frequency of $x$ can be computed where $s$ is replaced by $s_n/n$. We find
 \begin{equation}
 u(x) ={1-e^{-2\rho{S_n\over n}x}\over 1-e^{-2\rho{S_n\over n}}},
\label{57} \end{equation}
 and $T(x)$ can be computed similarly.

 In the  case of the $AkBl$ cycling environment, we write $n=k+l$, and if $w_A=W$, $w_B=w$ for $k$ generations, and $w_A=w$, $w_B=W$ for $l$ generations, we have
 \begin{equation}
 S_n=S_{k+l} =N(k-l)(W-w).
 \label{58}\end{equation}
 Figure 7 shows an example of how $(k-l)$, $\rho$, and $(W-w)$, which enter the formula for $u(x)$ in \eqref{57}, interact to affect fixation probabilities. More examples are illustrated in Supplementary Figure S11.

% Figure 7
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/lk_fix_prob_C.pdf}
\caption{Fixation in a finite population with different ratios of selection periods \(\frac{k}{l}\). Fixation probability of phenotype $A$ when starting with a single copy in a population of size $N$: $u(1/N) = (1-\exp(-2 \rho \frac{k-l}{k+l}(W-w))/(1-\exp(-2 N \rho \frac{k-l}{k+l}(W-w))$ (see \eqref{57} and \eqref{58}).
\emph{k} and \emph{l} are the number of
generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by
selection. Here, fitness of the favored phenotype is $W = 1$, fitness of the unfavored phenotype is $w=0.5$, and the population size is \(N=10,000\). 
See Figure S11 for additional examples.} \label{lk_fix_prob_C}
\end{figure}

 \section*{Discussion}

Non-chromosomal modes of phenotypic transmission are receiving increasing attention \cite{whiten2017extension,jaenisch2003epigenetic,allis2016molecular}, especially with respect to their potential role in adaptation and maintenance of diversity \cite{rivoire2014model}. Here we have focused on a dichotomous phenotype transmitted through a combination of parental and non-parental transmission. In addition to the roles that these transmission modes play in the dynamics of phenotypic diversity in large and small populations, we have also investigated a genetic model for the evolution of the transmission mode itself.

Our model differs markedly from that of Xue and Leibler \cite{xue2016evolutionary}, who took the individual phenotypic distribution (i.e., the probability that an individual develops one of a set of phenotypes) to be the inherited trait. In our model, the transmitted trait is the phenotype itself. Thus, with two phenotypic states $A$ and $B$, we track the frequency $x$ of $A$, whereas Xue and Leibler focus on the dynamics of the per-individual probability $\pi_A$ of learning the phenotype $A$. One interpretation of our model is as a mean-value approximation to the model of Xue and Leibler, where $x$, the state in our model, is the average of the population distribution of individual phenotype probabilities.

In a constant environment, the higher the vertical transmission rate $\rho$, the faster is the approach to fixation of the favored phenotype: $A$ if $w_A>w_B$ or $B$ if $w_B>w_A$. Here $1-\rho$, the oblique transmission rate, represents the added chance that an offspring becomes $A$ by learning from the parent's population after learning from its parents who have undergone selection (eq.\ 2). This simple phenotypic model does not allow a polymorphism to be achieved in a constant environment, but with more oblique transmission, approach to fixation is retarded.

 With fluctuating environments, the dynamics of the phenotype frequencies are, in general, much more complicated. In particular, with deterministically cycling symmetric fitness values (the $AkBl$ model), it is impossible for fixation in $A$ and $B$ to both be stable. If $k=l$, for example, neither fixation is stable, and there is a single stable polymorphic equilibrium (with phenotypes $A$ and $B$ present; see Result 5). In the $A1B1$ case this  polymorphism is globally stable. In the $AkBl$ case, bounds on $l/k$ that determine the instability of both fixations, and hence the protection of polymorphism are given by the inequalities of \eqref{20}, which depend on both the fitness differences and the rate $\rho$ of vertical transmission. We conjecture that with $k\ne l$ there is a unique stable polymorphism when both fixations are unstable.
This result is similar to the \emph{storage effect} \cite{chesson1981}, in which protection from selection maintains species coexistence: Consider a two species $A$ and $B$, with overlapping generations, an equal death rate $\rho$, and different growth rates $w_A$ and $w_B$; then \eqref{2} will model the change in frequency of species $A$. In our model, oblique transmission can be said to protect the unfavored phenotype from selection because it allows transmission without reproduction.

 In deterministic one-locus, two-allele diploid population genetic models with cycling fitness regimes, Haldane and Jayakar \cite{haldane1963polymorphism} first showed the relevance of the geometric mean of genotypic fitnesses (cf.\ \eqref{16} and \eqref{17}) for the maintenance (or loss) of polymorphism. However, with equal homozygote fitnesses, which alternated in strength as a two-generation cycle (cf.\ $A1B1$), Karlin and Liberman \cite{karlin1975random} extended the Haldane-Jayakar results and found conditions under which both allelic fixations and polymorphic equilibrium could all be stable, with the evolution depending on initial allele frequencies, as well as the homozygote fitness differences between alternate generations. Our haploid model appears not to produce such dependence on the initial conditions.

 When the fitnesses $w_A$ and $w_B$ are treated as random variables, rather than varying cyclically, stochastic local stability is the appropriate analog to local stability in the case of cyclic fitness variation. While fixations in phenotypes $A$ and $B$ cannot both be stochastically stable in this case, both may be unstable and a polymorphic distribution may result. The variance of this distribution is greater for larger values of $\rho$. This is because the  stochastic local stability conditions involve $E\bigl\{\log[1-\rho+\rho(w_A/w_B)]\bigr\}$ and the effect of the variance of $(w_A/w_B)$ will clearly increase as $\rho$ increases. In the finite population case, a greater level of vertical transmission makes selection more effective, increasing the probability $u(x)$ of fixation and reducing the expected time to fixation.

 We have shown that in the $A1B1$ case, the rate of vertical transmission tends to decrease when it is under the control of a genetic modifier. From numerical iteration, it appears that this is also true in the random selection case, when the fitnesses of $A$ and $B$ are identically distributed and independent between generations. However, for $AkBl$ selection regimes more complicated than $A1B1$, evolution of a modifier of vertical transmission is not straightforward. While reduction of $\rho$ occurs in the $A2B2$ regime, the uninvadable value $\rho^*$ is not zero for all the fitness values explored in $AkBk$ regimes with $k>2$ (Figure 6B, Table 1). In fact, $\rho^*$ increases sharply as $k$ increases beyond $k=2$. This is an unusual scenario for genetic modifiers, although it must be noted that a modifier of $\rho$ is not neutral; it affects primary selection, while neutral  modifiers of recombination, mutation, and migration affect induced or secondary selection.
To explain the marked difference in the value of the stable vertical transmission rate consider the probability $q$ that offspring receive phenotype $A$ when it is about to be the favored phenotype, and for simplicity, assume strong selection, such that parents are likely to be of the phenotype favored in the parental generation, a periodically changing environment $AkBl$, and a population at equilibrium.
With vertical transmission, $q_v=\frac{k-1}{k}$.
With oblique transmission, $q_o=x$ the frequency of the favored phenotype in the parental generation.
For $k=l$, the frequency of $A$ cycles around 0.5 so $q_o \approx 0.5$, whereas $q_v=0, {1 \over 2}, {2 \over 3}, {3 \over 4}, \ldots$ for $k=1, 2, 3, 4, \ldots$. Therefore, for $k>2$ vertical transmission increases $q$ whereas for $k=1$ and $2$, oblique transmission increases $q$.

 The dependence of the modifier dynamics on the strength of selection (that is $w$ when $W=1$) is complicated by the approach of the system to fixation. When the phenotype frequencies become exceedingly small, dependence of the dynamics of the modifier of $\rho$ becomes extremely difficult to detect due to numerical instability; this is especially true for larger values of $k$ in $AkBk$ regimes when $w$ is small (Table 1, Figure 6B with $k\ge 19$).

 Figure 6A (see also Table 1) shows that the value $\hat\rho$ that maximizes the geometric mean fitness is the same as the evolutionarily stable value $\rho^*$ in the $A1B1$ and $A2B2$ selection regimes. For $AkBk$ regimes with $k>2$, our numerical analysis shows the $\hat\rho$ depends strongly on the strength of selection; i.e., the value of $w$ relative to $W=1$. For $AkBl$ regimes with $w=0.1$, the difference between $\hat\rho$ and $\rho^*$ is seen even with the $A1B2$ environment. For $AkBk$ regimes and $w=0.9$, we find $\hat\rho=0$, while $\rho^*$ is close to $0.9$. For larger values of $k$, $\hat\rho$ lies between 0.15 and 0.25, while $\rho^*$ remains above 0.8 and can reach 0.99 for very large $k$. Comparing Figure 6A with the asymptotic growth rate (AGR) of Xue and Leibler \cite{xue2016evolutionary}, whose parameter $\eta$ is the rate at which an individual learns from its parental lineage, there is a similarity for our curves for $w=0.1$ and $0.5$ with their curve in the $AkBk$ environment. They show the AGR decreasing with $\eta$ in the $AkBk$ regime for small $k$, but larger values of $k$ entail that the AGR has a maximum for an intermediate value of $\eta$.

 Although the models of Xue and Leibler \cite{xue2016evolutionary} and that analyzed here both incorporate parental and non-parental transmission, they do so in qualitatively different ways. The model treated in the present paper is squarely in the tradition of gene-culture coevolutionary theory, together with modifier theory from population genetics. The different findings from the two classes of models are interesting and suggest that further exploration of the overlaps and discrepancies between the two approaches would be worthwhile.

\acknow{This manuscript was supported in part by the Stanford Center for Computational, Evolutionary and Human Genomics, and by the Morrison Institute for Population and Resources Studies, Stanford University.}

%\section*{Guide to using this template on Overleaf}

%Please note that whilst this template provides a preview of the typeset manuscript for submission, to help in this preparation, it will not necessarily be the final publication layout. For more detailed information please see the \href{http://www.pnas.org/site/authors/format.xhtml}{PNAS Information for Authors}.

%If you have a question while using this template on Overleaf, please use the help menu (``?'') on the top bar to search for \href{https://www.overleaf.com/help}{help and tutorials}. You can also \href{https://www.overleaf.com/contact}{contact the Overleaf support team} at any time with specific questions about your manuscript or feedback on the template.

%\subsection*{Author Affiliations}

%Include department, institution, and complete address, with the ZIP/postal code, for each author. Use lower case letters to match authors with institutions, as shown in the example. Authors with an ORCID ID may supply this information at submission.

%\subsection*{Submitting Manuscripts}

%All authors must submit their articles at \href{http://www.pnascentral.org/cgi-bin/main.plex}{PNAScentral}. If you are using Overleaf to write your article, you can use the ``Submit to PNAS'' option in the top bar of the editor window.

%\subsection*{Format}

%Many authors find it useful to organize their manuscripts with the following order of sections;  Title, Author Affiliation, Keywords, Abstract, Significance Statement, Results, Discussion, Materials and methods, Acknowledgments, and References. Other orders and headings are permitted.

%\subsection*{Manuscript Length}

%PNAS generally uses a two-column format averaging 67 characters, including spaces, per line. The maximum length of a Direct Submission research article is six pages and a PNAS PLUS research article is ten pages including all text, spaces, and the number of characters displaced by figures, tables, and equations.  When submitting tables, figures, and/or equations in addition to text, keep the text for your manuscript under 39,000 characters (including spaces) for Direct Submissions and 72,000 characters (including spaces) for PNAS PLUS.

%\subsection*{References}

%References should be cited in numerical order as they appear in text; this will be done automatically via bibtex, e.g. \cite{belkin2002using} and \cite{berard1994embedding,coifman2005geometric}. All references, including for the SI, should be included in the main manuscript file. References appearing in both sections should not be duplicated.  SI references included in tables should be included with the main reference section.

%\subsection*{Data Archival}

%PNAS must be able to archive the data essential to a published article. Where such archiving is not possible, deposition of data in public databases, such as GenBank, ArrayExpress, Protein Data Bank, Unidata, and others outlined in the Information for Authors, is acceptable.

%\subsection*{Language-Editing Services}
%Prior to submission, authors who believe their manuscripts would benefit from professional editing are encouraged to use a language-editing service (see list at www.pnas.org/site/authors/language-editing.xhtml). PNAS does not take responsibility for or endorse these services, and their use has no bearing on acceptance of a manuscript for publication.

%\begin{figure}%[tbhp]
%\centering
%\includegraphics[width=.8\linewidth]{frog}
%\caption{Placeholder image of a frog with a long example caption to show justification setting.}
%\label{fig:frog}
%\end{figure}

%\subsection*{Digital Figures}
%\label{sec:figures}

%Only TIFF, EPS, and high-resolution PDF for Mac or PC are allowed for figures that will appear in the main text, and images must be final size. Authors may submit U3D or PRC files for 3D images; these must be accompanied by 2D representations in TIFF, EPS, or high-resolution PDF format.  Color images must be in RGB (red, green, blue) mode. Include the font files for any text.

%Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.

%Figure \ref{fig:frog} shows an example of how to insert a column-wide figure. To insert a figure wider than one column, please use the \verb|\begin{figure*}...\end{figure*}| environment. Figures wider than one column should be sized to 11.4 cm or 17.8 cm wide.

%\subsection*{Single column equations}

%Authors may use 1- or 2-column equations in their article, according to their preference.

%To allow an equation to span both columns, options are to use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures, or to use the \verb|\begin{widetext}...\end{widetext}| environment as shown in equation \ref{eqn:example} below.

%Please note that this option may run into problems with floats and footnotes, as mentioned in the \href{http://texdoc.net/pkg/cuted}{cuted package documentation}. In the case of problems with footnotes, it may be possible to correct the situation using commands \verb|\footnotemark| and \verb|\footnotetext|.

%% Do not use widetext if paper is in single column.
%\begin{widetext}
%\begin{align*}
%(x+y)^3&=(x+y)(x+y)^2\\
%       &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
%       &=x^3+3x^2y+3xy^3+x^3. 
%\end{align*}
%\end{widetext}

%\begin{table}%[tbhp]
%\centering
%\caption{Comparison of the fitted potential energy surfaces and ab initio benchmark electronic energy calculations}
%\begin{tabular}{lrrr}
%Species & CBS & CV & G3 \\
%\midrule
%1. Acetaldehyde & 0.0 & 0.0 & 0.0 \\
%2. Vinyl alcohol & 9.1 & 9.6 & 13.5 \\
%3. Hydroxyethylidene & 50.8 & 51.2 & 54.0\\
%\bottomrule
%\end{tabular}

%\addtabletext{nomenclature for the TSs refers to the numbered species in the table.}
%\end{table}

%\subsection*{Supporting Information (SI)}

%The main text of the paper must stand on its own without the SI. Refer to SI in the manuscript at an appropriate point in the text. Number supporting figures and tables starting with S1, S2, etc. Authors are limited to no more than 10 SI files, not including movie files. Authors who place detailed materials and methods in SI must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the online methods. If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text. Because PNAS edits SI and composes it into a single PDF, authors must provide the following file formats only.

%\subsubsection*{SI Text}

%Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference).

%\subsubsection*{SI Figures}

%Provide a brief legend for each supporting figure after the supporting text. Provide figure images in TIFF, EPS, high-resolution PDF, JPEG, or GIF format; figures may not be embedded in manuscript text. When saving TIFF files, use only LZW compression; do not use JPEG compression. Do not save figure numbers, legends, or author names as part of the image. Composite figures must be pre-assembled.

%\subsubsection*{3D Figures}

%Supply a composable U3D or PRC file so that it may be edited and composed. Authors may submit a PDF file but please note it will be published in raw format and will not be edited or composed.

%\subsubsection*{SI Tables}

%Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference); include only one table per file. Do not use tabs or spaces to separate columns in Word tables.

%\subsubsection*{SI Datasets}

%Supply Excel (.xls), RTF, or PDF files. This file type will be published in raw format and will not be edited or composed.

%\subsubsection*{SI Movies}

%Supply Audio Video Interleave (avi), Quicktime (mov), Windows Media (wmv), animated GIF (gif), or MPEG files and submit a brief legend for each movie in a Word or RTF file. All movies should be submitted at the desired reproduction size and length. Movies should be no more than 10 MB in size.

%\subsubsection*{Still images}

%Authors must provide a still image from each video file. Supply TIFF, EPS, high-resolution PDF, JPEG, or GIF files.

%\subsubsection*{Appendices}

%PNAS prefers that authors submit individual source files to ensure readability. If this is not possible, supply a single PDF file that contains all of the SI associated with the paper. This file type will be published in raw format and will not be edited or composed.

%\matmethods{Please describe your materials and methods here. This can be more than one paragraph, and may contain subsections and equations as required. Authors should include a statement in the methods section describing how readers will be able to access the data in the paper.

%\subsection*{Subsection for Method}
%Example text for subsection.
%}

%\showmatmethods % Display the Materials and Methods section

%\acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}

%\showacknow % Display the acknowledgments section

% \pnasbreak splits and balances the columns before the references.
% If you see unexpected formatting errors, try commenting out this line
% as it can run into problems with floats and footnotes on the final page.
%\pnasbreak

% Bibliography
\bibliography{ramrefs}

\showacknow

\end{document}

### 2017-12-11 ms/PNAS/submission.tex
pnas branch, middle of revision

--- 2017-12-11 ms/PNAS/revision.tex
@@
 \begin{abstract}
-Conditions for the evolution and maintenance of social learning, in competition with individual learning, under fluctuating selection, is a well-studied problem in the theory of cultural evolution.
-Here we study competition between vertical and oblique cultural transmission of a dichotomous phenotype under constant, periodically cycling, and randomly fluctuating selection.
+Vertical and oblique cultural transmission of a dichotomous phenotype is studied under constant, periodic cycling, and randomly fluctuating selection.
 Conditions are derived for the existence of a stable polymorphism in a periodically cycling selection regime.
@@
 Similarly, some phenotypes are determined by genes that are commonly converted by uptake of foreign DNA, i.e., transformation \cite{milkman1990molecular}.
-In these cases, inheritance of a phenotype may combine vertical transmission from the parent cell, and oblique transmission from other cells, even if the mechanism for oblique transmission evolved to facilitate some other function \cite{redfield1993evolution}.
+In these cases, inheritance of a phenotype may combine vertical transmission from the parent cell, and oblique transmission from other cells, even if the latter did not originally evolve for that purpose \cite{redfield1993evolution}.
 
@@
 
-The linear approximation of $F(x)$ near $x^*=0$ (\eqref{11}) does not depend on the order in which phenotypes $A$ and $B$ are favored within a cycle of $k+l$ generations.
-Therefore, the local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations, $A$ is favored $k$ times and $B$ is favored $l$ times, and not their order in the cycle.
+The local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations $A$ is favored $k$ times and $B$ is favored $l$ times, and not their order in the cycle.
 When neither fixation in $A$ or $B$ is stable, there is a protected polymorphism, and we expect to have one or more polymorphic equilibria.
@@
 
-% Figure 1
 \begin{figure}[h]
 \centering
-\includegraphics[width=\linewidth]{../../figures/A1B2_polymorphism_vert.pdf}
+\includegraphics[width=.8\linewidth]{../pnas_submission/A1B2_polymorphism.pdf}
 \caption{Stable frequency of phenotype \emph{A} and geometric mean fitness in selection regime
@@
 In all cases, \emph{W=1}.}\label{fig:A1B2_polymorphism}
+%\label{fig:frog}
 \end{figure}
@@
 \centering
-\includegraphics[width=\linewidth]{../../figures/stochastic_env_x_t.pdf}
+\includegraphics[width=.8\linewidth]{../pnas_submission/stochastic_env_x_t.pdf}
 \caption{Stochastic local stability. 
@@
 \centering
-\includegraphics[width=\linewidth]{../../figures/stoch_two_iid_rvs.pdf}
+\includegraphics[width=.8\linewidth]{../pnas_submission/stoch_two_iid_rvs.pdf}
 \caption{Effect of vertical transmission rate $\rho$ on phenotype polymorphism in a randomly changing environment.
@@
 Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} \cite{feldman1986evolutionary,altenberg2017unified} to invasion by allele $M$.
-A constant environment always leads to fixation of the favored type, independent of $\rho$.
+Constant environment always leads to fixation of the favored type, independent of $\rho$.
 We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$ (but see Supplementary Material SP6 for computational analysis of the general $AkBl$ case).
@@
 \label{34}\end{equation}
-where the non-linear transformation $\underline x'=T_1\underline x$ is given by \eqref{32} with $w_A=W$, $w_B=w$, and the non-linear transformation $\underline x''=T_2\underline x'$ is given by \eqref{32} with $w_A=w$, $w_B=W$.
+where $\underline x'=T_1\underline x$ is given by \eqref{32} with $w_A=W$, $w_B=w$, and $\underline x''=T_2\underline x'$ is given by [32] with $w_A=w$, $w_B=W$.
 Here $\underline x$, $\underline x'$, $\underline x''$ are the frequency vectors.
@@
  %Figure 4
-\begin{figure*}[h]
+\begin{figure}[h]
 \centering
-\includegraphics[width=\linewidth]{../../figures/{A1B1_modifier_invasions_w0.5}.pdf}
-\caption{Consecutive fixation of modifiers that reduce the vertical transmission rate in selection regime \emph{A1B1}.
-The figure shows results of numerical simulations of evolution with two modifier alleles (\eqref{32}).
-When a modifier allele fixes (frequency\textgreater{}99.9\%), a new modifier allele is introduced with a vertical transmission rate one order of magnitude lower (vertical dashed lines).
-\textbf{(A)} The frequency of phenotype \emph{A} in the population over time.
-\textbf{(B)} The frequency of the invading modifier allele over time.
-\textbf{(C)} The population geometric mean fitness over time; insets zoom in to show that the mean fitness increases slightly with each invasion.
-Invading alleles are introduced at frequency 0.01\%; whenever their frequency drops below 0.01\% they are re-introduced.
-Parameters: vertical transmission rate of the initial resident modifier allele, \(\rho_0 =0.1\); fitness values: $W=1$; $w=0.5$.
-The x-axis is on a log-scale, as each sequential invasion takes an order of magnitude longer to complete.
-See Figure S12 for $w=0.1$ and $0.9$.}
-\label{fig:A1B1_modifier_invasions}
-\end{figure*}
+\includegraphics[width=.8\linewidth]{../pnas_submission/A1B1_modifier_invasions.pdf}
+\caption{Consecutive fixation of modifiers that reduce the vertical
+transmission rate in selection regime \emph{A1B1}. The figure shows
+results of numerical simulations of evolution with two modifier alleles
+(\eqref{32}). When a modifier allele fixes
+(frequency\textgreater{}99.9\%), a new modifier allele is introduced
+with a vertical transmission rate one order of magnitude lower (vertical
+dashed lines). \textbf{(A,D,G)} The frequency of phenotype \emph{A} in
+the population over time. \textbf{(B,E,H)} The frequency of the invading
+modifier allele over time. \textbf{(C,F,I)} The population geometric mean
+fitness over time; insets zoom in to show that the mean fitness 
+decreases slightly with each invasion. Parameters: vertical transmission rate of
+the initial resident modifier allele, \(\rho_0 =0.1\); fitness values:
+$W=1$; $w=0.1$ (\textbf{A-C}), 0.5 (\textbf{D-F}), and 0.9
+(\textbf{G-I}). The x-axis is on a log-scale, as each sequential invasion
+takes an order of magnitude longer to
+complete.}\label{fig:A1B1_modifier_invasions}
+\end{figure}
 
@@
 \centering
-\includegraphics[width=\linewidth]{../../figures/stoch_modifier_invasions.pdf}
+\includegraphics[width=.8\linewidth]{../pnas_submission/stoch_modifier_invasions.pdf}
 \caption{Consecutive fixation of modifiers that reduce the vertical transmission rate $\rho$ under symmetric randomly changing selection.
@@
 \textbf{(B)} The frequency of the invading modifier allele over time.
-Invading alleles are introduced at frequency 0.01\%; whenever their frequency drops below 0.01\% they are re-introduced.
 Parameters: vertical transmission rate of the initial resident modifier allele is \(\rho_0 =0.1\) and the ratio of fitness values  is $w_A/w_B=10$ with probability 0.5 and $w_A/w_B=0.1$ also with probability 0.5.
@@
 
-Values of the evolutionarily stable vertical transmission rate, $\rho^*$, for some $AkBl$ examples (see Supplementary Material SP7 and  Figure S10 for analytical details) are recorded in Table 1 for different values of $w$ relative to $W=1$.
+Values of $\rho^*$ for some $AkBl$ examples (see Supplementary Material SP7 and  Figure S10 for analytical details) are recorded in Table 1 for different values of $w$ relative to $W=1$.
 Interestingly, for the $A1B2$ regime, with $w=0.1$, the evolutionarily stable value of $\rho$ is substantially greater than zero, as it is in the $A3B10$ and $A5B30$ regimes.
@@
 For the evolution of mutation rates that are controlled by genetic modifiers, the stable mutation rate and the mutation rate that maximizes the geometric mean fitness of the population appear to be the same when the period of environmental fluctuation is low enough \cite{liberman2011evolution}.
-We can ask the same question here: is the stable rate $\rho^*$ the same as the rate $\hat\rho$ that maximizes the equilibrium value of the geometric mean fitness under fluctuating selection?
-First consider the $A1B1$ selection regime. We have the following result.
+We can ask the same question here: is the stable rate $\rho^*$ the same as the rate $\hat\rho$ that maximizes the equilibrium value of the geometric mean fitness under fluctuating selection.
+First consider the $A1B1$ selection regime.
+We have the following result.
 \medskip
@@
 
-Figure 4 illustrates the increase over time of the geometric mean fitness with decreasing $\rho$ at a polymorphic equilibrium in the $A1B1$ regime.
+  Figure 4 illustrates the decrease over time of the geometric mean fitness with increasing $\rho$ at a polymorphic equilibrium in the $A1B1$ regime.
 The values of $\hat\rho$ and $\rho^*$ are the same in $A1B1$ and $A2B2$ regimes, namely both are zero.
-Figure 1B shows the geometric mean fitness in the $A1B2$ regime, and we see that for small values of $w$, this mean fitness increases as $\rho$ decreases.
+Figure 1B shows the geometric mean fitness in the $A1B2$ regime, and we see that for small values of $w$, this mean fitness decreases as $\rho$ increases.
 At $w=0.1$, Table 1 shows that $\hat\rho$ is 0.00065, whereas $\rho^*=0.821$.
@@
 
-\begin{table}[h]
+\begin{table}%[h]
 \centering
@@
 \begin{tabular}{rrcll}
-$k$ & $l$ & $w^{\dagger}$ & $\rho^*$ & $\hat\rho$ \\
+$k$&$l$&$w^{\dagger}$&$\hat\rho$&$\rho^*$ \\
 \midrule
- 1	&	1	&	0.1		&	0.000000	&	0.000000	\cr
- 1	&	1	&	0.5		&	0.000000	&	0.000000	\cr
- 1	&	1	&	0.9		&	0.000000	&	0.000000	\cr
- 1	&	2	&	0.1		&	0.000000	&	0.00065		\cr
- 1	&	2	&	0.5		&	fixation	&	fixation	\cr
- 1	&	2	&	0.9		&	fixation	&	fixation	\cr
- 2	&	2	&	0.1		&	0.000000	&	0.000000	\cr
- 2	&	2	&	0.5		&	0.000000	&	0.000000	\cr
- 2	&	2	&	0.9		&	0.000000	&	0.000000	\cr
- 3	&	10	&	0.1		&	>0.4265		&	0.00031		\cr
- 3	&	10	&	0.5		&	fixation	&	fixation	\cr
- 3	&	10	&	0.9		&	fixation	&	fixation	\cr
- 5	&	30	&	0.1		&	>0.1489		&	0.00027     \cr
- 5	&	30	&	0.5		&	fixation	&	fixation	\cr
- 5	&	30	&	0.9		&	fixation	&	fixation	\cr
- 12	&	12	&	0.1	    &	0.84924		&	0.24347		\cr
- 12	&	12	&	0.5		&	0.91209		&	0.000000	\cr
- 12	&	12	&	0.9		&	0.95686		&	0.000000	\cr
- 20	&	20	&	0.1	    &				&	0.223925	\cr
- 20	&	20	&	0.5		&	0.94643		&	0.000000	\cr
- 20	&	20	&	0.9		&	0.98304		&	0.000000	\cr
- 30	&	30	&	0.1	    &				&	0.193280	\cr
- 30	&	30	&	0.5		&	0.96331		&	0.000000	\cr
- 30	&	30	&	0.9		&	0.99136		&	0.000000	\cr
- 50	&	50	&	0.1	    &				&	0.15419		\cr
- 50	&	50	&	0.5		&	0.9768		&	0.22107		\cr
- 50	&	50	&	0.9		&	0.99581		&	0.000000	\cr
+1&1&0.1&0.0000000&0.0000000\cr
+ 1&1&0.5&0.0000000&0.0000000\cr
+ 1&1&0.9&0.0000000&0.0000000\cr
+ 1&2&0.1&0.00065&0.821\cr
+ 1&2&0.5&fixation&fixation\cr
+ 1&2&0.9&fixation&fixation\cr
+ 2&2&0.1&0.0000000&0.0000000\cr
+ 2&2&0.5&0.0000000&0.0000000\cr
+ 2&2&0.9&0.0000000&0.0000000\cr
+ 3&10&0.1&0.00031&0.619\cr
+ 3&10&0.5&fixation&fixation\cr
+ 3&10&0.9&fixation&fixation\cr
+ 5&30&0.1&0.00027&0.6503\cr
+ 5&30&0.5&fixation&fixation\cr
+ 5&30&0.9&fixation&fixation\cr
+ 12&12&0.1&0.24347&0.84924\cr
+ 12&12&0.5&0.0000000&0.91209\cr
+ 12&12&0.9&0.0000000&0.95686\cr
+ 20&20&0.1&0.223925\cr
+ 20&20&0.5&0.0000000&0.94643\cr
+ 20&20&0.9&0.0000000&0.98304\cr
+ 30&30&0.1&0.1932801\cr
+ 30&30&0.5&0.0000000&0.96331\cr
+ 30&30&0.9&0.0000000&0.0000000\cr
+ 50&50&0.1&0.15419\cr
+ 50&50&0.5&0.22107&0.9768\cr
+ 50&50&0.9&0.0000000&0.99581\cr
 \bottomrule
@@
 \centering
-\includegraphics[width=\linewidth]{../../figures/AkBk_stable_optimal_rate.pdf}
+\includegraphics[width=.8\linewidth]{../pnas_submission/AkBk_stable_optimal_rate.pdf}
 \caption{Fitness ``optimal'' and evolutionary stable vertical transmission rate in $AkBk$ selection regime. 
@@
  \begin{equation}
- x' = \rho {w_A \over \overline w} x +(1-\rho)x.
+ x'={w_Ax\over\overline w}\rho +(1-\rho)x.
  \label{47}\end{equation}
@@
  This Markov chain  has two absorbing states, $X_t =N$ and $X_t=0$, corresponding to the two fixations in $A$ and $B$, respectively, and we are interested in the fixation probabilities and the time to fixation of these two absorbing states as functions of the initial frequency $x$ and also of $\rho$, $w_A$, and $w_B$.
- To these ends we use a diffusion approximation to the process $\{X_t\}$, which
-% Specifically, we compute the mean $\mu(x)$ and the variance $\sigma^2(x)$ of the change in one generation  in the frequency of phenotype $A$ given that at the beginning of the generation $X_t=Nx$.
-% 
-% To compute $\mu(x)$, observe that by \eqref{47}   
-% \begin{equation}
-% \an{
-% x' -x &={w_Ax\over\overline w}\rho +(1-\rho)x -x =\rho x\left[{w_A\over \overline w}-1\right]\cr
-% &= \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},}
-% \label{49}\end{equation}
-% since $\overline w =w_A x +w_B(1-x)$.
-% For the diffusion approximation, it is essential that the differential selection does not have a large effect per individual in each time period  $\Delta t$ $\bigl(\Delta t\simeq{1\over N}\bigr)$.
+ To these ends we use a diffusion approximation  to the process $\{X_t\}$.
+Specifically, we compute the mean $\mu(x)$ and the variance $\sigma^2(x)$ of the change in one generation  in the frequency of phenotype $A$ given that at the beginning of the generation $X_t=Nx$.
+
+ To compute $\mu(x)$, observe that by \eqref{47}   
+ \begin{equation}
+ \an{
+ x' -x &={w_Ax\over\overline w}\rho +(1-\rho)x -x =\rho x\left[{w_A\over \overline w}-1\right]\cr
+ &= \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},}
+ \label{49}\end{equation}
+ since $\overline w =w_A x +w_B(1-x)$.
+ For the diffusion approximation, it is essential that the differential selection does not have a large effect per individual in each time period  $\Delta t$ $\bigl(\Delta t\simeq{1\over N}\bigr)$.
 That is, we assume
-% \begin{equation}
-% w_A -w_B ={s\over N}.
-% \label{50}\end{equation}
-% Then
-% \begin{equation}
-% x'-x\simeq {1\over N}\rho sx(1-x)
-% \label{51}\end{equation}
-% up to terms of order small than ${1\over N}$. Since one generation corresponds to $\Delta t\simeq {1\over N}$, we  conclude that
-% \begin{equation}
-% \mu(x) =\rho sx(1-x);\quad 0\le x\le 1.
-% \label{52}\end{equation}
-% In the same way, we can compute $\sigma^2(x)$, which is $x(1-x)$. These values of $\mu(x)$ and $\sigma^2(x)$ 
- allows us to compute $u(x)$, the probability that phenotype $A$ goes to fixation when it's initial frequency is $x$, namely
+ \begin{equation}
+ w_A -w_B ={s\over N}.
+ \label{50}\end{equation}
+ Then
+ \begin{equation}
+ x'-x\simeq {1\over N}\rho sx(1-x)
+ \label{51}\end{equation}
+ up to terms of order small than ${1\over N}$. Since one generation corresponds to $\Delta t\simeq {1\over N}$, we  conclude that
+ \begin{equation}
+ \mu(x) =\rho sx(1-x);\quad 0\le x\le 1.
+ \label{52}\end{equation}
+ In the same way, we can compute $\sigma^2(x)$, which is $x(1-x)$. These values of $\mu(x)$ and $\sigma^2(x)$ allow us to compute $u(x)$, namely
   \begin{equation}
@@
  \label{53} \end{equation}
- The expected time to fixation in $A$ starting from a initial frequency of $x$ is given by
+ The expected time to fixation in $A$ starting from a initial frequency of $x$ is is given by
  \begin{equation}
@@
  \label{54}\end{equation}
- where $u(x)$ is given in \eqref{53}, and in generations, $T(x)$ is multiplied by $N$ (the derivation is in Supplementary Material SPX).
+ where $u(x)$ is given in \eqref{53}, and in generations, $T(x)$ is multiplied by $N$.
   Unfortunately the integrals in \eqref{54} cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ as Ewens points out on p.\ 160 \cite{ewens2004mathematical}, and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.
@@
 The fit is seen to be very good.
-Note that when $N$ is large, the Wright-Fisher model exhibits persistent fluctuation around the deterministic expectation, as shown by the orange traces in Supplemental Figure S2.
+Note that when $N$ is large, the Wright-Fisher model exhibits persistent fluctuation around the deterministic expectation, as shown by the orange diagram in Supplemental Figure S2.
 
@@
 
-% Figure 7
+%New Figure from old Fig 7 Panel C only
 \begin{figure}[h]
 \centering
-\includegraphics[width=\linewidth]{../../figures/lk_fix_prob_C.pdf}
+\includegraphics[width=.8\linewidth]{../pnas_submission/lk_fix_prob_C.pdf}
 \caption{Fixation in a finite population with different ratios of selection periods \(\frac{k}{l}\).
@@
 We conjecture that with $k\ne l$ there is a unique stable polymorphism when both fixations are unstable.
-This result is similar to the \emph{storage effect} \cite{chesson1981}, in which protection from selection maintains species coexistence: Consider a two species $A$ and $B$, with overlapping generations, an equal death rate $\rho$, and different growth rates $w_A$ and $w_B$; then \eqref{2} will model the change in frequency of species $A$.
-In our model, oblique transmission can be said to protect the unfavored phenotype from selection because it allows transmission without reproduction.
 
@@
 This is an unusual scenario for genetic modifiers, although it must be noted that a modifier of $\rho$ is not neutral; it affects primary selection, while neutral  modifiers of recombination, mutation, and migration affect induced or secondary selection.
-To explain the marked difference in the value of the stable vertical transmission rate consider the probability $q$ that offspring receive phenotype $A$ when it is about to be the favored phenotype, and for simplicity, assume strong selection, such that parents are likely to be of the phenotype favored in the parental generation, a periodically changing environment $AkBl$, and a population at equilibrium.
-With vertical transmission, $q_v=\frac{k-1}{k}$.
-With oblique transmission, $q_o=x$ the frequency of the favored phenotype in the parental generation.
-For $k=l$, the frequency of $A$ cycles around 0.5 so $q_o \approx 0.5$, whereas $q_v=0, {1 \over 2}, {2 \over 3}, {3 \over 4}, \ldots$ for $k=1, 2, 3, 4, \ldots$.
-Therefore, for $k>2$ vertical transmission increases $q$ whereas for $k=1$ and $2$, oblique transmission increases $q$.
 

### 2017-12-11 ms/figures.tex
pnas branch, middle of revision

@@
 \textbf{(B)} The evolutionary stable rate $\rho^*$, which cannot be invaded by modifiers with either higher or lower vertical transmission rate $P$, rapidly increases from zero when selection cycles are short ($k=1$ or $2$) to roughly 1 when selection cycles are longer.
-The stable rate was found by finding the root of $\frac{\partial \lambda_1}{\partial P}$, the derivative of the leading eigenvalue of the external stability matrix w.r.t to the rate of the invading modifier (see \emph{Supplemental Material SP6} for details).
 The dashed line shows $1-\frac{1}{k-1}$, which fit the values for $w=0.5$ (Carja et al.
@@
 The values for $w=0.1$ (blue) could not be calculated for $k > 19$ due to numerical instability when selection is strong and the duration between selection fluctuations is long.
-In all cases, $W=1$.}\label{fig:AkBk_stable_optimal_rate}
+In all cases, $W=1$. See \emph{Supplemental Material SP6} for details on how we calculated the stable rate.}\label{fig:AkBk_stable_optimal_rate}
 \end{figure}
@@
 
+%Figure 10
+\begin{figure}
+\centering
+\includegraphics{../figures/lk_fix_prob_C.pdf}
+\caption{Fixation in a finite population with different ratios of selection periods \(\frac{k}{l}\).
+Fixation probability of phenotype $A$ when starting with a single copy in a population of size $N$: $u(1/N) = (1-\exp(-2 \rho \frac{k-l}{k+l}(W-w))/(1-\exp(-2 N \rho \frac{k-l}{k+l}(W-w))$ (see eqs.
+59--60).
+\emph{k} and \emph{l} are the number of
+generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by
+selection. Here, fitness of the favored phenotype is $W = 1$, fitness of the unfavored phenotype is $w=0.5$, and the population size is \(N=10,000\). 
+See Figure SX for additional parameters.} \label{lk_fix_prob_C}
+\end{figure}
+
 \end{document}

### 2017-12-11 ms/ms.tex
pnas branch, middle of revision

@@
 \medskip
-\centerline{[FIGURE 1 HERE]}  
+
+\pdfximage width 14cm {figures/A1B2_polymorphism.pdf}
+\pdfrefximage\pdflastximage
+\centerline{
+Stable frequency of phenotype {\sl A} and geometric mean fitness in selection regime
+$A1B2$ as a function of the vertical transmission rate \(\rho\) and the fitness of the disfavored phenotype \(w\).
+{\bf (A)} Stable frequency of phenotype $A$ at the end of each three generations cycle.
+{\bf (B)} Geometric average of the stable population mean fitness over the three generations cycle: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
+Gray contour lines join \(\rho\) and \(w\) combinations that result in the same stable value.
+In all cases, \emph{W=1}.
+}
 \bigskip

### 2017-12-11 ms/stability_analysis.tex
pnas branch, middle of revision

@@
 \pagestyle{headings}
-\markright{Yoav Ram, \today \; v2\hfill}
+\markright{Yoav Ram, \today \; v1\hfill}
+
+\newcommand{\beginsupplement}{%
+	\setcounter{table}{0}
+	\renewcommand{\thetable}{A\arabic{table}}%
+	\setcounter{figure}{0}
+	\renewcommand{\thefigure}{A\arabic{figure}}%
+	\setcounter{equation}{0}
+	\renewcommand{\theequation}{A\arabic{equation}}%
+}
 
 \begin{document}
+\beginsupplement
+
 Here I describe how I analyzed the stability of a modifier allele $m$ with vertical transmission rate $\rho$ to invasion by a modifier $M$ with a vertical transmission rate $P$, as described in the main text, eqs.
@@
 
-\item Define $F_A(x)$ by eq. 35 of main text with $w_A=W$ and $w_B=w$ ($W>w$), and similarly $F_B(x)$ with $w_B=W$ and $w_A=w$.
+\item Define $F_A(x)$ by eq. 32 of main text with $w_A=W$ and $w_B=w$ ($W>w$), and similarly $F_B(x)$ with $w_B=W$ and $w_A=w$.
 
-\item Define, similar to eq. 37 of main text,
+\item Define, similar to eq. 34 of main text,
 \begin{equation}
-\vec{F} = \underbrace{F_A \circ \ldots \circ F_A}_{k \text{ times}} \circ
-	\underbrace{F_B \circ \ldots \circ F_B}_{l \text{ times}}.
+\vec{F} = \underbrace{F_B \circ \ldots \circ F_B}_{l \text{ times}} \circ
+	\underbrace{F_A \circ \ldots \circ F_A}_{k \text{ times}}.
 \label{eq:F}\end{equation}

### 2017-12-12 ms/PNAS/Supplementary_revision.tex
more changes after speaking with MWF and UL

\documentclass[9pt, a4paper]{extarticle}
\usepackage[margin=1in]{geometry} % See geometry.pdf to learn the layout options.
\geometry{a4paper}
\usepackage[parfill]{parskip}
\usepackage{graphicx}
\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage{longtable}
\usepackage[hyphens]{url}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\usepackage{authblk}

% messes up nameref
%\usepackage{titlesec}
%\titleformat{\subsection}[runin]
%  {\bf}{\thesection}{1em}{}

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
\newtheorem{result}{Result}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{I}

\newcommand{\beginsupplement}{%
	\setcounter{table}{0}
	\renewcommand{\thetable}{S\arabic{table}}%
	\setcounter{equation}{0}
	\renewcommand{\theequation}{S\arabic{equation}}%
	\setcounter{figure}{0}
	\renewcommand{\thefigure}{S\arabic{figure}}%
}

% NatBib
\usepackage[round,colon,authoryear]{natbib}

\title{Evolution of vertical and oblique transmission under fluctuating selection}
\author[a]{Yoav Ram}
\author[b]{Uri Liberman}
\author[a]{Marcus W. Feldman}
\affil[a]{Department of Biology, Stanford University, Stanford, CA}
\affil[b]{School of Mathematical Sciences, Tel Aviv University, Israel}
\date{\today}

\begin{document}
\beginsupplement
\maketitle

% SI Appendix
\section*{SI Appendix}
% Appendix A
\subsection*{SI Appendix A: Proof of Uniqueness in Result 5}

Following eq.~(10) the transformation of the frequency $x$ of phenotype $A$ is
\begin{equation}
x'= F(x) = \underbrace{F_B\circ\cdots\circ F_B}_{l\ \hbox{\sevenrm times}}\;\circ\; \underbrace{F_A\circ\cdots\circ F_A}_{k\ \hbox{\sevenrm times}}(x).
\label{eq:S1}
\end{equation}
 Using eqs.~(25-26) we can write
 \begin{align} 
 \label{eq:S2}
 F'_A(x) =\rho{wW\over \bigl[\overline w(x)\bigr]^2} +(1-\rho),\quad \overline w(x)=(W-w)x + w \\
  \label{eq:S3}
 F'_B(x) =\rho{wW\over \bigl[\widetilde w(x)\bigr]^2} +(1-\rho),\quad \widetilde w(x)=(w-W)x +W.
 \end{align}
Since $F'_A(x) >0$, $F'_B(x)>0$ for  $0\le x\le 1$,  all the functions $F_A$, $F_B$, $F_A\circ\cdots\circ F_A$, $F_B\circ\cdots\circ F_B$, and $F$ are monotone increasing for  $0\le x\le 1$.

From Result 2, the two fixations $x=0$ and $x=1$ are not stable because $F'(0)>1$ and $F'(1)>1$. Therefore
\begin{align}
F(x)-x>0\ \text{for}\ x>0\ \text{``near''}\ x=0, \\
F(x)-x<0\ \text{for}\ x<1\ \text{``near''}\ x=1.
\end{align}
Hence, as $F(x)-x$ is a continuous function of $x$ for $0\le x\le 1$, there exists (at least one) polymorphic equilibrium $x^*$ with $0<x^*<1$ such that $F(x^*)=x^*$.

If there is more than one polymorphic equilibrium, and as there is a finite number of equilibria, let $x^*$ be the ``closest'' polymorphic equilibrium to $x=0$.
Since $F(x)>x$ for $0<x<x^*$, $F(x)<x$ for $x>x^*$ (at least ``near'' $x^*$), and $F(x)$ is a monotone increasing function in $[0,1]$,  $x^*$ must be locally stable.

Let $\hat x=F(\hat x)$ with $0<\hat x<1$ be any polymorphic equilibrium, then from (\ref{eq:S1}) its evolution in the $k+k$ generations is 
\begin{equation}
\vbox{\halign{
\hfil$#$\hfil &&\hfil\ $#$\hfil\cr
&A&&A&&\cdots&&A&&B&&B&&\cdots&&B&&A\cr
\hat x=&\hat x_0&\to&\hat x_1&\to&\cdots&\to&x_{k-1}&\to&\hat y_0&\to&\hat y_1&\to&\cdots&\to&\hat y_{k-1}&\to&\hat x_0&=\hat x.\cr
}}
\label{eq:S6}
\end{equation}
Due to the symmetry between phenotypes $A$ and $B$ we have
\begin{equation}
\hat y_t=1-\hat x_t,\qquad \overline w\bigl(\hat x_t\bigr) =\widetilde w\bigl(\hat y_t\bigr)
\label{eq:S7}
\end{equation}
for all $t=0,1,2,\dots,k-1$.

The polymorphic equilibrium $\hat x$ is locally stable if $F'(\hat x)<1$, or from (\ref{eq:S1}), (\ref{eq:S2}), and (\ref{eq:S3}), if
\begin{equation}
\prod_{t=0}^{k-1}\left\{\rho{wW\over\bigl[\overline w(\hat x_t)\bigr]^2} +(1-\rho)\right\}\cdot\prod_{t=0}^{k-1} \left\{\rho{wW\over\bigl[\widetilde w(\hat y_t)\bigr]^2} +(1-\rho)\right\}<1.
\end{equation}
Applying (\ref{eq:S7}), we conclude that $\hat x$ is locally stable if
\begin{equation}
\prod_{t=0}^{k-1}\left\{\rho{wW\over\bigl[\overline w(\hat x_t)\bigr]^2} +(1-\rho)\right\}<1.
\label{eq:S9}
\end{equation}

As $x^*$, the ``closest'' polymorhpic equilibrium to $x=0$, is stable, then (\ref{eq:S9}) implies that
\begin{equation}
\prod_{t=0}^{k-1}\left\{\rho{wW\over\bigl[\overline w(x_t^*)\bigr]^2} +(1-\rho)\right\}\le 1,
\end{equation}
 where $x_t^*$ for $t=0,1,\dots,k-1$ is defined as in (\ref{eq:S6}).

 If $\hat x$ is any  polymorphic equilibrium other than $x^*$, then $\hat x>x^*$ or $\hat x_0>x_0^*$ by (S4). Since $F_A$ is a monotone increasing function and
\begin{equation}
\hat x_{t+1} =F_A\bigl(\hat x_t\bigr),\quad x_{t+1}^* =F_A\bigl(x_t^*\bigr)\quad t=0,1,\dots,k-1,
\end{equation}
 then by induction we have $\hat x_t>x_t^*$ for all $t=0,1,2,\dots,k-1$. In addition, as $\overline w(x) =(W-w)x+w$ and $W>w$, we also have $\overline w(\hat x_t)>\overline w(x_t^*)$ for all $t=0,1,2,\dots,k-1$, and
 \begin{equation}
\prod_{t=0}^{k-1}\left\{\rho{wW\over\bigl[\overline w(\hat x_t)\bigr]^2} +(1-\rho)\right\}<\prod_{t=0}^{k-1} \left\{\rho{wW\over\bigl[\overline w(x_t^*)\bigr]^2} +(1-\rho)\right\}\le 1.
 \end{equation}
Hence $\hat x$ is also locally stable.
But it is impossible that {\sl all} polymorphic equilibria are stable unless there is only one stable polymorphic equilibrium. Therefore $x^*$ is the unique stable polymorphic equilibrium, and since $F(x)>x$ for $0<x<x^*$ with $F(x)<x$ for $x^*<x<1$, and $F(x)$ is monotone increasing in $[0,1]$, therefore $x^*$ is globally stable.

% Appendix B
\subsection*{SI Appendix B: Proof of Result 6}

 Rewrite recursion eq.~(28) as
 \begin{equation}
{x_t+1\over x_t} =(1+\rho s_t)\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right].
\end{equation}
 Then
 \begin{equation}
\log x_{t+1} -\log x_t =\log(1+\rho s_t) +\log\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right].
\end{equation}
 Summation yields
  \begin{equation}
{1\over t}\left[\log x_t-\log x_0\right] ={1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n) +{1\over t}\sum_{n=0}^{t-1}\log\left[1-x_n{\rho s_n(1+s_n)\over (1+\rho s_n)(1+x_ns_n)}\right].
\label{eq:S15}
 \end{equation}
Let $\mu=E\left[\log(1+\rho s_t)\right]$. As $\{s_t\}_{t\ge 0}$ are independent and identically distributed random variables, the {\sl strong law of large numbers} applies and
 \begin{equation}
\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)=\mu
 \end{equation}
almost surely.

Let $\zeta$ be such that ${1\over t}\sum_{n=0}^{t-1}\log[1+\rho s_n(\zeta)]=\mu$ and assume that $\lim_{t\to\infty}x_t(\zeta)=0$. As the random variables $\{s_t\}_{t\ge 0}$ are uniformally bounded,
 \begin{equation}
x_t(\zeta){\rho s_t(\zeta)[1+s_t(\zeta)]\over [1+\rho s_t(\zeta)][1+x_t(\zeta)s_t(\zeta)]}\;\;\harr{}{t\to\infty}\;\;0
 \end{equation}
and
 \begin{equation}
\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log\left[1-x_n(\zeta){\rho x_n(\zeta)[1+s_n(\zeta)]\over [1+\rho s_n(\zeta)][1+x_t(\zeta)s_n(\zeta)]}\right]=0.
 \end{equation}
Thus (\ref{eq:S15}) implies that
 \begin{equation}
\lim_{t\to\infty}{1\over t}\left[\log x_t(\zeta) -\log x_0(\zeta)\right]=\mu.
 \label{eq:S19}
 \end{equation}
If $\mu =E\left[\log(1+s_t)\right]>0$, then from (\ref{eq:S19}) we deduce that $\lim_{t\to\infty}x_t(\zeta)=\infty$, a contradiction.
Therefore when $\mu>0$, $P\left(\lim_{t\to\infty}x_t=0\right)=0$, and fixation of $B$ ($x^*=0$) is stochastically locally unstable.

 Thus by Result 6, for $x^*=0$ to be stochastically locally stable it is necessary that $E[\log(1+\rho s_t)]\le 0$. In fact, the strict inequality is sufficient.

 Figure 3 presents a numerical example of the dynamics of recursion eq.~(28) with a specific random selection coefficient $s_t$.

% Appendix C
\subsection*{SI Appendix C: Proof of Result 7}

Let $\mu=E[\log(1+\rho s_t)]$. Then as $\{s_t\}_{t\ge 0}$ are independent and identically distributed random variables, the strong law of large number applies and almost surely
\begin{equation}
\lim_{t\to\infty}{1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)=\mu<0.
\end{equation} % S20
 Appealing to the Egoroff Theorem, for any $\epsilon>0$ there exists $T$ such that 
 \begin{equation}
 P\left({1\over t}\sum_{n=0}^{t-1}\log(1+\rho s_n)<{\mu\over 2}\ \text{for all}\ t\ge T\right)\ge 1-\epsilon.
 \end{equation}
 As $0\le \rho\le 1$ and the $\{s_t\}_{t\ge 0}$ are uniformly bounded, we can find a $\delta'>0$ such that
  \begin{equation}
x_t<\delta'\Longrightarrow\left|\log\left[1-x_t{\rho s_t(1+s_t)\over (1+\rho s_t)(1+x_ts_t)}\right]\right|<-{\mu\over 4}.
 \end{equation}
 Also, as $0\le x_t\le 1$ for all $t$,
  \begin{equation}
x_{t+1}=x_t{1+\rho s_t +x_t(1-\rho)s_t\over 1+x_ts_t}< Kx_t,
 \end{equation}
 where $K$ is independent of $t$. It follows that there exists a $\delta$ with $0<\delta<\delta'$ such that
   \begin{equation}
x_o<\delta\Longrightarrow x_t<\delta'\ \text{for all}\ t=0,1,2,\dots,T-1. 
\end{equation}
 Let $\xi$ be a realization of the evolutionary process such that
   \begin{equation}
{1\over t}\sum_{n=0}^{t-1}\log[1+\rho s_n(\xi)]<{\mu\over 2}\ \text{for all}\  t\ge T
\end{equation}
 and assume $x_0<\delta$. Then
 \begin{multline}
 {1\over T}[\log x_T(\xi) -\log x_0(\xi)] = \\
 {1\over T}\sum_{n=0}^{T-1}\log[1+\rho s_n(\xi)]+{1\over T}\sum_{n=0}^{T-1}\log[1-x_n(\xi){\rho s_n(\xi)[1+s_n(\xi)]\over [1+\rho s_n(\xi)][1+x_n(\xi)s_n(\xi)]}
 &<{\mu\over 2}- {\mu\over 4} ={\mu\over 4}<0, 
 \end{multline}
  and therefore $x_T(\xi) <x_0(\xi) <\delta'$. Invoking induction we get that for $t\ge T$
\begin{equation}
{1\over t}\log{x_t(\xi)\over x_0}\le {\mu\over 4},
\end{equation}
 or for all $t\ge T$
 \begin{equation}
 x_t(\xi)\le x_0\exp\left({\mu\over 4}t\right).
 \end{equation}

 As $\mu<0$, this implies that $x_t(\xi)\,\harr{}{t\to\infty}\,0$. Therefore we have shown that for given $\epsilon>0$ there is a $\delta>0$ such that if $0<x_0<\delta$, then $P\left(\lim_{t\to\infty}x_t =0\right)\ge 1-\epsilon$, and therefore $x^*=0$, the fixation in $B$ is stochastically locally stable. The second statement of Result 6 follows from the convexity of the log function and Jensen's inequality.

% Appendix D
\subsection*{SI Appendix D: Proof of Result 8}

The external stability of $\underline x^*$ (eqs.~(35-36)) to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ near $x^*$, which is derived from eqs.~(32) and given by
 \begin{equation}
 \overline w^*\cl_1 =\left[\begin{array}{cc}{W\bigl[(1-P)x_1^* +P\bigr] & w(1-P)x_1^*\\
 W(1-P)x_2^* & w\bigl[(1-P)x_2^* +P\bigr]\end{array}\right]
 \label{37}\end{equation}
 and
 \begin{equation}
 \overline w^{**}\cl_2 =\left[\begin{array}{cc}w\bigl[(1-P)x_1^{**} +P\bigr] & W(1-P)x_1^{**}\\
 w(1-P)x_2^{**} & W\bigl[(1-P)x_2^{**} +P\bigr]\end{array}\right],
 \label{38}\end{equation}
where $\underline x^{**} =T_1\underline x^*$, $x_1^{**}=x_2^*=1-x_1^*$, $x_2^{**}=x_1^*$,  and
\begin{equation}
\overline w^* =Wx_1^* +wx_2^*,\qquad \overline w^{**} =wx_1^{**} +Wx_2^{**}.\label{39}\end{equation}
Due to the symmetry between the two phenotypes $A$ and $B$ in the $A1B1$ case, we have $x_1^{**} =x_2^*$ and $x_2^{**} =x_1^*$ so that $\overline w^{**} =\overline w^*$ and in fact
\begin{equation}
\overline w^*\cl_2 = \left[\begin{array}{cc}w\bigl[(1-P)x_2^* +P\bigr] & W(1-P)x_2^*\\
 w(1-P)x_1^* & W\bigl[(1-P)x_1^* +P\bigr]\end{array}\right].
 \label{40}\end{equation}
 Note that as $\underline x^* =T_2(T_1\underline x^*)$ with $x_3^*=x_4^*=0$, from (\ref{37}) and (\ref{38}) with $P=\rho$ we have
 \begin{equation}\left[\begin{array}{c}x_1^*\\ x_2^*\end{array}\right]= \cl_2\cdot\cl_1\left[\begin{array}{c}x_1^*\cr x_2^*\end{array}\right] =\cl\left[\begin{array}{c}x_1^*\\ x_2^*\end{array}\right].
 \label{41}\end{equation}
 Hence when $P=\rho$ one of the eigenvalues of $\cl$ is 1.
 In general $\cl=\cl_2\cdot\cl_1$, and using (\ref{37}) and (\ref{40}) we have
  \begin{equation}
  \begin{aligned}
 \left(\overline w^*\right)^2\cl_{11} &=Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[w(1-P)x_1^*\bigr]^2 \\
 \left(\overline w^*\right)^2\cl_{12} &=W(1-P)\bigl[P+x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr]\\
 \left(\overline w^*\right)^2\cl_{21} &=w(1-P)\bigl[1-x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr]\\
 \left(\overline w^*\right)^2\cl_{22} &= Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[W(1-P)x_2^*\bigr]^2
 \end{aligned}
 \label{eq:42}\end{equation}
 The external stability of $\underline x^*$ is determined by the eigenvalues of $\cl$, namely the roots of its characteristic polynomial $R(\lambda) =\det(\cl-\lambda\ci)$, with $\ci$ the $2\times 2$ identity matrix. From \eqref{eq:42}, $R(\lambda)=a_2\lambda^2 +a_1\lambda +a_0$, where
 \begin{equation}
 a_0={P^2W^2w^2\over\left(\overline w^*\right)^4}, \quad
 a_1=-{2PWw +\left(1-P\right)^2\left[Wx_2^* +wx_1^*\right]^2\over \left(\overline w^*\right)^2},\quad
 a_2=1.
 \label{eq:43}\end{equation}

As $\cl$ is a positive matrix, by the Perron-Frobenius theorem $\cl$ has a positive eigenvalue, and as $a_0>0$ and $a_2=1$ the product of the two eigenvalues of $\cl$ is positive. Thus $\cl$ has two positive eigenvalues. Let $R(1)=R(1;P)$, then from (\ref{eq:43})
  \begin{equation}
  R(1;P) = {W^2w^2-\left(\overline w^*\widetilde w^*\right)^2\over\left(\overline w^*\right)^4}P^2 +2P{\left(\widetilde w^*\right)^2 -Ww\over \left(\overline w^*\right)^2} +{\left(\overline w^*\right)^2-\left(\widetilde w^*\right)^2\over\left(\overline w^*\right)^2},
  \end{equation}
where $\widetilde w^* =Wx_2^* +wx_1^*$.

By eq.~(36), $\bigl(\sqrt{Ww}-w\bigr)/(W-w) <x_1^* <{1\over 2}$, from which it is easily seen that
  \begin{equation} \label{eq:S30}
\sqrt{Ww}< \overline w^*<\widetilde w^* .
\end{equation}

 When $P=\rho$ one of the eigenvalues of $\cl$ is 1; hence $R(1;\rho)=0$. Another root of $R(1;P)=0$ is $\bigl[(\overline w^*)^2 +\overline w^*\widetilde w^*\bigr]/\bigl[Ww +\overline w^*\widetilde w^*\bigr]$, which by \eqref{eq:S30} is larger than 1.  As $R(1;0)=\bigl[(\overline w^*)^2 -(\widetilde w^*)^2\bigr]/(\overline w^*)^2 <0$ by \eqref{eq:S30}, we deduce that when $0<P<\rho$, $R(1;P)<0$, whereas when $\rho<P<1$, $R(1;P)>0$.
 Hence, when $P<\rho$, $R(1)<0$, and since $a_2=1$, $R(+\infty)>0$, we conclude that $R(\lambda)=0$ has a positive root larger than 1 and the largest positive eigenvalue of $\cl$ is larger than 1.

 When $P>\rho$, we have $R(1)>0$ and also $R(0)=a_0>0$. As $R(\lambda)=0$ has two positive roots and as $a_2>0$, $R(\lambda)$ is convex, either the two positive roots are less than 1 or both larger than one. But the product of the two roots is $P^2W^2w^2/(\overline w^*)^2<1$ by \eqref{eq:S30}; thus when $P>\rho$ the two positive eigenvalues of $\cl$ are less than 1.

% Appendix E
\subsection*{SI Appendix E: Proof of Result 10}

Without loss of generality and for the ease of representation, we will show that for $t>0$, 
\begin{equation}
	v (x;t) ={1-e^{-tx}\over 1-e^{-t}}
\end{equation}
  is monotone increasing as a function of $t$. Observe that
  \begin{equation}
  {\partial v\over\partial t} = {\left(1-e^{-t}\right)xe^{-tx} -\left(1-e^{-tx}\right)e^{-t}\over \left(1-e^{-t}\right)^2}.
  \end{equation}

\noindent For the monotonicity we have to show that
  \begin{equation} \label{eq:S33}
  f(x;t) =\left(1-e^{-t}\right)xe^{-tx} -\left(1 -e^{-tx}\right)e^{-t} \ge 0
  \end{equation}
  when $t>0$ and $0\le x\le 1$. Note that $f(0;t)=0$ and $f(1;t)=0$. Also
  \begin{equation}
	{\partial f\over\partial x} =\left(1-e^{-t}\right)\left(e^{-tx}-txe^{-tx}\right) -te^{-tx}e^{-t},
  \end{equation}
  or
  \begin{equation}
  {\partial f\over\partial x} =e^{-tx}\left[\left(1-e^{-t}\right)\left(1-tx\right) -te^{-t}\right] =e^{-tx}g(x;t),
  \end{equation}
 say, where for fixed $t$, $g(x;t)$ is a linear function of $x$, which vanishes at $x_0 =(1-e^{-t} -te^{-t})/t(1-e^{-t})$. If $t>0$, $e^t >1+t$, so $1>e^{-t}(1+t)$ and $x_0>0$. Also if $t>0$, $e^{-t}>1-t$, and so $1-e^{-t} -te^{-t} <t(1-e^{-t})$ and $x_0<1$. Since $g(0,t) =1 -e^{-t} -te^{-t} >0$ and $g(1;t) =(1-e^{-t})(1-t) -te^{-t} <0$ for $t>0$, we deduce that ${\partial f\over\partial x}(x,t)>0$ for $ 0<x<x_0$ and ${\partial f\over\partial x}(x,t)<0$ for $x_0<x<1$ for all $t>0$. These facts, combined with $f(0,t) =f(1;t) =0$, prove that $f(x;t)\ge 0$ for $0\le x\le 1$ (in fact, $f(x;t)>0$ for $0<x<1$), and inequality (\ref{eq:S33}) is satisfied as desired.

% Appendix F
\subsection*{SI Appendix F: Proof of Result 11}

The proof is based on induction on $n$, where in order to prove eq.~(45), we show that if $X_t$ is the number of individuals with phenotype $A$ at stage $t$ of the cycle, and $x$ is the initial frequency of $A$, then
 \begin{equation} \label{eq:S36}
 E\left({X_t\over N} -x\right) \simeq {1\over N}\rho S_t x(1-x),\qquad V\left({X_t\over N}\right)\simeq {1\over N}tx(1-x),
 \end{equation} % S36
 where $N$ is the size of the population. When $t=1$, (\ref{eq:S36}) coincides with the constant environment case. Assuming (\ref{eq:S36}), we go to $t+1$. Now $X_{t+1}$ given $X_t=Ny$ has a binomial distribution with parameters $(N,y')$. Hence
  \begin{equation} 
  E\left({X_{t+1}\over N} -{X_t\over N} \mid X_t=Ny\right) =y'-y.
  \end{equation}
 Following~\citet[ch.~5]{ewens2004}, $y'-y\simeq (1/N)\rho s_{t+1}y(1-y)$, and so
  \begin{equation} 
  E\left({X_{t+1}\over N} -{X_t\over N}\mid X_t\right)\simeq {1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right).
  \end{equation}
 Observe that
 \begin{equation} \label{eq:S39}
 E\left[{X_t\over N}\left(1-{X_t\over N}\right)\right] &= E\left({X_t\over N}\right) -E\left[\left({X_t\over N}\right)^2\right]
  = E\left({X_t\over N}\right) -V\left({X_t\over N}\right) -\left[E\left({X_t\over N}\right)\right]^2.\cr}  
 \end{equation}
 By the induction assumption, $V(X_t/N)\simeq (1/N)tx(1-x)$, and ignoring terms of order $1/N^2$ we have
 \begin{equation} 
 E\left({X_{t+1}\over N} -{X_t\over N}\right)\simeq {1\over N}\rho s_{t+1} E\left({X_t\over N}\right)\left[1-E\left({X_t\over N}\right)\right].
\end{equation} % S40
Applying (\ref{eq:S36}) we have
\begin{equation}\begin{aligned}
E\left({X_t\over N}\right)  &\simeq x+{1\over N}\rho S_tx(1-x), \\
1-E\left({X_t\over N}\right) &\simeq 1-x-{1\over N}\rho S_tx(1-x),
\end{aligned}\end{equation}
and ignoring terms $O(1/N^2)$, we find
\begin{equation} \label{eq:S42}
E\left({X_{t+1}\over N} -{X_t\over N}\right) \simeq {1\over N}\rho s_{t+1}x(1-x).
\end{equation}
Thus
\begin{equation}\begin{aligned}
E\left({X_{t+1}\over N} -x\right) &= E\left({X_{t+1}\over N} -{X_t\over N}\right) +E\left({X_t\over N} -x\right) \\
&\simeq {1\over N}\rho s_{t+1}x(1-x) +{1\over N}\rho S_t x(1-x),
\end{aligned}\end{equation}
and since $S_t +s_{t+1} =S_{t+1}$,
 \begin{equation}
 E\left({X_{t+1}\over N} -x\right)\simeq {1\over N}\rho S_{t+1}x(1-x)
 \end{equation}
 as desired.

 We now compute $V(X_{t+1}/N)$ using the induction assumption and the formula
 \begin{equation}
 V\left({X_{t+1}\over N}\right) =E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] +V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right],
 \end{equation}
 where by (\ref{eq:S36})
 \begin{equation}
 E\left({X_{t+1}\over N}\mid X_t\right)\simeq {X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)
 \end{equation}
 and
 \begin{equation}
 V\left({X_{t+1}\over N}\mid X_t\right)\simeq{1\over N}{X_t\over N}\left(1-{X_t\over N}\right).
 \end{equation}
 Here we used the fact that  $y'(1-y')\simeq y(1-y)$. Now
 \begin{equation} \label{eq:S48}
 E\left[V\left({X_{t+1}\over N}\mid X_t\right)\right] \simeq {1\over N}E\left[{X_t\over N}\left(1-{X_t\over N}\right)\right]\simeq {1\over N}x(1-x),
 \end{equation}
 where we use the same computations as led from (\ref{eq:S39}) to (\ref{eq:S42}).
  \begin{equation}
  V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right] =V\left[{X_t\over N} +{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right].
  \end{equation}
 Since $(X_t/N)\bigl[1-(X_t/N)\bigr]$ is a random variable taking values in [0,1],  its variance is less than 1/4 and
 \begin{equation}
 V\left[{1\over N}\rho s_{t+1}{X_t\over N}\left(1-{X_t\over N}\right)\right] \le {1\over 4N^2}\rho^2 s_{t+1}^2.
 \end{equation}
 We ignore terms $O(1/N^2)$ so that the random variable $(1/N)\rho s_{t+1}(X_t/N)\bigl[1-(X_t/N)\bigr]$ is almost constant. As a result,
 \begin{equation} \label{eq:S51}
 V\left[E\left({X_{t+1}\over N}\mid X_t\right)\right]\simeq V\left({X_t\over N}\right) \simeq {1\over N}tx(1-x),
 \end{equation}
 by the induction assumption. Combining (\ref{eq:S48}) and (\ref{eq:S51}) gives
 \begin{equation}
 V\left({X_{t+1}\over N}\right)\simeq {1\over N}x(1-x) +{1\over N}tx(1-x) ={1\over N}(t+1)x(1-x)
 \end{equation}
 as expected.

% Appendix G
\subsection*{SI Appendix G: Calculation of stable vertical transmission rate in $AkBk$}

Here we describe the analysis of the stability of a modifier allele $m$ with vertical transmission rate $\rho$ to invasion by a modifier $M$ with a vertical transmission rate $P$, as described in eqs.~(32), in environmental regime $AkBl$.
The analysis is similar to that used in Result 8 to analyze stability in $A1B1$, but it is numerical because the cases where $k>1$ or $l>1$ require solving polynomials of degree $>6$  in order to obtain closed form expressions.

The analysis includes the following steps for fixed $W, w, k,$ and $l$.
First, we find the stable frequency of phenotype $A$ with a single modifier $x^*$. This is done by minimizing the expression $|x_{k+l} - x_{0}|$ where $x_{t}$ is defined in eq.~(9). The minimization is done by iterating the recurrence in eq.~(9) until it converges, i.e. until the difference $|x_{k+l} - x_{0}|$ is smaller than available machine precision (roughly $10^{-8}$ when subtracting similar small numbers).
Next, we set the frequency vector with two modifiers to $\underline x^*=(x^*, 1-x^*, 0, 0)$, that is, to the stable frequencies in the absence of modifier $M$.

Now we define $F_A(\underline x)$ by eqs.~(32) with $w_A=W$ and $w_B=w$ ($W>w$), and similarly $F_B(\underline x)$ with $w_B=W$ and $w_A=w$.
Also, we define, similar to eq.~(34), $F(\underline x)$ by composition
\begin{equation} \label{eq:S53}
F=\underbrace{F_B\circ \cdots\circ F_B}_{l\ \hbox{\sevenrm times}} \circ\underbrace{F_A\circ \cdots\circ F_A}_{k\ \hbox{\sevenrm times}}.
\end{equation}

To obtain a linear approximation of $F(\underline x)$ "near" $\underline x^*$, we calculate the Jacobian matrix of $F(\underline x)$ at $\underline x = \underline x^*$,
\begin{equation} \label{eq:S54}
\cj_{ij} = \cj(\underline x^*)_{ij} = {\partial F(\underline x^*)_i \over \partial x_j},
\end{equation} 
and the $2\times 2$ external stability matrix $\cl=\cl_{ex}$ is as in \eqref{37} and \eqref{38} (note that the upper-right block is $\underline 0$ because $x^*_3=x^*_4=0$)
\begin{equation}
\cj = \begin{bmatrix}
\cl_{in} & \underline 0 \\
 * & \cl_{ex}
\end{bmatrix}
\end{equation}

We calculate the eigenvalues $\lambda_1 > \lambda_2$ of $\cl$ using the quadratic formula as the characteristic polynomial of $\cl$ has degree 2.
By the Perron-Frobenius theorem, the leading eigenvalue $\lambda_1$ is real and positive. Denote by $\lambda_1(\rho, P)$ the resulting leading eigenvalue with resident rate $\rho$ and invader rate $P$. Note that for any $\rho \in (0,1)$ 
\begin{equation} \label{eq:S56}
\lambda_1(\rho, \rho) = 1.
\end{equation}
The evolutionarily stable rate $\rho^*$ is defined to be stable to invasion; that is, for a small enough value $\partial P>0$ we have
\begin{equation} 
\lambda_1(\rho^*,\rho^* \pm \partial P) < 1 = \lambda_1(\rho^*,\rho^*),
\end{equation}
where the equality is given by eq.~(\ref{eq:S56}).
Therefore, 
\begin{equation} \label{eq:S58}
{\partial \lambda_1 \over \partial P} \big(\rho^*,\rho^*\big) = 0. 
\end{equation}

We use Brent's root-finding method~\citep{brent1971} to find $\rho^*$ that satisfies eq.~(\ref{eq:S58}).
If, due to numerical instability of the described numerical process, we have
\begin{equation}
\partial {\lambda_1 \over \partial P} \big(0,0\big) \cdot {\partial \lambda_1 \over \partial P} \big(1,1) > 0,
\end{equation}
i.e., the partial derivative sign is identical at $\rho=P=0$ and $\rho=P=1$, then we cannot use Brent's method.
In these cases we assume that the partial derivative doesn't have a root in $(0,1)$ and we determine the stable rate $\rho^*$ by the rule
\begin{equation}
\rho^* = \begin{cases}
0 & \text{if}\ {\partial \lambda_1 \over \partial P} \big(0,0\big) \le 0 \\
1 & \text{if}\ {\partial \lambda_1 \over \partial P} \big(0,0\big) > 0.
\end{cases}
\end{equation}

Supplementary Figure S10 shows the sensitivity of the leading eigenvalue $\lambda_1$ of the external stability matrix $\cl$ to changes in the invader rate $P$ as a function of the resident rate $\rho$, for different choices of environmental cycles $AkBk$.

The  numerical analysis above is fine for small $k$, but for large $k$ and especially for $w=0.1$ the calculation is unstable.
This is probably because when the environment is constant for a long period of time, most of the time the frequencies $x_i$ are very close to the boundaries (i.e., 0 and 1).

Crucially, the Jacobian $\cj$ in eq.~(\ref{eq:S54}) is calculated using {\sl automatic differentiation} from a function that iteratively calculates $F(\underline x^*)$ according to eq.~(\ref{eq:S53}). Similarly, the partial derivative ${\partial \lambda_1 \over \partial P}$ in eq.~(\ref{eq:S58}) is calculated from a function that calculates $\lambda_1$ using simple arithmetic operations.
Note that {\sl automatic differentiation} does not mean {\sl symbolic} or {\sl numerical differentiation}, which can lead to inefficient or inaccurate estimation of $\cj$ when $k$ is not very small. Rather, from~\citet{bartholomew2000}: 
{\sl ``Automatic differentiation is a set of techniques for transforming a program that calculates numerical values of a function, into a program which calculates numerical values for derivatives of that function with about the same accuracy and efficiency as the function values themselves.''}

% Appendix H
\subsection*{SI Appendix H: Diffusion Approximation}

We compute the mean $\mu(x)$ and the variance $\sigma^2(x)$ of the change in one generation  in the frequency of phenotype $A$ given that at the beginning of the generation $X_t=Nx$.
To compute $\mu(x)$, observe that by eq.~(40)
 \begin{equation}
 \begin{aligned}
 x' -x =
 {w_Ax\over\overline w}\rho +(1-\rho)x -x =
 \rho x\left[{w_A\over \overline w}-1\right] = \\
 \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},
 \label{49}\end{aligned}\end{equation}
 since $\overline w =w_A x +w_B(1-x)$.
 For the diffusion approximation, it is essential that the differential selection does not have a large effect per individual in each time period  $\Delta t$ $\bigl(\Delta t\simeq{1\over N}\bigr)$. That is, we assume
 \begin{equation}
 w_A -w_B ={s\over N}.
 \label{50}\end{equation}
 Then
 \begin{equation}
 x'-x\simeq {1\over N}\rho sx(1-x)
 \label{51}\end{equation}
 up to terms of order small than ${1\over N}$. Since one generation corresponds to $\Delta t\simeq {1\over N}$, we  conclude that
 \begin{equation}
 \mu(x) =\rho sx(1-x);\quad 0\le x\le 1.
 \label{52}\end{equation}
 In the same way, we can compute 
  \begin{equation}
  \sigma^2(x) = x(1-x).
 \end{equation}

% Bibliography
\bibliographystyle{agsm}
\bibliography{supplementary}

\end{document}

### 2017-12-12 ms/PNAS/revision.tex
more changes after speaking with MWF and UL

@@
 \begin{abstract}
-Conditions for the evolution and maintenance of social learning, in competition with individual learning, under fluctuating selection, is a well-studied problem in the theory of cultural evolution.
+The evolution and maintenance of social learning, in competition with individual learning, under fluctuating selection, have been well-studied in the theory of cultural evolution.
 Here we study competition between vertical and oblique cultural transmission of a dichotomous phenotype under constant, periodically cycling, and randomly fluctuating selection.
@@
 Similarly, some phenotypes are determined by genes that are commonly converted by uptake of foreign DNA, i.e., transformation \cite{milkman1990molecular}.
-In these cases, inheritance of a phenotype may combine vertical transmission from the parent cell, and oblique transmission from other cells, even if the mechanism for oblique transmission evolved to facilitate some other function \cite{redfield1993evolution}.
+In these cases, inheritance of a phenotype may combine vertical transmission from the parent cell, and oblique transmission from other cells.
 
@@
  \begin{equation}
- \an{
- x'&=\rho{w_A\over \overline w}x +(1-\rho)x\\
- &=     {w_A\over \overline w}x\bigl[(1-\rho)x +\rho\bigr] +{w_B\over \overline w}(1-x)\bigl[(1-\rho)x\bigr],}
+ \begin{aligned}
+ x' &=\rho{w_A\over \overline w}x +(1-\rho)x \\
+    &= {w_A\over \overline w}x\bigl[(1-\rho)x +\rho\bigr] +{w_B\over \overline w}(1-x)\bigl[(1-\rho)x\bigr],
+ \end{aligned}
  \label{2}\end{equation}
@@
  \an{
- x' &=& x\left[1+\rho(1-x){w_A-w_B\over \overline w}\right]\\
-     &=&x\cdot{x(1-\rho)(w_A-w_B) +\rho w_A +(1-\rho)w_B\over x(w_A-w_B) +w_B}.}
+ x' &= x\left[1+\rho(1-x){w_A-w_B\over \overline w}\right] \\
+     &=x\cdot{x(1-\rho)(w_A-w_B) +\rho w_A +(1-\rho)w_B\over x(w_A-w_B) +w_B}.}
     \label{4} \end{equation}
@@
 \begin{equation}\begin{array}{ccc}
-F'_A(x) &=& {x^2(1-\rho)\left(W-w\right)^2 +2xw(1-\rho)(W-w) +w\left[\rho W +(1-\rho)w\right]\over \left[x(W-w) +w\right]^2}\\
+F'_A(x) = {x^2(1-\rho)\left(W-w\right)^2 +2xw(1-\rho)(W-w) +w\left[\rho W +(1-\rho)w\right]\over \left[x(W-w) +w\right]^2}\\
  \end{array}\label{25}\end{equation}
@@
 \begin{equation}\begin{array}{ccc}
-F'_B(x) &=  &{x^2(1-\rho)\left(w-W\right)^2 +2xW(1-\rho)(w-W) +W\left[\rho w +(1-\rho)W\right]\over \left[x(w-W) +W\right]^2}.\\
+F'_B(x) = {x^2(1-\rho)\left(w-W\right)^2 +2xW(1-\rho)(w-W) +W\left[\rho w +(1-\rho)W\right]\over \left[x(w-W) +W\right]^2}.\\
 \end{array}\label{26}\end{equation}
@@
 
-The proof of Result 5 is in Supplementary Material SP1.  Supplementary Figure S3A shows the stable equilibrium frequencies $x^*$ as a function of $\rho$, $W$, and $w$ in the $A1B1$ regime.
+The proof of Result 5 is in SI Appendix A. Supplementary Figure S3A shows the stable equilibrium frequencies $x^*$ as a function of $\rho$, $W$, and $w$ in the $A1B1$ regime.
 For $AkBk$ selection regimes from $k=1$ to $k=40$, Supplementary Figure S4 illustrates the convergence to a single stable polymorphism.
@@
 
- The proofs of Results 6 and  7 can be found in Supplementary Materials SP2 and SP3.
+ The proofs of Results 6 and 7 can be found in SI Appendices B and C.
 
@@
 A constant environment always leads to fixation of the favored type, independent of $\rho$.
-We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$ (but see Supplementary Material SP6 for computational analysis of the general $AkBl$ case).
+We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$ (but see SI Appendix F for computational analysis of the general $AkBl$ case).
 Specifically, from \eqref{32} with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
@@
 
- The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl=\cl_2\cdot\cl_1$ near $x^*$, which is derived from \eqref{32} and given by
- \begin{equation}
- \overline w^*\cl_1 =\left[\begin{array}{cc}{W\bigl[(1-P)x_1^* +P\bigr] & w(1-P)x_1^*\\
- W(1-P)x_2^* & w\bigl[(1-P)x_2^* +P\bigr]\end{array}\right]
- \label{37}\end{equation}
- and
- \begin{equation}
- \overline w^{**}\cl_2 =\left[\begin{array}{cc}w\bigl[(1-P)x_1^{**} +P\bigr] & W(1-P)x_1^{**}\\
- w(1-P)x_2^{**} & W\bigl[(1-P)x_2^{**} +P\bigr]\end{array}\right],
- \label{38}\end{equation}
-where $\underline x^{**} =T_1\underline x^*$, $x_1^{**}=x_2^*=1-x_1^*$, $x_2^{**}=x_1^*$,  and
-\begin{equation}
-\overline w^* =Wx_1^* +wx_2^*,\qquad \overline w^{**} =wx_1^{**} +Wx_2^{**}.\label{39}\end{equation}
-Due to the symmetry between the two phenotypes $A$ and $B$ in the $A1B1$ case, we have $x_1^{**} =x_2^*$ and $x_2^{**} =x_1^*$ so that $\overline w^{**} =\overline w^*$ and in fact
-\begin{equation}
-\overline w^*\cl_2 = \left[\begin{array}{cc}w\bigl[(1-P)x_2^* +P\bigr] & W(1-P)x_2^*\\
- w(1-P)x_1^* & W\bigl[(1-P)x_1^* +P\bigr]\end{array}\right].
- \label{40}\end{equation}
- Note that as $\underline x^* =T_2(T_1\underline x^*)$ with $x_3^*=x_4^*=0$, from \eqref{37} and \eqref{38} with $P=\rho$ we have
- \begin{equation}\left[\begin{array}{c}x_1^*\\ x_2^*\end{array}\right]= \cl_2\cdot\cl_1\left[\begin{array}{c}x_1^*\cr x_2^*\end{array}\right] =\cl\left[\begin{array}{c}x_1^*\\ x_2^*\end{array}\right].
- \label{41}\end{equation}
- Hence when $P=\rho$ one of the eigenvalues of $\cl$ is 1.
+ The external stability of $\underline x^*$ to the introduction of the modifier allele $M$ with rate $P$ is determined by the linear approximation matrix $\cl$.
+ %=\cl_2\cdot\cl_1$.
+%  near $x^*$, which is derived from \eqref{32} and given by
+% \begin{equation}
+% \overline w^*\cl_1 =\left[\begin{array}{cc}{W\bigl[(1-P)x_1^* +P\bigr] & w(1-P)x_1^*\\
+% W(1-P)x_2^* & w\bigl[(1-P)x_2^* +P\bigr]\end{array}\right]
+% \label{37}\end{equation}
+% and
+% \begin{equation}
+% \overline w^{**}\cl_2 =\left[\begin{array}{cc}w\bigl[(1-P)x_1^{**} +P\bigr] & W(1-P)x_1^{**}\\
+% w(1-P)x_2^{**} & W\bigl[(1-P)x_2^{**} +P\bigr]\end{array}\right],
+% \label{38}\end{equation}
+%where $\underline x^{**} =T_1\underline x^*$, $x_1^{**}=x_2^*=1-x_1^*$, $x_2^{**}=x_1^*$,  and
+%\begin{equation}
+%\overline w^* =Wx_1^* +wx_2^*,\qquad \overline w^{**} =wx_1^{**} +Wx_2^{**}.\label{39}\end{equation}
+%Due to the symmetry between the two phenotypes $A$ and $B$ in the $A1B1$ case, we have $x_1^{**} =x_2^*$ and $x_2^{**} =x_1^*$ so that $\overline w^{**} =\overline w^*$ and in fact
+%\begin{equation}
+%\overline w^*\cl_2 = \left[\begin{array}{cc}w\bigl[(1-P)x_2^* +P\bigr] & W(1-P)x_2^*\\
+% w(1-P)x_1^* & W\bigl[(1-P)x_1^* +P\bigr]\end{array}\right].
+% \label{40}\end{equation}
+% Note that as $\underline x^* =T_2(T_1\underline x^*)$ with $x_3^*=x_4^*=0$, from \eqref{37} and \eqref{38} with $P=\rho$ we have
+% \begin{equation}\left[\begin{array}{c}x_1^*\\ x_2^*\end{array}\right]= \cl_2\cdot\cl_1\left[\begin{array}{c}x_1^*\cr x_2^*\end{array}\right] =\cl\left[\begin{array}{c}x_1^*\\ x_2^*\end{array}\right].
+% \label{41}\end{equation}
+% Hence when $P=\rho$ one of the eigenvalues of $\cl$ is 1.
+% 
+% In general $\cl=\cl_2\cdot\cl_1$, and using \eqref{37} and \eqref{40} we have
+%  \begin{equation}
+%  \an{
+% \left(\overline w^*\right)^2\cl_{11} &=Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[w(1-P)x_1^*\bigr]^2 \\
+% \left(\overline w^*\right)^2\cl_{12} &=W(1-P)\bigl[P+x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr]\\
+% \left(\overline w^*\right)^2\cl_{21} &=w(1-P)\bigl[1-x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr]\\
+% \left(\overline w^*\right)^2\cl_{22} &= Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[W(1-P)x_2^*\bigr]^2
+% }
+% \label{42}\end{equation}
+% The external stability of $\underline x^*$ is determined by the eigenvalues of $\cl$, namely the roots of its characteristic polynomial $R(\lambda) =\det(\cl-\lambda\ci)$, with $\ci$ the $2\times 2$ identity matrix.
+From \eqref{42}, $R(\lambda)=a_2\lambda^2 +a_1\lambda +a_0$, where
+% \begin{equation}
+% \an{
+% &a_0={P^2W^2w^2\over\left(\overline w^*\right)^4},\\
+% & a_1=-{2PWw +\left(1-P\right)^2\left[Wx_2^* +wx_1^*\right]^2\over \left(\overline w^*\right)^2},\quad a_2=1.}
+% \label{43}\end{equation}
 
- In general $\cl=\cl_2\cdot\cl_1$, and using \eqref{37} and \eqref{40} we have
-  \begin{equation}
-  \an{
- \left(\overline w^*\right)^2\cl_{11} &=Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[w(1-P)x_1^*\bigr]^2 \\
- \left(\overline w^*\right)^2\cl_{12} &=W(1-P)\bigl[P+x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr]\\
- \left(\overline w^*\right)^2\cl_{21} &=w(1-P)\bigl[1-x_1^*(1-P)\bigr]\bigl[Wx_2^* +wx_1^*\bigr]\\
- \left(\overline w^*\right)^2\cl_{22} &= Ww\left[(1-P)^2x_1^*x_2^* +P\right] +\bigl[W(1-P)x_2^*\bigr]^2
- }
- \label{42}\end{equation}
- The external stability of $\underline x^*$ is determined by the eigenvalues of $\cl$, namely the roots of its characteristic polynomial $R(\lambda) =\det(\cl-\lambda\ci)$, with $\ci$ the $2\times 2$ identity matrix.
-From \eqref{42}, $R(\lambda)=a_2\lambda^2 +a_1\lambda +a_0$, where
- \begin{equation}
- \an{
- &a_0={P^2W^2w^2\over\left(\overline w^*\right)^4},\\
- & a_1=-{2PWw +\left(1-P\right)^2\left[Wx_2^* +wx_1^*\right]^2\over \left(\overline w^*\right)^2},\quad a_2=1.}
- \label{43}\end{equation}
-
-We then have the following result, which is proved in Supplementary Material SP4.
- \medskip
+%
+%We then have the following result, which is proved in SI Appendix D.
+% \medskip
+We have the following result, which is proved in SI Appendix D.
 
@@
 The x-axis is on a log-scale, as each sequential invasion takes an order of magnitude longer to complete.
-See Figure S12 for $w=0.1$ and $0.9$.}
+See Supplementary Figure S12 for $w=0.1$ and $0.9$.}
 \label{fig:A1B1_modifier_invasions}
@@
 
-Values of the evolutionarily stable vertical transmission rate, $\rho^*$, for some $AkBl$ examples (see Supplementary Material SP7 and  Figure S10 for analytical details) are recorded in Table 1 for different values of $w$ relative to $W=1$.
-Interestingly, for the $A1B2$ regime, with $w=0.1$, the evolutionarily stable value of $\rho$ is substantially greater than zero, as it is in the $A3B10$ and $A5B30$ regimes.
+Values of the evolutionarily stable vertical transmission rate, $\rho^*$, for some $AkBl$ examples (see SI Appendix G and  Figure S10 for analytical details) are recorded in Table 1 for different values of $w$ relative to $W=1$.
+Interestingly, with $w=0.1$, the evolutionarily stable value of $\rho$ is zero for the $A1B2$ regime, but not for the $A3B10$ and $A5B30$ regimes, in which the only stable values are those that lead to fixation of phenotype $B$ (e.g.
+$\rho>0.4625$ and $\rho>0.1489$, respectively), and are therefore neutrally-stable (Supplementary Figure S10).
 $AkBk$ results are plotted in Figure 6B.
@@
 where $Z =\sqrt{(1-\rho)^2(W-w)^2 +4Ww}>0$.
-The mean fitness at the stable equilibrium  is $\overline w^*\cdot\overline w^{**}$, and as $\overline w^* =\overline w^{**}$  by  \eqref{39}, this allows us to reduce the problem to properties of $\overline w^*$.
+The geometric mean fitness at the stable equilibrium is $\overline w^*\cdot \overline w^{**}$, where $\overline w^{**}$ is the mean fitness at the middle of the $A1B1$ cycle.
+$\overline w^* =\overline w^{**}$ due to the symmetry between the two phenotypes $A$ and $B$ in this case, which allows us to reduce the problem to properties of $\overline w^*$.
 Now since $W>w$,  $\overline w^*$ is an increasing linear function of $x^*$:
@@
 Figure 1B shows the geometric mean fitness in the $A1B2$ regime, and we see that for small values of $w$, this mean fitness increases as $\rho$ decreases.
-At $w=0.1$, Table 1 shows that $\hat\rho$ is 0.00065, whereas $\rho^*=0.821$.
+At $w=0.1$, Table 1 shows that $\hat\rho$ and $\rho^*$ are roughly 0.
 In all $AkBk$ regimes that we tested with $w=0.9$, the value of $\hat\rho$ was also zero, substantially different from the values of $\rho^*$, as shown in Figures 6A and 6B.
@@
 In Figure 6, with $w=0.1$, $\hat\rho$ is between 0.15 and 0.24 for $12\le k\le 50$, while with $w=0.5$, $\hat\rho$ exceeds 0.2 for $31\le k\le 50$.
-More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Table 1, Supplementary Figures S7 and S8, and Supplemental Material SP7.
+More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Table 1, Supplementary Figures S7 and S8, and SI Appendix G.
+
+%Figure 6
+\begin{figure}[h]
+\centering
+\includegraphics[width=\linewidth]{../../figures/AkBk_stable_optimal_rate.pdf}
+\caption{Fitness ``optimal'' and evolutionary stable vertical transmission rate in $AkBk$ selection regime. 
+\textbf{(A)} The vertical transmission rate $\hat{\rho}$ that maximized the geometric average of the population mean fitness is zero (complete oblique transmission) when selection cycles quickly between favoring phenotype $A$ and $B$, and then abruptly transitions to $\approx 0.2$, followed by a slow decrease (see Fig.
+S5 and S6 for details on the abrupt transition).
+\textbf{(B)} The evolutionary stable rate $\rho^*$, which cannot be invaded by modifiers with either higher or lower vertical transmission rate $P$, rapidly increases from zero when selection cycles are short ($k=1$ or $2$) to roughly 1 when selection cycles are longer.
+The dashed line shows $1-\frac{1}{k-1}$, which fit the values for $w=0.5$ (Carja et al.
+2011).
+The values for $w=0.1$ (blue) could not be calculated for $k > 19$ due to numerical instability when selection is strong and the duration between selection fluctuations is long.
+In all cases, $W=1$. See SI Appendix F for details on how we calculated the stable rate.}\label{fig:AkBk_stable_optimal_rate}
+\end{figure}
 
@@
 \centering
-\caption{Values of $\rho^*$ (stable $\rho$)$^{\hbox{\sevenrm (a)}}$ and $\hat\rho$ (``optimal'' $\rho$)$^{\hbox{\sevenrm (b)}}$}
+\caption{Values of $\rho^*$ (stable $\rho$)$^{\text{(a)}}$ and $\hat\rho$ (``optimal'' $\rho$)$^{\text{(b)}}$}
 \begin{tabular}{rrcll}
@@
 
-\addtabletext{\noindent $^\dagger$Note: $W=1$.\hfill\par
+\addtabletext{
+\noindent $^\dagger$Note: $W=1$.\hfill\par
 \noindent (a): $\rho^*$ is the uninvadable value of the vertical transmission rate.\hfill\par
-(b): $\hat\rho$ is the geometric mean value of $\rho$ at the stable equilibrium of the $AkBl$ cycle.\hfill}
+\noindent (b): $\hat\rho$ is the rate that maximizes the geometric mean fitness at the stable equilibrium of the $AkBl$ cycle.\hfill}
 \end{table}
-
-%Figure 6
-\begin{figure}[h]
-\centering
-\includegraphics[width=\linewidth]{../../figures/AkBk_stable_optimal_rate.pdf}
-\caption{Fitness ``optimal'' and evolutionary stable vertical transmission rate in $AkBk$ selection regime. 
-\textbf{(A)} The vertical transmission rate $\hat{\rho}$ that maximized the geometric average of the population mean fitness is zero (complete oblique transmission) when selection cycles quickly between favoring phenotype $A$ and $B$, and then abruptly transitions to $\approx 0.2$, followed by a slow decrease (see Fig.
-S5 and S6 for details on the abrupt transition).
-\textbf{(B)} The evolutionary stable rate $\rho^*$, which cannot be invaded by modifiers with either higher or lower vertical transmission rate $P$, rapidly increases from zero when selection cycles are short ($k=1$ or $2$) to roughly 1 when selection cycles are longer.
-The dashed line shows $1-\frac{1}{k-1}$, which fit the values for $w=0.5$ (Carja et al.
-2011).
-The values for $w=0.1$ (blue) could not be calculated for $k > 19$ due to numerical instability when selection is strong and the duration between selection fluctuations is long.
-In all cases, $W=1$. See \emph{Supplemental Material SP6} for details on how we calculated the stable rate.}\label{fig:AkBk_stable_optimal_rate}
-\end{figure}
 
@@
  This Markov chain  has two absorbing states, $X_t =N$ and $X_t=0$, corresponding to the two fixations in $A$ and $B$, respectively, and we are interested in the fixation probabilities and the time to fixation of these two absorbing states as functions of the initial frequency $x$ and also of $\rho$, $w_A$, and $w_B$.
- To these ends we use a diffusion approximation to the process $\{X_t\}$, which
-% Specifically, we compute the mean $\mu(x)$ and the variance $\sigma^2(x)$ of the change in one generation  in the frequency of phenotype $A$ given that at the beginning of the generation $X_t=Nx$.
-% 
-% To compute $\mu(x)$, observe that by \eqref{47}   
-% \begin{equation}
-% \an{
-% x' -x &={w_Ax\over\overline w}\rho +(1-\rho)x -x =\rho x\left[{w_A\over \overline w}-1\right]\cr
-% &= \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},}
-% \label{49}\end{equation}
-% since $\overline w =w_A x +w_B(1-x)$.
-% For the diffusion approximation, it is essential that the differential selection does not have a large effect per individual in each time period  $\Delta t$ $\bigl(\Delta t\simeq{1\over N}\bigr)$.
-That is, we assume
-% \begin{equation}
-% w_A -w_B ={s\over N}.
-% \label{50}\end{equation}
-% Then
-% \begin{equation}
-% x'-x\simeq {1\over N}\rho sx(1-x)
-% \label{51}\end{equation}
-% up to terms of order small than ${1\over N}$. Since one generation corresponds to $\Delta t\simeq {1\over N}$, we  conclude that
-% \begin{equation}
-% \mu(x) =\rho sx(1-x);\quad 0\le x\le 1.
-% \label{52}\end{equation}
-% In the same way, we can compute $\sigma^2(x)$, which is $x(1-x)$. These values of $\mu(x)$ and $\sigma^2(x)$ 
- allows us to compute $u(x)$, the probability that phenotype $A$ goes to fixation when it's initial frequency is $x$, namely
+ To these ends we use a diffusion approximation to the process $\{X_t\}$, which allows us to compute $u(x)$, the probability that phenotype $A$ goes to fixation when it's initial frequency is $x$, namely
   \begin{equation}
@@
  \label{54}\end{equation}
- where $u(x)$ is given in \eqref{53}, and in generations, $T(x)$ is multiplied by $N$ (the derivation is in Supplementary Material SPX).
+ where $u(x)$ is given in \eqref{53}, and in generations, $T(x)$ is multiplied by $N$ (the derivation is in SI Appendix H).
   Unfortunately the integrals in \eqref{54} cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ as Ewens points out on p.\ 160 \cite{ewens2004mathematical}, and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.
@@
 
- The proof of Result 10 is in Supplementary Material SP5.  
+ The proof of Result 10 is in SI Appendix E.  
   Supplementary Figure S9 compares the fixation probability and time to fixation derived numerically from  simulating the Wright-Fisher Markov chain with the diffusion-derived values of $u(x)$ and $T(x)$.
@@
 
-\noindent The proof of Result 11, based on induction on $n$, is given in Supplemental Materials SP6.
+\noindent The proof of Result 11, based on induction on $n$, is given in SI Appendix F.
 
@@
 We conjecture that with $k\ne l$ there is a unique stable polymorphism when both fixations are unstable.
-This result is similar to the \emph{storage effect} \cite{chesson1981}, in which protection from selection maintains species coexistence: Consider a two species $A$ and $B$, with overlapping generations, an equal death rate $\rho$, and different growth rates $w_A$ and $w_B$; then \eqref{2} will model the change in frequency of species $A$.
+This result is similar to the \emph{storage effect} \cite{chesson1981}, in which protection from selection maintains species coexistence: consider two species $A$ and $B$, with overlapping generations, an equal death rate $\rho$, and different growth rates $w_A$ and $w_B$; then \eqref{2} will model the change in frequency of species $A$.
 In our model, oblique transmission can be said to protect the unfavored phenotype from selection because it allows transmission without reproduction.
@@
 This is an unusual scenario for genetic modifiers, although it must be noted that a modifier of $\rho$ is not neutral; it affects primary selection, while neutral  modifiers of recombination, mutation, and migration affect induced or secondary selection.
-To explain the marked difference in the value of the stable vertical transmission rate consider the probability $q$ that offspring receive phenotype $A$ when it is about to be the favored phenotype, and for simplicity, assume strong selection, such that parents are likely to be of the phenotype favored in the parental generation, a periodically changing environment $AkBl$, and a population at equilibrium.
-With vertical transmission, $q_v=\frac{k-1}{k}$.
-With oblique transmission, $q_o=x$ the frequency of the favored phenotype in the parental generation.
-For $k=l$, the frequency of $A$ cycles around 0.5 so $q_o \approx 0.5$, whereas $q_v=0, {1 \over 2}, {2 \over 3}, {3 \over 4}, \ldots$ for $k=1, 2, 3, 4, \ldots$.
-Therefore, for $k>2$ vertical transmission increases $q$ whereas for $k=1$ and $2$, oblique transmission increases $q$.
+To explain the marked difference in the value of the stable vertical transmission rate consider the probability $q$ that offspring receive phenotype $A$ when it is about to be the favored phenotype, and for simplicity, assume strong selection, such that parents are likely to be of the phenotype favored in the parental generation, a periodically changing environment $AkBk$, and a population at equilibrium.
+With oblique transmission, $q$ roughly equals the frequency of the favored phenotype in the parental generation.
+With vertical transmission, $q=\frac{k-1}{k} = 0, {1 \over 2}, {2 \over 3}, {3 \over 4}, \ldots$ for $k=1, 2, 3, 4, \ldots$.
+Therefore, for large $k$, $q$ is likely to increase with the vertical transmission rate.
 

### 2017-12-14 ms/PNAS/revision.tex
MWF and UL changes

@@
 In a recent analysis of evolution under fluctuating selection, Xue and Leibler \cite{xue2016evolutionary} allowed an organism to absorb information about the distribution of possible environments by learning the phenotypes of members of its parental lineage from previous generations.
-They describe this as ``positive feedback that enhances the probability of the offspring to express the same phenotype as the parent.'' In this formulation there was ``reinforcement of the parent phenotype'' in an offspring, such as might occur through epigenetic inheritance.
+They describe this as ``positive feedback that enhances the probability the probability that the offspring expresses the same phenotype as the parent.'' In this formulation there was ``reinforcement of the parent phenotype'' in an offspring, such as might occur through epigenetic inheritance.
 Although their analysis was not couched in terms of oblique and vertical transmission, as defined by Cavalli-Sforza and Feldman \cite{cavalli1981cultural}, we have been stimulated by their analysis to develop a model in which oblique transmission, at a rate dependent on the trait frequency in the parental generation, occurs in addition to classical vertical transmission.
@@
 
-If $\rho=0$ or $w_A=w_B>0$, the phenotype frequencies do not change over time.
-With {\sl extreme selection},  for example $w_A=1$ and $w_B=0$ (or $w_B=1$ and $w_A=0$),  \eqref{4} reduces to
-$x'=(1-\rho)x+\rho$, and therefore 
-$x_t=1-(1-\rho)^t(1-x_0)$.
-Hence, for $\rho>0$,  $x_t\to 1$ as $t\to\infty$ for all initial values $0\le x_0\le 1$, and  the favored phenotype fixes.
+%If $\rho=0$ or $w_A=w_B>0$, the phenotype frequencies do not change over time.
+%With {\sl extreme selection},  for example $w_A=1$ and $w_B=0$ (or $w_B=1$ and $w_A=0$),  \eqref{4} reduces to
+%$x'=(1-\rho)x+\rho$, and therefore 
+%$x_t=1-(1-\rho)^t(1-x_0)$.
+%Hence, for $\rho>0$,  $x_t\to 1$ as $t\to\infty$ for all initial values $0\le x_0\le 1$, and  the favored phenotype fixes.
 
@@
 
- Suppose the environment is not constant but rather changes periodically, such that the favored phenotype changes after a fixed number of generations.
+ Suppose the environment changes periodically, such that the favored phenotype changes after a fixed number of generations.
 Simple examples are $A1B1=ABABAB\dots$, in which the favored phenotype switches every generation, or $A2B1=AABAABAAB\dots$, where every two generations in which selection favors $A$ are followed by a single generation in which selection favors $B$.
@@
 
-\noindent 1.\ \ Since $k=l$ and the above result also holds  when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is also not stable.
+\noindent 1.\ \ Since $k=l$ and the above result also holds  when $0<W<w$, there is total symmetry between the two fixations in $A$ and $B$, and fixation in $A$ is also unstable.
 Thus neither phenotype can be lost, and there is a {\sl protected polymorphism} \cite{prout1968sufficient}.
@@
 Let $a=1+\rho{W-w\over w}$ and $b=1+\rho{w-W\over W}$, and our assumption entails  $a>1$ and $0<b<1$.
-Following \eqref{11}, fixation in $B$ is not stable if $a^kb^l>1$, and similarly  fixation in $A$ is unstable if $a^lb^k>1$.
+Following \eqref{11}, fixation in $B$ is unstable if $a^kb^l>1$, and similarly  fixation in $A$ is unstable if $a^lb^k>1$.
 Therefore both fixations are unstable if
@@
 \begin{equation}k\log a +l\log b>0\quad\hbox{\tenrm and}\quad l\log a+k\log b>0.\label{19}\end{equation}
-Now \eqref{19} holds if and only if
+Now the inequalities of \eqref{19} hold if and only if
 \begin{equation}k\,{\log(1/b)\over\log a} < l < k\,{\log a\over\log(1/b)}.\label{20}\end{equation}
 
-The inequalities of \eqref{20} are consistent if and only if $\log(1/b)<\log a$, i.e., $ab>1$, which according to \eqref{14}, is true.
+These inequalities are consistent if and only if $\log(1/b)<\log a$, i.e., $ab>1$, which is true by \eqref{14}.
 
@@
 
-\noindent {\bf Result 4}. {\sl In the case $A1B1$ with $0<\rho<1$ and $0<w<W$ the two fixations are unstable and there exists a unique stable polymorphism.}\par
+\noindent {\bf Result 4}. {\sl In the case $A1B1$ with $0<\rho<1$ and $0<w<W$, the two fixations are unstable and there exists a unique stable polymorphism.}\par
 \medskip
 
-\noindent{\it Proof}. Let $x$ be the initial frequency of $A$ and $x'$ its frequency  after one cycle of the $A1B1$ selection. Then $x' =F_B\bigl(F_A(x)\bigr)$ where by \eqref{8}
+\noindent{\it Proof}. Let $x$ be the initial frequency of $A$ and $x'$ its frequency  after one cycle of  $A1B1$ selection. Then $x' =F_B\bigl(F_A(x)\bigr)$ where by \eqref{8}
 \begin{equation}
@@
 The equilibrium equation is $x=F_B\bigl(F_A(x)\bigr)$, which reduces to a fourth degree polynomial equation in $x$.
-Since the two fixations in $B$ and $A$ are equilibria corresponding to the two solutions $x=0$ and $x=1$, the other equilibria correspond to solutions of a quadratic equation $Q(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$ with
-\begin{equation}\alpha_2=1,\  \alpha_1={W+w\over (2-\rho)(W-w)}-1,\  \alpha_0={-w\over (2-\rho)(W-w)}.\label{22}\end{equation}
-As $0<\rho<1$ and $0<w<W$ we have
+Since the  fixations in $B$ and $A$ are equilibria corresponding to the  solutions $x=0$ and $x=1$, the other equilibria correspond to solutions of a quadratic equation $Q(x) =\alpha_2x^2 +\alpha_1x +\alpha_0=0$, with $\alpha_2=1$ and
+\begin{equation}\alpha_1={W+w\over (2-\rho)(W-w)}-1,\  \alpha_0={-w\over (2-\rho)(W-w)}.\label{22}\end{equation}
+As $0<\rho<1$ and $0<w<W$, we have
 \begin{equation}
 \an{
-Q(0) &= {-w\over (2-\rho)(W-w)}<0\\
+Q(0) &= {-w\over (2-\rho)(W-w)}<0,\\%MOVE "AND" TO MIDDLE
 \hbox{\tenrm and}\ \ Q(1) &={W\over (2-\rho)(W-w)} >0.}\label{23}\end{equation}
@@
 Starting from any initial value $0<x_0 <1$ we have  $x_t\to x^*$ as $t\to\infty$.
-Supplementary Figure S2, panels A, C, and E, illustrate examples of how the frequency of $A$ changes over time in the $A1B1$ regime of cycling selection.
+Supplementary Figure S2, panels A, C, and E, illustrate  how the frequency of $A$ changes over time in the $A1B1$ regime of cycling selection.
 
@@
 \centering
-\includegraphics[width=\linewidth]{../../figures/A1B2_polymorphism_vert.pdf}
+\includegraphics[width=\linewidth]{A1B2_polymorphism_vert.pdf}
 \caption{Stable frequency of phenotype \emph{A} and geometric mean fitness in selection regime
 $A1B2$ as a function of the vertical transmission rate \(\rho\) and the fitness of the disfavored phenotype \(w\).
-\textbf{(A)} Stable frequency of phenotype $A$ at the end of each three generations cycle.
-\textbf{(B)} Geometric average of the stable population mean fitness over the three generations cycle: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
+\textbf{(A)} Stable frequency of phenotype $A$ at the end of each three-generation cycle.
+\textbf{(B)} Geometric average of the stable population mean fitness over the three-generation cycle: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
 Gray contour lines join \(\rho\) and \(w\) combinations that result in the same stable value.
@@
 However, the numerical examples in Supplementary Figure S1 (for $AkBl$) and in Figure 1 and Supplementary Figure S5 for the special case $A1B2$ all exhibit a single stable polymorphic equilibrium when fixations in $A$ and $B$ are unstable.
-These numerical results suggest that for $W>w>0$ and $0<\rho<1$ the high order polynomial that gives the equilibria has only a single root corresponding to a globally stable polymorphism.
+These numerical results suggest that for $W>w>0$ and $0<\rho<1$ the high order equilibrium polynomial has only a single root corresponding to a globally stable polymorphism.
 Supplementary Figure S6 shows that this is the case for the $A3B10$ regime.
@@
 Without loss of generality, assume that the fitness parameters at generation $t$ ($t=0,1,2,\dots$) are $1+s_t$ for phenotype $A$ and $1$ for phenotype $B$, where the random variables $s_t$ for $t=0,1,2,\dots$ are independent and identically distributed.
-Also we assume that there are positive constants $C$ and $D$ such that $P(-1+C<s_t <D)=1$.
+Also assume that there are positive constants $C$ and $D$ such that $P(-1+C<s_t <D)=1$.
 
@@
 
- The proofs of Results 6 and 7 can be found in SI Appendices B and C.
+ The proofs of Results 6 and 7 are in SI Appendices B and C.
 
- Using the general notation for the fitness parameters $w_A$ and $w_B$, the stochastic local stability of the $B$ fixation is determined by the sign of $E\left[\log\left(1-\rho+\rho{w_A\over w_B}\right)\right]$ and that of the $A$ fixation by the sign of $E\left[\log\left(1-\rho+\rho{w_B\over w_A}\right)\right]$.
-For example, if the  sign of the first  is negative, fixation in $B$ is stochastically locally stable, and when it is positive, with probability one convergence to fixation in phenotype $B$ does not occur.
+ Using the general notation for the fitness parameters $w_A$ and $w_B$, the stochastic local stability of fixation in $B$ is determined by the sign of $E\left[\log\left(1-\rho+\rho{w_A\over w_B}\right)\right]$ and that of fixation in $A$  by the sign of $E\left[\log\left(1-\rho+\rho{w_B\over w_A}\right)\right]$.
+For example, if the  sign of the first  is negative, fixation in $B$ is stochastically locally stable, and when it is positive, with probability one convergence to fixation in  $B$ does not occur.
 It is also true  that if $E(w_A/w_B)\le 1$, then fixation of $B$ is stochastically locally stable.
@@
 \centering
-\includegraphics[width=\linewidth]{../../figures/stochastic_env_x_t.pdf}
+\includegraphics[width=\linewidth]{stochastic_env_x_t.pdf}
 \caption{Stochastic local stability. 
@@
 \centering
-\includegraphics[width=\linewidth]{../../figures/stoch_two_iid_rvs.pdf}
+\includegraphics[width=\linewidth]{stoch_two_iid_rvs.pdf}
 \caption{Effect of vertical transmission rate $\rho$ on phenotype polymorphism in a randomly changing environment.
@@
 Let the vertical transmission rates determined by $m$ and $M$ be $\rho$ and $P$, respectively.
-Thus there are four pheno-genotypes $mA$, $mB$, $MA$, $MB$ whose frequencies at a given generation are denoted by $x_1$, $x_2$, $x_3$, $x_4$, respectively.
+Thus there are four pheno-genotypes: $mA$, $mB$, $MA$, $MB$, whose frequencies at a given generation are denoted by $x_1$, $x_2$, $x_3$, $x_4$, respectively.
 As the fitnesses are determined by the two phenotypes $A$ and $B$, and the modifier locus is selectively neutral, we have the following table.
@@
  \end{array}\label{31}\end{equation}
-Following the rationale leading to equation [2], the next generation pheno-genotype frequencies $x'_1$, $x'_2$, $x'_3$, $x'_4$ are
+Following the rationale leading to \eqref{2}, the next generation pheno-genotype frequencies $x'_1$, $x'_2$, $x'_3$, $x'_4$ are
 \begin{equation}
@@
 % \label{43}\end{equation}
-
  %
@@
 % \medskip
-We have the following result, which is proved in SI Appendix D.
+We prove the following result in SI Appendix D.
+\medskip
 
@@
 \centering
-\includegraphics[width=\linewidth]{../../figures/{A1B1_modifier_invasions_w0.5}.pdf}
+\includegraphics[width=\linewidth]{{A1B1_modifier_invasions_w0.5}.pdf}
 \caption{Consecutive fixation of modifiers that reduce the vertical transmission rate in selection regime \emph{A1B1}.
@@
 \centering
-\includegraphics[width=\linewidth]{../../figures/stoch_modifier_invasions.pdf}
+\includegraphics[width=\linewidth]{stoch_modifier_invasions.pdf}
 \caption{Consecutive fixation of modifiers that reduce the vertical transmission rate $\rho$ under symmetric randomly changing selection.
@@
 Values of the evolutionarily stable vertical transmission rate, $\rho^*$, for some $AkBl$ examples (see SI Appendix G and  Figure S10 for analytical details) are recorded in Table 1 for different values of $w$ relative to $W=1$.
-Interestingly, with $w=0.1$, the evolutionarily stable value of $\rho$ is zero for the $A1B2$ regime, but not for the $A3B10$ and $A5B30$ regimes, in which the only stable values are those that lead to fixation of phenotype $B$ (e.g.
-$\rho>0.4625$ and $\rho>0.1489$, respectively), and are therefore neutrally-stable (Supplementary Figure S10).
+Interestingly, with $w=0.1$, the evolutionarily stable value of $\rho$ is zero for the $A1B2$ regime, but not for the $A3B10$ and $A5B30$ regimes, in which the only stable values are those that lead to fixation of phenotype $B$ (e.g., $\rho>0.4625$ and $\rho>0.1489$, respectively); these are therefore neutrally-stable (Supplementary Figure S10).
 $AkBk$ results are plotted in Figure 6B.
@@
 We can ask the same question here: is the stable rate $\rho^*$ the same as the rate $\hat\rho$ that maximizes the equilibrium value of the geometric mean fitness under fluctuating selection?
-First consider the $A1B1$ selection regime. We have the following result.
+First, for  the $A1B1$ selection regime, we have the following result.
 \medskip
@@
 \centering
-\includegraphics[width=\linewidth]{../../figures/AkBk_stable_optimal_rate.pdf}
+\includegraphics[width=\linewidth]{AkBk_stable_optimal_rate.pdf}
 \caption{Fitness ``optimal'' and evolutionary stable vertical transmission rate in $AkBk$ selection regime. 
@@
 
- In order to include the effect of random drift due to finite population in the above deterministic model, we use the Wright-Fisher model.
+ In order to include the effect of random drift due to finite population size in the above deterministic model, we use the Wright-Fisher model.
 Let $X_t$ denote the number of individuals with phenotype $A$ in a population of fixed size $N$ at the $t$-th generation, and suppose $X_t=Nx$.
@@
  \begin{equation}
- P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\matrix{N\cr j\cr}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},
+ P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\begin{array}{c} N\\j\\ \end{array}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},
  \label{48}\end{equation}
@@
  where $u(x)$ is given in \eqref{53}, and in generations, $T(x)$ is multiplied by $N$ (the derivation is in SI Appendix H).
-  Unfortunately the integrals in \eqref{54} cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ as Ewens points out on p.\ 160 \cite{ewens2004mathematical}, and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.
+  Unfortunately the integrals in \eqref{54} cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ (see \cite{ewens2004mathematical}, p.\ 160), and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.
 
-  It is important to note that the fixation probability $u(x)$ is a monotone increasing function of $\rho$ when $s>0$. In fact we have the following result.
+  For the fixation probability $u(x)$,  we have the following result.
  \medskip
@@
  \label{58}\end{equation}
- Figure 7 shows an example of how $(k-l)$, $\rho$, and $(W-w)$, which enter the formula for $u(x)$ in \eqref{57}, interact to affect fixation probabilities.
-More examples are illustrated in Supplementary Figure S11.
+ Figure 7 shows an example of how $(k-l)$, $\rho$, and $(W-w)$  in \eqref{57} for $u(x)$, interact to affect fixation probabilities. More examples are illustrated in Supplementary Figure S11.
 
@@
 \centering
-\includegraphics[width=\linewidth]{../../figures/lk_fix_prob_C.pdf}
+\includegraphics[width=\linewidth]{lk_fix_prob_C.pdf}
 \caption{Fixation in a finite population with different ratios of selection periods \(\frac{k}{l}\).
@@
 In the $AkBl$ case, bounds on $l/k$ that determine the instability of both fixations, and hence the protection of polymorphism are given by the inequalities of \eqref{20}, which depend on both the fitness differences and the rate $\rho$ of vertical transmission.
-We conjecture that with $k\ne l$ there is a unique stable polymorphism when both fixations are unstable.
-This result is similar to the \emph{storage effect} \cite{chesson1981}, in which protection from selection maintains species coexistence: consider two species $A$ and $B$, with overlapping generations, an equal death rate $\rho$, and different growth rates $w_A$ and $w_B$; then \eqref{2} will model the change in frequency of species $A$.
-In our model, oblique transmission can be said to protect the unfavored phenotype from selection because it allows transmission without reproduction.
+We conjecture that with $k\ne l$ there is a unique stable polymorphism if both fixations are unstable.
+This result is similar to the \emph{storage effect} \cite{chesson1981}, in which protection from selection maintains species coexistence: consider two species $A$ and $B$, with overlapping generations, an equal death rate $\rho$, and different growth rates $w_A$ and $w_B$; then \eqref{2} describe the change in frequency of species $A$.
+In our model, oblique transmission can be said to protect the disfavored phenotype from selection because it allows transmission without reproduction.
 
  In deterministic one-locus, two-allele diploid population genetic models with cycling fitness regimes, Haldane and Jayakar \cite{haldane1963polymorphism} first showed the relevance of the geometric mean of genotypic fitnesses (cf.\ \eqref{16} and \eqref{17}) for the maintenance (or loss) of polymorphism.
-However, with equal homozygote fitnesses, which alternated in strength as a two-generation cycle (cf.\ $A1B1$), Karlin and Liberman \cite{karlin1975random} extended the Haldane-Jayakar results and found conditions under which both allelic fixations and polymorphic equilibrium could all be stable, with the evolution depending on initial allele frequencies, as well as the homozygote fitness differences between alternate generations.
+However, with equal homozygote fitnesses, alternating in strength as a two-generation cycle (cf.\ $A1B1$), Karlin and Liberman \cite{karlin1975random} extended the Haldane-Jayakar results and found conditions under which both allelic fixations and polymorphic equilibrium could all be stable, with the evolution depending on initial allele frequencies, as well as the homozygote fitness differences between alternate generations.
 Our haploid model appears not to produce such dependence on the initial conditions.
@@
 This is an unusual scenario for genetic modifiers, although it must be noted that a modifier of $\rho$ is not neutral; it affects primary selection, while neutral  modifiers of recombination, mutation, and migration affect induced or secondary selection.
-To explain the marked difference in the value of the stable vertical transmission rate consider the probability $q$ that offspring receive phenotype $A$ when it is about to be the favored phenotype, and for simplicity, assume strong selection, such that parents are likely to be of the phenotype favored in the parental generation, a periodically changing environment $AkBk$, and a population at equilibrium.
-With oblique transmission, $q$ roughly equals the frequency of the favored phenotype in the parental generation.
-With vertical transmission, $q=\frac{k-1}{k} = 0, {1 \over 2}, {2 \over 3}, {3 \over 4}, \ldots$ for $k=1, 2, 3, 4, \ldots$.
-Therefore, for large $k$, $q$ is likely to increase with the vertical transmission rate.
+%To explain the marked difference in the value of the stable vertical transmission rate consider the probability $q$ that offspring receive phenotype $A$ when it is about to be the favored phenotype, and for simplicity, assume strong selection, such that parents are likely to be of the phenotype favored in the parental generation, a periodically changing environment $AkBk$, and a population at equilibrium.
+%With oblique transmission, $q$ roughly equals the frequency of the favored phenotype in the parental generation.
+%With vertical transmission, $q=\frac{k-1}{k} = 0, {1 \over 2}, {2 \over 3}, {3 \over 4}, \ldots$ for $k=1, 2, 3, 4, \ldots$.
+%Therefore, for large $k$, $q$ is likely to increase with the vertical transmission rate.
 

### 2017-12-19 ms/PNAS/revision.tex
MWF final minor changes before resubmission

@@
 In a recent analysis of evolution under fluctuating selection, Xue and Leibler \cite{xue2016evolutionary} allowed an organism to absorb information about the distribution of possible environments by learning the phenotypes of members of its parental lineage from previous generations.
-They describe this as ``positive feedback that enhances the probability the probability that the offspring expresses the same phenotype as the parent.'' In this formulation there was ``reinforcement of the parent phenotype'' in an offspring, such as might occur through epigenetic inheritance.
+They describe this as ``positive feedback that enhances the probability that the offspring expresses the same phenotype as the parent.'' In this formulation there was ``reinforcement of the parent phenotype'' in an offspring, such as might occur through epigenetic inheritance.
 Although their analysis was not couched in terms of oblique and vertical transmission, as defined by Cavalli-Sforza and Feldman \cite{cavalli1981cultural}, we have been stimulated by their analysis to develop a model in which oblique transmission, at a rate dependent on the trait frequency in the parental generation, occurs in addition to classical vertical transmission.
@@
 \begin{equation}k\log a +l\log b>0\quad\hbox{\tenrm and}\quad l\log a+k\log b>0.\label{19}\end{equation}
-Now the inequalities of \eqref{19} hold if and only if
+Now the inequalities of (19) hold if and only if
 \begin{equation}k\,{\log(1/b)\over\log a} < l < k\,{\log a\over\log(1/b)}.\label{20}\end{equation}
@@
  In our case there are two constant equilibria $x^*=0$ and $x^*=1$ corresponding to fixation in $B$ and $A$, respectively.
-We can characterize the stochastic local stability of these fixations as follows.
+We can characterize the stochastic local stability of these fixations with the following results, whose proofs are in SI Appendices B and C.
  \medskip
@@
 In particular, if $E(s_t)\le 0$, $x^*=0$ is stochastically locally stable.}
-\smallskip
-
- The proofs of Results 6 and 7 are in SI Appendices B and C.
+%\smallskip
 
@@
  \label{58}\end{equation}
- Figure 7 shows an example of how $(k-l)$, $\rho$, and $(W-w)$  in \eqref{57} for $u(x)$, interact to affect fixation probabilities. More examples are illustrated in Supplementary Figure S11.
+ Figure 7 shows an example of how $(k-l)$, $\rho$, and $(W-w)$  in \eqref{57} for $u(x)$ interact to affect fixation probabilities. More examples are illustrated in Supplementary Figure S11.
 

### 2017-12-19 ms/PNAS/abbr_bib.py
abbreviate journal names

@@ -1,0 +1,30 @@
+#!/usr/bin/env python
+#-*- coding: utf-8 -*-
+
+import sys, os
+import re
+
+try:
+	bibtexdb = open(sys.argv[1]).read()
+except IndexError:
+	print("Error: specify the file to be processed!")
+	exit()
+
+if not os.path.isfile('journalList.txt'):
+    import urllib.request
+    urllib.request.urlretrieve("https://raw.githubusercontent.com/JabRef/jabref/master/src/main/resources/journals/journalList.txt", 
+            filename="journalList.txt")
+rulesfile = open('journalList.txt')
+
+for rule in rulesfile.readlines()[::-1]:           ## reversed alphabetical order matches extended journal names first
+    pattern1, pattern2 = rule.strip().split(" = ")
+    if pattern1 != pattern1.upper() and (' ' in pattern1):        ## avoid mere abbreviations
+    #bibtexdb = bibtexdb.replace(pattern1.strip(), pattern2.strip())    ## problem - this is case sensitive
+        repl = re.compile(re.escape(pattern1), re.IGNORECASE)               ## this is more robust, although ca. 10x slower
+        (bibtexdb, num_subs) = repl.subn(pattern2, bibtexdb)
+        if num_subs > 0:
+            print("Replacing '%s' FOR '%s'" % (pattern1, pattern2))
+
+with open('abbreviated.bib', 'w') as outfile:
+    outfile.write(bibtexdb)
+    print("Bibtex database with abbreviated files saved into 'abbreviated.bib'")

### 2017-12-19 ms/PNAS/collect_figures.py
collect figure and update zip on dropbox

@@ -1,0 +1,39 @@
+# import tempfile
+import shutil
+import os.path
+from os.path import join as join_path
+from os.path import split as split_path
+import re
+from zipfile import ZipFile
+
+if __name__ == '__main__':
+	ms_dir = '.'
+	figures_dir = '../../figures'
+	ms_figures_dir = join_path(ms_dir, 'figures')
+	zip_fname = '/Users/yoavram/Dropbox/Milpitas/PNAS_figures.zip'
+	tex_files = ['revision.tex', 'Supplementary_Figures.tex']
+	pattern = re.compile(r'\\includegraphics(?:\[.*\])?\{(.*\.\w{3})\}')
+
+	if os.path.exists(ms_figures_dir):
+		shutil.rmtree(ms_figures_dir)
+		os.mkdir(ms_figures_dir)
+
+	figures = []
+	for fn in tex_files:
+		with open(join_path(ms_dir, fn)) as f:
+			matches = (pattern.match(line) for line in f)
+			matches = (m for m in matches if m is not None)
+			filenames = (m.groups()[0] for m in matches)
+			filenames = (split_path(fn)[-1] for fn in filenames)
+			filenames = (join_path(figures_dir, fn) for fn in filenames)
+			figures.extend(filenames)
+
+	with ZipFile(zip_fname, 'w') as z:
+		for fn in figures:
+			fn = fn.replace('{', '').replace('}', '')
+			print(fn)
+			shutil.copy(fn, join_path(ms_figures_dir, split_path(fn)[-1]))
+			z.write(fn)
+
+	print("{} figures copied to {} and zipped to {}".format(
+		len(figures), ms_figures_dir, zip_fname))

### 2017-12-20 ms/PNAS/revision.tex
table 1 caption alignment

@@
 \addtabletext{
-\noindent $^\dagger$Note: $W=1$.\hfill\par
-\noindent (a): $\rho^*$ is the uninvadable value of the vertical transmission rate.\hfill\par
-\noindent (b): $\hat\rho$ is the rate that maximizes the geometric mean fitness at the stable equilibrium of the $AkBl$ cycle.\hfill}
+\noindent (a): \ $\rho^*$ is the uninvadable value of the vertical transmission rate.\hfill\par
+\noindent (b): \ $\hat\rho$ is the rate that maximizes the geometric mean fitness at \hfill\par
+\qquad\  the stable equilibrium of the $AkBl$ cycle.\hfill\par
+\noindent \ \ $\dagger$:\hskip.105truein Note: $W=1$.\hfill\par
+}
 \end{table}

### 2017-12-22 ms/PNAS/Supplementary_revision.tex
merged SI and SI figures into SI

@@
 \usepackage[parfill]{parskip}
-\usepackage{graphicx}
+\usepackage{graphicx,grffile}
 \usepackage{amssymb,amsmath,amsthm}
@@
 
-% messes up nameref
-%\usepackage{titlesec}
-%\titleformat{\subsection}[runin]
-%  {\bf}{\thesection}{1em}{}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{0}
 
@@
 
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\@fpsep\textheight
+\makeatother
+
+% supplement
 \newcommand{\beginsupplement}{%
@@
 \begin{document}
+\pagestyle{headings}
 \beginsupplement
@@
 
+% 1
+\begin{figure}
+\centering
+\includegraphics[width=0.9\linewidth]{../../figures/lk_phase_plane.pdf}
+\caption{Ratios of selection periods \(\frac{k}{l}\) that lead to
+fixation of phenotype \emph{A} (red) or polymorphism of phenotypes
+\emph{A} and \emph{B} (blue). \emph{k} and \emph{l} are the number of
+generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by selection.
+In all cases, \emph{W} = 1, \emph{w} = \(1-s\).}\label{fig:lk_phase_plane}
+\end{figure}
+
+% 2
+\begin{figure}
+\centering
+\includegraphics{../../figures/env_A1B1.pdf}
+\caption{Frequency of phenotype \emph{A} after every two generations in
+selection regime \(A1B1\). The orange line is the finite population model
+(eqs. 52-53; average of 100 simulations). The blue line is the infinite population model
+(eq.\ 21), and the green line is the  solution of \(Q(x)=0\) (eq.\ 22).
+In all cases, \(W=1\); for the finite population model (orange lines), population size is \(N=10,000\) and initial frequency of \(A\) is \(x_0=0.5\).}\label{fig:env_A1B1}
+\end{figure}
+
+% 3
+\begin{figure}
+\centering
+\includegraphics{../../figures/A1B1_equilibrium.pdf}
+\caption{Properties of stability in \emph{A1B1} selection regime.
+\textbf{(A)} Stable frequency of phenotype $A$ and \textbf{(B)} stable mean fitness as functions of the vertical transmission rate \(\rho\) and the fitness of the disfavored phenotype \(w\).
+Black contour lines join \(\rho\) and \(w\) combinations that result in the same stable value.
+In all cases, fitness of the favored phenotype is \emph{W=1}.
+}\label{fig:A1B1_equilibrium}
+\end{figure}
+
+% 4
+\begin{figure}
+\centering
+\includegraphics[width=0.9\linewidth]{../../figures/AkBk_x0.pdf}
+\caption{Convergence of the frequency of phenotype \emph{A} to a stable polymorphism in selection regime \emph{AkBk}.
+Comparison of dynamics starting with different initial frequencies of phenotype \(A\) (0.01--0.99), and different $k$, $\rho$ and $w$ values.
+The lines show the $x$ frequency of phenotype \emph{A} at the end of each period, after every $2k$ generations.
+In all cases, $W=1$.}\label{fig:AkBk_x0}
+\end{figure}
+
+% 5
+\begin{figure}
+\centering
+\includegraphics{../../figures/env_A1B2.pdf}
+\caption{Frequency of phenotype \emph{A} after every three generations in
+selection regime $A1B2$. Comparison of dynamics starting with
+different initial frequency of phenotype \(A\) (0.01--0.99).
+See also Figures 1 and  7. In all cases, $W=1$.}\label{fig:env_A1B2}
+\end{figure}
+
+% 6
+\begin{figure}
+\centering
+\includegraphics{../../figures/env_A3B10.pdf}
+\caption{Frequency of phenotype \emph{A} after every thirteen generations in
+selection regime $A3B10$. Comparison of dynamics starting with
+different initial frequency of phenotype \(A\) (0.01--0.99).
+In all cases, $W=1$.}\label{fig:env_A3B10}
+\end{figure}
+
+% 7
+\begin{figure}
+\centering
+\includegraphics{../../figures/AkBk_stable_wbar.pdf}
+\caption{Stable population mean fitness in selection regime \emph{AkBk} as a function of the vertical transmission rate \(\rho\) and the number \(k\) of generations in which phenotypes \emph{A} and \emph{B} are favored by selection, for different selection intensities: \textbf{(A)} \(w=0.1\), \textbf{(B)} \(w=0.5\), and \textbf{(C)} \(w=0.9\).
+Colors represent the geometric average of the stable population mean fitness over \(2k\) generations, calculated by iterating eq.
+10 until phenotype frequencies stabilized, and for at least \(1,000 \) generations.
+Blue markers show the maximum average mean fitness for each period \(k\).
+For example, with \(w=0.1\), \(\hat{\rho}=0\) maximizes the average fitness for \(k \le 11\), then \(\hat{\rho}\) increases to \(\hat{\rho} \approx 0.24\), and then continues to decrease as \(k\) increases, down to \(\hat{\rho} \approx 0.15\) for \(k=50\) (see also Fig.
+8A).
+Contour lines represent \(\rho\) and \(k\) combinations that produce the same average mean fitness. 
+In all cases, $W=1$.}\label{fig:AkBk_stable_wbar}
+\end{figure}
+
+% 8
+\begin{figure}
+\centering
+\includegraphics{../../figures/AkBk_w0.5_geomwbar.pdf}
+\caption{The geometric average of the stable population mean fitness over the $2k$ generation period peaks at $\rho=0$ for $k \le 30$ (red, blue and green lines) and at $\rho \approx 0.23$ for $k=31$ and $32$ (purple and orange lines).
+See also Figs. 8A, S5B. The inset zooms out to show that the geometric mean fitness is strictly and significantly decreasing for $\rho>0.3$ (reaching $\approx 0.7$ for $\rho=1$).
+In all cases, $W=1$, $w=0.5$.}\label{fig:AkBk_w0.5_geomwbar}
+\end{figure}
+
+% 9
+\begin{figure}
+\centering
+\includegraphics{../../figures/fixation_prob_time.pdf}
+\caption{Fixation probability and time in a finite population.
+\textbf{(A)} Fixation probability \(u(x)\) of phenotype \(A\) (eq.\ 55), and \textbf{(B)}
+Expected time to fixation \(T(x)\) of phenotype \(A\) (eq.\ 56) conditioned on its
+fixation, starting with a single copy in a population of size \(N\). The
+figure compares two estimates: Wright-Fisher simulations (blue circles) and diffusion
+equation approximation (green solid line). Parameters: selection coefficient, \(s=w_A-w_B=0.1\),
+population size, \(N=10,000\).}\label{fixation_prob_time}
+\end{figure}
+
+% 10
+\begin{figure}
+\centering
+\includegraphics{../../figures/AkBl_stable_modifier_w_0.1.pdf}
+ \caption{Evolutionarily stable vertical transmission rate in $AkBl$ selection regime.
+ The figure shows $\frac{\partial \lambda_1}{\partial P}$ the sensitivity of the leading eigenvalue of the external stability matrix $\mathbf{L}$ to changes in $P$ the vertical transmission rate of the invader allele as a function of $\rho$ the vertical transmission rate of the resident allele (see \emph{Section 2} in main text and \emph{Supplemental Material SP6} for details).
+ The shaded area marks $\rho$ values for which phenotype $B$ fixes and there is no polymorphism (Eq.~(20) in main text).
+ Without polymorphism, selection does not affect the transmission rate, so any rate in the shaded area is neutrally stable.
+ In panels A, B, D, G, J, and M, $\frac{\partial \lambda_1}{\partial P} < 0$ at the vicinity of $\rho=0$ and therefore the stable rate is $\rho^*=0$.
+ In panels B, C, E, F, H, I, K, and L, the stable rate $\rho^*$ can be identified as the $\rho$ value at which $\frac{\partial \lambda_1}{\partial P}$ changes from positive to negative.
+ In panels N and O, $\frac{\partial \lambda_1}{\partial P} > 0$ for all $\rho$ values that protect polymorphism, so there are only neutrally stable rates (in the shaded area).
+   Here, $W=1$ and $w=0.1$.}\label{fig:AkBl_stable_modifier_w_0.1}
+\end{figure}
+
+% 11
+\begin{figure}
+\centering
+\includegraphics[width=0.9\linewidth]{../../figures/lk_fix_prob.pdf}
+\caption{Fixation in a finite population with different ratios of selection periods \(\frac{k}{l}\).
+Fixation probability of phenotype $A$ when starting with a single copy in a population of size $N$: $u(1/N) = (1-\exp(-2 \rho \frac{k-l}{k+l}(W-w))/(1-\exp(-2 N \rho \frac{k-l}{k+l}(W-w))$ (see eqs.
+59--60).
+\emph{k} and \emph{l} are the number of
+generations in which phenotypes \emph{A} and \emph{B}, respectively, are favored by
+selection. In all cases, fitness of the favored phenotype, $W = 1$; fitness of the unfavored phenotype is $w=1-s$ and the population size is \(N=10,000\).} \label{lk_fix_prob}
+\end{figure}
+
+% 12
+\begin{figure}
+\centering
+\includegraphics{../../figures/A1B1_modifier_invasions.pdf}
+\caption{Consecutive fixation of modifiers that reduce the vertical
+transmission rate in selection regime \emph{A1B1}. The figure shows
+results of numerical simulations of evolution with two modifier alleles
+(Eq. (32) in main text).
+When a modifier allele fixes (frequency\textgreater{}99.9\%), a new modifier allele is introduced with a vertical transmission rate one order of magnitude lower (vertical
+dashed lines). \textbf{(A,D,G)} The frequency of phenotype \emph{A} in
+the population over time. \textbf{(B,E,H)} The frequency of the invading
+modifier allele over time. \textbf{(C,F,I)} The population geometric mean
+fitness over time; insets zoom in to show that the mean fitness 
+decreases slightly with each invasion. 
+Invading alleles are introduced at frequency 0.01\%; whenever their frequency drops below 0.01\% they are re-introduced.
+Parameters: vertical transmission rate of
+the initial resident modifier allele, \(\rho_0 =0.1\); fitness values:
+$W=1$; $w=0.1$ (\textbf{A-C}), 0.5 (\textbf{D-F}), and 0.9
+(\textbf{G-I}). The x-axis is on a log-scale, as each sequential invasion
+takes an order of magnitude longer to complete.
+Panels D-F are the same as in Figure 4 in the main text.
+}\label{fig:A1B1_modifier_invasions}
+\end{figure}
+
 % Bibliography
